---
title: "Clean and Analyze Compensation Data Provided by Schertz"
subtitle: "FY20-FY25"
description: ""
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true     # Set to 'false' for drafts. Override if you're NOT sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: false
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: false
    error: false
    warning: false
    message: false
    cache: false


---


```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED

source(here::here("R", "download_button_helpers.R"))

# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)



# Flextable built-in themes:
  # flextable::theme_alafoli()	|>  # BLAH
  # flextable::theme_apa()  # THIS IS NICE
  # flextable::theme_booktabs() |>  # NICE, MORE COMPACT
  # flextable::theme_box() |>   # OK, INCLUDES CELL BORDERS
  # flextable::theme_tron() |>  # 'DARK MODE' BLUE TEXT
  # flextable::theme_tron_legacy() |>   # 'DARK MODE' YELLOW TEXT
  # flextable::theme_vader() |>    # 'DARK MODE' WHITE TEXT
  # flextable::theme_vanilla() |>   # NOT SPECIAL
  # flextable::theme_zebra()	|>
  #


# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 28,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 22,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 20,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 18,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)



  
# Set seed for reproducibility
base::set.seed(123)

```


```{r}
#|label: helper-functions

# -----------------------------------------------------------------------------
# Download button helper (HTML only)
# -----------------------------------------------------------------------------
ds_download_data_btn <- function(df, filename_base, label = "Download Data (CSV)",
                                  clean_df = NULL) {
  write_df <- if (!base::is.null(clean_df)) clean_df else df
  tmp <- base::tempfile(fileext = ".csv")
  readr::write_excel_csv(write_df, tmp)
  downloadthis::download_file(
    path         = tmp,
    output_name  = filename_base,
    button_label = label,
    button_type  = "primary",
    has_icon     = TRUE,
    icon         = "fa fa-download"
  )
}


# -----------------------------------------------------------------------------
# Plot download button helper (HTML only)
# -----------------------------------------------------------------------------
ds_download_plot_btn <- function(plot_obj, filename_base, label = "Download Plot (PNG)",
                                  width = 13, height = 8, dpi = 300) {
  tmp <- base::tempfile(fileext = ".png")
  ggplot2::ggsave(tmp, plot = plot_obj, width = width, height = height,
                  dpi = dpi, bg = "white")
  downloadthis::download_file(
    path         = tmp,
    output_name  = filename_base,
    button_label = label,
    button_type  = "warning",
    has_icon     = TRUE,
    icon         = "fa fa-image"
  )
}

```

\

#### **Filename:** 
clean_and_analyze_fy20_fy25_compensation_data_provided_by_
schertz_with_buttons.qmd

\

## Introduction

This analysis implements **explicit column handling** for the Schertz compensation data pipeline. Every earnings component is captured distinctly, validated independently, and then summed to verify that totals match.

**Key Philosophy**: Each column name is specified **exactly** as it appears after `janitor::clean_names()`. No regex patterns, no guessing.

**Explicit Compensation Components** (FY23 structure):

| Component               | Excel Column Name        | Goes Into          |
|-------------------------|--------------------------|--------------------|
| Leave Payout            | FY23 LEAVE PAYOUT        | Total Earnings     |
| Regular Earnings        | FY23 REGULAR EARNINGS    | Total Earnings     |
| Overtime Earnings       | FY23 OVERTIME EARNINGS   | Total Earnings     |
| Additional Earnings     | FY23 ADDITIONAL EARNINGS | Total Earnings     |
| Deployment Earnings     | FY23 DEPLOYMENT EARNINGS | Total Earnings     |
| Arbitration/Settlements | ARBITRATION/ SETTLEMENTS | Total Earnings     |
| **Total Earnings**      | FY23 TOTAL EARNINGS      | (calculated)       |
| Additional Benefits     | FY23 ADDITIONAL BENEFITS | Total Compensation |
| **Total Compensation**  | FY23 TOTAL COMPENSATION  | (final)            |

**Validation Formula**:

```         
TOTAL_EARNINGS = LEAVE_PAYOUT + REGULAR_EARNINGS + OVERTIME_EARNINGS + 
                 ADDITIONAL_EARNINGS + DEPLOYMENT_EARNINGS + ARBITRATION_SETTLEMENTS

TOTAL_COMPENSATION = TOTAL_EARNINGS + ADDITIONAL_BENEFITS
```

------------------------------------------------------------------------

# PHASE 1: DISCOVERY

Run this section first to see exact column names after `clean_names()`.

## Step 1: Define Your Files

**Edit these filenames to match your actual files:**

```{r}
#| label: define-raw-data-files

# =============================================================================
# EDIT THESE FILENAMES TO MATCH YOUR ACTUAL FILES
# =============================================================================

files <- base::list(
  FY20 = "FY20_Earnings_CLEANED.xlsx",
  FY21 = "FY21_Earnings_CLEANED.xlsx",
  FY22 = "FY22_Earnings_CLEANED.xlsx",
  FY23 = "FY23Employee_Comp_FY23_CLEANED.xlsx",
  FY24 = "FY24Employee_Comp_FY24_All_CLEANED.xlsx",
  FY25 = "FY25Employee_Comp_FY25_CLEANED.xlsx"
)

data_folder <- here::here("data_raw")

# Configuration paths for export
paths <- base::list(
  data_raw = here::here("data_raw"),
  data_clean = here::here("data_clean"),
  reports = here::here("reports"),
  output_test = here::here("output_test")
)

# Report settings
report_settings <- base::list(
  test_mode = FALSE
)
```

## Step 2: Read Files and Show Column Names

This reads each file, applies `clean_names()`, and shows you the exact column names to use.

```{r}
#| label: import-and-show-columns


# Read each file and capture column names
file_info <- purrr::map_dfr(base::names(files), function(fy) {
  filepath <- base::file.path(data_folder, files[[fy]])
  
  if (!base::file.exists(filepath)) {
    return(dplyr::tibble(
      fiscal_year = fy,
      filename = files[[fy]],
      status = "FILE NOT FOUND",
      row_count = NA_integer_,
      columns = NA_character_
    ))
  }
  
  df <- readxl::read_excel(filepath) |>
    janitor::clean_names()
  
  dplyr::tibble(
    fiscal_year = fy,
    filename = files[[fy]],
    status = "OK",
    row_count = base::nrow(df),
    columns = base::paste(base::names(df), collapse = ", ")
  )
})

# Display summary
file_info |>
  dplyr::select(fiscal_year, filename, status, row_count) |>
  flextable::flextable() |>
  flextable::add_header_lines("File Import Summary") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: file-info-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_data_btn(
  file_info,
  filename_base = "schertz_file_import_summary_fy20_fy25",
  label = "Download File Import Summary (CSV)"
)
```
:::

## Step 3: Exact Column Names by File

**Run Step 3 code and copy the exact column names by year shown in the Console into Step 4 below.**

```{r}
#| label: show-exact-columns

# Show columns for each file in a usable format
for (fy in base::names(files)) {
  filepath <- base::file.path(data_folder, files[[fy]])
  
  if (base::file.exists(filepath)) {
    df <- readxl::read_excel(filepath) |>
      janitor::clean_names()
    
    base::cat("\n")
    base::cat("# =========================================================\n")
    base::cat("#", fy, "- Columns after clean_names():\n")
    base::cat("# =========================================================\n")
    
    cols <- base::names(df)
    for (i in base::seq_along(cols)) {
      base::cat("#", base::sprintf("%2d", i), ":", cols[i], "\n")
    }
  }
}
```

------------------------------------------------------------------------

# ⛔ STOP HERE FIRST TIME

Run Phase 1 above. Then map the actual column names shown to fill in the exact mappings in Phase 2 below.  Actual column names on right; standardized column names on left.

------------------------------------------------------------------------

# PHASE 2: PROCESSING

**After reviewing Phase 1 output, edit the column mappings below.**

## Step 4: Define Exact Column Mappings

**Edit rows below.  DO NOT change the standardized names (left side).  Edit right side to match the exact column names from Phase 1:**

```{r}
#| label: define-column-mappings

# =============================================================================
# COLUMN MAPPINGS - EDIT THESE TO MATCH YOUR FILES EXACTLY
# =============================================================================
# Use the exact column names shown in Phase 1 (after clean_names)
# Set to NA if that column doesn't exist in a given year

column_maps <- base::list(
  
  # ----- FY20 -----
  FY20 = base::list(
    # Status column
    status = "status",
    # Identity columns
    name = "name",                    # Single name column (or NA if split)
    last_name =NA,                   # NA if using single name column
    first_name = NA,                  # NA if using single name column
    department = "department",
    job_title = "job_title",
    hire_date = "hire_date",
    termination_date = "termination_date",
    
    # Earnings columns (set to NA if not present)
    leave_payout = NA,
    regular_earnings = "regular_earnings",
    overtime_earnings = "overtime_earnings",
    additional_earnings = "additional_earnings",
    deployment_earnings = NA,
    arbitration_settlements = NA,
    
    # Benefits (set to NA if not present)
    additional_benefits = NA,
    
    # Totals for validation
    total_earnings = "total_earnings",
    total_compensation = NA,          # NA if no total_comp column
    misc_earnings = NA
  ),
  
  # ----- FY21 -----
  FY21 = base::list(
    status = "status",
    name = "name",
    last_name =NA,
    first_name = NA,
    department = "home_department",
    job_title = "job_title",
    hire_date = "hire_date",
    termination_date = "termination_date",
    leave_payout = NA,
    regular_earnings = "regular_earnings",
    overtime_earnings = "overtime_earnings",
    additional_earnings = "additional_earnings",
    deployment_earnings = NA,
    arbitration_settlements = NA,
    additional_benefits = NA,
    total_earnings = "total_earnings",
    total_compensation = NA,
    misc_earnings = NA
  ),
  
  # ----- FY22 -----
  FY22 = base::list(
    status = "status",
    name = "name",
    last_name =NA,
    first_name = NA,
    department = "department",
    job_title = "job_title",
    hire_date = "hire_date",
    termination_date = "termination_date",
    leave_payout = NA,
    regular_earnings = "regular_earnings",
    overtime_earnings = "overtime_earnings",
    additional_earnings = "additional_earnings",
    deployment_earnings = NA,
    arbitration_settlements = NA,
    additional_benefits = NA,
    total_earnings = "total_earnings",
    total_compensation = NA,
    misc_earnings = NA
  ),
  
  # ----- FY23 -----
  FY23 = base::list(
    name = NA,                                    # NA - using split names
    last_name ="last_name",
    first_name = "first_name",
    department = "department",
    job_title = "job_title",
    employee_category = "employee_category",
    hire_date = "hire_date",
    separation_date = "separation_date",
    annual_salary = NA,
    leave_payout = "fy23_leave_payout",
    regular_earnings = "fy23_regular_earnings",
    overtime_earnings = "fy23_overtime_earnings",
    additional_earnings = "fy23_additional_earnings",
    deployment_earnings = "fy23_deployment_earnings",
    arbitration_settlements = "arbitration_settlements",
    additional_benefits = "fy23_additional_benefits",
    total_earnings = "fy23_total_earnings",
    total_compensation = "fy23_total_compensation",
    misc_earnings = NA
  ),
  
  # ----- FY24 -----
  FY24 = base::list(
    name = NA,
    last_name ="last_name",
    first_name = "first_name",
    department = "home_department",
    job_title = "job_title",
    employee_category = "employee_category",
    hire_date = "hire_date",
    separation_date = "separation_date",
    annual_salary = NA,
    leave_payout = "fy24_leave_payout",
    regular_earnings = "fy24_regular_earnings",
    overtime_earnings = "fy24_overtime_earnings",
    additional_earnings = "fy24_additional",
    deployment_earnings = "fy24_deployment",
    arbitration_settlements = "arbitration_settlements",
    additional_benefits = "fy24_additional_2",
    total_earnings = "fy24_total_earnings",
    total_compensation = "fy24_total_compensation",
    misc_earnings = NA
  ),

# ----- FY25 -----
FY25 = base::list(
  name = NA,
  last_name ="last_name",
  first_name = "first_name",
  position_id = "position_id",
  department = "department",
  job_title = "job_title",
  employee_category = "employee_category",
  hire_date = "hire_date",
  separation_date = "separation_date",
  leave_payout = "fy25_leave_payout",
  regular_earnings = "fy25_regular_earnings",
  overtime_earnings = "fy25_overtime_earnings",
  additional_earnings = "fy25_additional_earnings",
  deployment_earnings = "fy25_deployment_earnings",
  arbitration_settlements = "arbitration_settlements_fy25",
  additional_benefits = "fy25_additional_benefits",
  total_earnings = "total_earnings_fy25",
  total_compensation = "fy25_total_compensation",
  misc_earnings = NA
  )
)

base::cat("Column mappings defined for", base::length(column_maps), "fiscal years\n")
```

------------------------------------------------------------------------

## Step 5: Utility Functions

```{r}
#| label: utility-functions

# =============================================================================
# UTILITY FUNCTIONS - RUN THIS AND PUT THEM INTO GLOBAL ENVIRONMENT
# =============================================================================

# Parse currency to numeric
parse_currency <- function(x) {
  if (base::is.numeric(x)) return(x)
  x <- stringr::str_remove_all(x, "[$,\\s]")
  base::suppressWarnings(base::as.numeric(x))
}

# Fix invisible spaces
fix_spaces <- function(x) {
  if (!base::is.character(x)) return(x)
  stringr::str_squish(x)
}

# Standardize name for matching
standardize_name <- function(x) {
  x <- fix_spaces(x)
  x <- stringr::str_to_upper(x)
  stringr::str_remove_all(x, "[^A-Z ]")
}

# Safe column extract - returns 0 if column is NA or doesn't exist
safe_extract <- function(df, col_name) {
  if (base::is.na(col_name) || !col_name %in% base::names(df)) {
    return(base::rep(0, base::nrow(df)))
  }
  result <- parse_currency(df[[col_name]])
  dplyr::coalesce(result, 0)
}
```

------------------------------------------------------------------------

## Step 6: Process Each File for Each Year

```{r}
#| label: process-all-files-into-wide

# =============================================================================
# PROCESS EACH FILE INTO WIDE FORMAT
# =============================================================================

process_file <- function(fy, filename, col_map) {
  
  filepath <- base::file.path(data_folder, filename)
  
  if (!base::file.exists(filepath)) {
    base::warning("File not found: ", filepath)
    return(NULL)
  }
  
  df <- readxl::read_excel(filepath) |>
    janitor::clean_names()
  
  # Handle column name - either single column or split
  if (!base::is.na(col_map$name)) {
    # Single name column
    name_values <- fix_spaces(df[[col_map$name]])
    last_name <- NA_character_
    first_name <- NA_character_
  } else {
    # Handle split name columns
    last_name <- fix_spaces(df[[col_map$last_name]])
    first_name <- fix_spaces(df[[col_map$first_name]])
    name_values <- base::paste(last_name, first_name, sep = ", ")
  }
  
  # Build result
  result <- dplyr::tibble(
    fiscal_year = fy,
    name = name_values,
    name_std = standardize_name(name_values),
    last_name =last_name,
    first_name = first_name,
    department = fix_spaces(df[[col_map$department]]),
    job_title = fix_spaces(df[[col_map$job_title]]),
    
    # Earnings
    leave_payout = safe_extract(df, col_map$leave_payout),
    regular_earnings = safe_extract(df, col_map$regular_earnings),
    overtime_earnings = safe_extract(df, col_map$overtime_earnings),
    additional_earnings = safe_extract(df, col_map$additional_earnings),
    deployment_earnings = safe_extract(df, col_map$deployment_earnings),
    arbitration_settlements = safe_extract(df, col_map$arbitration_settlements),
    additional_benefits = safe_extract(df, col_map$additional_benefits),
    
    # Source totals for validation
    source_total_earnings = safe_extract(df, col_map$total_earnings),
    source_total_compensation = safe_extract(df, col_map$total_compensation)
  )
  
  # Calculate misc_earnings (gap between source earnings and itemized earnings)
  result <- result |>
    dplyr::mutate(
      itemized_earnings = leave_payout + regular_earnings + overtime_earnings +
                          additional_earnings + deployment_earnings + arbitration_settlements,
      misc_earnings = dplyr::if_else(
        source_total_earnings > 0,
        pmax(source_total_earnings - itemized_earnings, 0),
        0
      )
    )
  
  # Calculate totals
  result <- result |>
    dplyr::mutate(
      calc_total_earnings = itemized_earnings + misc_earnings,
      calc_total_compensation = calc_total_earnings + additional_benefits,
      employee_key = base::paste(name_std, department, sep = "_")
    )
  
  # For files without source_total_compensation, calculate what it SHOULD be
  # (source_total_earnings + additional_benefits)
  result <- result |>
    dplyr::mutate(
      source_total_compensation = dplyr::if_else(
        source_total_compensation == 0 & source_total_earnings > 0,
        source_total_earnings + additional_benefits,
        source_total_compensation
      )
    )
  
  # Drop helper column
  result <- result |>
    dplyr::select(-itemized_earnings)
  
  return(result)
}

# Process all files
all_data <- purrr::map2_dfr(
  base::names(files),
  files,
  function(fy, filename) {
    process_file(fy, filename, column_maps[[fy]])
  }
)

base::cat("Processed", base::nrow(all_data), "total xlsx rows\n")
base::cat("For Fiscal years:", base::paste(base::unique(all_data$fiscal_year), collapse = ", "), "\n")
```

------------------------------------------------------------------------

## Step 7a: Validation - Compare Calculated vs xlsx Source Totals

```{r}
#| label: validation

# =============================================================================
# VALIDATION: COMPARE CALCULATED VS SOURCE TOTALS
# =============================================================================

validation <- all_data |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    rows = dplyr::n(),
    calc_total = base::sum(calc_total_compensation, na.rm = TRUE),
    source_total = base::sum(source_total_compensation, na.rm = TRUE),
    difference = calc_total - source_total,
    .groups = "drop"
  ) |>
  dplyr::mutate(
    pct_diff = abs(difference) / source_total * 100,
    status = dplyr::if_else(abs(difference) <= 500 | pct_diff < 0.01, "PASS", "FAIL")
  )

validation |>
  dplyr::mutate(
    calc_total = scales::dollar(calc_total, accuracy = 1),
    source_total = scales::dollar(source_total, accuracy = 1),
    difference = scales::dollar(difference, accuracy = 0.01),
    pct_diff = sprintf("%.4f%%", pct_diff)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Validation: Compare Calculated vs Raw Source Totals") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::bg(i = ~ status == "FAIL", bg = "#ffcccc") |>
  flextable::bg(i = ~ status == "PASS", bg = "#ccffcc") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: validation-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_data_btn(
  validation,
  filename_base = "schertz_comp_validation_fy20_fy25",
  label = "Download Validation Results (CSV)"
)
```
:::

```{r}
#| label: validation-interpretation

# Report on validation status
validation_passed <- base::all(validation$status == "PASS")

if (!validation_passed) {
  base::cat("\n=== VALIDATION FAILED ===\n\n")
  
  failures <- validation |>
    dplyr::filter(status == "FAIL")
  
  for (i in base::seq_len(base::nrow(failures))) {
    row <- failures[i, ]
    base::cat("MISMATCH in", row$fiscal_year, ":\n")
    base::cat("  - Calculated:", scales::dollar(row$calc_total), "\n")
    base::cat("  - Source:", scales::dollar(row$source_total), "\n")
    base::cat("  - Difference:", scales::dollar(row$difference), "\n\n")
  }
  
  base::cat("ACTION REQUIRED:\n")
  base::cat("1. Check the column_maps for missing or incorrect column names\n")
  base::cat("2. Review the column names shown in Phase 1\n")
  base::cat("3. Update the column mappings to match the exact column names\n")
  
} else {
  base::cat("Validation: PASSED\n")
  base::cat("All calculated totals match source file totals within $1.00\n")
}
```

## Step 7b: Impute Benefits Component for FY20-FY22 From FY23-FY25 Data

```{r}
#| label: impute-benefits-fy20-fy22

# =============================================================================
# IMPUTE ADDITIONAL BENEFITS FOR FY20-FY22
# =============================================================================
# FY20-FY22 files don't have additional_benefits data
# We estimate it based on the benefits-to-regular-earnings ratio from FY23-FY25

# Step 1: Calculate benefits as % of regular earnings for FY23-FY25
benefits_ratios <- all_data |>
  dplyr::filter(fiscal_year %in% c("FY23", "FY24", "FY25")) |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    total_regular = sum(regular_earnings, na.rm = TRUE),
    total_benefits = sum(additional_benefits, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    benefits_pct = total_benefits / total_regular * 100,
    benefits_ratio = total_benefits / total_regular
  )

# Display the ratios
benefits_ratios |>
  dplyr::mutate(
    total_regular = scales::dollar(total_regular, accuracy = 1),
    total_benefits = scales::dollar(total_benefits, accuracy = 1),
    benefits_pct = base::sprintf("%.2f%%", benefits_pct)
  ) |>
  dplyr::select(fiscal_year, total_regular, total_benefits, benefits_pct) |>
  flextable::flextable() |>
  flextable::add_header_lines("Calculate Benefits as % of Regular Earnings (FY23-FY25)") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()

# Step 2: Calculate average ratio to use for imputation
# (or you could use a trend/interpolation if the ratios vary significantly)
avg_benefits_ratio <- base::mean(benefits_ratios$benefits_ratio)
base::cat("\nAverage benefits ratio (FY23-FY25):\n", 
          base::sprintf("%.4f (%.2f%%)", avg_benefits_ratio, avg_benefits_ratio * 100), "\n")

# Step 3: Impute benefits for FY20-FY22 and add to the all_data object
all_data <- all_data |>
  dplyr::mutate(
    # Flag imputed rows
    benefits_imputed = fiscal_year %in% c("FY20", "FY21", "FY22"),
    
    # Impute additional_benefits for early years
    additional_benefits = dplyr::if_else(
      benefits_imputed & additional_benefits == 0,
      regular_earnings * avg_benefits_ratio,
      additional_benefits
    ),
    
    # Recalculate total compensation with imputed benefits
    calc_total_compensation = calc_total_earnings + additional_benefits
  )

# Step 4: Verify the imputation
imputation_summary <- all_data |>
  dplyr::group_by(fiscal_year, benefits_imputed) |>
  dplyr::summarise(
    employees = dplyr::n(),
    total_regular = sum(regular_earnings, na.rm = TRUE),
    total_benefits = sum(additional_benefits, na.rm = TRUE),
    total_comp = sum(calc_total_compensation, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    benefits_pct = total_benefits / total_regular * 100
  )

base::cat("\n--- Imputation Summary ---\n")
imputation_summary |>
  dplyr::mutate(
    total_regular = scales::dollar(total_regular, accuracy = 1),
    total_benefits = scales::dollar(total_benefits, accuracy = 1),
    total_comp = scales::dollar(total_comp, accuracy = 1),
    benefits_pct = base::sprintf("%.2f%%", benefits_pct),
    benefits_imputed = dplyr::if_else(benefits_imputed, "Yes (Imputed)", "No (Actual)")
  ) |>
  dplyr::rename(
    FY = fiscal_year,
    'Benefits Imputed?' = benefits_imputed,
    `Employee Count` = employees,
    `Total Regular Earnings` = total_regular,
    `Added Benefits` = total_benefits,
    `Total Comp with Benefits` = total_comp,
    `Benefits as % of Total Reg Earnings` = benefits_pct
  )|>
  flextable::flextable() |>
  flextable::add_header_lines("Revised Compensation Summary with Imputed Benefits Included") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::bg(i = ~ grepl("Imputed", `Benefits Imputed?`), bg = "#ffffcc") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: benefits-imputation-downloads
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_data_btn(
  benefits_ratios,
  filename_base = "schertz_benefits_ratios_fy23_fy25",
  label = "Download Benefits Ratios (CSV)"
)
ds_download_data_btn(
  imputation_summary,
  filename_base = "schertz_imputation_summary_fy20_fy25",
  label = "Download Imputation Summary (CSV)"
)
```
:::


## Step 7c: Add inflation rates for all years covered in analysis

```{r}
#| label: add-inflation-rates

# =============================================================================
# ADD PUBLISHED INFLATION RATES (FRED Series: FPCPITOTLZGUSA)
# =============================================================================

# Official annual inflation rates from FRED
# FY20 is base year, so we only need FY21-FY25 rates
inflation_rates <- dplyr::tibble(
  fiscal_year = c("FY20", "FY21", "FY22", "FY23", "FY24", "FY25"),
  inflation_rate = c(NA_real_, 4.70, 8.00, 4.12, 2.95, 2.52)  # FY20 = base, no rate
)

# Join to all_data
all_data <- all_data |>
  dplyr::left_join(inflation_rates, by = "fiscal_year")

base::cat("Added inflation rates to all_data\n")
```

------------------------------------------------------------------------

## Step 8: Total Compensation Components Breakdown by Year

```{r}
#| label: component-breakdown

# =============================================================================
# COMPONENT BREAKDOWN BY FISCAL YEAR
# =============================================================================

component_summary <- all_data |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    `Leave Payout` = base::sum(leave_payout, na.rm = TRUE),
    `Regular` = base::sum(regular_earnings, na.rm = TRUE),
    `Overtime` = base::sum(overtime_earnings, na.rm = TRUE),
    `Additional` = base::sum(additional_earnings, na.rm = TRUE),
    `Deployment` = base::sum(deployment_earnings, na.rm = TRUE),
    `Arbitration` = base::sum(arbitration_settlements, na.rm = TRUE),
    `Benefits` = base::sum(additional_benefits, na.rm = TRUE),
    `TOTAL` = base::sum(calc_total_compensation, na.rm = TRUE),
    .groups = "drop"
  )

component_summary |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.numeric), ~ scales::dollar(.x, accuracy = 1))
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Total Compensaton Breakdown by Component by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: component-summary-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_data_btn(
  component_summary,
  filename_base = "schertz_comp_component_summary_fy20_fy25",
  label = "Download Component Summary (CSV)"
)
```
:::

------------------------------------------------------------------------

## Step 9: Convert to Long Format

```{r}
#| label: convert-to-long

# =============================================================================
# CONVERT TO LONG FORMAT FOR ANALYSIS
# =============================================================================

final_df <- all_data |>
  dplyr::select(
    fiscal_year, name, name_std, last_name, first_name,
    department, job_title, employee_key,
    leave_payout, regular_earnings, overtime_earnings,
    additional_earnings, deployment_earnings, arbitration_settlements,
    additional_benefits
  ) |>
  tidyr::pivot_longer(
    cols = c(leave_payout, regular_earnings, overtime_earnings,
             additional_earnings, deployment_earnings, arbitration_settlements,
             additional_benefits),
    names_to = "earnings_type",
    values_to = "amount"
  ) |>
  dplyr::mutate(
    earnings_type = dplyr::case_when(
      earnings_type == "leave_payout" ~ "Leave",
      earnings_type == "regular_earnings" ~ "Regular",
      earnings_type == "overtime_earnings" ~ "Overtime",
      earnings_type == "additional_earnings" ~ "Additional",
      earnings_type == "deployment_earnings" ~ "Deployment",
      earnings_type == "arbitration_settlements" ~ "Arbitration",
      earnings_type == "additional_benefits" ~ "Benefits",
      TRUE ~ earnings_type
    )
  )

base::cat("Long format data created:", base::nrow(final_df), "rows\n")
base::cat("Unique employees:", dplyr::n_distinct(final_df$employee_key), "\n")
```

------------------------------------------------------------------------

## Step 10: Checkpoint H - Compare Components of Total Compensation to Original Data Provided

This checkpoint shows exactly how much of each component was captured, proving nothing was missed.

```{r}
#| label: checkpoint-h-component-breakdown

# =============================================================================
# CHECKPOINT H: EXPLICIT COMPONENT BREAKDOWN
# =============================================================================

# Create a detailed breakdown showing each component's contribution
component_breakdown <- final_df |>
  dplyr::group_by(fiscal_year, earnings_type) |>
  dplyr::summarise(
    amount = base::sum(amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(
    names_from = earnings_type,
    values_from = amount,
    values_fill = 0
  )

# Add totals row
component_breakdown <- component_breakdown |>
  dplyr::mutate(
    `Total Earnings` = `Leave` + Regular + Overtime + Additional + Deployment + Arbitration,
    `Total Compensation` = `Total Earnings` + Benefits
  )

# Display
component_breakdown |>
  dplyr::rename(FY = fiscal_year) |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.numeric), ~ scales::dollar(.x, accuracy = 1))
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Checkpoint H: Earnings with Benefits by Component by Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: component-breakdown-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_data_btn(
  component_breakdown,
  filename_base = "schertz_comp_component_breakdown_fy20_fy25",
  label = "Download Component Breakdown (CSV)"
)
```
:::

------------------------------------------------------------------------

## Step 11a: Summary Statistics

```{r}
#| label: summary-statistics

# =============================================================================
# SUMMARY STATISTICS
# =============================================================================

# Summary by year
yearly_summary <- final_df |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    total_compensation = base::sum(amount, na.rm = TRUE),
    total_employees = dplyr::n_distinct(employee_key),
    avg_compensation = total_compensation / total_employees,
    .groups = "drop"
  ) |>
  dplyr::left_join(inflation_rates, by = "fiscal_year") |>
  dplyr::arrange(fiscal_year) |>
  dplyr::mutate(
    tot_comp_growth_rate = (total_compensation - dplyr::lag(total_compensation)) /
                            dplyr::lag(total_compensation) * 100
  )

yearly_summary |>
  dplyr::mutate(
    total_compensation = scales::dollar(total_compensation, accuracy = 1),
    avg_compensation = scales::dollar(avg_compensation, accuracy = 1),
    tot_comp_growth_rate = dplyr::if_else(
      base::is.na(tot_comp_growth_rate),
      "—",
      base::paste0(base::round(tot_comp_growth_rate, 1), "%")
    )
    ) |>
  dplyr::rename(
    Year = fiscal_year,
    'Total Comp' = total_compensation,
    EEs = total_employees,
    'Avg Tot Comp' = avg_compensation,
    'Total Comp Growth Rate' = tot_comp_growth_rate
  ) |>
  dplyr::select(Year, 'Total Comp', EEs, 'Avg Tot Comp', 'Total Comp Growth Rate', inflation_rate) |>

  flextable::flextable() |>
  flextable::add_header_lines("Total Compensation, Headcount, Avg Total Comp, Total Compensation Growth Rate, and Inflation Rates for  FY20 - FY25") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: yearly-summary-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_data_btn(
  yearly_summary,
  filename_base = "schertz_comp_yearly_summary_fy20_fy25",
  label = "Download Yearly Summary (CSV)"
)
```
:::

------------------------------------------------------------------------

## Step 11b: Total Regular Earnings Summary by Fiscal Year, Including Time-to-Double (in Years)

Regular Earnings here means all earnings *excluding* Benefits: 

Specifically:  Additional Earnings + Arbitration + Leave + Overtime + Regular pay.

Growth rates and time-to-double are calculated year-over-year between adjacent fiscal years.

```{r}
#| label: reg-earnings-summary

# =============================================================================
# REGULAR EARNINGS SUMMARY (excludes Benefits)
# Components: Additional, Arbitration, Leave, Overtime, Regular
# =============================================================================

reg_earnings_summary <- component_breakdown |>
  dplyr::mutate(
    total_reg_earnings = Additional + Arbitration + Leave + Overtime + Regular
  ) |>
  dplyr::select(fiscal_year, total_reg_earnings) |>
  dplyr::left_join(
    yearly_summary |> dplyr::select(fiscal_year, total_employees),
    by = "fiscal_year"
  ) |>
  dplyr::arrange(fiscal_year) |>
  dplyr::mutate(
    avg_reg_earnings   = total_reg_earnings / total_employees,
    # YOY growth rate: (current - prior) / prior * 100
    annual_growth_rate = (total_reg_earnings - dplyr::lag(total_reg_earnings)) /
                          dplyr::lag(total_reg_earnings) * 100,
    # Rule of 70: years to double = 70 / growth_rate
    time_to_double     = dplyr::if_else(
      !base::is.na(annual_growth_rate) & annual_growth_rate > 0,
      70 / annual_growth_rate,
      NA_real_
    )
  ) |>
  dplyr::select(
    fiscal_year, total_reg_earnings, avg_reg_earnings,
    annual_growth_rate, time_to_double
  )

reg_earnings_summary |>
  dplyr::rename(
    FY                        = fiscal_year,
    `Total Reg Earnings`      = total_reg_earnings,
    `Avg Reg Earnings`        = avg_reg_earnings,
    `Annual Growth Rate (%)`  = annual_growth_rate,
    `Time to Double (yrs)`    = time_to_double
  ) |>
  dplyr::mutate(
    `Total Reg Earnings`     = scales::dollar(`Total Reg Earnings`, accuracy = 1),
    `Avg Reg Earnings`       = scales::dollar(`Avg Reg Earnings`, accuracy = 1),
    `Annual Growth Rate (%)` = dplyr::if_else(
      base::is.na(`Annual Growth Rate (%)`),
      "—",
      base::paste0(base::round(`Annual Growth Rate (%)`, 1), "%")
    ),
    `Time to Double (yrs)`   = dplyr::if_else(
      base::is.na(`Time to Double (yrs)`),
      "—",
      base::as.character(base::round(`Time to Double (yrs)`, 1))
    )
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines(
    "Total Regular Earnings Summary Including Time-to-Double by Fiscal Year (Additional + Arbitration + Leave + Overtime + Total Regular)"
  ) |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: reg-earnings-summary-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_data_btn(
  reg_earnings_summary,
  filename_base = "schertz_reg_earnings_summary_fy20_fy25",
  label = "Download Regular Earnings Summary (CSV)"
)
```
:::

------------------------------------------------------------------------


## Step 12: Export Data

```{r}
#| label: export-data

# =============================================================================
# EXPORT DATA - CORRECT FILENAMES AND UNCOMMENT THEN RUN
# =============================================================================

# Ensure output directory exists
if (!base::dir.exists(paths$data_clean)) {
  base::dir.create(paths$data_clean, recursive = TRUE)
}

# Export
test_suffix <- base::ifelse(report_settings$test_mode, "_TEST", "")

rds_path <- base::file.path(paths$data_clean, base::paste0("compensation_long_fy20_fy25", test_suffix, ".rds"))
base::saveRDS(final_df, file = rds_path)

csv_path <- base::file.path(paths$data_clean, base::paste0("compensation_long_fy20_fy25", test_suffix, ".csv"))
readr::write_csv(final_df, file = csv_path)

# Also export the wide format for convenience
wide_rds_path <- base::file.path(paths$data_clean, base::paste0("compensation_wide_fy20_fy25", test_suffix, ".rds"))
base::saveRDS(all_data, file = wide_rds_path)

wide_csv_path <- base::file.path(paths$data_clean, base::paste0("compensation_wide_fy20_fy25", test_suffix, ".csv"))
readr::write_csv(all_data, file = wide_csv_path)

base::cat("Data exported:\n")
base::cat("  Long format:\n")
base::cat("    -", rds_path, "\n")
base::cat("    -", csv_path, "\n")
base::cat("  Wide format:\n")
base::cat("    -", wide_rds_path, "\n")
base::cat("    -", wide_csv_path, "\n")
```

------------------------------------------------------------------------



## Appendix A: Column Structure Comparison Chart

This shows exactly what columns exist in each year's file, helping identify when the spreadsheet structure changes.


------------------------------------------------------------------------

```{r}
#| label: column-structure-comparison

# =============================================================================
# APPENDIX A: COLUMN STRUCTURE COMPARISON
# =============================================================================


# Reset data_folder to raw data location
data_folder <- here::here("data_raw")

# Rest of your code...
# Build a comparison of actual column names across years
col_comparison <- purrr::map_dfr(
  base::names(files),
  function(fy) {
    filepath <- base::file.path(data_folder, files[[fy]])
    
    if (base::file.exists(filepath)) {
      df <- readxl::read_excel(filepath) |>
        janitor::clean_names()
      
      dplyr::tibble(
        fiscal_year = fy,
        column_name = base::names(df)
      )
    } else {
      NULL
    }
  }
)

# Check what we got
base::cat("Rows:", nrow(col_comparison), "\n")
base::cat("Columns:", paste(names(col_comparison), collapse = ", "), "\n")

# Now pivot
col_comparison <- col_comparison |>
  dplyr::mutate(present = TRUE) |>
  tidyr::pivot_wider(
    names_from = fiscal_year,
    values_from = present,
    values_fill = FALSE
  )

col_comparison |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.logical), ~ dplyr::if_else(.x, "✓", ""))
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Source File Columns by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()

```

::: {.content-visible when-format="html"}
```{r}
#| label: col-comparison-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_data_btn(
  col_comparison,
  filename_base = "schertz_source_column_structure_fy20_fy25",
  label = "Download Column Structure (CSV)"
)
```
:::

```{r}
#| label: processed-data-structure

# =============================================================================
# APPENDIX A-2: PROCESSED DATA STRUCTURE
# =============================================================================

# Show columns in the processed all_data dataframe
processed_cols <- dplyr::tibble(
  column_name = base::names(all_data),
  data_type = purrr::map_chr(all_data, ~ base::class(.x)[1])
)

processed_cols |>
  flextable::flextable() |>
  flextable::add_header_lines("Processed Data Structure (all_data) Columns") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: processed-cols-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_data_btn(
  processed_cols,
  filename_base = "schertz_processed_data_structure",
  label = "Download Processed Data Structure (CSV)"
)
```
:::



## Appendix B: Data Quality Checks

```{r}
#| label: data-quality-checks

# =============================================================================
# APPENDIX B: DATA QUALITY CHECKS
# =============================================================================

# Check for missing names
missing_names <- all_data |>
  dplyr::filter(base::is.na(name) | name == "" | name == "NA, NA") |>
  dplyr::select(fiscal_year, name, department, job_title)

if (base::nrow(missing_names) > 0) {
  base::cat("WARNING: Found", base::nrow(missing_names), "records with missing names:\n")
  base::print(missing_names)
} else {
  base::cat("No missing names found.\n")
}

# Check for duplicate employee keys within a year
duplicates <- all_data |>
  dplyr::group_by(fiscal_year, employee_key) |>
  dplyr::filter(dplyr::n() > 1) |>
  dplyr::arrange(fiscal_year, employee_key)

if (base::nrow(duplicates) > 0) {
  base::cat("\nWARNING: Found", base::nrow(duplicates), "duplicate employee keys:\n")
  duplicates |>
    dplyr::select(fiscal_year, name, department, job_title) |>
    base::print()
} else {
  base::cat("No duplicate employee keys found within fiscal years.\n")
}

# Check for negative amounts
negative_amounts <- all_data |>
  dplyr::filter(
    leave_payout < 0 | regular_earnings < 0 | overtime_earnings < 0 |
    additional_earnings < 0 | deployment_earnings < 0 | arbitration_settlements < 0 |
    additional_benefits < 0
  )

if (base::nrow(negative_amounts) > 0) {
  base::cat("\nNOTE: Found", base::nrow(negative_amounts), "records with negative amounts (may be corrections/adjustments)\n")
} else {
  base::cat("No negative amounts found.\n")
}
```


------------------------------------------------------------------------




## Appendix C: Total Compensation and Headcount Year-over-Year Changes

```{r}
#| label: yoy-comparison

# =============================================================================
# APPENDIX C: YEAR-OVER-YEAR COMPARISON OF EE TOTALS, COMPENSATION, PCT, AVG COMP
# =============================================================================

yoy_summary <- yearly_summary |>
  dplyr::mutate(
    prev_total = dplyr::lag(total_compensation),
    prev_employees = dplyr::lag(total_employees),
    total_change = total_compensation - prev_total,
    total_pct_change = (total_change / prev_total) * 100,
    employee_change = total_employees - prev_employees
  )

yoy_summary |>
  dplyr::mutate(
    total_compensation = scales::dollar(total_compensation, accuracy = 1),
    avg_compensation = scales::dollar(avg_compensation, accuracy = 1),
    total_change = scales::dollar(total_change, accuracy = 1),
    total_pct_change = base::sprintf("%.1f%%", total_pct_change)
  ) |>
  dplyr::select(
    fiscal_year, total_employees, employee_change,
    total_compensation, total_change, total_pct_change, avg_compensation
  ) |>
  dplyr::rename(FY = fiscal_year, EEs = total_employees, 'EE Chnge' = employee_change, 'Total Comp' = total_compensation, 'Total Chnge' = total_change, 'Pct Chnge' = total_pct_change, 'Avg Total Comp' = avg_compensation) |>
  flextable::flextable() |>
  flextable::add_header_lines("COMPARISON of Headcount, Headcount Change, Total Comp, Change in Total Comp, Pct Change in Total Comp, Avg Total Comp for FY20-FY25") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: yoy-summary-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_data_btn(
  yoy_summary,
  filename_base = "schertz_comp_yoy_comparison_fy20_fy25",
  label = "Download YoY Comparison (CSV)"
)
```
:::


------------------------------------------------------------------------



## Appendix D: Bump Chart - Top 3 in Total Compensation

```{r}
#| label: top-3-bump-chart
top_3_earners <- all_data |>
  dplyr::slice_max(
    order_by = calc_total_compensation, 
    n = 3, 
    by = fiscal_year
  ) |>
  dplyr::arrange(fiscal_year, dplyr::desc(calc_total_compensation)) |>
  dplyr::mutate(rank = dplyr::row_number(), .by = fiscal_year)

p_top3_bump <- ggplot2::ggplot(
  top_3_earners, 
  ggplot2::aes(x = fiscal_year, y = rank, group = name, color = name)
) +
  ggplot2::geom_line(linewidth = 1.2) +
  ggplot2::geom_point(size = 3) +
  ggrepel::geom_text_repel(
    data = dplyr::filter(top_3_earners, fiscal_year == max(fiscal_year)),
    ggplot2::aes(label = name),
    direction = "y",
    hjust = 0,
    nudge_x = 0.3,
    segment.color = "gray70",
    size = 3
  ) +
  ggplot2::scale_y_reverse(breaks = 1:5) +
  ggplot2::scale_x_discrete(expand = ggplot2::expansion(add = c(0.2, 1.5))) +
  ggplot2::labs(
    title = "Top 3 in Total Compensation by Rank Over Time",
    x = NULL,
    y = "Rank"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "none",
    panel.grid.minor = ggplot2::element_blank()
  )
p_top3_bump
```

::: {.content-visible when-format="html"}
```{r}
#| label: top-3-bump-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_plot_btn(
  p_top3_bump,
  filename_base = "schertz_top3_earners_bump_chart",
  label = "Download Top 3 Bump Chart (PNG)"
)
```
:::


------------------------------------------------------------------------


## Appendix E: Department Bump Chart

```{r}
#| label: department-bump

# =============================================================================
# APPENDIX E: DEPARTMENT SUMMARY - BUMP CHART
# =============================================================================

dept_summary <- all_data |>
  dplyr::summarise(
    total_comp = sum(calc_total_compensation, na.rm = TRUE),
    .by = c(fiscal_year, department)
  ) |>
  dplyr::arrange(fiscal_year, dplyr::desc(total_comp)) |>
  dplyr::mutate(rank = dplyr::row_number(), .by = fiscal_year)

# Keep only top N departments for readability (adjust as needed)
top_n_depts <- 8
top_depts <- dept_summary |>
  dplyr::filter(fiscal_year == max(fiscal_year), rank <= top_n_depts) |>
  dplyr::pull(department)

dept_bump <- dept_summary |>
  dplyr::filter(department %in% top_depts)

p_dept_bump <- ggplot2::ggplot(
  dept_bump,
  ggplot2::aes(x = fiscal_year, y = rank, group = department, color = department)
) +
  ggplot2::geom_line(linewidth = 1.2) +
  ggplot2::geom_point(size = 3) +
  ggrepel::geom_text_repel(
    data = dplyr::filter(dept_bump, fiscal_year == max(fiscal_year)),
    ggplot2::aes(label = department),
    direction = "y",
    hjust = 0,
    nudge_x = 0.3,
    segment.color = "gray70",
    size = 3.5,
    fontface = "bold",
    color = "black"
  ) +
  ggplot2::scale_y_reverse(breaks = seq_len(top_n_depts)) +
  ggplot2::scale_x_discrete(expand = ggplot2::expansion(add = c(0.2, 2))) +
  ggplot2::labs(
    title = "Department Rankings by Total Compensation",
    x = NULL,
    y = "Rank"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "none",
    panel.grid.minor = ggplot2::element_blank()
  )
p_dept_bump
```

::: {.content-visible when-format="html"}
```{r}
#| label: department-bump-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_plot_btn(
  p_dept_bump,
  filename_base = "schertz_department_bump_chart",
  label = "Download Department Bump Chart (PNG)"
)
```
:::


------------------------------------------------------------------------


## Appendix F: Overtime Analysis by Fiscal Year

```{r}
#| label: overtime-analysis

# =============================================================================
# APPENDIX F: OVERTIME ANALYSIS
# =============================================================================

overtime_summary <- all_data |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    total_overtime = base::sum(overtime_earnings, na.rm = TRUE),
    total_regular = base::sum(regular_earnings, na.rm = TRUE),
    overtime_pct = (total_overtime / total_regular) * 100,
    employees_with_ot = base::sum(overtime_earnings > 0),
    total_employees = dplyr::n(),
    pct_with_ot = (employees_with_ot / total_employees) * 100,
    .groups = "drop"
  )

overtime_summary |>
  dplyr::mutate(
    total_overtime = scales::dollar(total_overtime, accuracy = 1),
    total_regular = scales::dollar(total_regular, accuracy = 1),
    overtime_pct = base::sprintf("%.1f%%", overtime_pct),
    pct_with_ot = base::sprintf("%.1f%%", pct_with_ot)
  ) |>
  dplyr::rename(
    FY = fiscal_year,
    Overtime = total_overtime,
    'Total Reg Pay' =  total_regular,
    'Pct Reg Pay is OT' = overtime_pct,
    'EEs with OT' = employees_with_ot,
    'Total EEs' = total_employees,
    'Pct EEs Rec OT' = pct_with_ot
    ) |>
  
  flextable::flextable() |>
  flextable::add_header_lines("Overtime Analysis by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: overtime-summary-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_data_btn(
  overtime_summary,
  filename_base = "schertz_comp_overtime_analysis_fy20_fy25",
  label = "Download Overtime Analysis (CSV)"
)
```
:::



# Appendix F2: Overtime by Department - Top 3 

```{r}
#| label: fig-overtime-by-dept
#| fig-cap: "Top 3 Departments by Overtime Usage"

# ======================================================================
# OVERTIME BY DEPARTMENT - Top 3 LINE CHART
# ======================================================================

# STEP 1: Identify the top 3 departments by total overtime across all years
top_3_depts <- all_data |>
  dplyr::summarise(
    total_ot = sum(overtime_earnings, na.rm = TRUE),
    .by = department
  ) |>
  dplyr::slice_max(order_by = total_ot, n = 3) |>
  dplyr::pull(department)

# STEP 2: Summarize overtime by fiscal year for only those Top 3 departments
dept_overtime <- all_data |>
  dplyr::filter(department %in% top_3_depts) |>
  dplyr::summarise(
    overtime_dollars = sum(overtime_earnings, na.rm = TRUE),
    .by = c(fiscal_year, department)
  ) |>
  dplyr::mutate(
    label_text = scales::dollar(overtime_dollars, accuracy = 1, scale = 1e-3, suffix = "K")
  )

# STEP 3: Build the line chart
p_overtime_by_dept <- ggplot2::ggplot(
  dept_overtime, 
  ggplot2::aes(x = fiscal_year, y = overtime_dollars, color = department, group = department)
) +
  ggplot2::geom_line(linewidth = 4) +
  ggplot2::geom_point(size = 6) +
  ggplot2::geom_text(
    ggplot2::aes(label = label_text),
    vjust = -1.2,
    size = 6,
    color = "black",
    show.legend = FALSE
  ) +
  # Add endpoint labels instead of legend
  ggrepel::geom_text_repel(
    data = dplyr::filter(dept_overtime, fiscal_year == max(fiscal_year)),
    ggplot2::aes(label = department),
    direction = "y",
    hjust = 0,
    nudge_x = 0.2,
    segment.color = "gray70",
    size = 6,
    fontface = "bold",
    color = "black"
  ) +
  ggplot2::scale_y_continuous(
    labels = scales::dollar_format(scale = 1e-3, suffix = "K"),
    expand = ggplot2::expansion(mult = c(0.05, 0.15))
  ) +
  ggplot2::scale_x_discrete(expand = ggplot2::expansion(add = c(0.2, 1.5))) +
  ggplot2::labs(
    title = "Overtime Usage by Department",
    subtitle = "Top 3 Departments by Total Overtime Used",
    x = "Fiscal Year",
    y = "Overtime ($)"
  ) +
  ggplot2::theme(
  legend.position = "none",
  axis.text.y = ggplot2::element_blank(),
  axis.ticks.y = ggplot2::element_blank()
)
p_overtime_by_dept
```

::: {.content-visible when-format="html"}
```{r}
#| label: fig-overtime-by-dept-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_plot_btn(
  p_overtime_by_dept,
  filename_base = "schertz_overtime_by_dept",
  label = "Download Overtime by Dept Chart (PNG)"
)
```
:::

------------------------------------------------------------------------


## Appendix G: Department Total Compenstation Ranking Over Time (Bump Chart)

```{r}
#| label: dept-rank-bump-chart
#| fig-width: 13
#| fig-height: 8

# =============================================================================
# APPENDIX G: DEPARTMENT RANKING BUMP CHART
# =============================================================================

# Calculate ranks by year for all departments
dept_ranks_all <- final_df |>
  dplyr::group_by(department, fiscal_year) |>
  dplyr::summarise(
    total_compensation = base::sum(amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::group_by(fiscal_year) |>
  dplyr::mutate(
    rank = dplyr::min_rank(dplyr::desc(total_compensation))
  ) |>
  dplyr::ungroup()

# Get departments that were ever in top 10
top_10_ever <- dept_ranks_all |>
  dplyr::filter(rank <= 10) |>
  dplyr::distinct(department) |>
  dplyr::pull(department)

# Filter to only those departments
dept_ranks_top_10 <- dept_ranks_all |>
  dplyr::filter(department %in% top_10_ever)

okabe_ito <- c(
  "#000000", # black
  "#E69F00", # orange
  "#56B4E9", # sky blue
  "#009E73", # bluish green
  "red", 
  "#0072B2", # blue
  "#D55E00", # vermilion
  "#CC79A7", # reddish purple
  "firebrick",
  "navy",
  "darkgreen",
  "brown",
  "purple"
)

# Create bump chart
p_dept_rank_bump <- ggplot2::ggplot(dept_ranks_top_10, ggplot2::aes(x = fiscal_year, y = rank, 
                                          group = department, color = department)) +
  ggplot2::geom_line(linewidth = 1.5) +
  ggplot2::geom_point(size = 4) +
  ggplot2::geom_text(
    data = dplyr::filter(dept_ranks_top_10, fiscal_year == "FY20"),
    ggplot2::aes(label = department),
    hjust = 1.1,
    size = 3,
    color = "black"
  ) +
  ggplot2::scale_color_manual(values = okabe_ito) +
  ggplot2::scale_y_reverse(breaks = 1:15) +
  ggplot2::scale_x_discrete(expand = ggplot2::expansion(mult = c(0.25, 0.05))) +
  ggplot2::labs(
    title = "Department Total Compensation Rankings Over Time",
    subtitle = "Rank 1 = highest total compensation. Lines show how departments moved in rankings relative to each other.",
    x = "Fiscal Year",
    y = "Rank (1 = Highest)"
  ) +
  ggplot2::theme(legend.position = "none")
p_dept_rank_bump
```

::: {.content-visible when-format="html"}
```{r}
#| label: dept-rank-bump-chart-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_plot_btn(
  p_dept_rank_bump,
  filename_base = "schertz_dept_rank_bump_chart",
  label = "Download Dept Ranking Bump Chart (PNG)",
  width = 13, height = 8
)
```
:::


# Appendix G2:  Total Comp Rankings by Department in Table Format

```{r}
#| label: dept-rank-table

# Show the underlying data
dept_ranks_top_10 |>
  tidyr::pivot_wider(
    id_cols = department,
    names_from = fiscal_year,
    values_from = rank
  ) |>
  dplyr::arrange(FY24) |>
  flextable::flextable() |>
  flextable::add_header_lines("Total Comp Rankings by Dept by Fiscal Year (1 = Highest TOTAL Compensation)") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: dept-rank-table-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
dept_ranks_wide <- dept_ranks_top_10 |>
  tidyr::pivot_wider(id_cols = department, names_from = fiscal_year, values_from = rank)
ds_download_data_btn(
  dept_ranks_wide,
  filename_base = "schertz_comp_dept_rankings_fy20_fy25",
  label = "Download Department Rankings (CSV)"
)
```
:::


------------------------------------------------------------------------


## Appendix H: Line Chart - City Manager Total Compensation Over Time

```{r}
#| label: plot-city-manager-total-comp
#| fig-width: 13
#| fig-height: 8

# =============================================================================
# APPENDIX H: CITY MANAGER COMPENSATION OVER TIME
# =============================================================================

comp_types <- c(
  "Leave Payout", "Regular", "Overtime", "Additional",
  "Deployment", "Arbitration", "Benefits"
)

city_manager_data <- final_df |>
  dplyr::filter(
    job_title == "City Manager",
    earnings_type %in% comp_types
  ) |>
  dplyr::summarise(
    total_compensation = sum(amount, na.rm = TRUE),
    .by = c(fiscal_year, name)
  )

if (nrow(city_manager_data) > 0) {
  
  # Find each person's last fiscal year
  endpoint_labels <- city_manager_data |>
    dplyr::slice_max(order_by = fiscal_year, n = 1, by = name)
  
  p_city_mgr <- ggplot2::ggplot(
    city_manager_data,
    ggplot2::aes(x = fiscal_year, y = total_compensation, group = name, color = name)
  ) +
    ggplot2::geom_line(linewidth = 1) +
    ggplot2::geom_point(size = 6) +
    ggplot2::geom_text(
      ggplot2::aes(label = scales::dollar(total_compensation, accuracy = 1)),
      vjust = -1.5,
      size = 6.0,
      color = "black",
      show.legend = FALSE
    ) +
    ggrepel::geom_text_repel(
      data = endpoint_labels,
      ggplot2::aes(label = name),
      direction = "y",
      hjust = 0,
      nudge_x = 0.2,
      segment.color = "gray70",
      size = 6,
      fontface = "bold",
      color = "black"
    ) +
    ggplot2::scale_y_continuous(
      expand = ggplot2::expansion(mult = c(0, 0.15))
    ) +
    ggplot2::scale_x_discrete(expand = ggplot2::expansion(add = c(0.2, 1.2))) +
    ggplot2::labs(
      title = "City Manager - Total Compensation by Fiscal Year",
      subtitle = "City of Schertz",
      x = "Fiscal Year",
      y = NULL
    ) +
    ggplot2::theme(
      legend.position = "none",
      axis.text.y = ggplot2::element_blank(),
      axis.ticks.y = ggplot2::element_blank(),
      panel.grid.major.y = ggplot2::element_blank()
    )
  print(p_city_mgr)
} else {
  cat("No City Manager records found in the data.\n")
  p_city_mgr <- NULL
}
```

::: {.content-visible when-format="html"}
```{r}
#| label: plot-city-manager-total-comp-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
if (!base::is.null(p_city_mgr)) {
  ds_download_plot_btn(
    p_city_mgr,
    filename_base = "schertz_city_manager_total_comp",
    label = "Download City Manager Comp Chart (PNG)",
    width = 13, height = 8
  )
}
```
:::




------------------------------------------------------------------------


## Appendix I: Total Compensation by Year
```{r}
#| label: plot-total-compensation-by-year
#| fig-width: 10
#| fig-height: 6

# =============================================================================
# APPENDIX I: TOTAL COMPENSATION GROWTH BY FISCAL YEAR - ALL DEPTS
# =============================================================================
p_total_comp_by_year <- ggplot2::ggplot(validation, 
                ggplot2::aes(x = fiscal_year, y = source_total, group = 1)) +
  ggplot2::geom_line(linewidth = 1.5, color = "steelblue") +
  ggplot2::geom_point(size = 4, color = "steelblue") +
  ggplot2::geom_text(
    ggplot2::aes(label = scales::dollar(source_total, scale = 1e-6, suffix = "M")),
    vjust = -2
  ) +
  ggplot2::theme(
    axis.text.y  = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    panel.grid.major.y = ggplot2::element_blank()
  ) +
  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = c(0.05, 0.15))
  ) +
  ggplot2::labs(
    title = "Total Employee Compensation by Fiscal Year - All Depts",
    subtitle = "City of Schertz",
    x = "Fiscal Year",
    y = "Total Compensation"
  )
p_total_comp_by_year
```

::: {.content-visible when-format="html"}
```{r}
#| label: plot-total-compensation-by-year-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_plot_btn(
  p_total_comp_by_year,
  filename_base = "schertz_total_comp_by_year",
  label = "Download Total Comp by Year Chart (PNG)",
  width = 10, height = 6
)
```
:::


------------------------------------------------------------------------


## Appendix J: Total Compensation Year-over-Year Growth Rate Bar Chart


```{r}
#| label: plot-growth-rate
#| fig-width: 10
#| fig-height: 6

# =============================================================================
# APPENDIX J: TOTAL COMP YEAR-OVER-YEAR GROWTH RATE BAR CHART
# =============================================================================

growth_data <- yearly_summary |>
  dplyr::arrange(fiscal_year) |>
  dplyr::mutate(
    prior_total = dplyr::lag(total_compensation),
    growth_rate = (total_compensation - prior_total) / prior_total,
    growth_pct = growth_rate * 100
  )

p_growth_rate <- growth_data |>
  dplyr::filter(!base::is.na(growth_rate)) |>
  ggplot2::ggplot(ggplot2::aes(x = fiscal_year, y = growth_pct)) +
  ggplot2::geom_col(fill = "steelblue") +
  ggplot2::geom_text(
    ggplot2::aes(label = base::paste0(base::round(growth_pct, 1), "%")),
    vjust = -0.5
  ) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed") +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.15))) +
  ggplot2::labs(
    title = "Year-over-Year Total Compensation Growth",
    subtitle = "City of Schertz",
    x = "Fiscal Year",
    y = "Growth Rate (%)"
  )
p_growth_rate
```

::: {.content-visible when-format="html"}
```{r}
#| label: plot-growth-rate-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_plot_btn(
  p_growth_rate,
  filename_base = "schertz_yoy_growth_rate_chart",
  label = "Download YoY Growth Rate Chart (PNG)",
  width = 10, height = 6
)
```
:::


```{r}
#| label: growth-rate-table

# APPENDIX J: YEAR-OVER-YEAR GROWTH RATE TABLE

growth_data |>
  dplyr::mutate(
    change_num = total_compensation - prior_total,
    total_compensation_disp = scales::dollar(total_compensation, accuracy = 1),
    prior_total_disp        = scales::dollar(prior_total, accuracy = 1),
    change_disp             = scales::dollar(change_num, accuracy = 1),
    growth_pct_disp         = sprintf("%.1f%%", growth_pct)
  ) |>
  dplyr::select(
    fiscal_year, total_employees,
    total_compensation = total_compensation_disp,
    prior_total        = prior_total_disp,
    change             = change_disp,
    growth_pct         = growth_pct_disp
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Annual Compensation Growth Details") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()

```

::: {.content-visible when-format="html"}
```{r}
#| label: growth-rate-table-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_data_btn(
  growth_data,
  filename_base = "schertz_comp_annual_growth_fy20_fy25",
  label = "Download Annual Growth Data (CSV)"
)
```
:::




## Appendix K: Growth Rate Comparisons for Each Element of Total Compensation vs Inflation

#| label: growth-rates-comparison-data

# =============================================================================
# APPENDIX K: GROWTH RATES COMPARISON - ALL METRICS VS INFLATION
# =============================================================================



```{r}

# Build comprehensive growth rates dataset
growth_rates_all <- all_data |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    employees = dplyr::n(),
    total_comp = sum(calc_total_compensation, na.rm = TRUE),
    regular_earnings = sum(regular_earnings, na.rm = TRUE),
    additional_benefits = sum(additional_benefits, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::left_join(inflation_rates, by = "fiscal_year") |>
  dplyr::left_join(
    imputation_summary |> dplyr::select(fiscal_year, benefits_imputed),
    by = "fiscal_year"
  ) |>
  dplyr::arrange(fiscal_year) |>
  dplyr::mutate(
    # Calculate year-over-year growth rates
    prior_employees = dplyr::lag(employees),
    prior_total_comp = dplyr::lag(total_comp),
    prior_regular = dplyr::lag(regular_earnings),
    prior_benefits = dplyr::lag(additional_benefits),
    
    # Growth rates as percentages
    employee_growth = (employees - prior_employees) / prior_employees * 100,
    total_comp_growth = (total_comp - prior_total_comp) / prior_total_comp * 100,
    regular_growth = (regular_earnings - prior_regular) / prior_regular * 100,
    benefits_growth = (additional_benefits - prior_benefits) / prior_benefits * 100
  )

# Display the raw growth rates data
growth_rates_all |>
  dplyr::filter(!base::is.na(total_comp_growth)) |>
  dplyr::select(
    fiscal_year, benefits_imputed,
    inflation_rate, total_comp_growth, regular_growth, 
    benefits_growth, employee_growth
  ) |>
  dplyr::mutate(
    benefits_imputed = dplyr::if_else(benefits_imputed, "Yes*", "No"),
    inflation_rate = base::sprintf("%.2f%%", inflation_rate),
    total_comp_growth = base::sprintf("%.2f%%", total_comp_growth),
    regular_growth = base::sprintf("%.2f%%", regular_growth),
    benefits_growth = base::sprintf("%.2f%%", benefits_growth),
    employee_growth = base::sprintf("%.2f%%", employee_growth)
  ) |>
  dplyr::rename(
    FY = fiscal_year,
    `Benefits Imputed?` = benefits_imputed,
    `U.S. Inflation` = inflation_rate,
    `Growth Rate of Total Comp` = total_comp_growth,
    `Growth Rate of Regular Earnings` = regular_growth,
    `Growth Rate of Add'l Benefits` = benefits_growth,
    `Growth Rate of Employees` = employee_growth
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Year-over-Year Comparisons of Growth Rates by Earnings Category") |>
  flextable::add_footer_lines("* Benefits imputed for FY20-FY22 based on FY23-FY25 ratios") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::italic(part = "footer") |>
  flextable::fontsize(size = 11, part = "footer") |>
  flextable::autofit()

```

::: {.content-visible when-format="html"}
```{r}
#| label: growth-rates-all-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_data_btn(
  growth_rates_all,
  filename_base = "schertz_comp_growth_rates_all_fy20_fy25",
  label = "Download Growth Rates Data (CSV)"
)
```
:::


```{r}
#| label: fig-growth-rates-comparison
#| fig-cap: "Growth Rates Comparison: Compensation Metrics vs Published Inflation"
#| fig-width: 13
#| fig-height: 8

# =============================================================================
# LINE CHART: ALL GROWTH RATES COMPARED
# =============================================================================

# Reshape data for plotting
growth_rates_long <- growth_rates_all |>
  dplyr::filter(!base::is.na(total_comp_growth)) |>
  dplyr::select(
    fiscal_year,
    `Inflation (FRED)` = inflation_rate,
    `Total Compensation` = total_comp_growth,
    `Regular Earnings` = regular_growth,
    `Additional Benefits` = benefits_growth,
    `Employee Count` = employee_growth
  ) |>
  tidyr::pivot_longer(
    cols = -fiscal_year,
    names_to = "metric",
    values_to = "growth_rate"
  ) |>
  dplyr::mutate(
    metric = base::factor(
      metric,
      levels = c("Inflation (FRED)", "Total Compensation", "Regular Earnings", 
                 "Additional Benefits", "Employee Count")
    )
  )

# Define colors for each metric
metric_colors <- c(
  "Inflation (FRED)" = "firebrick",
  "Total Compensation" = "steelblue",
  "Regular Earnings" = "darkgreen",
  "Additional Benefits" = "purple",
  "Employee Count" = "darkorange"
)

# Find each metric's FIRST fiscal year for LEFT-side endpoint labels
endpoint_labels <- growth_rates_long |>
  dplyr::group_by(metric) |>
  dplyr::slice_min(order_by = fiscal_year, n = 1) |>
  dplyr::ungroup()

# Create the plot
p_growth_rates_comparison <- ggplot2::ggplot(
  growth_rates_long,
  ggplot2::aes(x = fiscal_year, y = growth_rate, group = metric, color = metric)
) +
  ggplot2::geom_hline(yintercept = 0, linetype = "solid", color = "gray50") +
  ggplot2::geom_line(linewidth = 3.0) +
  ggplot2::geom_point(size = 6) +
  ggplot2::geom_text(
    ggplot2::aes(label = base::sprintf("%.1f%%", growth_rate)),
    vjust = -1.2,
    size = 6,
    show.legend = FALSE
  ) +
  ggrepel::geom_text_repel(
    data = endpoint_labels,
    ggplot2::aes(label = metric),
    direction = "y",
    hjust = 1,
    nudge_x = -0.3,
    segment.color = "gray70",
    size = 6,
    fontface = "bold",
    color = "black"
  ) +
  ggplot2::scale_color_manual(values = metric_colors) +
  ggplot2::scale_y_continuous(
    labels = function(x) base::paste0(x, "%"),
    breaks = base::seq(-10, 30, by = 5)
  ) +
  ggplot2::scale_x_discrete(expand = ggplot2::expansion(add = c(2.0, 0.3))) +
  ggplot2::labs(
    title = "Year-over-Year Growth Rates: Compensation Metrics vs Inflation",
    subtitle = "City of Schertz | FY20-FY22 benefits are imputed estimates",
    x = "Fiscal Year",
    y = "Growth Rate (%)",
    caption = "Inflation data: FRED Series FPCPITOTLZGUSA | FY20 is base year (no growth rate)"
  ) +
  ggplot2::theme(
    legend.position = "none",
    panel.grid.minor = ggplot2::element_blank()
  )
p_growth_rates_comparison
```

::: {.content-visible when-format="html"}
```{r}
#| label: fig-growth-rates-comparison-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_plot_btn(
  p_growth_rates_comparison,
  filename_base = "schertz_growth_rates_comparison_chart",
  label = "Download Growth Rates Comparison Chart (PNG)",
  width = 13, height = 8
)
```
:::


```{r}
#| label: rule-of-70-table

# =============================================================================
# RULE OF 70: DOUBLING TIME BASED ON MOST RECENT GROWTH RATES
# =============================================================================

# Get the most recent year's growth rates AND amounts (FY25)
latest_data <- growth_rates_all |>
  dplyr::filter(fiscal_year == "FY25") |>
  dplyr::select(
    inflation_rate, total_comp_growth, regular_growth, 
    benefits_growth, employee_growth,
    total_comp, regular_earnings, additional_benefits, employees
  )

# Calculate Rule of 70 doubling times
rule_of_70 <- dplyr::tibble(
  Metric = c(
    "Inflation (CPI)",
    "Total Compensation",
    "Regular Earnings",
    "Additional Benefits",
    "Employee Count"
  ),
  growth_rate_num = c(
    latest_data$inflation_rate,
    latest_data$total_comp_growth,
    latest_data$regular_growth,
    latest_data$benefits_growth,
    latest_data$employee_growth
  ),
  current_amount = c(
    NA_real_,
    latest_data$total_comp,
    latest_data$regular_earnings,
    latest_data$additional_benefits,
    latest_data$employees
  )
) |>
  dplyr::mutate(
    doubling_time_num = dplyr::case_when(
      growth_rate_num <= 0 ~ NA_real_,
      TRUE ~ 70 / growth_rate_num
    ),
    doubled_amount = current_amount * 2
  ) |>
  # Sort by doubling time ascending (fastest on top)
  dplyr::arrange(doubling_time_num) |>
  dplyr::mutate(
    `FY25 Growth Rate` = base::sprintf("%.2f%%", growth_rate_num),
    `Doubling Time (Years)` = dplyr::if_else(
      base::is.na(doubling_time_num),
      "N/A (no growth)",
      base::sprintf("%.1f years", doubling_time_num)
    ),
    `FY25 Doubled` = dplyr::case_when(
      Metric == "Inflation (CPI)" ~ "N/A",
      Metric == "Employee Count" ~ base::format(base::as.integer(doubled_amount), big.mark = ","),
      TRUE ~ scales::dollar(doubled_amount, accuracy = 1)
    )
  ) |>
  dplyr::select(Metric, `FY25 Growth Rate`, `Doubling Time (Years)`, `FY25 Doubled`)

rule_of_70 |>
  flextable::flextable() |>
  flextable::add_header_lines("Rule of 70: Doubling Time at Current Growth Rates (FY25)") |>
  flextable::add_footer_lines("Rule of 70: Years to double = 70 ÷ Growth Rate (%) | Sorted by fastest doubling time") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::italic(part = "footer") |>
  flextable::fontsize(size = 11, part = "footer") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: rule-of-70-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_data_btn(
  rule_of_70,
  filename_base = "schertz_comp_rule_of_70_fy25",
  label = "Download Rule of 70 (Current Rates) (CSV)"
)
```
:::

```{r}
#| label: rule-of-70-average-table

# =============================================================================
# RULE OF 70: USING AVERAGE GROWTH RATES (FY21-FY25)
# =============================================================================

# Get FY25 amounts for doubling calculation
latest_amounts <- growth_rates_all |>
  dplyr::filter(fiscal_year == "FY25") |>
  dplyr::select(total_comp, regular_earnings, additional_benefits, employees)

# Calculate average growth rates across all years
avg_rates <- growth_rates_all |>
  dplyr::filter(!base::is.na(total_comp_growth)) |>
  dplyr::summarise(
    inflation_rate = base::mean(inflation_rate, na.rm = TRUE),
    total_comp_growth = base::mean(total_comp_growth, na.rm = TRUE),
    regular_growth = base::mean(regular_growth, na.rm = TRUE),
    benefits_growth = base::mean(benefits_growth, na.rm = TRUE),
    employee_growth = base::mean(employee_growth, na.rm = TRUE)
  )

# Build Rule of 70 table with averages
rule_of_70_avg <- dplyr::tibble(
  Metric = c(
    "Inflation (CPI)",
    "Total Compensation",
    "Regular Earnings",
    "Additional Benefits",
    "Employee Count"
  ),
  growth_rate_num = c(
    avg_rates$inflation_rate,
    avg_rates$total_comp_growth,
    avg_rates$regular_growth,
    avg_rates$benefits_growth,
    avg_rates$employee_growth
  ),
  current_amount = c(
    NA_real_,
    latest_amounts$total_comp,
    latest_amounts$regular_earnings,
    latest_amounts$additional_benefits,
    latest_amounts$employees
  )
) |>
  dplyr::mutate(
    doubling_time_num = dplyr::case_when(
      growth_rate_num <= 0 ~ NA_real_,
      TRUE ~ 70 / growth_rate_num
    ),
    doubled_amount = current_amount * 2
  ) |>
  # Sort by doubling time ascending (fastest on top)
  dplyr::arrange(doubling_time_num) |>
  dplyr::mutate(
    `Avg Growth Rate (FY21-FY25)` = base::sprintf("%.2f%%", growth_rate_num),
    `Doubling Time (Years)` = dplyr::if_else(
      base::is.na(doubling_time_num),
      "N/A (no growth)",
      base::sprintf("%.1f years", doubling_time_num)
    ),
    `FY25 Doubled` = dplyr::case_when(
      Metric == "Inflation (CPI)" ~ "N/A",
      Metric == "Employee Count" ~ base::format(base::as.integer(doubled_amount), big.mark = ","),
      TRUE ~ scales::dollar(doubled_amount, accuracy = 1)
    )
  ) |>
  dplyr::select(Metric, `Avg Growth Rate (FY21-FY25)`, `Doubling Time (Years)`, `FY25 Doubled`)

rule_of_70_avg |>
  flextable::flextable() |>
  flextable::add_header_lines("Rule of 70: Doubling Time at Average Growth Rates (FY21-FY25)") |>
  flextable::add_footer_lines("Rule of 70: Years to double = 70 ÷ Growth Rate (%) | Sorted by fastest doubling time") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::italic(part = "footer") |>
  flextable::fontsize(size = 11, part = "footer") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: rule-of-70-avg-download
#| eval: !expr knitr::pandoc_to() == "html"
#| echo: false
ds_download_data_btn(
  rule_of_70_avg,
  filename_base = "schertz_comp_rule_of_70_avg_fy21_fy25",
  label = "Download Rule of 70 (Avg Rates) (CSV)"
)
```
:::




## How to Add a New Fiscal Year

When you receive a new fiscal year's file:

1.  **Add the CLEANED filename** to the `files` list in Step 1

2.  **Run Phase 1** to see the exact column names:

``` r
new_file <- readxl::read_excel(here::here("data_raw", "NEW_FILE.xlsx"))
base::names(janitor::clean_names(new_file))
```

3.  **Add a new entry to `column_maps`** in Step 4 with the exact column names from the console

4.  **Run the pipeline** - Validation will tell you if any columns are missing

5.  **Update column names** if the spreadsheet structure has changed

This explicit approach means you'll never silently lose data when the spreadsheet structure changes.

------------------------------------------------------------------------

## Summary

| Year    | Structure                              | Notes                    |
|---------|----------------------------------------|--------------------------|
| FY20-22 | Single `name` column, no benefits      | Total = Total Earnings   |
| FY23-24 | Split `last_name`/`first_name`, has benefits | Full Total Compensation |

If validation shows **FAIL**, check the column mappings in Step 4.
