---
title: "City of Schertz Base Salary vs Additional Earnings for Directors, Managers, and Public Safety Leaders-Searchable DT Tables"
subtitle: "FY20-FY25"
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    echo: true
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true     # Set to 'false' for drafts. Override if you're NOT sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    echo: false
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    error: false
    warning: false
    message: false
    cache: false


params:
  data_updated: !expr format(Sys.Date(), "%B %Y")
---

```{r}
#| label: setup
#| include: false

# Capture current filename for use in document
file_name <- if (!base::is.null(knitr::current_input())) {
  paste0(tools::file_path_sans_ext(knitr::current_input()), ".qmd")
} else {
  "interactive_session.qmd"
}

file_stem <- paste0(tools::file_path_sans_ext(file_name), "_chat")



# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED

# Download button helpers live here; knitr::is_html_output() guard is inside
# the helpers so they silently no-op when rendering to typst

source(here::here("R", "download_button_helpers.R"))
source(here::here("R", "report_format_helpers.R"))
source(here::here("R", "flextable_helpers.R"))
source(here::here("R", "download_button_helpers.R"))

# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  theme_fun = base::identity,          # <-- add this
  post_process_html = identity,
  post_process_docx = identity
)

# Standard table styler: blue italic title row + palegreen column header row.
# Defined here so all QMDs that source this setup get the same presentation
# without depending on census_helpers.R for a purely cosmetic function.
ds_style_flextable <- function(ft, title = NULL, title_size = 16) {
  # Fixed layout helps long text wrap instead of expanding columns
  ft <- flextable::set_table_properties(ft, layout = "fixed")

  if (!base::is.null(title)) {
    ft <- flextable::add_header_lines(ft, values = title)
    ft <- flextable::color(ft, i = 1, part = "header", color = "blue")
    ft <- flextable::italic(ft, i = 1, part = "header")
    ft <- flextable::align(ft, i = 1, part = "header", align = "left")
    ft <- flextable::bold(ft, i = 1, part = "header")
    ft <- flextable::fontsize(ft, i = 1, part = "header", size = title_size)
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "white")
    ft <- flextable::bg(ft, i = 2, part = "header", bg = "palegreen")

    # Optional: make wrapped titles read better
    ft <- flextable::line_spacing(ft, i = 1, part = "header", space = 1.2)
  } else {
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "palegreen")
  }

  ft <- flextable::autofit(ft)
  ft
}



# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 28,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 22,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 18,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 18,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)


# claude <- ellmer::chat_anthropic(model = "claude-opus-4-6-20260205")
  
# Set seed for reproducibility
base::set.seed(123)

```

\

#### **Filename:**

::: {.content-visible when-format="html"}
[`r file_name`]{style="word-break: break-all; font-size: 1.00em;"}
:::

```{=typst}
#let wrap_underscores(s) = s.replace("_", "_\u{200B}")
#text(size: 16pt)[#raw(wrap_underscores("`r file_name`"))]
```

\


## Introduction
This analysis breaks down **total compensation** into its component parts for the City of Schertz's top-paid employees:

1. **Base Salary** (Regular Earnings) - Core salary/wages
2. **Additional Earnings** - Overtime, leave payout, deployment, arbitration, etc.
3. **Additional Benefits** - Healthcare, pension contributions, etc.

This breakdown reveals:
- The portion of pay increases go to base salary vs. benefits
- How overtime and other earnings affect total compensation
- The true cost of employment beyond base salary

::: {.callout-note}
## Interactive Tables
**Data last updated:** `r params$data_updated`

- **Sort**: Click any column header to sort
- **Search**: Use the search box to find specific employees  
- **Filter**: Use column filters to narrow results
- **Export**: Use the Copy, CSV, Excel, or Print buttons
:::

---

## Step 1: Load Cached Data
```{r}
#| label: load-cached-data
#| cache: true

# =============================================================================
# LOAD CACHED DATA FROM OMNIBUS ANALYSIS
# =============================================================================

data_path <- here::here("data_clean", "compensation_wide_fy20_fy25.rds")

if (base::file.exists(data_path)) {
  all_data <- base::readRDS(data_path)

  data_load_df <- data.frame(
    Item = c("Data file", "Rows loaded", "Fiscal years found"),
    Value = c(
      data_path,
      base::format(base::nrow(all_data), big.mark = ","),
      base::paste(base::sort(base::unique(all_data$fiscal_year)), collapse = ", ")
    ),
    stringsAsFactors = FALSE
  )

  data_load_spec <- list(
    title = "Loaded Compensation Data Summary",
    labels = list(Item = "Item", Value = "Value")
  )

  ds_ft(data_load_df, data_load_spec)
} else {
  base::stop(
    "Data file not found at: ", data_path, "
",
    "Please run the omnibus analysis first to generate the cleaned data."
  )
}
```

---

## Step 2: Filter to Top Employees
```{r}
#| label: filter-top-employees
#| cache: true

# =============================================================================
# FILTER DATA TO TOP EMPLOYEES ONLY
# =============================================================================

leadership_pattern <- base::paste(
  "city manager", "assistant city manager", "deputy city manager",
  "director", "assistant director", "deputy director",
  "manager", "superintendent",
  "chief", "commander", "captain", "battalion chief", "marshal",
  sep = "|"
)

top_employees_full <- all_data |>
  dplyr::filter(
    stringr::str_detect(
      job_title,
      stringr::regex(leadership_pattern, ignore_case = TRUE)
    )
  ) |>
  dplyr::select(
    fiscal_year,
    name,
    name_std,
    department,
    job_title,
    regular_earnings,
    leave_payout,
    overtime_earnings,
    additional_earnings,
    deployment_earnings,
    arbitration_settlements,
    additional_benefits,
    calc_total_compensation
  ) |>
  dplyr::arrange(fiscal_year, dplyr::desc(calc_total_compensation))

filter_results_df <- data.frame(
  Item = c("Top employees identified", "Unique individuals"),
  Value = c(
    base::format(base::nrow(top_employees_full), big.mark = ","),
    base::format(dplyr::n_distinct(top_employees_full$name_std), big.mark = ",")
  ),
  stringsAsFactors = FALSE
)

filter_results_spec <- list(
  title = "Top Employees Filter Results",
  labels = list(Item = "Item", Value = "Value"),
  align_right = "Value"
)

ds_ft(filter_results_df, filter_results_spec)
```

---

## Step 3: Calculate Component Totals
```{r}
#| label: calculate-components
#| cache: true

# =============================================================================
# CALCULATE COMPENSATION COMPONENTS
# =============================================================================

top_employees_components <- top_employees_full |>
  dplyr::mutate(
    base_salary = regular_earnings,
    add_earnings = dplyr::coalesce(leave_payout, 0) +
                   dplyr::coalesce(overtime_earnings, 0) +
                   dplyr::coalesce(additional_earnings, 0) +
                   dplyr::coalesce(deployment_earnings, 0) +
                   dplyr::coalesce(arbitration_settlements, 0),
    add_benefits = dplyr::coalesce(additional_benefits, 0),
    calc_total = base_salary + add_earnings + add_benefits,
    discrepancy = base::abs(calc_total - calc_total_compensation)
  )

max_discrepancy <- base::max(top_employees_components$discrepancy, na.rm = TRUE)

validation_df <- data.frame(
  Item = c("Maximum discrepancy between components and total", "Validation status"),
  Value = c(
    scales::dollar(max_discrepancy),
    ifelse(max_discrepancy > 1, "Warning: discrepancy exceeds $1", "OK")
  ),
  stringsAsFactors = FALSE
)

validation_spec <- list(
  title = "Compensation Components Validation",
  labels = list(Item = "Item", Value = "Value"),
  color_cells = list(
    list(i = 2, j = "Value", color = ifelse(max_discrepancy > 1, "firebrick", "darkgreen"))
  )
)

ds_ft(validation_df, validation_spec)

if (max_discrepancy > 1) {
  base::warning("Discrepancy exceeds $1 - check component calculations")
}
```

---

## Step 4: Create Wide Tables for Each Component
```{r}
#| label: create-component-tables
#| cache: true

# =============================================================================
# CREATE WIDE FORMAT TABLES FOR EACH COMPONENT
# =============================================================================

most_recent_info <- top_employees_components |>
  dplyr::group_by(name_std) |>
  dplyr::slice_max(order_by = fiscal_year, n = 1) |>
  dplyr::ungroup() |>
  dplyr::select(name_std, name, department, job_title)

total_comp_wide <- top_employees_components |>
  dplyr::select(name_std, fiscal_year, calc_total_compensation) |>
  tidyr::pivot_wider(
    names_from = fiscal_year,
    values_from = calc_total_compensation,
    names_prefix = "Tot_",
    values_fill = NA_real_
  ) |>
  dplyr::left_join(most_recent_info, by = "name_std")

base_salary_wide <- top_employees_components |>
  dplyr::select(name_std, fiscal_year, base_salary) |>
  tidyr::pivot_wider(
    names_from = fiscal_year,
    values_from = base_salary,
    names_prefix = "Base_",
    values_fill = NA_real_
  ) |>
  dplyr::left_join(most_recent_info, by = "name_std")

add_earnings_wide <- top_employees_components |>
  dplyr::select(name_std, fiscal_year, add_earnings) |>
  tidyr::pivot_wider(
    names_from = fiscal_year,
    values_from = add_earnings,
    names_prefix = "AddE_",
    values_fill = NA_real_
  ) |>
  dplyr::left_join(most_recent_info, by = "name_std")

add_benefits_wide <- top_employees_components |>
  dplyr::select(name_std, fiscal_year, add_benefits) |>
  tidyr::pivot_wider(
    names_from = fiscal_year,
    values_from = add_benefits,
    names_prefix = "Ben_",
    values_fill = NA_real_
  ) |>
  dplyr::left_join(most_recent_info, by = "name_std")

table_counts_df <- data.frame(
  Table = c("Total Comp", "Base Salary", "Add Earnings", "Add Benefits"),
  Employees = c(
    base::nrow(total_comp_wide),
    base::nrow(base_salary_wide),
    base::nrow(add_earnings_wide),
    base::nrow(add_benefits_wide)
  ),
  stringsAsFactors = FALSE
)

table_counts_spec <- list(
  title = "Wide Tables Created",
  labels = list(Table = "Table", Employees = "Employees"),
  align_right = "Employees"
)

ds_ft(table_counts_df, table_counts_spec)
```

---

## Step 5: Create Combined Breakdown Table
```{r}
#| label: create-combined-breakdown
#| cache: true

# =============================================================================
# CREATE COMBINED BREAKDOWN TABLE
# =============================================================================

combined_breakdown <- top_employees_components |>
  dplyr::select(
    name_std, name, department, job_title,
    fiscal_year,
    base_salary, add_earnings, add_benefits, calc_total_compensation
  ) |>
  dplyr::rename(
    `Base Salary` = base_salary,
    `Add'l Earnings` = add_earnings,
    `Add'l Benefits` = add_benefits,
    `Total Comp` = calc_total_compensation
  ) |>
  dplyr::arrange(name, fiscal_year)

combined_breakdown_df <- data.frame(
  Item = "Combined breakdown rows",
  Value = base::format(base::nrow(combined_breakdown), big.mark = ","),
  stringsAsFactors = FALSE
)

combined_breakdown_spec <- list(
  title = "Combined Breakdown Table Summary",
  labels = list(Item = "Item", Value = "Value"),
  align_right = "Value"
)

ds_ft(combined_breakdown_df, combined_breakdown_spec)
```

---

## Table 3A: Total Compensation by Fiscal Year

::: {.content-visible when-format="html"}
```{r}
#| label: display-total-comp-dt
#| cache: false

# =============================================================================
# HTML OUTPUT: TOTAL COMPENSATION - INTERACTIVE DT::DATATABLE
# =============================================================================

total_comp_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("Tot_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "Tot_"),
    dplyr::starts_with("Tot_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("FY"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  DT::datatable(
    caption = "Table 3A: Total Compensation by Fiscal Year",
    extensions = "Buttons",
    options = base::list(
      pageLength = 25,
      scrollX = TRUE,
      dom = "Bfrtip",
      buttons = base::list(
        "copy", "csv", "excel",
        base::list(extend = "print", title = "Total Compensation - City of Schertz")
      )
    ),
    rownames = FALSE,
    filter = "top"
  )
```
:::

::: {.content-visible when-format="typst"}
```{r}
#| label: display-total-comp-flextable
#| cache: false

# =============================================================================
# PDF/TYPST OUTPUT: TOTAL COMPENSATION - FLEXTABLE
# =============================================================================

total_comp_display <- total_comp_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("Tot_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "Tot_"),
    dplyr::starts_with("Tot_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(dplyr::starts_with("FY"), ~ scales::dollar(.x, accuracy = 1))
  )

total_comp_spec <- list(
  title = "Table 3A: Total Compensation by Fiscal Year",
  footer_lines = "Total Comp = Base Salary + Additional Earnings + Additional Benefits",
  footer_fontsize = 10,
  footer_italic = TRUE
)

ds_ft(total_comp_display, total_comp_spec)

ds_download_data_btn(
  total_comp_wide,
  filename_base = base::paste0(file_stem, "_schertz_total_comp_by_fy"),
  label = "Download Table CSV"
)
```
:::

---

## Table 3B: Base Salary (Regular Earnings) by Fiscal Year

::: {.content-visible when-format="html"}
```{r}
#| label: display-base-salary-dt
#| cache: false

# =============================================================================
# HTML OUTPUT: BASE SALARY - INTERACTIVE DT::DATATABLE
# =============================================================================

base_salary_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("Base_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "Base_"),
    dplyr::starts_with("Base_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("FY"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  DT::datatable(
    caption = "Table 3B: Base Salary (Regular Earnings) by Fiscal Year",
    extensions = "Buttons",
    options = base::list(
      pageLength = 25,
      scrollX = TRUE,
      dom = "Bfrtip",
      buttons = base::list(
        "copy", "csv", "excel",
        base::list(extend = "print", title = "Base Salary - City of Schertz")
      )
    ),
    rownames = FALSE,
    filter = "top"
  )
```
:::

::: {.content-visible when-format="typst"}
```{r}
#| label: display-base-salary-flextable
#| cache: false

# =============================================================================
# PDF/TYPST OUTPUT: BASE SALARY - FLEXTABLE
# =============================================================================

base_salary_display <- base_salary_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("Base_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "Base_"),
    dplyr::starts_with("Base_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(dplyr::starts_with("FY"), ~ scales::dollar(.x, accuracy = 1))
  )

base_salary_spec <- list(
  title = "Table 3B: Base Salary (Regular Earnings) by Fiscal Year",
  footer_lines = "Base Salary = Regular Earnings only (excludes OT, bonuses, benefits)",
  footer_fontsize = 10,
  footer_italic = TRUE
)

ds_ft(base_salary_display, base_salary_spec)

ds_download_data_btn(
  base_salary_wide,
  filename_base = base::paste0(file_stem, "_schertz_base_salary_by_fy"),
  label = "Download Table CSV"
)
```
:::

---

## Table 3C: Additional Earnings by Fiscal Year

::: {.content-visible when-format="html"}
```{r}
#| label: display-add-earnings-dt
#| cache: false

# =============================================================================
# HTML OUTPUT: ADDITIONAL EARNINGS - INTERACTIVE DT::DATATABLE
# =============================================================================

add_earnings_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("AddE_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "AddE_"),
    dplyr::starts_with("AddE_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("FY"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  DT::datatable(
    caption = "Table 3C: Additional Earnings (OT, Leave, Deployment, etc.) by Fiscal Year",
    extensions = "Buttons",
    options = base::list(
      pageLength = 25,
      scrollX = TRUE,
      dom = "Bfrtip",
      buttons = base::list(
        "copy", "csv", "excel",
        base::list(extend = "print", title = "Additional Earnings - City of Schertz")
      )
    ),
    rownames = FALSE,
    filter = "top"
  )
```
:::

::: {.content-visible when-format="typst"}
```{r}
#| label: display-add-earnings-flextable
#| cache: false

# =============================================================================
# PDF/TYPST OUTPUT: ADDITIONAL EARNINGS - FLEXTABLE
# =============================================================================

add_earnings_display <- add_earnings_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("AddE_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "AddE_"),
    dplyr::starts_with("AddE_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(dplyr::starts_with("FY"), ~ scales::dollar(.x, accuracy = 1))
  )

add_earnings_spec <- list(
  title = "Table 3C: Additional Earnings by Fiscal Year",
  footer_lines = "Additional Earnings = Leave Payout + Overtime + Additional + Deployment + Arbitration",
  footer_fontsize = 10,
  footer_italic = TRUE
)

ds_ft(add_earnings_display, add_earnings_spec)

ds_download_data_btn(
  add_earnings_wide,
  filename_base = base::paste0(file_stem, "_schertz_add_earnings_by_fy"),
  label = "Download Table CSV"
)
```
:::

---

## Table 3D: Additional Benefits by Fiscal Year

::: {.content-visible when-format="html"}
```{r}
#| label: display-add-benefits-dt
#| cache: false

# =============================================================================
# HTML OUTPUT: ADDITIONAL BENEFITS - INTERACTIVE DT::DATATABLE
# =============================================================================

add_benefits_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("Ben_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "Ben_"),
    dplyr::starts_with("Ben_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("FY"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  DT::datatable(
    caption = "Table 3D: Additional Benefits (Healthcare, Pension, etc.) by Fiscal Year",
    extensions = "Buttons",
    options = base::list(
      pageLength = 25,
      scrollX = TRUE,
      dom = "Bfrtip",
      buttons = base::list(
        "copy", "csv", "excel",
        base::list(extend = "print", title = "Additional Benefits - City of Schertz")
      )
    ),
    rownames = FALSE,
    filter = "top"
  )
```
:::

::: {.content-visible when-format="typst"}
```{r}
#| label: display-add-benefits-flextable
#| cache: false

# =============================================================================
# PDF/TYPST OUTPUT: ADDITIONAL BENEFITS - FLEXTABLE
# =============================================================================

add_benefits_display <- add_benefits_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("Ben_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "Ben_"),
    dplyr::starts_with("Ben_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(dplyr::starts_with("FY"), ~ scales::dollar(.x, accuracy = 1))
  )

add_benefits_spec <- list(
  title = "Table 3D: Additional Benefits by Fiscal Year",
  footer_lines = "Additional Benefits = Healthcare, pension contributions, etc. | FY20-FY22 benefits are imputed",
  footer_fontsize = 10,
  footer_italic = TRUE
)

ds_ft(add_benefits_display, add_benefits_spec)

ds_download_data_btn(
  add_benefits_wide,
  filename_base = base::paste0(file_stem, "_schertz_add_benefits_by_fy"),
  label = "Download Table CSV"
)
```
:::

---

## Earnings Elements Summary by Fiscal Year

```{r}
#| label: component-summary
#| cache: true

# =============================================================================
# AGGREGATE COMPONENT SUMMARY BY FISCAL YEAR
# =============================================================================

component_summary <- top_employees_components |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    n_employees = dplyr::n(),
    total_base = base::sum(base_salary, na.rm = TRUE),
    total_add_earnings = base::sum(add_earnings, na.rm = TRUE),
    total_benefits = base::sum(add_benefits, na.rm = TRUE),
    total_comp = base::sum(calc_total_compensation, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    pct_base = total_base / total_comp * 100,
    pct_add_earnings = total_add_earnings / total_comp * 100,
    pct_benefits = total_benefits / total_comp * 100
  )

component_summary_display <- component_summary |>
  dplyr::mutate(
    total_base = scales::dollar(total_base, accuracy = 1),
    total_add_earnings = scales::dollar(total_add_earnings, accuracy = 1),
    total_benefits = scales::dollar(total_benefits, accuracy = 1),
    total_comp = scales::dollar(total_comp, accuracy = 1),
    pct_base = base::sprintf("%.1f%%", pct_base),
    pct_add_earnings = base::sprintf("%.1f%%", pct_add_earnings),
    pct_benefits = base::sprintf("%.1f%%", pct_benefits)
  ) |>
  dplyr::select(
    fiscal_year, n_employees,
    total_base, pct_base,
    total_add_earnings, pct_add_earnings,
    total_benefits, pct_benefits,
    total_comp
  ) |>
  dplyr::rename(
    FY = fiscal_year,
    N = n_employees,
    `Base Salary` = total_base,
    `% Base` = pct_base,
    `Add'l Earnings` = total_add_earnings,
    `% Add'l` = pct_add_earnings,
    Benefits = total_benefits,
    `% Ben` = pct_benefits,
    `Total Comp` = total_comp
  )

component_summary_spec <- list(
  title = "Earning Elements for All Top Employees by Fiscal Year with Percentages",
  footer_lines = "Percentages show each component's share of total compensation",
  footer_fontsize = 10,
  footer_italic = TRUE
)

ds_ft(component_summary_display, component_summary_spec)

ds_download_data_btn(
  component_summary,
  filename_base = base::paste0(file_stem, "_schertz_component_summary_by_fy"),
  label = "Download Table CSV"
)
```

---

## Object Reference

| Object Name | Description | Rows | Columns |
|-------------|-------------|------|---------|
| `all_data` | Full compensation dataset (cached) | `r base::nrow(all_data)` | `r base::ncol(all_data)` |
| `top_employees_full` | Filtered with all earnings columns | `r base::nrow(top_employees_full)` | `r base::ncol(top_employees_full)` |
| `top_employees_components` | With calculated component totals | `r base::nrow(top_employees_components)` | `r base::ncol(top_employees_components)` |
| `total_comp_wide` | Wide format: Total comp by FY | `r base::nrow(total_comp_wide)` | `r base::ncol(total_comp_wide)` |
| `base_salary_wide` | Wide format: Base salary by FY | `r base::nrow(base_salary_wide)` | `r base::ncol(base_salary_wide)` |
| `add_earnings_wide` | Wide format: Add'l earnings by FY | `r base::nrow(add_earnings_wide)` | `r base::ncol(add_earnings_wide)` |
| `add_benefits_wide` | Wide format: Benefits by FY | `r base::nrow(add_benefits_wide)` | `r base::ncol(add_benefits_wide)` |
| `combined_breakdown` | Long format: All components | `r base::nrow(combined_breakdown)` | `r base::ncol(combined_breakdown)` |
| `component_summary` | Aggregate by FY | `r base::nrow(component_summary)` | `r base::ncol(component_summary)` |

---

## Data Notes

| Component | Includes |
|-----------|----------|
| **Base Salary** | Regular Earnings only |
| **Add'l Earnings** | Leave Payout + Overtime + Additional + Deployment + Arbitration |
| **Add'l Benefits** | Healthcare, pension, other employer contributions |
| **Total Comp** | Base + Add'l Earnings + Add'l Benefits |

- **FY20-FY22**: Benefits were imputed based on FY23-FY25 ratios (not in source files)
- **Blank cells**: Employee not in position during that fiscal year

---

*Report generated: `r base::format(base::Sys.time(), "%B %d, %Y at %I:%M %p")`*
*Data source: City of Schertz Employee Compensation Records*
