---
title: "Test New Functions"
format: html
---

```{r}
#| label: load-libraries-and-source

library(tidycensus)
library(dplyr)
library(ggplot2)
library(purrr)

# All R files at once
purrr::walk(list.files("R", full.names = TRUE, pattern = "\\.R$"), source)

# # One at a time
# source("R/get_city_pep.R")
# source("R/plot_city_pop.R")

```

##✔️ get_city_pep.R is a file

It’s just a container / script

It does not define a function named get_city_pep()

##✔️ Inside that 'library' file are two separate functions

get_city_pep_2010_2024()

get_city_pep_2010_2024_with_change()


## A Table

| Function name                          | Returns                                         | Use case                                   |
| -------------------------------------- | ----------------------------------------------- | ------------------------------------------ |
| `get_city_pep_2010_2024()`             | raw population values                           | simplest population series                 |
| `get_city_pep_2010_2024_with_change()` | population **plus** lag, change, percent change | plotting comparisons & rate-of-change work |


| object returned | what it represents                   |
| --------------- | ------------------------------------ |
| tibble          | year-by-year population for a city   |
| tibble          | same, with change & pct change added |


## Fill in the city and state arguments

```{r}
df <- get_city_pep_2010_2024_with_change("Schertz","TX")

```



| object returned | what it represents       |
| --------------- | ------------------------ |
| ggplot object   | population over time     |
| ggplot object   | percent change over time |




## Plot the results of get_city_pep_2010_2024_with_change()

Typical usage:


```{r}
plot_city_pop("Schertz","TX", type="pct_change")

```


