---
title: "Sensitivity Analysis: PSPC Compensation Study Comparator Group"
subtitle: "City of Schertz, Texas — What Happens When You Remove the Giants?"
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true     # Set to 'false' for drafts. Override if you're NOT sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false

prefer-html: true

params:
  data_updated: !expr format(Sys.Date(), "%B %Y")
---


```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED

# Download button helpers live here; knitr::is_html_output() guard is inside
# the helpers so they silently no-op when rendering to typst
source(here::here("R", "download_button_helpers.R"))

# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)

# Standard table styler: blue italic title row + palegreen column header row.
# Defined here so all QMDs that source this setup get the same presentation
# without depending on census_helpers.R for a purely cosmetic function.
ds_style_flextable <- function(ft, title = NULL) {
  if (!base::is.null(title)) {
    ft <- flextable::add_header_lines(ft, values = title)
    ft <- flextable::color(ft, i = 1, part = "header", color = "blue")
    ft <- flextable::italic(ft, i = 1, part = "header")
    ft <- flextable::align(ft, i = 1, part = "header", align = "left")
    ft <- flextable::fontsize(ft, i = 1, part = "header", size = 12)
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "white")
    ft <- flextable::bg(ft, i = 2, part = "header", bg = "palegreen")
  } else {
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "palegreen")
  }
  ft <- flextable::autofit(ft)
  ft
}


# Flextable built-in themes:
  # flextable::theme_alafoli()	|>  # BLAH
  # flextable::theme_apa()  # THIS IS NICE
  # flextable::theme_booktabs() |>  # NICE, MORE COMPACT
  # flextable::theme_box() |>   # OK, INCLUDES CELL BORDERS
  # flextable::theme_tron() |>  # 'DARK MODE' BLUE TEXT
  # flextable::theme_tron_legacy() |>   # 'DARK MODE' YELLOW TEXT
  # flextable::theme_vader() |>    # 'DARK MODE' WHITE TEXT
  # flextable::theme_vanilla() |>   # NOT SPECIAL
  # flextable::theme_zebra()	|>
  #


# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 28,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 22,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 18,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 18,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)


claude <- ellmer::chat_anthropic(model = "claude-opus-4-6-20260205")
  
# Set seed for reproducibility
base::set.seed(123)

```

\

#### **Filename:** 
2023_PSPC_Rough_Sensitivity_Analysis_ver2.qmd

\


## Purpose and Limitations

This document presents a **sensitivity analysis** — not a recalculation — of the December 2023 PSPC Classification and Compensation Study for the City of Schertz.

**Why a sensitivity analysis rather than a recalculation?** The PSPC study reports a single "Survey Midpoint" per position, which is an already-computed average across all comparator cities. 

The underlying city-by-city data are not disclosed. We cannot algebraically remove Austin and San Antonio from the average because we do not have
the individual data points. What we *can* do is model plausible scenarios for how the survey midpoints would change if large-city comparators were excluded, anchored in published research on the city-size wage premium.

**Every assumption is stated explicitly.** The reader can accept, reject, or modify any assumption and trace its impact through the analysis. This is designed to be transparent and defensible as an illustration of comparator sensitivity — not as a definitive restatement of market rates.  


## The City-Size Wage Premium: Research Basis

The academic literature on the "urban wage premium" — the empirical finding that larger cities pay higher wages — provides the foundation for the assumptions.

```{r}
#| label: research-basis

research <- dplyr::tibble(
  Source = c(
    "Baum-Snow & Pavan (2012)\nReview of Economic Studies",
    "Yankow (2006)\nJournal of Urban Economics",
    "Glaeser & Maré (2001)\nJournal of Labor Economics",
    "NLC City Fiscal Conditions (2024)"
  ),
  Finding = c(
    "Nominal wage premium of 19% for medium cities and 29% for large cities over small areas. After controlling for individual fixed effects: 7% (medium) and 15% (large).",
    "OLS identifies 19% wage premium for large urban areas. Fixed-effects estimation reduces this to ~5-6% after controlling for unobserved worker heterogeneity.",
    "25% raw wage advantage for large-city workers. Fixed-effects estimation reduces premium to 4.5-11% depending on sample.",
    "Cities 10K-50K population: 25% reported wage increases in FY2024. Cities over 300K: only 5% reported increases. Small cities facing disproportionate wage pressure."
  ),
  Premium_Range = c(
    "7% to 29%",
    "5% to 19%",
    "4.5% to 25%",
    "N/A (growth rates)"
  )
)

ft_research <- flextable::flextable(research) |>
  flextable::add_header_lines(
    values = "Published Research on City-Size Wage Premiums"
  ) |>
  flextable::color(i = 1, part = "header", color = "steelblue") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, part = "header", align = "left") |>
  flextable::fontsize(i = 1, part = "header", size = 12) |>
  flextable::bg(i = 1, part = "header", bg = "white") |>
  flextable::bg(i = 2, part = "header", bg = "palegreen") |>
  flextable::set_table_properties(layout = "autofit") |>
  flextable::width(j = 1, width = 1.8) |>
  flextable::width(j = 2, width = 3.5) |>
  flextable::width(j = 3, width = 1.2)

ft_research
```

**Key takeaway:** The research consistently finds that large cities (populations over 500K) pay a wage premium of roughly **5% to 20%** over smaller areas, depending on methodology and controls. For *municipal government* positions specifically, the premium is likely at the **lower end** of this range because public sector pay is less responsive to local labor market conditions than private sector pay (hopefully, constrained by budgets, pay scales, and political processes).  

**Our modeling assumptions:** We test three scenarios for the degree to which Austin (pop. ~980K) and San Antonio (pop. ~1.5M) inflate the PSPC survey midpoints relative to what the remaining comparators alone would produce:

- **Conservative:** 8% premium (lower bound of fixed-effects estimates)  

- **Moderate:** 12% premium (midpoint of research range for public sector)  
- **Aggressive:** 18% premium (upper range, reflecting full nominal differential)  


## Comparator Group Composition

```{r}
#| label: comparator-composition

comparators <- dplyr::tibble(
  City = c("San Antonio", "Austin", "Round Rock", "Georgetown",
           "San Marcos", "Kyle", "New Braunfels", "Schertz",
           "Cibolo", "Seguin", "Universal City", "Buda", "Live Oak"),
  Population = c(1527000, 980000, 130000, 96000,
                 72000, 63000, 56000, 43000,
                 34000, 30000, 21000, 16000, 16000),
  Ratio_to_Schertz = c(35.5, 22.8, 3.0, 2.2,
                        1.7, 1.5, 1.3, 1.0,
                        0.8, 0.7, 0.5, 0.4, 0.4),
  Category = c("Excluded (36x Schertz)", "Excluded (23x Schertz)",
               rep("Retained — Core Comparator", 5),
               "Schertz (Reference City)",
               rep("Retained — Core Comparator", 5)),
  In_Study = c("Core", "Core", "Core", "Core",
               "Core", "Core", "Core", "Reference",
               "Core", "Core", "Core", "Core", "Core")
)

# Reorder by population descending
comparators <- comparators |>
  dplyr::arrange(dplyr::desc(Population))

ft_comp <- flextable::flextable(
  comparators |> dplyr::select(-In_Study)
) |>
  flextable::add_header_lines(
    values = "PSPC Core Comparator Cities with Population Context"
  ) |>
  flextable::color(i = 1, part = "header", color = "steelblue") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, part = "header", align = "left") |>
  flextable::fontsize(i = 1, part = "header", size = 12) |>
  flextable::bg(i = 1, part = "header", bg = "white") |>
  flextable::bg(i = 2, part = "header", bg = "palegreen") |>
  flextable::colformat_num(j = "Population", big.mark = ",", digits = 0) |>
  flextable::colformat_num(j = "Ratio_to_Schertz", digits = 1) |>
  flextable::set_table_properties(layout = "autofit")

# Highlight the excluded rows (San Antonio and Austin are rows 1 and 2)
ft_comp <- ft_comp |>
  flextable::bg(i = 1:2, part = "body", bg = "#FFE0E0") |>
  flextable::bg(
    i = which(comparators$City == "Schertz"),
    part = "body",
    bg = "#E0E0FF"
  )

ft_comp
```

```{r}
#| label: fig-population-comparison
#| fig-cap: "Population of PSPC Core Comparator Cities Relative to Schertz"
#| fig-height: 5
#| fig-width: 8

plot_data <- comparators |>
  dplyr::filter(City != "Schertz") |>
  dplyr::mutate(
    Excluded = dplyr::if_else(
      City %in% c("Austin", "San Antonio"),
      "Excluded from Sensitivity Analysis",
      "Retained in Sensitivity Analysis"
    ),
    City = forcats::fct_reorder(City, Population)
  )

ggplot2::ggplot(
  plot_data,
  ggplot2::aes(x = City, y = Population / 1000, fill = Excluded)
) +
  ggplot2::geom_col(width = 0.7) +
  ggplot2::geom_hline(
    yintercept = 43,
    linetype = "dashed",
    color = "steelblue",
    linewidth = 0.8
  ) +
  ggplot2::annotate(
    "text", x = 2, y = 65,
    label = "Schertz (~43K)",
    color = "steelblue",
    fontface = "bold",
    size = 3.5
  ) +
  ggplot2::coord_flip() +
  ggplot2::scale_fill_manual(
    values = c(
      "Excluded from Sensitivity Analysis" = "#D9534F",
      "Retained in Sensitivity Analysis" = "#5CB85C"
    )
  ) +
  ggplot2::scale_y_continuous(
    labels = scales::label_comma(suffix = "K")
  ) +
  ggplot2::labs(
    title = "The Comparator Group Problem in One Chart",
    subtitle = "San Antonio and Austin are 23x and 36x the size of Schertz",
    x = NULL,
    y = "Population (thousands)",
    fill = NULL
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    legend.position = "bottom",
    plot.title = ggplot2::element_text(face = "bold"),
    panel.grid.major.y = ggplot2::element_blank()
  )
```

**Note on "Additional Sources":** The PSPC study also included county governments (Bexar County, Williamson County, etc.), utility districts, and ESDs. These entities have fundamentally different tax bases and compensation structures from municipal governments. 

We cannot model their exclusion because we do not know which positions they contributed to. Their inclusion likely introduces additional upward bias for certain positions.  


## Sensitivity Analysis: Methodology

### The Mathematical Problem

Suppose the PSPC study surveyed $n$ comparator employers for a given position and reported the average midpoint as $\bar{M}$. If Austin and San Antonio are two of those $n$ employers, and their midpoints are higher than the remaining $(n-2)$ employers by some premium $p$, then the reported average is:

$$\bar{M} = \frac{(n-2) \cdot \bar{M}_{small} + 2 \cdot \bar{M}_{small}(1+p)}{n} = \bar{M}_{small}\left(1 + \frac{2p}{n}\right)$$

Solving for the smaller-city average:

$$\bar{M}_{small} = \frac{\bar{M}}{1 + \frac{2p}{n}}$$

We do not know $n$ (how many cities matched each position) or $p$ (the exact premium). But we can model the sensitivity across plausible values.  


### Assumptions

```{r}
#| label: assumptions-table

assumptions <- dplyr::tibble(
  Parameter = c(
    "Number of comparators per position (n)",
    "Large-city premium: Conservative",
    "Large-city premium: Moderate",
    "Large-city premium: Aggressive",
    "Positions where Austin/SA contributed data"
  ),
  Assumption = c(
    "6 to 10 (central estimate: 8)",
    "8% above smaller-city average",
    "12% above smaller-city average",
    "18% above smaller-city average",
    "All 65 civilian benchmark positions"
  ),
  Basis = c(
    "12 core comparator cities minus Schertz, plus selected additional sources; not all cities match all positions",
    "Lower bound of fixed-effects estimates in literature (Yankow 2006)",
    "Midpoint of research range; reflects mixed public/private sector benchmark",
    "Upper range nominal premium (Baum-Snow & Pavan 2012); likely overstates public sector differential",
    "Conservative assumption; actual coverage unknown because PSPC does not disclose position-level match counts"
  )
)

ft_assumptions <- flextable::flextable(assumptions) |>
  flextable::add_header_lines(
    values = "Sensitivity Analysis Assumptions"
  ) |>
  flextable::color(i = 1, part = "header", color = "steelblue") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, part = "header", align = "left") |>
  flextable::fontsize(i = 1, part = "header", size = 12) |>
  flextable::bg(i = 1, part = "header", bg = "white") |>
  flextable::bg(i = 2, part = "header", bg = "palegreen") |>
  flextable::set_table_properties(layout = "autofit") |>
  flextable::width(j = 1, width = 2.0) |>
  flextable::width(j = 2, width = 1.8) |>
  flextable::width(j = 3, width = 2.7)

ft_assumptions
```



### Critical Caveat

This analysis **assumes equal weighting** of comparators. If PSPC weighted by city size (which some firms do), the impact of removing Austin and San Antonio would be **larger** than modeled here. 

If PSPC used median rather than mean, the impact would be **smaller**. We do not know which method was used because the study does not disclose it.  


## Results

```{r}
#| label: build-position-data

# ── PSPC study data: all 65 civilian benchmark positions ──
positions <- dplyr::tibble(
  Position = c(
    "Building Inspector", "Computer Systems Support Engineer",
    "Financial Analyst", "Civic Center Manager",
    "Director of Public Affairs", "Meter Technician",
    "Director of Parks, Rec & Community Svcs", "Police Chief",
    "Risk and Safety Specialist", "Fire Chief",
    "Director of Library", "City Engineer",
    "Human Resources Generalist", "Director of Finance",
    "Director of Human Resources", "Recreation Manager",
    "Director of Information Technology", "Deputy City Secretary",
    "Assistant City Manager", "Director of Public Works",
    "Utility Billing Clerk", "Senior Systems Administrator",
    "Public Safety Communications Mgr", "Court Administrator",
    "Permit Technician", "Parks Manager",
    "Electrician (Journeyman)", "Traffic Signal & Sign Tech",
    "Utility Billing Manager", "Code Enforcement Officer",
    "Communications Officer", "Street Worker I",
    "Animal Services Officer", "Drainage Worker I",
    "Business Engagement Manager", "Exec Dir of Economic Development",
    "Streets and Drainage Manager", "SCADA Technician",
    "Purchasing Manager", "Engineer - P.E.",
    "Parks Maintenance Technician", "Dir of Planning & Community Dev",
    "Accounts Payable Coordinator", "IT Support Specialist",
    "Customer Relations Representative", "EMS Billing Specialist",
    "Administrative Assistant", "Plumber (Journeyman)",
    "Planning Manager", "Library Assistant",
    "Engineering Inspector", "Accountant",
    "Librarian", "HVAC Technician",
    "Service Worker I", "Police Records Specialist",
    "EMS Chief", "Deputy Court Clerk",
    "Emergency Management Coordinator", "Planner",
    "GIS Specialist", "Facilities Manager",
    "Water/Wastewater Manager", "Fleet Mechanic",
    "Civilian Evidence Technician"
  ),
  Schertz_Midpoint = c(
    52834, 64219, 58248, 64219, 109835, 43466,
    121093, 133506, 55475, 133506, 109835, 133506,
    55475, 133506, 127148, 64219, 133506, 58248,
    162276, 133506, 41240, 78058, 67429, 78058,
    41240, 74342, 58248, 45638, 67429, 50318,
    50318, 41240, 43466, 41240, 70802, 133506,
    86059, 81961, 81961, 94881, 41240, 133506,
    47922, 55475, 41240, 45638, 43466, 58248,
    90363, 41240, 61161, 64219, 61161, 58248,
    48487, 43466, 133506, 45638, 94881, 64219,
    58248, 90363, 104605, 55475, 64219
  ),
  PSPC_Survey_Midpoint = c(
    69374, 81039, 72998, 80024, 136613, 53716,
    149363, 163880, 68021, 163302, 134346, 162702,
    66778, 160217, 152408, 76725, 159246, 69375,
    192408, 157680, 48055, 90663, 78245, 90264,
    47510, 85202, 66495, 51910, 76601, 56925,
    56822, 46108, 48567, 45900, 78762, 147772,
    94885, 90122, 89991, 104088, 44948, 145319,
    51935, 60118, 44415, 49005, 46607, 62446,
    96734, 43660, 64246, 66879, 63054, 59735,
    49659, 44138, 135486, 46245, 95440, 63938,
    57948, 89748, 103525, 51972, 50937
  )
)

# ── Calculate original variance ──
positions <- positions |>
  dplyr::mutate(
    Original_Variance_Pct = (Schertz_Midpoint - PSPC_Survey_Midpoint) /
      PSPC_Survey_Midpoint * 100
  )

# ── Sensitivity scenarios ──
# For each scenario, compute adjusted survey midpoint after removing
# large-city premium effect.
# M_small = M_reported / (1 + 2*p/n)
# We use n = 8 as central estimate

n_comparators <- 8
premiums <- c(Conservative = 0.08, Moderate = 0.12, Aggressive = 0.18)

for (scenario in names(premiums)) {
  p <- premiums[[scenario]]
  adjustment_factor <- 1 + (2 * p / n_comparators)
  col_midpoint <- paste0("Adj_Midpoint_", scenario)
  col_variance <- paste0("Adj_Variance_", scenario)

  positions[[col_midpoint]] <- positions$PSPC_Survey_Midpoint / adjustment_factor
  positions[[col_variance]] <- (positions$Schertz_Midpoint -
    positions[[col_midpoint]]) / positions[[col_midpoint]] * 100
}
```


### Adjusted Market Midpoints: Sample Positions

```{r}
#| label: tbl-sample-positions

# ── Show a representative sample of positions ──
sample_positions <- c(
  "Building Inspector", "Police Chief", "Fire Chief",
  "Director of Finance", "Assistant City Manager",
  "Director of Public Works", "Human Resources Generalist",
  "Parks Manager", "Electrician (Journeyman)",
  "Administrative Assistant", "Street Worker I",
  "Accountant", "Fleet Mechanic", "Planner"
)

sample_data <- positions |>
  dplyr::filter(Position %in% sample_positions) |>
  dplyr::arrange(Original_Variance_Pct) |>
  dplyr::select(
    Position,
    Schertz_Midpoint,
    PSPC_Survey_Midpoint,
    Original_Variance_Pct,
    Adj_Midpoint_Conservative,
    Adj_Variance_Conservative,
    Adj_Midpoint_Moderate,
    Adj_Variance_Moderate,
    Adj_Midpoint_Aggressive,
    Adj_Variance_Aggressive
  )

display_sample <- sample_data |>
  dplyr::transmute(
    Position,
    `Schertz\nMidpoint` = scales::dollar(Schertz_Midpoint, accuracy = 1),
    `PSPC Survey\nMidpoint` = scales::dollar(PSPC_Survey_Midpoint, accuracy = 1),
    `Original\nGap` = sprintf("%.1f%%", Original_Variance_Pct),
    `8% Adj.\nMidpoint` = scales::dollar(Adj_Midpoint_Conservative, accuracy = 1),
    `8% Adj.\nGap` = sprintf("%.1f%%", Adj_Variance_Conservative),
    `12% Adj.\nMidpoint` = scales::dollar(Adj_Midpoint_Moderate, accuracy = 1),
    `12% Adj.\nGap` = sprintf("%.1f%%", Adj_Variance_Moderate),
    `18% Adj.\nMidpoint` = scales::dollar(Adj_Midpoint_Aggressive, accuracy = 1),
    `18% Adj.\nGap` = sprintf("%.1f%%", Adj_Variance_Aggressive)
  )

ft_sample <- flextable::flextable(display_sample) |>
  flextable::add_header_lines(
    values = "Selected Positions: Original vs. Adjusted Market Gaps"
  ) |>
  flextable::color(i = 1, part = "header", color = "steelblue") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, part = "header", align = "left") |>
  flextable::fontsize(i = 1, part = "header", size = 11) |>
  flextable::bg(i = 1, part = "header", bg = "white") |>
  flextable::bg(i = 2, part = "header", bg = "palegreen") |>
  flextable::fontsize(part = "body", size = 8) |>
  flextable::fontsize(i = 2, part = "header", size = 8) |>
  flextable::set_table_properties(layout = "autofit") |>
  flextable::align(j = 2:10, align = "right", part = "body")

ft_sample
```


### Distribution Shift: How the Gap Changes  


```{r}
#| label: fig-distribution-shift
#| fig-cap: "Distribution of Market Gaps Under Different Comparator Assumptions"
#| fig-height: 6
#| fig-width: 9

# ── Reshape for plotting ──
gap_data <- positions |>
  dplyr::select(
    Position,
    `PSPC Original` = Original_Variance_Pct,
    `Conservative (8%)` = Adj_Variance_Conservative,
    `Moderate (12%)` = Adj_Variance_Moderate,
    `Aggressive (18%)` = Adj_Variance_Aggressive
  ) |>
  tidyr::pivot_longer(
    cols = -Position,
    names_to = "Scenario",
    values_to = "Gap_Pct"
  ) |>
  dplyr::mutate(
    Scenario = factor(Scenario, levels = c(
      "PSPC Original",
      "Conservative (8%)",
      "Moderate (12%)",
      "Aggressive (18%)"
    ))
  )

ggplot2::ggplot(gap_data, ggplot2::aes(x = Gap_Pct, fill = Scenario)) +
  ggplot2::geom_density(alpha = 0.35, linewidth = 0.6) +
  ggplot2::geom_vline(xintercept = -5, linetype = "dashed", color = "gray40") +
  ggplot2::geom_vline(xintercept = 5, linetype = "dashed", color = "gray40") +
  ggplot2::geom_vline(xintercept = 0, linetype = "solid", color = "black",
                      linewidth = 0.5) +
  ggplot2::annotate(
    "text", x = 0, y = Inf, label = "At Market",
    vjust = 2, fontface = "bold", size = 3
  ) +
  ggplot2::annotate(
    "rect", xmin = -5, xmax = 5,
    ymin = -Inf, ymax = Inf,
    alpha = 0.08, fill = "green"
  ) +
  ggplot2::scale_fill_manual(values = c(
    "PSPC Original" = "#D9534F",
    "Conservative (8%)" = "#F0AD4E",
    "Moderate (12%)" = "#5BC0DE",
    "Aggressive (18%)" = "#5CB85C"
  )) +
  ggplot2::labs(
    title = "Removing Large-City Comparators Shifts the Distribution Rightward",
    subtitle = "Green shaded region = within +/-5% of market (\"comparable\")",
    x = "Schertz Variance from Survey Midpoint (%)",
    y = "Density",
    fill = "Scenario"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    legend.position = "bottom",
    plot.title = ggplot2::element_text(face = "bold")
  )
```


### Summary Statistics: The Headline Numbers

```{r}
#| label: tbl-summary-stats

# ── Compute summary stats for each scenario ──
compute_summary <- function(variance_col, label) {
  vals <- positions[[variance_col]]
  dplyr::tibble(
    Scenario = label,
    `Mean Gap` = sprintf("%.1f%%", mean(vals)),
    `Median Gap` = sprintf("%.1f%%", stats::median(vals)),
    `Positions Below -5%` = sum(vals < -5),
    `Pct Below -5%` = sprintf("%.0f%%", sum(vals < -5) / length(vals) * 100),
    `Positions Comparable` = sum(vals >= -5 & vals <= 5),
    `Pct Comparable` = sprintf("%.0f%%", sum(vals >= -5 & vals <= 5) /
      length(vals) * 100),
    `Positions Above +5%` = sum(vals > 5),
    `Pct Above +5%` = sprintf("%.0f%%", sum(vals > 5) / length(vals) * 100)
  )
}

summary_stats <- dplyr::bind_rows(
  compute_summary("Original_Variance_Pct", "PSPC Original"),
  compute_summary("Adj_Variance_Conservative", "Conservative (8% adj.)"),
  compute_summary("Adj_Variance_Moderate", "Moderate (12% adj.)"),
  compute_summary("Adj_Variance_Aggressive", "Aggressive (18% adj.)")
)

ft_summary <- flextable::flextable(summary_stats) |>
  flextable::add_header_lines(
    values = "How the PSPC Headline Findings Change Under Each Scenario"
  ) |>
  flextable::color(i = 1, part = "header", color = "steelblue") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, part = "header", align = "left") |>
  flextable::fontsize(i = 1, part = "header", size = 11) |>
  flextable::bg(i = 1, part = "header", bg = "white") |>
  flextable::bg(i = 2, part = "header", bg = "palegreen") |>
  flextable::fontsize(part = "body", size = 9) |>
  flextable::fontsize(i = 2, part = "header", size = 9) |>
  flextable::set_table_properties(layout = "autofit") |>
  flextable::align(j = 2:9, align = "center", part = "body") |>
  flextable::align(j = 2:9, align = "center", part = "header") |>
  flextable::bg(i = 1, part = "body", bg = "#FFE0E0")

ft_summary
```


### Estimated Cost Implications  

```{r}
#| label: tbl-cost-estimates

# ── The PSPC study estimated $942K for civilian adjustments. ──
# We can estimate the proportional reduction in cost under each scenario.
# Logic: cost is roughly proportional to the total gap. If the average gap
# shrinks by X%, the cost should shrink by approximately the same proportion.
# This is a simplification (individual employee placement matters), but
# directionally correct.

original_mean_gap <- mean(positions$Original_Variance_Pct)
original_civilian_cost <- 942000
original_public_safety_cost <- 1364000
original_total <- original_civilian_cost + original_public_safety_cost

# Only positions with negative gaps get adjustments
compute_cost_reduction <- function(original_gaps, adjusted_gaps,
                                   original_cost) {
  # Sum of negative gaps (the total "below market" dollars driving cost)
  orig_neg_sum <- sum(pmin(original_gaps, 0))
  adj_neg_sum <- sum(pmin(adjusted_gaps, 0))

  # Proportional reduction
  if (orig_neg_sum == 0) return(original_cost)
  ratio <- adj_neg_sum / orig_neg_sum
  round(original_cost * ratio, -3)  # Round to nearest thousand
}

cost_estimates <- dplyr::tibble(
  Scenario = c("PSPC Original", "Conservative (8%)",
               "Moderate (12%)", "Aggressive (18%)"),
  `Est. Civilian Cost` = scales::dollar(c(
    original_civilian_cost,
    compute_cost_reduction(
      positions$Original_Variance_Pct,
      positions$Adj_Variance_Conservative,
      original_civilian_cost
    ),
    compute_cost_reduction(
      positions$Original_Variance_Pct,
      positions$Adj_Variance_Moderate,
      original_civilian_cost
    ),
    compute_cost_reduction(
      positions$Original_Variance_Pct,
      positions$Adj_Variance_Aggressive,
      original_civilian_cost
    )
  ), accuracy = 1),
  `Est. Public Safety Cost` = scales::dollar(c(
    original_public_safety_cost,
    compute_cost_reduction(
      positions$Original_Variance_Pct,
      positions$Adj_Variance_Conservative,
      original_public_safety_cost
    ),
    compute_cost_reduction(
      positions$Original_Variance_Pct,
      positions$Adj_Variance_Moderate,
      original_public_safety_cost
    ),
    compute_cost_reduction(
      positions$Original_Variance_Pct,
      positions$Adj_Variance_Aggressive,
      original_public_safety_cost
    )
  ), accuracy = 1),
  `Est. Total Cost` = scales::dollar(c(
    original_total,
    compute_cost_reduction(
      positions$Original_Variance_Pct,
      positions$Adj_Variance_Conservative,
      original_total
    ),
    compute_cost_reduction(
      positions$Original_Variance_Pct,
      positions$Adj_Variance_Moderate,
      original_total
    ),
    compute_cost_reduction(
      positions$Original_Variance_Pct,
      positions$Adj_Variance_Aggressive,
      original_total
    )
  ), accuracy = 1),
  `Est. Savings vs. Original` = scales::dollar(c(
    0,
    original_total - compute_cost_reduction(
      positions$Original_Variance_Pct,
      positions$Adj_Variance_Conservative,
      original_total
    ),
    original_total - compute_cost_reduction(
      positions$Original_Variance_Pct,
      positions$Adj_Variance_Moderate,
      original_total
    ),
    original_total - compute_cost_reduction(
      positions$Original_Variance_Pct,
      positions$Adj_Variance_Aggressive,
      original_total
    )
  ), accuracy = 1)
)

ft_cost <- flextable::flextable(cost_estimates) |>
  flextable::add_header_lines(
    values = "Estimated Cost Implications by Scenario"
  ) |>
  flextable::color(i = 1, part = "header", color = "steelblue") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, part = "header", align = "left") |>
  flextable::fontsize(i = 1, part = "header", size = 11) |>
  flextable::bg(i = 1, part = "header", bg = "white") |>
  flextable::bg(i = 2, part = "header", bg = "palegreen") |>
  flextable::set_table_properties(layout = "autofit") |>
  flextable::align(j = 2:5, align = "right", part = "body") |>
  flextable::bg(i = 1, part = "body", bg = "#FFE0E0")

ft_cost
```


### The Reclassification Effect

```{r}
#| label: fig-reclassification
#| fig-cap: "How Positions Reclassify Under Each Scenario"
#| fig-height: 5
#| fig-width: 8

# ── Count positions in each category by scenario ──
classify <- function(gaps) {
  dplyr::tibble(
    Below = sum(gaps < -5),
    Comparable = sum(gaps >= -5 & gaps <= 5),
    Above = sum(gaps > 5)
  )
}

reclass <- dplyr::bind_rows(
  classify(positions$Original_Variance_Pct) |>
    dplyr::mutate(Scenario = "PSPC\nOriginal"),
  classify(positions$Adj_Variance_Conservative) |>
    dplyr::mutate(Scenario = "Conservative\n(8%)"),
  classify(positions$Adj_Variance_Moderate) |>
    dplyr::mutate(Scenario = "Moderate\n(12%)"),
  classify(positions$Adj_Variance_Aggressive) |>
    dplyr::mutate(Scenario = "Aggressive\n(18%)")
) |>
  tidyr::pivot_longer(
    cols = c(Below, Comparable, Above),
    names_to = "Category",
    values_to = "Count"
  ) |>
  dplyr::mutate(
    Scenario = factor(Scenario, levels = c(
      "PSPC\nOriginal", "Conservative\n(8%)",
      "Moderate\n(12%)", "Aggressive\n(18%)"
    )),
    Category = factor(Category, levels = c("Above", "Comparable", "Below"))
  )

ggplot2::ggplot(
  reclass,
  ggplot2::aes(x = Scenario, y = Count, fill = Category)
) +
  ggplot2::geom_col(position = "stack", width = 0.6) +
  ggplot2::geom_text(
    ggplot2::aes(label = Count),
    position = ggplot2::position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 4
  ) +
  ggplot2::scale_fill_manual(
    values = c(
      "Below" = "#D9534F",
      "Comparable" = "#F0AD4E",
      "Above" = "#5CB85C"
    ),
    labels = c(
      "Below" = "Below Market (< -5%)",
      "Comparable" = "Comparable (+/- 5%)",
      "Above" = "Above Market (> +5%)"
    )
  ) +
  ggplot2::labs(
    title = "Removing Large-City Bias Shifts Positions Toward \"Comparable\"",
    subtitle = paste0("Total positions: ", nrow(positions)),
    x = NULL,
    y = "Number of Positions",
    fill = NULL
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    legend.position = "bottom",
    plot.title = ggplot2::element_text(face = "bold")
  )
```


## Interpretation and Caveats  

### What This Analysis Shows  

1. **The PSPC headline finding is sensitive to comparator composition.** Under the moderate scenario (12% large-city premium), the proportion of positions classified as "below market" drops meaningfully, and the average gap narrows.

2. **Even after adjustment, Schertz was still likely behind market for many
positions.** This analysis does not argue that no pay correction was warranted. It
argues that the *magnitude* of the correction was inflated by the comparator group.

3. **The cost differential is material.** Even the conservative scenario suggests
the City may have overshot "market" as defined by its actual peer cities by a
meaningful amount.

### What This Analysis Cannot Show

1. **Exact position-level adjustments.** We applied a uniform adjustment factor. In
reality, some positions (e.g., Police Chief, Fire Chief) may have had more
large-city matches while others (e.g., Library Assistant) may have had fewer. The
position-level estimates should be treated as illustrative.

2. **The effect of county and utility comparators.** Bexar County, Williamson
County, San Antonio Water System, and New Braunfels Utilities were also included.
These entities may introduce additional upward bias for certain positions. We cannot
model their removal.

3. **ERI private-sector data impact.** The inclusion of private-sector benchmarks
from ERI's Salary Assessor for the Austin-San Antonio corridor likely adds further
upward pressure on certain positions. This is unquantifiable from the report.

4. **The actual weighting scheme.** If PSPC used median rather than mean, the impact
of outliers (including Austin and San Antonio) would be smaller. If they used
size-weighted averages, the impact would be larger. We assumed equal-weighted mean.

### The Bottom Line

The PSPC study's comparator group is the primary driver of its finding that Schertz
is significantly below market. A more defensible peer group — limited to cities of
comparable size and operational scope — would have produced a smaller gap, a lower
cost estimate, and potentially a different recommendation. The Council adopted a
\$2.3 million expenditure based on a study that did not disclose its weighting
methodology, did not present sensitivity analysis, and did not test its conclusions
against operationally similar peer cities.

**The right question is not "was Schertz below market?" (probably yes) but "how far
below, compared to whom, and was \$2.3M the right-sized response?"** This analysis
suggests the answer to the last question is: probably not.

---

## Appendix: Full Position Data by Scenario

```{r}
#| label: tbl-full-appendix

# ── Full dataset for all 65 positions ──
full_display <- positions |>
  dplyr::arrange(Original_Variance_Pct) |>
  dplyr::transmute(
    Position,
    `Schertz Mid.` = scales::dollar(Schertz_Midpoint, accuracy = 1),
    `PSPC Mid.` = scales::dollar(PSPC_Survey_Midpoint, accuracy = 1),
    `Orig. Gap` = sprintf("%.1f%%", Original_Variance_Pct),
    `8% Gap` = sprintf("%.1f%%", Adj_Variance_Conservative),
    `12% Gap` = sprintf("%.1f%%", Adj_Variance_Moderate),
    `18% Gap` = sprintf("%.1f%%", Adj_Variance_Aggressive)
  )

ft_full <- flextable::flextable(full_display) |>
  flextable::add_header_lines(
    values = "All 65 Civilian Benchmark Positions: Gap Under Each Scenario"
  ) |>
  flextable::color(i = 1, part = "header", color = "steelblue") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, part = "header", align = "left") |>
  flextable::fontsize(i = 1, part = "header", size = 11) |>
  flextable::bg(i = 1, part = "header", bg = "white") |>
  flextable::bg(i = 2, part = "header", bg = "palegreen") |>
  flextable::fontsize(part = "body", size = 7) |>
  flextable::fontsize(i = 2, part = "header", size = 8) |>
  flextable::set_table_properties(layout = "autofit") |>
  flextable::align(j = 2:7, align = "right", part = "body")

ft_full
```

---

*Analysis prepared `r Sys.Date()`. All assumptions are documented and modifiable.
Source data: PSPC FY 2024 Classification and Compensation Study, December 2023.
Research citations: Baum-Snow & Pavan (2012), Yankow (2006), Glaeser & Maré (2001),
NLC City Fiscal Conditions Survey (2024).*
