---
title: "Schertz Compensation Review - Append New Data"
author: "Compensation Review Project"
date: today
format: html
editor: visual
---

## Overview

This pipeline processes City of Schertz employee compensation data from FY20 through FY25 (and beyond). It handles the different Excel file structures automatically.

**To add a new fiscal year, you only need ONE line of code.**

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup
#| message: false

library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(janitor)
library(purrr)
library(lubridate)
library(here)

# Load pipeline components
source(here::here("schertz_config.R"))
source(here::here("schertz_processing.R"))
```

------------------------------------------------------------------------
## Adding a New Fiscal Year (The Easy Way)

When you receive a new Excel file (e.g., FY26):

### Step 1: Place the file in your data_raw folder

Copy the Excel file to: `data_raw/`

### Step 2: Preview it (optional but recommended)

```{r}
#| label: preview-new-file
#| eval: false

# See what the file contains before adding it
preview_file("Employee_Compensation_Report_FY25.xlsx")
```

This shows you:

-   Detected fiscal year
-   Detected structure type
-   All column names
-   First/last rows (to check for totals)

### Step 3: Add and process with ONE command

```{r}
#| label: add-new-year
#| eval: false

# This does everything:
#   - Auto-detects fiscal year from filename
#   - Auto-detects structure type from columns
#   - Adds to file registry
#   - Runs the full pipeline
source(here::here("append_fy25.R"))
```

That's it! The pipeline:

1.  Detects whether it's FY26 from the filename or column names
2.  Figures out the column structure automatically
3.  Maps columns to standardized names
4.  Filters out total/subtotal rows
5.  Cleans names and creates employee keys
6.  Combines with all previous years

------------------------------------------------------------------------
