---
title: "Consumer Price Index (CPI) Analysis"
subtitle: "Fetching, Saving, and Visualizing Annual CPI Data (1970-Present)"
author: "Dan Swart"
date: today
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

This document demonstrates how to:

1. Fetch annual CPI data from the Bureau of Labor Statistics (BLS)
2. Save the data locally for future use
3. Load the saved data
4. Filter to selected years
5. Create a line chart of CPI over time

The CPI measures the average change over time in the prices paid by urban consumers for a market basket of consumer goods and services.

---

## Step 1: Setup and Load Required Packages

First, we load the packages we'll need. If you don't have these installed, run `install.packages(c("httr2", "jsonlite", "here", "ggplot2", "dplyr", "flextable", "scales"))` first.

```{r}
#| label: setup

# Set options
options(scipen = 999)

# We'll use these packages:
# - httr2: for making HTTP requests to the BLS API
# - jsonlite: for parsing JSON responses
# - here: for file path management
# - ggplot2: for plotting
# - dplyr: for data manipulation
# - flextable: for nice tables
# - scales: for formatting axis labels
```

---

## Step 2: Define the Data Storage Location

We'll save the CPI data locally so we don't have to fetch it every time.

```{r}
#| label: define-paths

# Define where to save the data
data_folder <- here::here("data_clean")

# Create the folder if it doesn't exist
if (!base::dir.exists(data_folder)) {
  base::dir.create(data_folder, recursive = TRUE)
  base::cat("Created folder:", data_folder, "\n")
} else {
  base::cat("Data folder exists:", data_folder, "\n")
}

# Define the file path for saved CPI data
cpi_file_path <- base::file.path(data_folder, "annual_cpi_data.rds")

base::cat("CPI data will be saved to:", cpi_file_path, "\n")
```

---
 
## Step 3: Fetch CPI Data from the Bureau of Labor Statistics

The BLS provides free API access to CPI data. We'll fetch the "All Urban Consumers" CPI (series ID: CUUR0000SA0).

**Note:** The BLS API v1 (no registration required) allows fetching 10 years at a time. We'll make multiple requests to get data from 1970 to present.

```{r}
#| label: fetch-cpi-data

fetch_cpi_from_bls <- function(start_year, end_year) {
  
  # BLS API endpoint (version 1 - no API key required)
  # Series ID CUUR0000SA0 = CPI-U All Items, US City Average, Not Seasonally Adjusted
  
  base_url <- "https://api.bls.gov/publicAPI/v1/timeseries/data/"
  series_id <- "CUUR0000SA0"
  
  # Build the request URL
  url <- base::paste0(
    base_url, series_id,
    "?startyear=", start_year,
    "&endyear=", end_year
  )
  
  base::cat("Fetching CPI data for", start_year, "-", end_year, "...\n")
  
  # Make the request
  response <- httr2::request(url) |>
    httr2::req_perform()
  
  # Parse JSON response
  json_data <- httr2::resp_body_json(response)
  
  # Check for errors
  if (json_data$status != "REQUEST_SUCCEEDED") {
    base::warning("BLS API request failed for ", start_year, "-", end_year)
    return(NULL)
  }
  
  # Extract the data
  series_data <- json_data$Results$series[[1]]$data
  
  # Convert to data frame
  df <- purrr::map_dfr(series_data, function(obs) {
    dplyr::tibble(
      year = base::as.integer(obs$year),
      period = obs$period,
      period_name = obs$periodName,
      value = base::as.numeric(obs$value)
    )
  })
  
  return(df)
}

# Function to get annual average CPI (average of all months in a year)
fetch_all_cpi_data <- function(start_year = 1970, end_year = NULL) {
  
  # Default to current year
  if (base::is.null(end_year)) {
    end_year <- base::as.integer(base::format(base::Sys.Date(), "%Y"))
  }
  
  # BLS API v1 allows max 10 years per request
  # We'll chunk the requests
  all_data <- dplyr::tibble()
  
  current_start <- start_year
  
  while (current_start <= end_year) {
    current_end <- base::min(current_start + 9, end_year)
    
    chunk <- fetch_cpi_from_bls(current_start, current_end)
    
    if (!base::is.null(chunk)) {
      all_data <- dplyr::bind_rows(all_data, chunk)
    }
    
    current_start <- current_end + 1
    
    # Be nice to the API - pause between requests
    if (current_start <= end_year) {
      base::Sys.sleep(1)
    }
  }
  
  # Filter to annual averages only (period M13 is the annual average)
  # If M13 not available, calculate from monthly data
  annual_data <- all_data |>
    dplyr::filter(period == "M13" | base::grepl("^M[0-1][0-9]$", period)) |>
    dplyr::group_by(year) |>
    dplyr::summarise(
      annual_cpi = dplyr::if_else(
        base::any(period == "M13"),
        value[period == "M13"][1],
        base::mean(value, na.rm = TRUE)
      ),
      .groups = "drop"
    ) |>
    dplyr::arrange(year)
  
  return(annual_data)
}
```

---

## Step 4: Check for Existing Data or Fetch New Data

We'll check if we already have saved data. If not (or if it's outdated), we'll fetch fresh data.

```{r}
#| label: load-or-fetch

# Check if we have saved data
if (base::file.exists(cpi_file_path)) {
  
  # Load existing data
  cpi_data <- base::readRDS(cpi_file_path)
  base::cat("Loaded existing CPI data from:", cpi_file_path, "\n")
  base::cat("Data covers years:", base::min(cpi_data$year), "-", base::max(cpi_data$year), "\n")
  
  # Check if data is current (includes current or previous year)
  current_year <- base::as.integer(base::format(base::Sys.Date(), "%Y"))
  
  if (base::max(cpi_data$year) < (current_year - 1)) {
    base::cat("Data is outdated. Fetching fresh data...\n")
    cpi_data <- fetch_all_cpi_data(start_year = 1970)
  }
  
} else {
  
  # No saved data - fetch from BLS
  base::cat("No saved data found. Fetching from BLS API...\n")
  cpi_data <- fetch_all_cpi_data(start_year = 1970)
  
}

# Show summary
base::cat("\nCPI data summary:\n")
base::cat("- Total years:", base::nrow(cpi_data), "\n")
base::cat("- Year range:", base::min(cpi_data$year), "-", base::max(cpi_data$year), "\n")
base::cat("- CPI range:", base::round(base::min(cpi_data$annual_cpi), 1), "-", 
          base::round(base::max(cpi_data$annual_cpi), 1), "\n")
```

---

## Step 5: Save the Data Locally

Save the fetched data so we can reuse it without hitting the API again.

```{r}
#| label: save-data

# Save the data
base::saveRDS(cpi_data, file = cpi_file_path)
base::cat("CPI data saved to:", cpi_file_path, "\n")

# Also save as CSV for easy viewing
csv_path <- base::file.path(data_folder, "annual_cpi_data.csv")
readr::write_csv(cpi_data, csv_path)
base::cat("CSV copy saved to:", csv_path, "\n")
```

---

## Step 6: Display the Full CPI Data Table

Let's see all the data we have in a nice table format.

```{r}
#| label: display-full-data

cpi_data |>
  dplyr::mutate(
    annual_cpi = base::round(annual_cpi, 1)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Annual CPI-U (All Urban Consumers) - 1970 to Present") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

---

## Step 7: Define the Years You Want to Plot

Here you can specify exactly which years to include in your chart. Edit the `selected_years` variable to customize.

```{r}
#| label: select-years

# =============================================================================
# EDIT THIS SECTION TO SELECT YOUR YEARS
# =============================================================================

# Option 1: Specify a range
selected_start_year <- 1970
selected_end_year <- 2024

# Option 2: Or specify individual years (uncomment to use)
# specific_years <- c(1970, 1980, 1990, 2000, 2010, 2020, 2024)

# Filter the data to selected years
# Using Option 1 (range):
cpi_filtered <- cpi_data |>
  dplyr::filter(year >= selected_start_year & year <= selected_end_year)

# Using Option 2 (specific years) - uncomment these lines to use:
# cpi_filtered <- cpi_data |>
#   dplyr::filter(year %in% specific_years)

base::cat("Selected", base::nrow(cpi_filtered), "years of data\n")
base::cat("Years:", base::min(cpi_filtered$year), "-", base::max(cpi_filtered$year), "\n")
```

---

## Step 8: Display the Filtered Data

Let's see the data we've selected for plotting.

```{r}
#| label: display-filtered-data

cpi_filtered |>
  dplyr::mutate(
    annual_cpi = base::round(annual_cpi, 1)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines(
    base::paste0("Selected CPI Data: ", selected_start_year, " - ", selected_end_year)
  ) |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

---

## Step 9: Create the CPI Line Chart

Now let's visualize the CPI trend over the selected years.

```{r}
#| label: plot-cpi
#| fig-width: 12
#| fig-height: 7

ggplot2::ggplot(
  cpi_filtered,
  ggplot2::aes(x = year, y = annual_cpi)
) +
  ggplot2::geom_line(linewidth = 1.2, color = "steelblue") +
  ggplot2::geom_point(size = 2, color = "steelblue") +
  ggplot2::scale_x_continuous(
    breaks = base::seq(
      base::floor(base::min(cpi_filtered$year) / 5) * 5,
      base::max(cpi_filtered$year),
      by = 5
    )
  ) +
  ggplot2::scale_y_continuous(
    labels = scales::comma_format()
  ) +
  ggplot2::labs(
    title = "Consumer Price Index (CPI-U) Over Time",
    subtitle = base::paste0(
      "Annual Average, All Urban Consumers, ",
      base::min(cpi_filtered$year), " - ", base::max(cpi_filtered$year)
    ),
    x = "Year",
    y = "CPI Index Value",
    caption = "Source: Bureau of Labor Statistics (BLS)"
  ) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", size = 16),
    plot.subtitle = ggplot2::element_text(color = "gray40"),
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    panel.grid.minor = ggplot2::element_blank()
  )
```

---

## Step 10: Calculate and Plot Year-over-Year Inflation Rate

The CPI index itself shows price levels. To see inflation (rate of change), we calculate the percentage change from year to year.

```{r}
#| label: calculate-inflation

cpi_with_inflation <- cpi_filtered |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    prior_cpi = dplyr::lag(annual_cpi),
    inflation_rate = (annual_cpi - prior_cpi) / prior_cpi * 100
  )

# Display the data with inflation rates
cpi_with_inflation |>
  dplyr::filter(!base::is.na(inflation_rate)) |>
  dplyr::mutate(
    annual_cpi = base::round(annual_cpi, 1),
    prior_cpi = base::round(prior_cpi, 1),
    inflation_rate = base::round(inflation_rate, 2)
  ) |>
  dplyr::select(year, prior_cpi, annual_cpi, inflation_rate) |>
  dplyr::rename(
    Year = year,
    `Prior Year CPI` = prior_cpi,
    `Current CPI` = annual_cpi,
    `Inflation Rate (%)` = inflation_rate
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Year-over-Year Inflation Rates") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

---

## Step 11: Plot the Inflation Rate Over Time

```{r}
#| label: plot-inflation
#| fig-width: 12
#| fig-height: 7

inflation_data <- cpi_with_inflation |>
  dplyr::filter(!base::is.na(inflation_rate))

ggplot2::ggplot(
  inflation_data,
  ggplot2::aes(x = year, y = inflation_rate)
) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  ggplot2::geom_line(linewidth = 1, color = "firebrick") +
  ggplot2::geom_point(size = 2, color = "firebrick") +
  ggplot2::geom_area(alpha = 0.2, fill = "firebrick") +
  ggplot2::scale_x_continuous(
    breaks = base::seq(
      base::floor(base::min(inflation_data$year) / 5) * 5,
      base::max(inflation_data$year),
      by = 5
    )
  ) +
  ggplot2::scale_y_continuous(
    labels = function(x) base::paste0(x, "%")
  ) +
  ggplot2::labs(
    title = "Annual Inflation Rate (Year-over-Year % Change in CPI)",
    subtitle = base::paste0(
      base::min(inflation_data$year), " - ", base::max(inflation_data$year)
    ),
    x = "Year",
    y = "Inflation Rate (%)",
    caption = "Source: Bureau of Labor Statistics (BLS)"
  ) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", size = 16),
    plot.subtitle = ggplot2::element_text(color = "gray40"),
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    panel.grid.minor = ggplot2::element_blank()
  )
```

---

## Step 12: Highlight Notable Periods

Let's add annotations for notable inflation periods.

```{r}
#| label: plot-inflation-annotated
#| fig-width: 12
#| fig-height: 7

ggplot2::ggplot(
  inflation_data,
  ggplot2::aes(x = year, y = inflation_rate)
) +
  # Shaded regions for notable periods
  ggplot2::annotate(
    "rect", xmin = 1973, xmax = 1975, ymin = -Inf, ymax = Inf,
    alpha = 0.2, fill = "orange"
  ) +
  ggplot2::annotate(
    "rect", xmin = 1979, xmax = 1982, ymin = -Inf, ymax = Inf,
    alpha = 0.2, fill = "orange"
  ) +
  ggplot2::annotate(
    "rect", xmin = 2021, xmax = 2023, ymin = -Inf, ymax = Inf,
    alpha = 0.2, fill = "orange"
  ) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  ggplot2::geom_hline(yintercept = 2, linetype = "dotted", color = "darkgreen", linewidth = 1) +
  ggplot2::geom_line(linewidth = 1, color = "firebrick") +
  ggplot2::geom_point(size = 2, color = "firebrick") +
  # Annotations
  ggplot2::annotate(
    "text", x = 1974, y = 12, label = "Oil Crisis\n1973-75",
    size = 3, fontface = "italic"
  ) +
  ggplot2::annotate(
    "text", x = 1980.5, y = 14.5, label = "Volcker Era\n1979-82",
    size = 3, fontface = "italic"
  ) +
  ggplot2::annotate(
    "text", x = 2022, y = 9, label = "Post-COVID\n2021-23",
    size = 3, fontface = "italic"
  ) +
  ggplot2::annotate(
    "text", x = 2000, y = 2.5, label = "Fed Target: 2%",
    size = 3, color = "darkgreen", fontface = "bold"
  ) +
  ggplot2::scale_x_continuous(
    breaks = base::seq(
      base::floor(base::min(inflation_data$year) / 5) * 5,
      base::max(inflation_data$year),
      by = 5
    )
  ) +
  ggplot2::scale_y_continuous(
    labels = function(x) base::paste0(x, "%")
  ) +
  ggplot2::labs(
    title = "Annual Inflation Rate with Notable Periods Highlighted",
    subtitle = "Orange shading indicates high-inflation periods",
    x = "Year",
    y = "Inflation Rate (%)",
    caption = "Source: Bureau of Labor Statistics (BLS) | Green dotted line = Fed's 2% target"
  ) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", size = 16),
    plot.subtitle = ggplot2::element_text(color = "gray40"),
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    panel.grid.minor = ggplot2::element_blank()
  )
```

---

## Summary

This document demonstrated:

1. **Fetching data**: Used the BLS API to retrieve CPI data
2. **Saving locally**: Stored data as RDS and CSV for reuse
3. **Loading data**: Checked for existing data before fetching
4. **Filtering**: Selected specific years for analysis
5. **Visualization**: Created line charts of CPI and inflation rates
6. **Annotation**: Added context for notable historical periods

### Key Takeaways from the Data

```{r}
#| label: summary-stats

summary_stats <- inflation_data |>
  dplyr::summarise(
    `Average Inflation` = base::paste0(base::round(base::mean(inflation_rate), 2), "%"),
    `Median Inflation` = base::paste0(base::round(stats::median(inflation_rate), 2), "%"),
    `Max Inflation` = base::paste0(base::round(base::max(inflation_rate), 2), "% (", 
                                    year[inflation_rate == base::max(inflation_rate)], ")"),
    `Min Inflation` = base::paste0(base::round(base::min(inflation_rate), 2), "% (",
                                    year[inflation_rate == base::min(inflation_rate)], ")"),
    `Years Above 5%` = base::sum(inflation_rate > 5),
    `Years Below 2%` = base::sum(inflation_rate < 2)
  ) |>
  tidyr::pivot_longer(
    cols = dplyr::everything(),
    names_to = "Statistic",
    values_to = "Value"
  )

summary_stats |>
  flextable::flextable() |>
  flextable::add_header_lines("Inflation Summary Statistics") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

---

## How to Customize

To change the years plotted, edit **Step 7**:

```r
# For a range:
selected_start_year <- 2000
selected_end_year <- 2024

# For specific years:
specific_years <- c(1980, 1990, 2000, 2010, 2020)
```

To refresh the data from BLS, delete the saved file:

```r
file.remove(here::here("data_clean", "annual_cpi_data.rds"))
```

Then re-run the document to fetch fresh data.
