---
title: "Get U.S. Inflation Rates"
subtitle: "Using Official Data from FRED (Federal Reserve Economic Data)"
author: "Dan Swart"
date: today
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

This document demonstrates how to:

1. Get a free FRED API key
2. Fetch official inflation rate data from the Federal Reserve
3. Save the data locally for future use
4. Filter to selected years
5. Create visualizations of inflation over time

**Why FRED?**

- Run by the Federal Reserve Bank of St. Louis
- Trusted by economists, academics, and policymakers worldwide
- Provides pre-calculated inflation rates (no manual calculation needed)
- Free API access
- Over 800,000 economic data series

---

## Step 1: Get Your Free FRED API Key

Before running this code, you need a free API key from FRED:
asdf
1. Go to: https://fred.stlouisfed.org/docs/api/api_key.html
2. Click "Request API Key"
3. Create a free account (or log in)
4. Your API key will be displayed

**Store your API key** by running this once in your R console:

```r
# Run this ONCE in your console (not in the document)
# Replace "your_api_key_here" with your actual key
fredr::fredr_set_key("your_api_key_here")
```

Or set it in your `.Renviron` file for permanent storage:

```r
# Run this ONCE to open your .Renviron file
usethis::edit_r_environ()

# Add this line to the file:
# FRED_API_KEY=your_api_key_here
```

---

## Step 2: Setup and Load Packages

```{r}
#| label: setup

# Set options
options(scipen = 999)

# Check if fredr is installed
if (!base::requireNamespace("fredr", quietly = TRUE)) {
  base::stop("Please install the fredr package: install.packages('fredr')")
}

# Set your FRED API key (uncomment and add your key, or use .Renviron)
# fredr::fredr_set_key("your_api_key_here")

base::cat("Setup complete.\n")
base::cat("If you haven't set your FRED API key, do so now before proceeding.\n")
```

---

## Step 3: Define Data Storage Location

```{r}
#| label: define-paths

# Define where to save the data
inflation_rates_folder <- here::here("inflation_rates")

# Create the folder if it doesn't exist
if (!base::dir.exists(inflation_rates_folder)) {
  base::dir.create(inflation_rates_folder, recursive = TRUE)
  base::cat("Created folder:", inflation_rates_folder, "\n")
} else {
  base::cat("Data folder exists:", inflation_rates_folder, "\n")
}

# Define file paths
inflation_file_path <- base::file.path(inflation_rates_folder, "fred_inflation_data.rds")

base::cat("Data will be saved to:", inflation_file_path, "\n")
```

---

## Step 4: Understand the FRED Series We'll Use

FRED has several inflation-related series. Here are the most useful:

| Series ID | Description | Frequency |
|-----------|-------------|-----------|
| `FPCPITOTLZGUSA` | Inflation rate (CPI-based), annual | Annual |
| `CPIAUCSL` | CPI index (all urban consumers) | Monthly |
| `CPILFESL` | Core CPI (excludes food & energy) | Monthly |
| `T10YIE` | 10-year breakeven inflation rate | Daily |

We'll use **`FPCPITOTLZGUSA`** — the official annual inflation rate for the United States, published by the World Bank and distributed through FRED.

```{r}
#| label: series-info

# The series we'll fetch
series_id <- "FPCPITOTLZGUSA"

base::cat("We will fetch series:", series_id, "\n")
base::cat("Description: Inflation, consumer prices for the United States\n")
base::cat("Source: World Bank via FRED\n")
base::cat("Frequency: Annual\n")
```

---

## Step 5: Fetch Inflation Data from FRED

```{r}
#| label: fetch-data

fetch_fred_inflation <- function(series_id = "FPCPITOTLZGUSA", 
                                  start_date = "1970-01-01",
                                  end_date = NULL) {
  
  # Default end date to today
  if (base::is.null(end_date)) {
    end_date <- base::Sys.Date()
  }
  
  base::cat("Fetching", series_id, "from FRED...\n")
  
  # Fetch data from FRED
  inflation_raw <- fredr::fredr(
    series_id = series_id,
    observation_start = base::as.Date(start_date),
    observation_end = base::as.Date(end_date)
  )
  
  # Clean up the data
  inflation_data <- inflation_raw |>
    dplyr::transmute(
      year = base::as.integer(base::format(date, "%Y")),
      date = date,
      inflation_rate = value,
      series_id = series_id
    ) |>
    dplyr::filter(!base::is.na(inflation_rate)) |>
    dplyr::arrange(year)
  
  base::cat("Retrieved", base::nrow(inflation_data), "observations\n")
  base::cat("Year range:", base::min(inflation_data$year), "-", 
            base::max(inflation_data$year), "\n")
  
  return(inflation_data)
}
```

---

## Step 6: Load Existing Data or Fetch Fresh Data

```{r}
#| label: load-or-fetch

# Check if we have saved data
if (base::file.exists(inflation_file_path)) {
  
  # Load existing data
  inflation_data <- base::readRDS(inflation_file_path)
  base::cat("Loaded existing inflation data from:", inflation_file_path, "\n")
  base::cat("Data covers years:", base::min(inflation_data$year), "-", 
            base::max(inflation_data$year), "\n")
  
  # Check if data is current
  current_year <- base::as.integer(base::format(base::Sys.Date(), "%Y"))
  
  if (base::max(inflation_data$year) < (current_year - 1)) {
    base::cat("Data may be outdated. Fetching fresh data...\n")
    inflation_data <- fetch_fred_inflation(start_date = "1960-01-01")
  }
  
} else {
  
  # No saved data - fetch from FRED
  base::cat("No saved data found. Fetching from FRED...\n")
  inflation_data <- fetch_fred_inflation(start_date = "1960-01-01")
  
}

# Summary
base::cat("\n--- Data Summary ---\n")
base::cat("Total years:", base::nrow(inflation_data), "\n")
base::cat("Year range:", base::min(inflation_data$year), "-", 
          base::max(inflation_data$year), "\n")
base::cat("Inflation range:", 
          base::round(base::min(inflation_data$inflation_rate), 2), "% to",
          base::round(base::max(inflation_data$inflation_rate), 2), "%\n")
```

---

## Step 7: Save the Data Locally

```{r}
#| label: save-data

# Save as RDS
base::saveRDS(inflation_data, file = inflation_file_path)
base::cat("Inflation data saved to:", inflation_file_path, "\n")

# Also save as CSV
csv_path <- base::file.path(inflation_rates_folder, "fred_inflation_data.csv")
readr::write_csv(inflation_data, csv_path)
base::cat("CSV copy saved to:", csv_path, "\n")
```

---

## Step 8: Display the Full Inflation Data

```{r}
#| label: display-full-data

inflation_data |>
  dplyr::select(year, inflation_rate) |>
  dplyr::mutate(
    inflation_rate = base::paste0(base::round(inflation_rate, 2), "%")
  ) |>
  dplyr::rename(
    Year = year,
    `Inflation Rate` = inflation_rate
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Annual U.S. Inflation Rate (FRED Series: FPCPITOTLZGUSA)") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

---

## Step 9: Select Years to Plot

**Edit this section to customize which years appear in your charts.**

```{r}
#| label: select-years

# =============================================================================
# EDIT THESE VALUES TO SELECT YOUR YEARS
# =============================================================================

# Option 1: Specify a range
selected_start_year <- 1970
selected_end_year <- 2024

# Option 2: Specific years (uncomment to use instead)
# specific_years <- c(1970, 1975, 1980, 1985, 1990, 1995, 2000, 2005, 2010, 2015, 2020, 2024)

# =============================================================================

# Filter the data using Option 1 (range)
inflation_filtered <- inflation_data |>
  dplyr::filter(year >= selected_start_year & year <= selected_end_year)

# To use Option 2 (specific years), uncomment these lines:
# inflation_filtered <- inflation_data |>
#   dplyr::filter(year %in% specific_years)

base::cat("Selected", base::nrow(inflation_filtered), "years\n")
base::cat("Range:", base::min(inflation_filtered$year), "-", 
          base::max(inflation_filtered$year), "\n")
```

---

## Step 10: Display Filtered Data

```{r}
#| label: display-filtered

inflation_filtered |>
  dplyr::select(year, inflation_rate) |>
  dplyr::mutate(
    inflation_rate = base::paste0(base::round(inflation_rate, 2), "%")
  ) |>
  dplyr::rename(
    Year = year,
    `Inflation Rate` = inflation_rate
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines(
    base::paste0("U.S. Inflation Rate: ", selected_start_year, " - ", selected_end_year)
  ) |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

---

## Step 11: Plot Inflation Rate Over Time

```{r}
#| label: plot-inflation
#| fig-width: 12
#| fig-height: 7

ggplot2::ggplot(
  inflation_filtered,
  ggplot2::aes(x = year, y = inflation_rate)
) +
  ggplot2::geom_hline(yintercept = 0, linetype = "solid", color = "gray30") +
  ggplot2::geom_hline(yintercept = 2, linetype = "dashed", color = "darkgreen", 
                      linewidth = 1) +
  ggplot2::geom_line(linewidth = 1.2, color = "firebrick") +
  ggplot2::geom_point(size = 2.5, color = "firebrick") +
  ggplot2::annotate(
    "text", 
    x = base::max(inflation_filtered$year) - 5, 
    y = 2.5, 
    label = "Fed Target: 2%",
    color = "darkgreen", 
    fontface = "bold",
    size = 4
  ) +
  ggplot2::scale_x_continuous(
    breaks = base::seq(
      base::floor(base::min(inflation_filtered$year) / 5) * 5,
      base::max(inflation_filtered$year),
      by = 5
    )
  ) +
  ggplot2::scale_y_continuous(
    labels = function(x) base::paste0(x, "%"),
    breaks = base::seq(-2, 16, by = 2)
  ) +
  ggplot2::labs(
    title = "U.S. Annual Inflation Rate",
    subtitle = base::paste0(
      base::min(inflation_filtered$year), " - ", 
      base::max(inflation_filtered$year),
      " | Source: FRED (Federal Reserve Economic Data)"
    ),
    x = "Year",
    y = "Inflation Rate (%)",
    caption = "Series: FPCPITOTLZGUSA | Green dashed line = Fed's 2% target"
  ) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", size = 18),
    plot.subtitle = ggplot2::element_text(color = "gray40", size = 12),
    plot.caption = ggplot2::element_text(color = "gray50", size = 10),
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    panel.grid.minor = ggplot2::element_blank()
  )
```

---

## Step 12: Inflation with Notable Periods Highlighted

```{r}
#| label: plot-inflation-annotated
#| fig-width: 13
#| fig-height: 8

ggplot2::ggplot(
  inflation_filtered,
  ggplot2::aes(x = year, y = inflation_rate)
) +
  # Shaded regions for high-inflation periods
  ggplot2::annotate(
    "rect", xmin = 1973, xmax = 1975, ymin = -Inf, ymax = Inf,
    alpha = 0.15, fill = "orange"
  ) +
  ggplot2::annotate(
    "rect", xmin = 1978, xmax = 1982, ymin = -Inf, ymax = Inf,
    alpha = 0.15, fill = "orange"
  ) +
  ggplot2::annotate(
    "rect", xmin = 2021, xmax = 2023, ymin = -Inf, ymax = Inf,
    alpha = 0.15, fill = "orange"
  ) +
  # Reference lines
  ggplot2::geom_hline(yintercept = 0, linetype = "solid", color = "gray30") +
  ggplot2::geom_hline(yintercept = 2, linetype = "dashed", color = "darkgreen", 
                      linewidth = 1) +
  # Data
  ggplot2::geom_area(alpha = 0.3, fill = "firebrick") +
  ggplot2::geom_line(linewidth = 1.2, color = "firebrick") +
  ggplot2::geom_point(size = 2, color = "firebrick") +
  # Annotations for periods
  ggplot2::annotate(
    "text", x = 1974, y = 12.5, 
    label = "Oil\nCrisis",
    size = 3.5, fontface = "bold.italic", color = "gray30"
  ) +
  ggplot2::annotate(
    "text", x = 1980, y = 14.5, 
    label = "Volcker\nEra",
    size = 3.5, fontface = "bold.italic", color = "gray30"
  ) +
  ggplot2::annotate(
    "text", x = 2022, y = 9.5, 
    label = "Post-\nCOVID",
    size = 3.5, fontface = "bold.italic", color = "gray30"
  ) +
  ggplot2::annotate(
    "text", 
    x = 1995, 
    y = 2.7, 
    label = "Fed Target: 2%",
    color = "darkgreen", 
    fontface = "bold",
    size = 4
  ) +
  # Scales
  ggplot2::scale_x_continuous(
    breaks = base::seq(
      base::floor(base::min(inflation_filtered$year) / 5) * 5,
      base::max(inflation_filtered$year),
      by = 5
    )
  ) +
  ggplot2::scale_y_continuous(
    labels = function(x) base::paste0(x, "%"),
    breaks = base::seq(-2, 16, by = 2)
  ) +
  ggplot2::labs(
    title = "U.S. Annual Inflation Rate with Notable Periods",
    subtitle = "Orange shading indicates high-inflation periods",
    x = "Year",
    y = "Inflation Rate (%)",
    caption = "Source: FRED (Series: FPCPITOTLZGUSA) | Federal Reserve Bank of St. Louis"
  ) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", size = 18),
    plot.subtitle = ggplot2::element_text(color = "gray40", size = 12),
    plot.caption = ggplot2::element_text(color = "gray50", size = 10),
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    panel.grid.minor = ggplot2::element_blank()
  )
```

---

## Step 13: Summary Statistics

```{r}
#| label: summary-stats

# Calculate summary statistics
stats_summary <- inflation_filtered |>
  dplyr::summarise(
    `Average Inflation` = base::paste0(base::round(base::mean(inflation_rate), 2), "%"),
    `Median Inflation` = base::paste0(base::round(stats::median(inflation_rate), 2), "%"),
    `Std Deviation` = base::paste0(base::round(stats::sd(inflation_rate), 2), "%"),
    `Maximum` = base::paste0(
      base::round(base::max(inflation_rate), 2), "% (", 
      year[inflation_rate == base::max(inflation_rate)], ")"
    ),
    `Minimum` = base::paste0(
      base::round(base::min(inflation_rate), 2), "% (",
      year[inflation_rate == base::min(inflation_rate)], ")"
    ),
    `Years Above 5%` = base::as.character(base::sum(inflation_rate > 5)),
    `Years Below 2%` = base::as.character(base::sum(inflation_rate < 2 & inflation_rate > 0)),
    `Years of Deflation` = base::as.character(base::sum(inflation_rate < 0))
  ) |>
  tidyr::pivot_longer(
    cols = dplyr::everything(),
    names_to = "Statistic",
    values_to = "Value"
  )

stats_summary |>
  flextable::flextable() |>
  flextable::add_header_lines(
    base::paste0("Inflation Summary: ", selected_start_year, " - ", selected_end_year)
  ) |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

---

## Step 14: Decade-by-Decade Comparison

```{r}
#| label: decade-comparison

decade_summary <- inflation_filtered |>
  dplyr::mutate(
    decade = base::paste0(base::floor(year / 10) * 10, "s")
  ) |>
  dplyr::group_by(decade) |>
  dplyr::summarise(
    years = dplyr::n(),
    avg_inflation = base::mean(inflation_rate),
    max_inflation = base::max(inflation_rate),
    min_inflation = base::min(inflation_rate),
    .groups = "drop"
  ) |>
  dplyr::arrange(decade)

decade_summary |>
  dplyr::mutate(
    avg_inflation = base::paste0(base::round(avg_inflation, 2), "%"),
    max_inflation = base::paste0(base::round(max_inflation, 2), "%"),
    min_inflation = base::paste0(base::round(min_inflation, 2), "%")
  ) |>
  dplyr::rename(
    Decade = decade,
    Years = years,
    `Avg Inflation` = avg_inflation,
    `Max Inflation` = max_inflation,
    `Min Inflation` = min_inflation
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Inflation by Decade") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

---

## Step 15: Bar Chart by Decade

```{r}
#| label: plot-decade-bars
#| fig-width: 10
#| fig-height: 6

decade_plot_data <- inflation_filtered |>
  dplyr::mutate(
    decade = base::paste0(base::floor(year / 10) * 10, "s")
  ) |>
  dplyr::group_by(decade) |>
  dplyr::summarise(
    avg_inflation = base::mean(inflation_rate),
    .groups = "drop"
  )

ggplot2::ggplot(
  decade_plot_data,
  ggplot2::aes(x = decade, y = avg_inflation, fill = avg_inflation)
) +
  ggplot2::geom_col(width = 0.7) +
  ggplot2::geom_hline(yintercept = 2, linetype = "dashed", color = "darkgreen", 
                      linewidth = 1) +
  ggplot2::geom_text(
    ggplot2::aes(label = base::paste0(base::round(avg_inflation, 1), "%")),
    vjust = -0.5,
    fontface = "bold",
    size = 4
  ) +
  ggplot2::scale_fill_gradient2(
    low = "steelblue", mid = "gray80", high = "firebrick",
    midpoint = 3
  ) +
  ggplot2::scale_y_continuous(
    labels = function(x) base::paste0(x, "%"),
    expand = ggplot2::expansion(mult = c(0, 0.15))
  ) +
  ggplot2::labs(
    title = "Average Inflation Rate by Decade",
    subtitle = "Green dashed line = Fed's 2% target",
    x = "Decade",
    y = "Average Annual Inflation (%)",
    caption = "Source: FRED"
  ) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", size = 16),
    legend.position = "none"
  )
```

---

## How to Customize

### Change the Year Range

Edit **Step 9**:

```r
selected_start_year <- 1990
selected_end_year <- 2024
```

### Use Specific Years Only

Uncomment in **Step 9**:

```r
specific_years <- c(1970, 1980, 1990, 2000, 2010, 2020)
```

### Refresh Data from FRED

Delete the saved file and re-run:

```r
file.remove(here::here("data_clean", "fred_inflation_data.rds"))
```

### Fetch a Different Series

Change `series_id` in **Step 4**. Other useful series:

- `CPIAUCSL` — CPI index (not inflation rate)
- `CPILFESL` — Core CPI (excludes food & energy)
- `PCEPI` — PCE Price Index (Fed's preferred measure)
- `T10YIE` — Market-implied inflation expectations

---

## References

- **FRED Website**: https://fred.stlouisfed.org/
- **FRED API Documentation**: https://fred.stlouisfed.org/docs/api/fred/
- **fredr R Package**: https://cran.r-project.org/package=fredr
- **Series Used**: https://fred.stlouisfed.org/series/FPCPITOTLZGUSA
