---
title: "Top Employee Compensation Changes"
subtitle: "City of Schertz | Year-over-Year $ and % Changes | FY20 - FY25"
description: "Directors, Managers, and Public Safety Leaders"
authors: 
  - name: "Dan Swart, CPA (ret)"
  - name: "Claude Sonnet 4.5"
date: today
date-format: long
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
    embed-resources: true
    toc: true
    toc-float: true
    fontsize: 16pt
    
  typst:
    toc: true
    fontsize: 14pt
    margin:
      x: 0.75in
      y: 0.75in

execute:
  echo: true
  message: false
  warning: false
  eval: true
  cache: true

params:
  data_updated: !expr format(Sys.Date(), "%B %Y")
---

```{r}
#| label: setup
#| include: false
#| cache: false

# =============================================================================
# SETUP: LOAD PACKAGES AND SET OPTIONS
# =============================================================================
# Purpose: Initialize environment with required packages and global settings
# Note: cache: false ensures setup always runs fresh

# Force dplyr's select/filter to take precedence over other packages
select <- dplyr::select
filter <- dplyr::filter

# Suppress scientific notation for cleaner number display
base::options(scipen = 999)

# Flextable defaults matching Dan's preferred styling
flextable::set_flextable_defaults(
  font.size = 12, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, 
  padding.bottom = 3,
  padding.left = 4, 
  padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = ",",
  na_str = "<na>"
)

# Set seed for reproducibility
base::set.seed(123)
```

## Introduction
This document presents **year-over-year compensation changes** (both dollar amounts and percentages) for the City of Schertz's top-paid employees. This analysis reveals the trajectory of pay increases for leadership positions from FY20 through FY25.

::: {.callout-note}
## Interactive Tables
**Data last updated:** `r params$data_updated`

- **Sort**: Click any column header to sort (e.g., find largest increases)
- **Search**: Use the search box to find specific employees
- **Filter**: Use column filters to narrow results
- **Export**: Use the Copy, CSV, Excel, or Print buttons
:::

---

## Step 1: Load Cached Data
```{r}
#| label: load-cached-data
#| cache: true

# =============================================================================
# LOAD CACHED DATA FROM OMNIBUS ANALYSIS
# =============================================================================
# Purpose: Load the pre-processed all_data object from the main analysis
# Cache: TRUE - data loads once, subsequent renders use cached version

data_path <- here::here("data_clean", "compensation_wide_fy20_fy25.rds")

if (base::file.exists(data_path)) {
  all_data <- base::readRDS(data_path)
  base::cat("Loaded data:", base::nrow(all_data), "rows\n")
  base::cat("Fiscal years:", base::paste(base::unique(all_data$fiscal_year), collapse = ", "), "\n")
} else {
  base::stop(
    "Data file not found at: ", data_path, "\n",
    "Please run the omnibus analysis first to generate the cleaned data."
  )
}
```

---

## Step 2: Filter to Top Employees
```{r}
#| label: filter-top-employees
#| cache: true

# =============================================================================
# FILTER DATA TO TOP EMPLOYEES ONLY
# =============================================================================
# Purpose: Apply leadership pattern to identify top-paid positions

# Pattern for leadership job titles (case-insensitive)
leadership_pattern <- base::paste(
  "city manager", "assistant city manager", "deputy city manager",
  "director", "assistant director", "deputy director",
  "manager", "superintendent",
  "chief", "commander", "captain", "battalion chief", "marshal",
  sep = "|"
)

# Filter to leadership positions
top_employees <- all_data |>
  dplyr::filter(
    stringr::str_detect(
      job_title, 
      stringr::regex(leadership_pattern, ignore_case = TRUE)
    )
  ) |>
  dplyr::select(
    fiscal_year, name, name_std, department, job_title,
    calc_total_compensation
  ) |>
  dplyr::arrange(fiscal_year, dplyr::desc(calc_total_compensation))

base::cat("Top employees identified:", base::nrow(top_employees), "\n")
base::cat("Unique individuals:", dplyr::n_distinct(top_employees$name_std), "\n")
```

---

## Step 3: Calculate Year-over-Year Changes
```{r}
#| label: calculate-yoy-changes
#| cache: true

# =============================================================================
# CALCULATE YEAR-OVER-YEAR CHANGES ($ AND %)
# =============================================================================
# Purpose: Compute dollar change and percent change for each employee's
#          total compensation between consecutive fiscal years
#
# Method:
#   1. Sort by employee and fiscal year
#   2. Use lag() to get prior year's compensation
#   3. Calculate: change_dollars = current - prior
#   4. Calculate: change_pct = (current - prior) / prior * 100
#
# Note: First year for each employee will have NA for changes

yoy_changes <- top_employees |>
  # Sort by employee and fiscal year for proper lag calculation
  dplyr::arrange(name_std, fiscal_year) |>
  # Group by employee to calculate changes within each person
  dplyr::group_by(name_std) |>
  dplyr::mutate(
    # Get prior year's total compensation
    prior_total_comp = dplyr::lag(calc_total_compensation),
    # Calculate dollar change
    change_dollars = calc_total_compensation - prior_total_comp,
    # Calculate percent change (handle divide by zero)
    change_pct = dplyr::if_else(
      prior_total_comp > 0,
      (calc_total_compensation - prior_total_comp) / prior_total_comp * 100,
      NA_real_
    ),
    # Track the fiscal year transition (e.g., "FY20→FY21")
    fy_transition = dplyr::if_else(
      !base::is.na(prior_total_comp),
      base::paste0(dplyr::lag(fiscal_year), "→", fiscal_year),
      NA_character_
    )
  ) |>
  dplyr::ungroup()

# Report calculation results
base::cat("\n=== YOY CHANGE CALCULATIONS ===\n")
base::cat("Total records:", base::nrow(yoy_changes), "\n")
base::cat("Records with changes calculated:", 
          base::sum(!base::is.na(yoy_changes$change_dollars)), "\n")
```

---

## Step 4: Create Wide Format - Dollar Changes
```{r}
#| label: create-dollar-changes-wide
#| cache: true

# =============================================================================
# CREATE WIDE FORMAT: DOLLAR CHANGES BY FISCAL YEAR TRANSITION
# =============================================================================
# Purpose: Pivot so each FY transition (e.g., FY20→FY21) is a column
# Shows dollar increase/decrease for each employee year-over-year

# Get most recent employee info
most_recent_info <- top_employees |>
  dplyr::group_by(name_std) |>
  dplyr::slice_max(order_by = fiscal_year, n = 1) |>
  dplyr::ungroup() |>
  dplyr::select(name_std, name, department, job_title)

# Create wide format for dollar changes
dollar_changes_wide <- yoy_changes |>
  # Remove rows without changes (first year for each employee)
  dplyr::filter(!base::is.na(fy_transition)) |>
  # Select columns for pivot
  dplyr::select(name_std, fy_transition, change_dollars) |>
  # Pivot transitions to columns
  tidyr::pivot_wider(
    names_from = fy_transition,
    values_from = change_dollars,
    values_fill = NA_real_
  ) |>
  # Join employee info
  dplyr::left_join(most_recent_info, by = "name_std") |>
  # Reorder columns
  dplyr::select(
    name, department, job_title,
    dplyr::any_of(c("FY20→FY21", "FY21→FY22", "FY22→FY23", "FY23→FY24", "FY24→FY25"))
  ) |>
  # Sort by most recent change amount
  dplyr::arrange(
    dplyr::desc(
      dplyr::coalesce(`FY24→FY25`, `FY23→FY24`, `FY22→FY23`, `FY21→FY22`, `FY20→FY21`)
    )
  )

base::cat("Dollar changes table:", base::nrow(dollar_changes_wide), "employees\n")
```

---

## Step 5: Create Wide Format - Percent Changes
```{r}
#| label: create-pct-changes-wide
#| cache: true

# =============================================================================
# CREATE WIDE FORMAT: PERCENT CHANGES BY FISCAL YEAR TRANSITION
# =============================================================================
# Purpose: Pivot so each FY transition is a column showing percent change
# Helps identify proportional increases relative to prior compensation

# Create wide format for percent changes
pct_changes_wide <- yoy_changes |>
  dplyr::filter(!base::is.na(fy_transition)) |>
  dplyr::select(name_std, fy_transition, change_pct) |>
  tidyr::pivot_wider(
    names_from = fy_transition,
    values_from = change_pct,
    values_fill = NA_real_
  ) |>
  dplyr::left_join(most_recent_info, by = "name_std") |>
  dplyr::select(
    name, department, job_title,
    dplyr::any_of(c("FY20→FY21", "FY21→FY22", "FY22→FY23", "FY23→FY24", "FY24→FY25"))
  ) |>
  dplyr::arrange(
    dplyr::desc(
      dplyr::coalesce(`FY24→FY25`, `FY23→FY24`, `FY22→FY23`, `FY21→FY22`, `FY20→FY21`)
    )
  )

base::cat("Percent changes table:", base::nrow(pct_changes_wide), "employees\n")
```

---

## Table 2A: Year-over-Year Dollar Changes

::: {.content-visible when-format="html"}
```{r}
#| label: display-dollar-changes-dt
#| cache: false

# =============================================================================
# HTML OUTPUT: DOLLAR CHANGES - INTERACTIVE DT::DATATABLE
# =============================================================================

dollar_changes_wide |>
  dplyr::mutate(
    dplyr::across(
      dplyr::contains("→"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  DT::datatable(
    caption = "Year-over-Year Dollar Change in Total Compensation",
    extensions = "Buttons",
    options = base::list(
      pageLength = 25,
      scrollX = TRUE,
      dom = "Bfrtip",
      buttons = base::list(
        "copy", 
        "csv", 
        "excel",
        base::list(
          extend = "print",
          title = "YOY Dollar Changes - City of Schertz Top Employees"
        )
      )
    ),
    rownames = FALSE,
    filter = "top"
  )
```
:::

::: {.content-visible when-format="typst"}
```{r}
#| label: display-dollar-changes-flextable
#| cache: false

# =============================================================================
# PDF/TYPST OUTPUT: DOLLAR CHANGES - FLEXTABLE
# =============================================================================

dollar_changes_wide |>
  dplyr::mutate(
    dplyr::across(
      dplyr::contains("→"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Year-over-Year Dollar Change in Total Compensation") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::add_footer_lines(
    "Positive = increase | Negative = decrease | Blank = not in position"
  ) |>
  flextable::italic(part = "footer") |>
  flextable::fontsize(size = 10, part = "footer") |>
  ftExtra::colformat_md() |>
  flextable::autofit()
```
:::

---

## Table 2B: Year-over-Year Percent Changes

::: {.content-visible when-format="html"}
```{r}
#| label: display-pct-changes-dt
#| cache: false

# =============================================================================
# HTML OUTPUT: PERCENT CHANGES - INTERACTIVE DT::DATATABLE
# =============================================================================

pct_changes_wide |>
  dplyr::mutate(
    dplyr::across(
      dplyr::contains("→"),
      ~ base::sprintf("%.1f%%", .x)
    )
  ) |>
  DT::datatable(
    caption = "Year-over-Year Percent Change in Total Compensation",
    extensions = "Buttons",
    options = base::list(
      pageLength = 25,
      scrollX = TRUE,
      dom = "Bfrtip",
      buttons = base::list(
        "copy", 
        "csv", 
        "excel",
        base::list(
          extend = "print",
          title = "YOY Percent Changes - City of Schertz Top Employees"
        )
      )
    ),
    rownames = FALSE,
    filter = "top"
  )
```
:::

::: {.content-visible when-format="typst"}
```{r}
#| label: display-pct-changes-flextable
#| cache: false

# =============================================================================
# PDF/TYPST OUTPUT: PERCENT CHANGES - FLEXTABLE
# =============================================================================

pct_changes_wide |>
  dplyr::mutate(
    dplyr::across(
      dplyr::contains("→"),
      ~ base::sprintf("%.1f%%", .x)
    )
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Year-over-Year Percent Change in Total Compensation") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::add_footer_lines(
    "Percent change = (Current - Prior) / Prior × 100 | Blank = not in position"
  ) |>
  flextable::italic(part = "footer") |>
  flextable::fontsize(size = 10, part = "footer") |>
  ftExtra::colformat_md() |>
  flextable::autofit()
```
:::

---

## Summary Statistics
```{r}
#| label: summary-statistics
#| cache: true

# =============================================================================
# SUMMARY STATISTICS FOR YEAR-OVER-YEAR CHANGES
# =============================================================================
# Purpose: Provide aggregate view of compensation changes across all top employees

change_summary <- yoy_changes |>
  dplyr::filter(!base::is.na(change_dollars)) |>
  dplyr::group_by(fy_transition) |>
  dplyr::summarise(
    n_employees = dplyr::n(),
    avg_dollar_change = base::mean(change_dollars, na.rm = TRUE),
    median_dollar_change = stats::median(change_dollars, na.rm = TRUE),
    min_change = base::min(change_dollars, na.rm = TRUE),
    max_change = base::max(change_dollars, na.rm = TRUE),
    avg_pct_change = base::mean(change_pct, na.rm = TRUE),
    .groups = "drop"
  )

change_summary |>
  dplyr::mutate(
    avg_dollar_change = scales::dollar(avg_dollar_change, accuracy = 1),
    median_dollar_change = scales::dollar(median_dollar_change, accuracy = 1),
    min_change = scales::dollar(min_change, accuracy = 1),
    max_change = scales::dollar(max_change, accuracy = 1),
    avg_pct_change = base::sprintf("%.1f%%", avg_pct_change)
  ) |>
  dplyr::rename(
    `FY Transition` = fy_transition,
    `N Employees` = n_employees,
    `Avg $ Change` = avg_dollar_change,
    `Median $ Change` = median_dollar_change,
    `Min Change` = min_change,
    `Max Change` = max_change,
    `Avg % Change` = avg_pct_change
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Summary: Year-over-Year Changes Across All Top Employees") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  ftExtra::colformat_md() |>
  flextable::autofit()
```

---

## Object Reference

| Object Name | Description | Rows | Columns |
|-------------|-------------|------|---------|
| `all_data` | Full compensation dataset (cached) | `r base::nrow(all_data)` | `r base::ncol(all_data)` |
| `top_employees` | Filtered to leadership positions | `r base::nrow(top_employees)` | `r base::ncol(top_employees)` |
| `yoy_changes` | With calculated YOY changes | `r base::nrow(yoy_changes)` | `r base::ncol(yoy_changes)` |
| `dollar_changes_wide` | Wide format: $ changes by FY transition | `r base::nrow(dollar_changes_wide)` | `r base::ncol(dollar_changes_wide)` |
| `pct_changes_wide` | Wide format: % changes by FY transition | `r base::nrow(pct_changes_wide)` | `r base::ncol(pct_changes_wide)` |
| `change_summary` | Aggregate statistics by transition | `r base::nrow(change_summary)` | `r base::ncol(change_summary)` |

---

## Data Notes

- **Dollar Change** = Current Year Total Comp − Prior Year Total Comp
- **Percent Change** = (Current − Prior) / Prior × 100
- **Positive values** = compensation increase; **Negative** = decrease
- **Blank cells** = employee not in position during one or both years of transition

---

*Report generated: `r base::format(base::Sys.time(), "%B %d, %Y at %I:%M %p")`*
*Data source: City of Schertz Employee Compensation Records*
