---
title: "Schertz Compensation Data Pipeline - Explicit Column Version"
subtitle: "Compensation Review Project"
description: "Each earnings component explicitly captured and validated"
authors: 
  - name: "Dan Swart, CPA (ret)"
  - name: "Claude Opus 4.5"
date: today
date-format: long
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: false
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0

editor: source

quarto:
  render:
    cache-refresh: true

execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8

knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false
---

```{r}
#| label: setup
#| include: false

# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(DT.options = list(dom = 'pBlfrti'))
options(shiny.maxRequestSize = 50 * 1024^2)
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 

# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)

# Set global ggplot2 theme (NO theme_minimal)
ggplot2::theme_set(
  ggplot2::theme_bw(base_size = 16) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 16,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 14,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 14,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 14),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 12,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)

# Set seed for reproducibility
base::set.seed(123)
```

## Introduction

This document implements **explicit column handling** for the Schertz compensation data pipeline. Every earnings component is captured distinctly, validated independently, and then summed to verify totals match.

**Key Philosophy**: The spreadsheet structure evolves year-to-year. Rather than assuming columns exist, we:

1.  **Profile each file** to discover exactly which columns are present
2.  **Map each column** to a canonical component name
3.  **Validate each component** was captured correctly
4.  **Sum components** to verify against source totals

**Explicit Compensation Components** (FY23 structure):

| Component               | Excel Column Name        | Goes Into          |
|-------------------------|--------------------------|--------------------|
| Leave Payout            | FY23 LEAVE PAYOUT        | Total Earnings     |
| Regular Earnings        | FY23 REGULAR EARNINGS    | Total Earnings     |
| Overtime Earnings       | FY23 OVERTIME EARNINGS   | Total Earnings     |
| Additional Earnings     | FY23 ADDITIONAL EARNINGS | Total Earnings     |
| Deployment Earnings     | FY23 DEPLOYMENT EARNINGS | Total Earnings     |
| Arbitration/Settlements | ARBITRATION/ SETTLEMENTS | Total Earnings     |
| **Total Earnings**      | FY23 TOTAL EARNINGS      | (calculated)       |
| Additional Benefits     | FY23 ADDITIONAL BENEFITS | Total Compensation |
| **Total Compensation**  | FY23 TOTAL COMPENSATION  | (final)            |

**Validation Formula**:

```         
TOTAL_EARNINGS = LEAVE_PAYOUT + REGULAR_EARNINGS + OVERTIME_EARNINGS + 
                 ADDITIONAL_EARNINGS + DEPLOYMENT_EARNINGS + ARBITRATION_SETTLEMENTS

TOTAL_COMPENSATION = TOTAL_EARNINGS + ADDITIONAL_BENEFITS
```

------------------------------------------------------------------------

## Step 1: Define Explicit Column Registry

This is the **master registry** of all possible column names across all fiscal years. Each column is mapped to a canonical component name.

```{r}
#| label: define-column-registry

# =============================================================================
# EXPLICIT COLUMN REGISTRY
# =============================================================================
# This defines EVERY possible column name for each component across all years.
# Patterns use regex to handle spacing variations (double spaces, etc.)

column_registry <- dplyr::tibble(
  component = c(
    # Identity columns
    "last_name", "first_name", "department", "job_title",
    # Earnings components (summed to total_earnings)
    "leave_payout",
    "regular_earnings",
    "overtime_earnings",
    "additional_earnings",
    "deployment_earnings",
    "arbitration_settlements",
    # Subtotal
    "total_earnings",
    # Benefits
    "additional_benefits",
    # Grand total
    "total_compensation"
  ),
  description = c(
    "Employee last name",
    "Employee first name", 
    "Department assignment",
    "Job title/position",
    "Leave payout amount",
    "Regular/base earnings",
    "Overtime earnings",
    "Additional earnings",
    "Deployment earnings (public safety)",
    "Arbitration/settlement payments",
    "Sum of all earnings components",
    "Additional benefits (insurance, retirement, etc.)",
    "Total compensation (earnings + benefits)"
  ),
  goes_into = c(
    NA, NA, NA, NA,
    "total_earnings",
    "total_earnings",
    "total_earnings",
    "total_earnings",
    "total_earnings",
    "total_earnings",
    "total_compensation",
    "total_compensation",
    NA
  )
)

# Display the registry
column_registry |>
  flextable::flextable() |>
  flextable::add_header_lines("Master Column Registry: All Compensation Components") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

------------------------------------------------------------------------

## Step 2: Define File-Specific Column Mappings

Each fiscal year's file may have different column names. This section defines the **exact mapping** from Excel column names (after `janitor::clean_names()`) to our canonical component names.

```{r}
#| label: define-file-mappings

# =============================================================================
# FILE REGISTRY WITH EXPLICIT COLUMN MAPPINGS
# =============================================================================
# Each file gets its own explicit list of which columns map to which components.
# This handles the evolving spreadsheet structure.

# Patterns for matching column names (after clean_names)
# These use regex to handle variations in spacing

file_mappings <- base::list(
  
  # ----- FY20 Structure (simpler, no FY prefix) -----
  FY20 = base::list(
    filename = "FY20 Earnings.xlsx",
    fiscal_year = "FY20",
    identity_columns = base::list(
      name = "^name$",
      department = "department",
      job_title = "job_title"
    ),
    earnings_columns = base::list(
      regular_earnings = "^regular_earnings$",
      overtime_earnings = "^overtime_earnings$",
      additional_earnings = "^additional_earnings"
    ),
    benefit_columns = base::list(
      # None in FY20
    ),
    total_columns = base::list(
      total_earnings = "^total_earnings$"
    )
  ),
  
  # ----- FY21 Structure -----
  FY21 = base::list(
    filename = "FY21 Earnings.xlsx",
    fiscal_year = "FY21",
    identity_columns = base::list(
      name = "^name$",
      department = "department",
      job_title = "job_title"
    ),
    earnings_columns = base::list(
      regular_earnings = "^regular_earnings$",
      overtime_earnings = "^overtime_earnings$",
      additional_earnings = "^additional_earnings"
    ),
    benefit_columns = base::list(),
    total_columns = base::list(
      total_earnings = "^total_earnings$"
    )
  ),
  
  # ----- FY22 Structure -----
  FY22 = base::list(
    filename = "FY22 Earnings.xlsx",
    fiscal_year = "FY22",
    identity_columns = base::list(
      name = "^name$",
      department = "department",
      job_title = "job_title"
    ),
    earnings_columns = base::list(
      regular_earnings = "^regular_earnings$",
      overtime_earnings = "^overtime_earnings$",
      additional_earnings = "^additional_earnings"
    ),
    benefit_columns = base::list(),
    total_columns = base::list(
      total_earnings = "^total_earnings$"
    )
  ),
  
  # ----- FY23 Structure -----
  FY23 = base::list(
    filename = "FY23Employee_Comp_FY23.xlsx",
    fiscal_year = "FY23",
    identity_columns = base::list(
      last_name = "last_name",
      first_name = "first_name",
      department = "department",
      job_title = "job_title"
    ),
    earnings_columns = base::list(
      leave_payout = "fy23.*leave.*payout",
      regular_earnings = "fy23.*regular.*earnings",
      overtime_earnings = "fy23.*overtime.*earnings",
      additional_earnings = "fy23.*additional.*earnings",
      deployment_earnings = "fy23.*deployment.*earnings",
      arbitration_settlements = "^arbitration"
    ),
    benefit_columns = base::list(
      additional_benefits = "fy23.*additional.*benefits"
    ),
    total_columns = base::list(
      total_earnings = "fy23.*total.*earnings",
      total_compensation = "fy23.*total.*compensation"
    )
  ),
  
  # ----- FY24 Structure (truncated column names) -----
  FY24 = base::list(
    filename = "FY24Employee_Comp_FY24_All.xlsx",
    fiscal_year = "FY24",
    identity_columns = base::list(
      last_name = "last_name",
      first_name = "first_name",
      department = "home_department",
      job_title = "job_title"
    ),
    earnings_columns = base::list(
      leave_payout = "fy24.*leave.*payout",
      regular_earnings = "fy24.*regular.*earnings",
      overtime_earnings = "fy24.*overtime.*earnings",
      additional_earnings = "^fy24_additional$",
      deployment_earnings = "^fy24_deployment$",
      arbitration_settlements = "^arbitration"
    ),
    benefit_columns = base::list(
      additional_benefits = "^fy24_additional_1$"
    ),
    total_columns = base::list(
      total_earnings = "fy24.*total.*earnings",
      total_compensation = "fy24.*total.*compensation"
    )
  )
)
```
------------------------------------------------------------------------

## Step 3: Utility Functions

```{r}
#| label: utility-functions


# =============================================================================
# UTILITY FUNCTIONS
# =============================================================================

#' Parse currency strings to numeric
#' Handles: "$1,234.56", "(1,234.56)", "$1 234.56", etc.
parse_currency <- function(x) {
  if (base::is.numeric(x)) return(x)
  
  # Remove $ and spaces
  cleaned <- stringr::str_remove_all(x, "[$\\s]")
  
  # Handle parentheses as negative
  is_negative <- stringr::str_detect(cleaned, "^\\(.*\\)$")
  cleaned <- stringr::str_remove_all(cleaned, "[()]")
  
  # Remove commas
  cleaned <- stringr::str_remove_all(cleaned, ",")
  
  # Convert to numeric
  result <- base::suppressWarnings(base::as.numeric(cleaned))
  
  # Apply negative sign
  result <- dplyr::if_else(is_negative, -result, result)
  
  return(result)
}


#' Fix invisible Unicode spaces and normalize whitespace
fix_spaces <- function(x) {
  if (!base::is.character(x)) return(x)
  
  # Replace various Unicode spaces with regular space
  x <- stringr::str_replace_all(x, "[\u00A0\u2000-\u200B\u202F\u205F\u3000]", " ")
  
  # Collapse multiple spaces
  x <- stringr::str_squish(x)
  
  return(x)
}


#' Find column by pattern (case-insensitive regex)
find_column <- function(col_names, pattern) {
  if (base::is.null(pattern)) return(NULL)
  
  matches <- stringr::str_subset(
    col_names, 
    stringr::regex(pattern, ignore_case = TRUE)
  )
  
  if (base::length(matches) >= 1) {
    return(matches[1])
  }
  return(NULL)
}


#' Standardize employee name for matching
standardize_name <- function(name) {
  name <- fix_spaces(name)
  name <- stringr::str_to_upper(name)
  name <- stringr::str_remove_all(name, "[^A-Z ]")
  name <- stringr::str_squish(name)
  return(name)
}

```

------------------------------------------------------------------------

## Step 4: Profile Each Excel File

This step reads each source file and documents **exactly which columns exist**. This creates an audit trail showing what was found vs. what was expected.

```{r}
#| label: profile-excel-files

# =============================================================================
# PROFILE EACH EXCEL FILE
# =============================================================================
# Discover which columns actually exist in each file

profile_excel_file <- function(filepath, fiscal_year, mapping) {
  
  # Read the raw Excel file
  raw_df <- readxl::read_excel(filepath)
  original_cols <- base::names(raw_df)
  
  # Apply clean_names
  raw_df <- janitor::clean_names(raw_df)
  cleaned_cols <- base::names(raw_df)
  
  # Build a profile of found columns
  found_columns <- dplyr::tibble(
    fiscal_year = fiscal_year,
    component = base::character(),
    pattern = base::character(),
    found_column = base::character(),
    original_name = base::character(),
    status = base::character()
  )
  
  # Check identity columns
  for (comp_name in base::names(mapping$identity_columns)) {
    pattern <- mapping$identity_columns[[comp_name]]
    found_col <- find_column(cleaned_cols, pattern)
    
    found_columns <- dplyr::bind_rows(found_columns, dplyr::tibble(
      fiscal_year = fiscal_year,
      component = comp_name,
      pattern = pattern,
      found_column = base::ifelse(base::is.null(found_col), NA_character_, found_col),
      original_name = NA_character_,
      status = base::ifelse(base::is.null(found_col), "MISSING", "FOUND")
    ))
  }
  
  # Check earnings columns
  for (comp_name in base::names(mapping$earnings_columns)) {
    pattern <- mapping$earnings_columns[[comp_name]]
    found_col <- find_column(cleaned_cols, pattern)
    
    found_columns <- dplyr::bind_rows(found_columns, dplyr::tibble(
      fiscal_year = fiscal_year,
      component = comp_name,
      pattern = pattern,
      found_column = base::ifelse(base::is.null(found_col), NA_character_, found_col),
      original_name = NA_character_,
      status = base::ifelse(base::is.null(found_col), "NOT_PRESENT", "FOUND")
    ))
  }
  
  # Check benefit columns
  for (comp_name in base::names(mapping$benefit_columns)) {
    pattern <- mapping$benefit_columns[[comp_name]]
    found_col <- find_column(cleaned_cols, pattern)
    
    found_columns <- dplyr::bind_rows(found_columns, dplyr::tibble(
      fiscal_year = fiscal_year,
      component = comp_name,
      pattern = pattern,
      found_column = base::ifelse(base::is.null(found_col), NA_character_, found_col),
      original_name = NA_character_,
      status = base::ifelse(base::is.null(found_col), "NOT_PRESENT", "FOUND")
    ))
  }
  
  # Check total columns
  for (comp_name in base::names(mapping$total_columns)) {
    pattern <- mapping$total_columns[[comp_name]]
    found_col <- find_column(cleaned_cols, pattern)
    
    found_columns <- dplyr::bind_rows(found_columns, dplyr::tibble(
      fiscal_year = fiscal_year,
      component = comp_name,
      pattern = pattern,
      found_column = base::ifelse(base::is.null(found_col), NA_character_, found_col),
      original_name = NA_character_,
      status = base::ifelse(base::is.null(found_col), "MISSING", "FOUND")
    ))
  }
  
  base::list(
    fiscal_year = fiscal_year,
    row_count = base::nrow(raw_df),
    original_columns = original_cols,
    cleaned_columns = cleaned_cols,
    column_profile = found_columns,
    raw_data = raw_df
  )
}

# Profile all files
file_profiles <- base::list()

for (fy in base::names(file_mappings)) {
  mapping <- file_mappings[[fy]]
  filepath <- base::file.path(paths$data_raw, mapping$filename)
  
  if (base::file.exists(filepath)) {
    file_profiles[[fy]] <- profile_excel_file(filepath, fy, mapping)
    base::cat("Profiled:", fy, "-", file_profiles[[fy]]$row_count, "rows,",
              base::length(file_profiles[[fy]]$cleaned_columns), "columns\n")
  } else {
    base::cat("WARNING: File not found:", filepath, "\n")
  }
}
```

```{r}
#| label: display-column-profiles

# Combine all profiles into one display table
all_profiles <- purrr::map_dfr(file_profiles, ~ .x$column_profile)

# Show which components were found in each year
all_profiles |>
  dplyr::select(fiscal_year, component, found_column, status) |>
  tidyr::pivot_wider(
    names_from = fiscal_year,
    values_from = c(found_column, status),
    names_glue = "{fiscal_year}_{.value}"
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Column Discovery: What Was Found in Each File") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

------------------------------------------------------------------------

## Step 5: Extract Explicit Component Totals from Each File

This is the **critical validation step**. We extract each component column's total directly from the Excel file, then verify our formulas match.

```{r}
#| label: extract-component-totals

# =============================================================================
# EXTRACT EXPLICIT COMPONENT TOTALS FROM EACH FILE
# =============================================================================

extract_component_totals <- function(profile, mapping) {
  
  df <- profile$raw_data
  col_names <- profile$cleaned_columns
  fy <- profile$fiscal_year
  
  # Initialize all components to 0
  totals <- dplyr::tibble(
    fiscal_year = fy,
    leave_payout = 0,
    regular_earnings = 0,
    overtime_earnings = 0,
    additional_earnings = 0,
    deployment_earnings = 0,
    arbitration_settlements = 0,
    calculated_total_earnings = 0,
    source_total_earnings = 0,
    additional_benefits = 0,
    calculated_total_compensation = 0,
    source_total_compensation = 0
  )
  
  # Extract each earnings component
  for (comp_name in base::names(mapping$earnings_columns)) {
    pattern <- mapping$earnings_columns[[comp_name]]
    found_col <- find_column(col_names, pattern)
    
    if (!base::is.null(found_col)) {
      col_values <- parse_currency(df[[found_col]])
      totals[[comp_name]] <- base::sum(col_values, na.rm = TRUE)
    }
  }
  
  # Calculate total earnings from components
  totals$calculated_total_earnings <- (
    totals$leave_payout +
    totals$regular_earnings +
    totals$overtime_earnings +
    totals$additional_earnings +
    totals$deployment_earnings +
    totals$arbitration_settlements
  )
  
  # Extract source total_earnings if present
  te_pattern <- mapping$total_columns$total_earnings
  te_col <- find_column(col_names, te_pattern)
  if (!base::is.null(te_col)) {
    totals$source_total_earnings <- base::sum(parse_currency(df[[te_col]]), na.rm = TRUE)
  }
  
  # Extract additional benefits
  for (comp_name in base::names(mapping$benefit_columns)) {
    pattern <- mapping$benefit_columns[[comp_name]]
    found_col <- find_column(col_names, pattern)
    
    if (!base::is.null(found_col)) {
      totals[[comp_name]] <- base::sum(parse_currency(df[[found_col]]), na.rm = TRUE)
    }
  }
  
  # Calculate total compensation
  totals$calculated_total_compensation <- totals$calculated_total_earnings + totals$additional_benefits
  
  # Extract source total_compensation
  tc_pattern <- mapping$total_columns$total_compensation
  tc_col <- find_column(col_names, tc_pattern)
  if (!base::is.null(tc_col)) {
    totals$source_total_compensation <- base::sum(parse_currency(df[[tc_col]]), na.rm = TRUE)
  }
  
  return(totals)
}

# Extract totals from all files
source_component_totals <- purrr::map_dfr(
  base::names(file_profiles),
  function(fy) {
    extract_component_totals(file_profiles[[fy]], file_mappings[[fy]])
  }
)
```

```{r}
#| label: display-component-totals

# Display the explicit component breakdown
source_component_totals |>
  dplyr::mutate(
    dplyr::across(
      dplyr::where(is.numeric),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  tidyr::pivot_longer(
    cols = -fiscal_year,
    names_to = "component",
    values_to = "amount"
  ) |>
  tidyr::pivot_wider(
    names_from = fiscal_year,
    values_from = amount
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Explicit Component Totals: Direct from Excel Files") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

------------------------------------------------------------------------

## Step 6: Checkpoint G - Component Validation

This checkpoint verifies that: 1. Our calculated totals match the source file's totals 2. Each component we expected to find was actually found

```{r}
#| label: checkpoint-g-component-validation

# =============================================================================
# CHECKPOINT G: COMPONENT-LEVEL VALIDATION
# =============================================================================

component_validation <- source_component_totals |>
  dplyr::mutate(
    # Compare calculated vs source
    earnings_diff = calculated_total_earnings - source_total_earnings,
    earnings_match = base::abs(earnings_diff) <= 1.00,
    
    comp_diff = calculated_total_compensation - source_total_compensation,
    comp_match = base::abs(comp_diff) <= 1.00,
    
    # Overall pass
    all_match = earnings_match & comp_match
  )

# Display validation results
component_validation |>
  dplyr::select(
    fiscal_year,
    calculated_total_earnings,
    source_total_earnings,
    earnings_diff,
    earnings_match,
    calculated_total_compensation,
    source_total_compensation,
    comp_diff,
    comp_match
  ) |>
  dplyr::mutate(
    calculated_total_earnings = scales::dollar(calculated_total_earnings, accuracy = 1),
    source_total_earnings = scales::dollar(source_total_earnings, accuracy = 1),
    earnings_diff = scales::dollar(earnings_diff, accuracy = 0.01),
    calculated_total_compensation = scales::dollar(calculated_total_compensation, accuracy = 1),
    source_total_compensation = scales::dollar(source_total_compensation, accuracy = 1),
    comp_diff = scales::dollar(comp_diff, accuracy = 0.01),
    earnings_status = dplyr::if_else(earnings_match, "PASS", "FAIL"),
    comp_status = dplyr::if_else(comp_match, "PASS", "FAIL")
  ) |>
  dplyr::select(-earnings_match, -comp_match) |>
  flextable::flextable() |>
  flextable::add_header_lines("Checkpoint G: Component-Level Validation") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::bg(i = ~ earnings_status == "FAIL", j = "earnings_status", bg = "#ffcccc") |>
  flextable::bg(i = ~ earnings_status == "PASS", j = "earnings_status", bg = "#ccffcc") |>
  flextable::bg(i = ~ comp_status == "FAIL", j = "comp_status", bg = "#ffcccc") |>
  flextable::bg(i = ~ comp_status == "PASS", j = "comp_status", bg = "#ccffcc") |>
  flextable::autofit()
```

```{r}
#| label: checkpoint-g-interpretation

# Report on validation status
checkpoint_g_passed <- base::all(component_validation$all_match)

if (!checkpoint_g_passed) {
  base::cat("\n=== CHECKPOINT G FAILED ===\n\n")
  
  failures <- component_validation |>
    dplyr::filter(!all_match)
  
  for (i in base::seq_len(base::nrow(failures))) {
    row <- failures[i, ]
    base::cat("MISMATCH in", row$fiscal_year, ":\n")
    
    if (!row$earnings_match) {
      base::cat("  - Total Earnings: Calculated", 
                scales::dollar(row$calculated_total_earnings),
                "vs Source", scales::dollar(row$source_total_earnings),
                "(diff:", scales::dollar(row$earnings_diff), ")\n")
    }
    
    if (!row$comp_match) {
      base::cat("  - Total Compensation: Calculated",
                scales::dollar(row$calculated_total_compensation),
                "vs Source", scales::dollar(row$source_total_compensation),
                "(diff:", scales::dollar(row$comp_diff), ")\n")
    }
  }
  
  base::cat("\nACTION REQUIRED:\n")
  base::cat("1. Check the file_mappings for missing column patterns\n")
  base::cat("2. Review the column profiles above to see what columns exist\n")
  base::cat("3. Update the mapping patterns to capture all earnings columns\n")
  
} else {
  base::cat("Checkpoint G: PASSED\n")
  base::cat("All calculated totals match source file totals within $1.00\n")
}
```

------------------------------------------------------------------------

## Step 7: Process Files into Long Format

Now that we've validated the components, we process each file into a tidy long format where each row represents one employee-year-component combination.

```{r}
#| label: process-to-long-format

# =============================================================================
# PROCESS FILES INTO LONG FORMAT
# =============================================================================

process_file_to_long <- function(profile, mapping, name_corrections = NULL) {
  
  df <- profile$raw_data
  col_names <- profile$cleaned_columns
  fy <- profile$fiscal_year
  
  # Detect identity column structure
  # FY20-22 have single "name" column; FY23-24 have last_name/first_name
  has_single_name <- "name" %in% base::names(mapping$identity_columns)
  
  # Extract identity columns
  dept_col <- find_column(col_names, mapping$identity_columns$department)
  title_col <- find_column(col_names, mapping$identity_columns$job_title)
  
  if (has_single_name) {
    # FY20-22 structure: single name column
    name_col <- find_column(col_names, mapping$identity_columns$name)
    
    result <- dplyr::tibble(
      fiscal_year = fy,
      name = fix_spaces(df[[name_col]]),
      department = fix_spaces(df[[dept_col]]),
      job_title = fix_spaces(df[[title_col]])
    ) |>
      dplyr::mutate(
        last_name = NA_character_,
        first_name = NA_character_,
        name_original = name,
        name_std = standardize_name(name)
      )
    
  } else {
    # FY23-24 structure: separate last_name/first_name
    last_name_col <- find_column(col_names, mapping$identity_columns$last_name)
    first_name_col <- find_column(col_names, mapping$identity_columns$first_name)
    
    result <- dplyr::tibble(
      fiscal_year = fy,
      last_name = fix_spaces(df[[last_name_col]]),
      first_name = fix_spaces(df[[first_name_col]]),
      department = fix_spaces(df[[dept_col]]),
      job_title = fix_spaces(df[[title_col]])
    ) |>
      dplyr::mutate(
        name = base::paste(last_name, first_name, sep = ", "),
        name_original = name,
        name_std = standardize_name(name)
      )
  }
  
  # Apply name corrections if provided
  if (!base::is.null(name_corrections) && base::nrow(name_corrections) > 0) {
    for (i in base::seq_len(base::nrow(name_corrections))) {
      wrong <- name_corrections$wrong_name[i]
      correct <- name_corrections$correct_name[i]
      result <- result |>
        dplyr::mutate(
          name = dplyr::if_else(name == wrong, correct, name),
          name_std = standardize_name(name)
        )
    }
  }
  
  # Create employee key
  result <- result |>
    dplyr::mutate(
      employee_key = base::paste(name_std, department, sep = "_")
    )
  
  # Extract each earnings component (with safe fallback to 0)
  earnings_wide <- dplyr::tibble(.rows = base::nrow(df))
  
  # Leave payout
  col <- find_column(col_names, mapping$earnings_columns$leave_payout)
  earnings_wide$leave_payout <- if (!base::is.null(col)) parse_currency(df[[col]]) else 0
  
  # Regular earnings
  col <- find_column(col_names, mapping$earnings_columns$regular_earnings)
  earnings_wide$regular_earnings <- if (!base::is.null(col)) parse_currency(df[[col]]) else 0
  
  # Overtime earnings
  col <- find_column(col_names, mapping$earnings_columns$overtime_earnings)
  earnings_wide$overtime_earnings <- if (!base::is.null(col)) parse_currency(df[[col]]) else 0
  
  # Additional earnings
  col <- find_column(col_names, mapping$earnings_columns$additional_earnings)
  earnings_wide$additional_earnings <- if (!base::is.null(col)) parse_currency(df[[col]]) else 0
  
  # Deployment earnings
  col <- find_column(col_names, mapping$earnings_columns$deployment_earnings)
  earnings_wide$deployment_earnings <- if (!base::is.null(col)) parse_currency(df[[col]]) else 0
  
  # Arbitration/settlements
  col <- find_column(col_names, mapping$earnings_columns$arbitration_settlements)
  earnings_wide$arbitration_settlements <- if (!base::is.null(col)) parse_currency(df[[col]]) else 0
  
  # Additional benefits
  col <- find_column(col_names, mapping$benefit_columns$additional_benefits)
  earnings_wide$additional_benefits <- if (!base::is.null(col)) parse_currency(df[[col]]) else 0
  
  # Replace NA with 0
  earnings_wide <- earnings_wide |>
    dplyr::mutate(dplyr::across(dplyr::everything(), ~ dplyr::coalesce(.x, 0)))
  
  # Calculate totals
  earnings_wide <- earnings_wide |>
    dplyr::mutate(
      total_earnings = leave_payout + regular_earnings + overtime_earnings +
                       additional_earnings + deployment_earnings + arbitration_settlements,
      total_compensation = total_earnings + additional_benefits
    )
  
  # Combine identity and earnings
  result <- dplyr::bind_cols(result, earnings_wide)
  
  # Pivot to long format
  result_long <- result |>
    tidyr::pivot_longer(
      cols = c(leave_payout, regular_earnings, overtime_earnings,
               additional_earnings, deployment_earnings, arbitration_settlements,
               additional_benefits),
      names_to = "earnings_type",
      values_to = "amount"
    ) |>
    dplyr::mutate(
      earnings_type = dplyr::case_when(
        earnings_type == "leave_payout" ~ "Leave Payout",
        earnings_type == "regular_earnings" ~ "Regular",
        earnings_type == "overtime_earnings" ~ "Overtime",
        earnings_type == "additional_earnings" ~ "Additional",
        earnings_type == "deployment_earnings" ~ "Deployment",
        earnings_type == "arbitration_settlements" ~ "Arbitration",
        earnings_type == "additional_benefits" ~ "Benefits",
        TRUE ~ earnings_type
      )
    )
  
  return(result_long)
}
```

------------------------------------------------------------------------

## Step 8: Validate Pipeline Output Against Source

```{r}
#| label: validate-pipeline-output

# =============================================================================
# VALIDATE PIPELINE OUTPUT
# =============================================================================

# Calculate totals from pipeline output
pipeline_totals <- final_df |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    pipeline_leave = base::sum(amount[earnings_type == "Leave Payout"], na.rm = TRUE),
    pipeline_regular = base::sum(amount[earnings_type == "Regular"], na.rm = TRUE),
    pipeline_overtime = base::sum(amount[earnings_type == "Overtime"], na.rm = TRUE),
    pipeline_additional = base::sum(amount[earnings_type == "Additional"], na.rm = TRUE),
    pipeline_deployment = base::sum(amount[earnings_type == "Deployment"], na.rm = TRUE),
    pipeline_arbitration = base::sum(amount[earnings_type == "Arbitration"], na.rm = TRUE),
    pipeline_benefits = base::sum(amount[earnings_type == "Benefits"], na.rm = TRUE),
    pipeline_total = base::sum(amount, na.rm = TRUE),
    .groups = "drop"
  )

# Compare to source totals
validation <- source_component_totals |>
  dplyr::left_join(pipeline_totals, by = "fiscal_year") |>
  dplyr::mutate(
    total_diff = pipeline_total - source_total_compensation,
    status = dplyr::if_else(base::abs(total_diff) <= 1.00, "PASS", "FAIL")
  )

validation |>
  dplyr::select(fiscal_year, source_total_compensation, pipeline_total, total_diff, status) |>
  dplyr::mutate(
    source_total_compensation = scales::dollar(source_total_compensation, accuracy = 1),
    pipeline_total = scales::dollar(pipeline_total, accuracy = 1),
    total_diff = scales::dollar(total_diff, accuracy = 0.01)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Pipeline Validation: Source vs Output Totals") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::bg(i = ~ status == "FAIL", bg = "#ffcccc") |>
  flextable::bg(i = ~ status == "PASS", bg = "#ccffcc") |>
  flextable::autofit()
```

------------------------------------------------------------------------

## Step 9: Checkpoint H - Explicit Component Breakdown

This checkpoint shows exactly how much of each component was captured, proving nothing was missed.

```{r}
#| label: checkpoint-h-component-breakdown

# =============================================================================
# CHECKPOINT H: EXPLICIT COMPONENT BREAKDOWN
# =============================================================================

# Create a detailed breakdown showing each component's contribution
component_breakdown <- final_df |>
  dplyr::group_by(fiscal_year, earnings_type) |>
  dplyr::summarise(
    amount = base::sum(amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(
    names_from = earnings_type,
    values_from = amount,
    values_fill = 0
  )

# Add totals row
component_breakdown <- component_breakdown |>
  dplyr::mutate(
    `Total Earnings` = `Leave Payout` + Regular + Overtime + Additional + Deployment + Arbitration,
    `Total Compensation` = `Total Earnings` + Benefits
  )

# Display
component_breakdown |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.numeric), ~ scales::dollar(.x, accuracy = 1))
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Checkpoint H: Explicit Component Breakdown by Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

------------------------------------------------------------------------

## Step 10: Summary Statistics

```{r}
#| label: summary-statistics

# Summary by year
yearly_summary <- final_df |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    total_compensation = base::sum(amount, na.rm = TRUE),
    total_employees = dplyr::n_distinct(employee_key),
    avg_compensation = total_compensation / total_employees,
    .groups = "drop"
  ) |>
  dplyr::arrange(fiscal_year)

yearly_summary |>
  dplyr::mutate(
    total_compensation = scales::dollar(total_compensation, accuracy = 1),
    avg_compensation = scales::dollar(avg_compensation, accuracy = 1)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Yearly Summary: Total Compensation and Headcount") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

------------------------------------------------------------------------

## Step 11: Export Data

```{r}
#| label: export-data

# Ensure output directory exists
if (!base::dir.exists(paths$data_clean)) {
  base::dir.create(paths$data_clean, recursive = TRUE)
}

# Export
test_suffix <- base::ifelse(report_settings$test_mode, "_TEST", "")

rds_path <- base::file.path(paths$data_clean, base::paste0("compensation_explicit", test_suffix, ".rds"))
base::saveRDS(final_df, file = rds_path)

csv_path <- base::file.path(paths$data_clean, base::paste0("compensation_explicit", test_suffix, ".csv"))
readr::write_csv(final_df, file = csv_path)

base::cat("Data exported:\n")
base::cat("  -", rds_path, "\n")
base::cat("  -", csv_path, "\n")
```

------------------------------------------------------------------------

## Appendix: Column Structure Comparison

This shows exactly what columns exist in each year's file, helping identify when the spreadsheet structure changes.

```{r}
#| label: column-structure-comparison

# Build a comparison of actual column names across years
col_comparison <- purrr::map_dfr(
  base::names(file_profiles),
  function(fy) {
    dplyr::tibble(
      fiscal_year = fy,
      column_name = file_profiles[[fy]]$cleaned_columns
    )
  }
) |>
  dplyr::mutate(present = TRUE) |>
  tidyr::pivot_wider(
    names_from = fiscal_year,
    values_from = present,
    values_fill = FALSE
  )

col_comparison |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.logical), ~ dplyr::if_else(.x, "✓", ""))
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Column Presence by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

------------------------------------------------------------------------

## How to Add a New Fiscal Year

When you receive a new fiscal year's file:

1.  **Profile the file** to see what columns exist:

``` r
new_file <- readxl::read_excel("path/to/new_file.xlsx")
base::names(janitor::clean_names(new_file))
```

2.  **Add a new entry to `file_mappings`** with the appropriate column patterns

3.  **Run the pipeline** - Checkpoint G will tell you if any columns are missing

4.  **Update patterns** if the column names have changed

This explicit approach means you'll never silently lose data when the spreadsheet structure changes.
