---
title: "Schertz Sample Data Visualizations"
subtitle: ""
description: ""
authors: 
  - name: "Dan Swart, CPA (ret)"
  - name: "Claude Sonnet 4.5"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: false     # Fast for drafts. Override for sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    # fig-width: 13
    E fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: false
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: false
    error: false
    warning: false
    message: false
    cache: false


---


```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED


# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)


# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 16,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 14,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 15, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 14,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 14),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 12,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)


# Set seed for reproducibility
base::set.seed(123)

```



### Data

```{r}
final_df_long <- readRDS(here::here("data_clean", "compensation_long_fy20_fy25.rds"))
```

#### Long format is best for:

* Bar charts comparing earnings types across years
* Faceted plots by earnings_type
* Grouped summaries (total by year, by department, by earnings type)
* Most ggplot2 work

#### Wide format is needed for:

* Calculating total compensation per employee (summing across columns)
* Tables showing one row per employee with all earnings side-by-side
* Scatter plots comparing two earnings types against each other
* Validation checks (comparing calculated vs source totals)

Pivot on the fly when needed


#### Load Long; Create Wide When Needed

```{r}
# Long to wide 
final_df_wide <- final_df_long |>
  tidyr::pivot_wider(
    names_from = earnings_type,
    values_from = amount,
    values_fill = 0
  ) |>
  dplyr::mutate(
    total_reg_earnings = Leave + Regular + Overtime + Additional + Deployment + Arbitration,
    total_compensation = total_reg_earnings + Benefits
  )
```



## Summary Statistics - Total Compensation

```{r}
#| label: summary-total-comp

# =============================================================================
# SUMMARY STATISTICS 
# =============================================================================

# Create small dataframe of total comp by year
# group_by(fiscal_year) — splits the dataframe into chunks, one per fiscal year
# summarise(...) — reduces each chunk to one row with your computed values
# Result: one row per fiscal year instead of one row per original observation

yearly_summary_total_comp <- final_df_long |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    total_compensation = base::sum(amount, na.rm = TRUE),
    total_employees = dplyr::n_distinct(employee_key),
    avg_total_compensation = total_compensation / total_employees,
    .groups = "drop"
  ) |>
  dplyr::arrange(fiscal_year)

# Format columns
yearly_summary_total_comp |>
 dplyr::mutate(
   total_compensation = scales::dollar(total_compensation, accuracy = 1),
   avg_total_compensation = scales::dollar(avg_total_compensation, accuracy = 1)
   ) |>

  flextable::flextable() |>
  flextable::add_header_lines("Yearly Summary: Total Compensation and Headcount") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```


## Summary Statistics - Regular Compensation

```{r}
#| label: summary-regular-comp

# =============================================================================
# SUMMARY STATISTICS 
# =============================================================================

# Summaries by year
yearly_summary_reg_comp <- final_df_long |>
  dplyr::filter(earnings_type != "Benefits") |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    total_reg_comp = base::sum(amount, na.rm = TRUE),
    total_employees = dplyr::n_distinct(employee_key),
    avg_reg_comp = total_reg_comp / total_employees,
    .groups = "drop"
  ) |>
  dplyr::arrange(fiscal_year)

# Format certain columns
yearly_summary_reg_comp |>
 dplyr::mutate(
   total_reg_comp = scales::dollar(total_reg_comp, accuracy = 1),
   avg_reg_comp = scales::dollar(avg_reg_comp, accuracy = 1)
   ) |>

  flextable::flextable() |>
  flextable::add_header_lines("Yearly Summary: All Comp, Excluding Benefits") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```


### Combine both Tables for Side-by-Side Comparison

```{r}
#| label: summaries-combined

# =============================================================================
# COMBINED SUMMARY TABLE
# =============================================================================
yearly_summaries_combined <- dplyr::left_join(
  yearly_summary_total_comp,
  yearly_summary_reg_comp |> dplyr::select(-total_employees),
  by = "fiscal_year"
) |>
  dplyr::mutate(
    total_compensation = scales::dollar(total_compensation, accuracy = 1),
    avg_total_compensation = scales::dollar(avg_total_compensation, accuracy = 1),
    total_reg_comp = scales::dollar(total_reg_comp, accuracy = 1),
    avg_reg_comp = scales::dollar(avg_reg_comp, accuracy = 1)
  ) |>
  dplyr::rename(
    Year = fiscal_year,
    `Total Comp` = total_compensation,
    EEs = total_employees,
    `Avg Tot Comp` = avg_total_compensation,
    `Total Reg` = total_reg_comp,
    `Avg Reg Pay` = avg_reg_comp
  )

yearly_summaries_combined |>
  flextable::flextable() |>
  flextable::add_header_lines("Yearly Summaris: Total vs Regular Compensation") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```



## City Manager Total Compensation Over Time

```{r}
#| label: plot-city-manager-total-comp
#| fig-width: 13
#| fig-height: 8

# =============================================================================
# APPENDIX H: CITY MANAGER TOTAL COMPENSATION OVER TIME
# =============================================================================


# City Manager TOTAL COMPENSATION by Fiscal Year
# Total compensation = Leave + Regular + Overtime + Additional + Deployment + Arbitration + Benefits

# If you've already recoded earnings_type to pretty labels (e.g., "Regular"),
# these are the labels to include:
comp_types <- c(
  "Leave Payout", "Regular", "Overtime", "Additional",
  "Deployment", "Arbitration", "Benefits"
)

city_manager_data <- final_df_long |>
  dplyr::filter(
    job_title == "City Manager",
    earnings_type %in% comp_types
  ) |>
  dplyr::group_by(fiscal_year, name) |>
  dplyr::summarise(
    total_compensation = sum(amount, na.rm = TRUE),
    .groups = "drop"
  )

if (base::nrow(city_manager_data) > 0) {

  ggplot2::ggplot(
    city_manager_data,
    ggplot2::aes(x = fiscal_year, y = total_compensation, group = name, color = name)
  ) +
    ggplot2::geom_line(linewidth = 1) +
    ggplot2::geom_point(size = 3) +

    # Value labels ABOVE points
    ggplot2::geom_text(
      ggplot2::aes(label = scales::dollar(total_compensation, accuracy = 1)),
      vjust = -1.5,
      size = 4.5,
      color = "black",
      show.legend = FALSE
    ) +

    # Keep headroom for labels
    ggplot2::scale_y_continuous(
      expand = ggplot2::expansion(mult = c(0, 0.15))
    ) +

    ggplot2::labs(
      title = "City Manager - Total Compensation by Fiscal Year",
      subtitle = "City of Schertz",
      x = "Fiscal Year",
      y = NULL,
      color = "Employee"
    ) +

    # Remove y-axis clutter
    ggplot2::theme(
      legend.position = "bottom",
      axis.text.y  = ggplot2::element_blank(),
      axis.ticks.y = ggplot2::element_blank(),
      panel.grid.major.y = ggplot2::element_blank()
    )

} else {
  base::cat("No City Manager records found in the data.\n")
}

```




------------------------------------------------------------------------


## Total Compensation Growth by Year

```{r}
#| label: total-comp-growth-by-year
#| fig-width: 10
#| fig-height: 6

# =============================================================================
# APPENDIX I: TOTAL COMPENSATION GROWTH BY FISCAL YEAR - ALL DEPTS
# =============================================================================

ggplot2::ggplot(yearly_summary_total_comp, 
                ggplot2::aes(x = fiscal_year, y = total_compensation, group = 1)) +
  ggplot2::geom_line(linewidth = 1.5, color = "steelblue") +
  ggplot2::geom_point(size = 4, color = "steelblue") +
  ggplot2::geom_text(
    ggplot2::aes(label = scales::dollar(total_compensation, scale = 1e-6, suffix = "M")),
    vjust = -2
  ) +
  
  ggplot2::theme(
    axis.text.y  = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    panel.grid.major.y = ggplot2::element_blank()
  ) +

  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = c(0.05, 0.25)
                                )
    ) +

  ggplot2::labs(
    title = "Total Employee Compensation by Fiscal Year - All Depts",
    subtitle = "City of Schertz",
    x = "Fiscal Year",
    y = "Total Compensation"
  )

```


------------------------------------------------------------------------







## Year-over-Year Comparison

```{r}
#| label: yoy-comparison

# =============================================================================
# APPENDIX C: YEAR-OVER-YEAR COMPARISON OF EE TOTALS, COMPENSATION, PCT, AVG COMP
# =============================================================================

yoy_summary <- yearly_summary_total_comp |>
  dplyr::mutate(
    prev_total = dplyr::lag(total_compensation),
    prev_employees = dplyr::lag(total_employees),
    total_change = total_compensation - prev_total,
    total_pct_change = (total_change / prev_total) * 100,
    employee_change = total_employees - prev_employees
  )

yoy_summary |>
  dplyr::mutate(
    total_compensation = scales::dollar(total_compensation, accuracy = 1),
    avg_compensation = scales::dollar(avg_total_compensation, accuracy = 1),
    total_change = scales::dollar(total_change, accuracy = 1),
    total_pct_change = base::sprintf("%.1f%%", total_pct_change)
  ) |>
  dplyr::select(
    fiscal_year, total_employees, employee_change,
    total_compensation, total_change, total_pct_change, avg_total_compensation
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Year-over-Year Comparison") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

```{r}
yoy_summary <- yearly_summary_total_comp |>
  dplyr::mutate(
    prev_total = dplyr::lag(total_compensation),
    prev_employees = dplyr::lag(total_employees),
    total_change = total_compensation - prev_total,
    total_pct_change = (total_change / prev_total) * 100,
    employee_change = total_employees - prev_employees
  )

yoy_summary |>
  dplyr::mutate(
    total_compensation = scales::dollar(total_compensation, accuracy = 1),
    avg_total_compensation = scales::dollar(avg_total_compensation, accuracy = 1),
    total_change = scales::dollar(total_change, accuracy = 1),
    total_pct_change = base::sprintf("%.1f%%", total_pct_change)
  ) |>
  dplyr::select(
    fiscal_year, total_employees, employee_change,
    total_compensation, total_change, total_pct_change, avg_total_compensation
  ) |>
  dplyr::rename(
    Year = fiscal_year,
    EEs = total_employees,
    `EE Chg` = employee_change,
    `Total Comp` = total_compensation,
    `Change` = total_change,
    `Chg %` = total_pct_change,
    `Avg Total Comp` = avg_total_compensation
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Year-over-Year Comparison") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```


------------------------------------------------------------------------


## Top 10 Earners by Year

```{r}
#| label: top-10--in-total-comp-per-year

# =============================================================================
# APPENDIX D: TOP 10 IN TOTAL COMPENSATION BY FY
# =============================================================================

top_10_earners <- final_df_wide |>
  dplyr::group_by(fiscal_year) |>
  dplyr::slice_max(order_by = total_compensation, n = 10) |>
  dplyr::select(
    fiscal_year, name, department, job_title, total_compensation
  ) |>
  dplyr::mutate(
    rank = dplyr::row_number()
  ) |>
  dplyr::ungroup()


top_10_earners |>
  dplyr::mutate(
    total_compensation = scales::dollar(total_compensation, accuracy = 1)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Top 10 in Total Compensation for Each Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```


------------------------------------------------------------------------


## Department Summary

```{r}
#| label: department-summary

# =============================================================================
# APPENDIX E: DEPARTMENT SUMMARY
# =============================================================================

dept_summary <- final_df_wide |>
  dplyr::group_by(fiscal_year, department) |>
  dplyr::summarise(
    employees = dplyr::n(),
    total_comp = base::sum(total_compensation, na.rm = TRUE),
    avg_comp = base::mean(total_compensation, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(fiscal_year, dplyr::desc(total_comp))

# Show FY25 as example
dept_summary |>
  dplyr::filter(fiscal_year == "FY25") |>
  dplyr::mutate(
    total_comp = scales::dollar(total_comp, accuracy = 1),
    avg_comp = scales::dollar(avg_comp, accuracy = 1)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Department Summary - FY25") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```


------------------------------------------------------------------------


## Overtime Analysis by Fiscal Year

```{r}
#| label: overtime-analysis-by-fy

# =============================================================================
# APPENDIX F: OVERTIME ANALYSIS BY FY
# =============================================================================

overtime_summary <- final_df_wide |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    total_overtime = base::sum(Overtime, na.rm = TRUE),
    total_regular = base::sum(Regular, na.rm = TRUE),
    overtime_pct = (total_overtime / total_regular) * 100,
    employees_with_ot = base::sum(Overtime > 0),
    total_employees = dplyr::n(),
    pct_with_ot = (employees_with_ot / total_employees) * 100,
    .groups = "drop"
  )

overtime_summary |>
  dplyr::mutate(
    total_overtime = scales::dollar(total_overtime, accuracy = 1),
    total_regular = scales::dollar(total_regular, accuracy = 1),
    overtime_pct = base::sprintf("%.1f%%", overtime_pct),
    pct_with_ot = base::sprintf("%.1f%%", pct_with_ot)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Overtime Analysis by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```



# OVERTIME BY DEPARTMENT - TOP 3 

```{r}
#| label: fig-overtime-by-dept
#| fig-cap: "Top 3 Departments by Overtime Usage"

# ======================================================================
# OVERTIME BY DEPARTMENT - TOP 3 LINE CHART
# ======================================================================

# STEP 1: Identify the top 3 departments by total overtime across all years
# We need to know which departments to include before we can chart them

top_3_depts <- final_df_wide |>
  # Group by department to aggregate across all fiscal years
  dplyr::group_by(department) |>
  # Calculate total overtime for each department (all years combined)
  # na.rm = TRUE ignores missing values in the sum
  dplyr::summarise(
    total_ot = base::sum(Overtime, na.rm = TRUE),
    .groups = "drop"
  ) |>
  # Keep only the 3 departments with highest total overtime
  # slice_max automatically sorts descending and takes top n rows
  dplyr::slice_max(order_by = total_ot, n = 3) |>
  # Extract just the department names as a character vector
  # This vector will be used to filter the main data
dplyr::pull(department)

# STEP 2: Summarize overtime by fiscal year for only those top 3 departments

dept_overtime <- final_df_wide |>
  # Keep only rows where department is in our top 5 list
  # %in% checks if each value exists in the vector
  dplyr::filter(department %in% top_3_depts) |>
  # Group by both fiscal_year and department
  # This gives us one row per year-department combination
  dplyr::group_by(fiscal_year, department) |>
  # Sum overtime for each year-department combination
  dplyr::summarise(
    overtime_dollars = base::sum(Overtime, na.rm = TRUE),
    .groups = "drop"
  ) |>
  # Create the label text that will appear above each point
  # paste0 concatenates strings with no separator
  # scales::dollar formats numbers as currency
  # scale = 1e-3 divides by 1000, suffix = "K" adds K for thousands
  # Result looks like "($150K)"
  dplyr::mutate(
    label_text = base::paste0(scales::dollar(overtime_dollars, 
                                                   accuracy = 1, 
                                                   scale = 1e-3, 
                                                   suffix = "K"))
  )

# STEP 3: Build the line chart using ggplot2

# Initialize the plot with data and aesthetic mappings
# aes() maps data columns to visual properties
# x = fiscal_year puts years on horizontal axis
# y = overtime_dollars puts dollar amounts on vertical axis
# color = department assigns a different color to each department
# group = department tells geom_line to connect points within each department
ggplot2::ggplot(dept_overtime, 
                ggplot2::aes(x = fiscal_year, 
                             y = overtime_dollars, 
                             color = department, 
                             group = department)) +

  # Add lines connecting points for each department
# linewidth controls thickness (1 is moderate)
ggplot2::geom_line(linewidth = 2) +

  # Add points at each data value
# size = 3 makes points visible but not overwhelming
ggplot2::geom_point(size = 4) +

  # Add text labels above each point
# aes(label = label_text) specifies which column contains the label
# vjust = -1.2 moves labels up (negative = above, positive = below)
# size = 3 controls text size
# show.legend = FALSE prevents "a" appearing in legend
ggplot2::geom_text(ggplot2::aes(label = label_text),
                   vjust = -1.2,
                   size = 5,
                   color = "black",
                   show.legend = FALSE) +

 # Format the y-axis labels as dollars in thousands
# expand = expansion(mult = c(0.05, 0.15)) adds padding:
#   - 0.05 = 5% padding at bottom
#   - 0.15 = 15% padding at top (makes room for labels)
ggplot2::scale_y_continuous(
  labels = scales::dollar_format(scale = 1e-3, suffix = "K"),
  expand = ggplot2::expansion(mult = c(0.05, 0.15))
) +

  # Add titles and axis labels
# title = main chart title
# subtitle = secondary descriptive text
# x = x-axis label
# y = y-axis label
# color = legend title (matches the aes color mapping)
ggplot2::labs(
  title = "Overtime Usage by Department",
  subtitle = "Top 3 Departments by Total Overtime Used",
  x = "Fiscal Year",
  y = "Overtime ($)",
  color = "Department"
) +

  # Apply a clean, minimal theme
# Removes gray background, uses light gridlines
ggplot2::theme_minimal() +
ggplot2::theme(legend.position = "bottom"
               )

  # Customize specific theme elements
# legend.position = "bottom" # moves legend below chart (saves horizontal space)
# element_text(face = "bold") makes the title bold
# ggplot2::theme(
#   legend.position = "bottom",
#   plot.title = ggplot2::element_text(face = "bold")
# )

```
------------------------------------------------------------------------


## Dept Total Compenstation Ranking Over Time (Bump Chart)
```{r}
#| label: dept-rank-bump-chart
#| fig-width: 13
#| fig-height: 8

# =============================================================================
# APPENDIX G: DEPARTMENT RANKING BUMP CHART
# =============================================================================

# Calculate ranks by year for all departments
dept_ranks_all <- final_df_long |>
  dplyr::group_by(department, fiscal_year) |>
  dplyr::summarise(
    total_compensation = base::sum(amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::group_by(fiscal_year) |>
  dplyr::mutate(
    rank = dplyr::min_rank(dplyr::desc(total_compensation))
  ) |>
  dplyr::ungroup()

# Get departments that were ever in top 10
top_10_ever <- dept_ranks_all |>
  dplyr::filter(rank <= 10) |>
  dplyr::distinct(department) |>
  dplyr::pull(department)

# Filter to only those departments
dept_ranks_top_10 <- dept_ranks_all |>
  dplyr::filter(department %in% top_10_ever)

okabe_ito <- c(
  "#000000", # black
  "#E69F00", # orange
  "#56B4E9", # sky blue
  "#009E73", # bluish green
  "black",   # black
  "#0072B2", # blue
  "#D55E00", # vermilion
  "#CC79A7",  # reddish purple
  "firebrick",
  "navy",
  "darkgreen",
  "brown",
  "purple"
  
)


# Create bump chart
ggplot2::ggplot(dept_ranks_top_10, ggplot2::aes(x = fiscal_year, y = rank, 
                                          group = department, color = department)) +
  ggplot2::geom_line(linewidth = 1.5) +
  ggplot2::geom_point(size = 4) +
  ggplot2::geom_text(
    data = dplyr::filter(dept_ranks_top_10, fiscal_year == "FY25"),
    ggplot2::aes(label = department),
    hjust = -0.1,
    nudge_x = 0.15,
    size = 3
  ) +
  ggplot2::scale_color_manual(values = okabe_ito) +
  ggplot2::scale_y_reverse(breaks = 1:15) +
  ggplot2::scale_x_discrete(expand = ggplot2::expansion(mult = c(0.02, 0.35))) +
  ggplot2::labs(
    title = "Dept Total Compensation Rankings Over Time",
    subtitle = "Rank 1 = highest total compensation. Lines show how departments moved in rankings relative to each other.",
    x = "Fiscal Year",
    y = "Rank (1 = Highest)"
  ) +
  ggplot2::theme(legend.position = "none")
```


# Calculate compensation rankings by year for all departments (in table form)

```{r}
#| label: dept-rank-table

# Show the underlying data
dept_ranks_top_10 |>
  tidyr::pivot_wider(
    id_cols = department,
    names_from = fiscal_year,
    values_from = rank
  ) |>
  dplyr::arrange(FY25) |>
  flextable::flextable() |>
  flextable::add_header_lines("Department Rankings by Fiscal Year (1 = Highest TOTAL Compensation)") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```


------------------------------------------------------------------------


## City Manager Total Compensation Over Time

```{r}
#| label: city-manager-total-comp-fy20-fy25
#| fig-width: 13
#| fig-height: 8

# =============================================================================
# APPENDIX H: CITY MANAGER TOTAL COMPENSATION OVER TIME
# =============================================================================


# City Manager TOTAL COMPENSATION by Fiscal Year
# Total compensation = Leave + Regular + Overtime + Additional + Deployment + Arbitration + Benefits

# If you've already recoded earnings_type to pretty labels (e.g., "Regular"),
# these are the labels to include:
comp_types <- c(
  "Leave Payout", "Regular", "Overtime", "Additional",
  "Deployment", "Arbitration", "Benefits"
)

city_manager_data <- final_df_long |>
  dplyr::filter(
    job_title == "City Manager",
    earnings_type %in% comp_types
  ) |>
  dplyr::group_by(fiscal_year, name) |>
  dplyr::summarise(
    total_compensation = sum(amount, na.rm = TRUE),
    .groups = "drop"
  )

if (base::nrow(city_manager_data) > 0) {

  ggplot2::ggplot(
    city_manager_data,
    ggplot2::aes(x = fiscal_year, y = total_compensation, group = name, color = name)
  ) +
    ggplot2::geom_line(linewidth = 1) +
    ggplot2::geom_point(size = 3) +

    # Value labels ABOVE points
    ggplot2::geom_text(
      ggplot2::aes(label = scales::dollar(total_compensation, accuracy = 1)),
      vjust = -1.5,
      size = 4.5,
      color = "black",
      show.legend = FALSE
    ) +

    # Keep headroom for labels
    ggplot2::scale_y_continuous(
      expand = ggplot2::expansion(mult = c(0, 0.15))
    ) +

    ggplot2::labs(
      title = "City Manager - Total Compensation by Fiscal Year",
      subtitle = "City of Schertz",
      x = "Fiscal Year",
      y = NULL,
      color = "Employee"
    ) +

    # Remove y-axis clutter
    ggplot2::theme(
      legend.position = "bottom",
      axis.text.y  = ggplot2::element_blank(),
      axis.ticks.y = ggplot2::element_blank(),
      panel.grid.major.y = ggplot2::element_blank()
    )

} else {
  base::cat("No City Manager records found in the data.\n")
}

```




------------------------------------------------------------------------


## Total Compensation Growth by Year

```{r}
#| label: plot-total-compensation-growth-by-year
#| fig-width: 10
#| fig-height: 6

# =============================================================================
# APPENDIX I: TOTAL COMPENSATION GROWTH BY FISCAL YEAR - ALL DEPTS
# =============================================================================

ggplot2::ggplot(yearly_summary_total_comp,
                ggplot2::aes(x = fiscal_year, y = total_compensation, group = 1)) +
  ggplot2::geom_line(linewidth = 1.5, color = "steelblue") +
  ggplot2::geom_point(size = 4, color = "steelblue") +
  ggplot2::geom_text(
    ggplot2::aes(label = scales::dollar(total_compensation, scale = 1e-6, suffix = "M")),
    vjust = -2
  ) +

  ggplot2::theme(
    axis.text.y  = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    panel.grid.major.y = ggplot2::element_blank()
  ) +

  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = c(0.05, 0.25)
                                )
    ) +

  ggplot2::labs(
    title = "Total Employee Compensation by Fiscal Year - All Depts",
    subtitle = "City of Schertz",
    x = "Fiscal Year",
    y = "Total Compensation"
  )

```


------------------------------------------------------------------------


## Total Compensation Growth Rate Over Time


```{r}
#| label: plot-growth-rate
#| fig-width: 10
#| fig-height: 6

# =============================================================================
# APPENDIX J: YEAR-OVER-YEAR GROWTH RATE BAR CHART
# =============================================================================

growth_data <- yearly_summary_total_comp |>
  dplyr::arrange(fiscal_year) |>
  dplyr::mutate(
    prior_total = dplyr::lag(total_compensation),
    growth_rate = (total_compensation - prior_total) / prior_total,
    growth_pct = growth_rate * 100
  )

growth_data |>
  dplyr::filter(!base::is.na(growth_rate)) |>
  ggplot2::ggplot(ggplot2::aes(x = fiscal_year, y = growth_pct)) +
  ggplot2::geom_col(fill = "steelblue") +
  ggplot2::geom_text(
    ggplot2::aes(label = base::paste0(base::round(growth_pct, 1), "%")),
    vjust = -0.5
  ) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed") +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.15))) +
  ggplot2::labs(
    title = "Total Compensation Growth Rate FY20 - FY25",
    subtitle = "City of Schertz",
    x = "Fiscal Year",
    y = "Growth Rate (%)"
  )
```



```{r}
#| label: growth-rate-table

# APPENDIX J: YEAR-OVER-YEAR GROWTH RATE TABLE

growth_data |>
  dplyr::mutate(
    change_num = total_compensation - prior_total,
    total_compensation_disp = scales::dollar(total_compensation, accuracy = 1),
    prior_total_disp        = scales::dollar(prior_total, accuracy = 1),
    change_disp             = scales::dollar(change_num, accuracy = 1),
    growth_pct_disp         = sprintf("%.1f%%", growth_pct)
  ) |>
  dplyr::select(
    fiscal_year, total_employees,
    total_compensation = total_compensation_disp,
    prior_total        = prior_total_disp,
    change             = change_disp,
    growth_pct         = growth_pct_disp
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Total Compensation Growth Details - All Depts") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()

```







