---
title: "Schertz_Data_Viz"
format: html
---


### Load Long Data

```{r}
final_df <- readRDS(here::here("data_clean", "compensation_long_fy20_fy25.rds"))
```

### Long format (what you have) is best for:

* Bar charts comparing earnings types across years
* Faceted plots by earnings_type
* Grouped summaries (total by year, by department, by earnings type)
* Most ggplot2 work

### Wide format is needed for:

* Calculating total compensation per employee (summing across columns)
* Tables showing one row per employee with all earnings side-by-side
* Scatter plots comparing two earnings types against each other
* Validation checks (comparing calculated vs source totals)

You can always pivot on the fly when needed:

### Create a Wide Format DF (When Needed)

```{r}
# Long to wide (when you need it)
wide_df <- final_df |>
  tidyr::pivot_wider(
    names_from = earnings_type,
    values_from = amount,
    values_fill = 0
  ) |>
  dplyr::mutate(
    total_earnings = Leave + Regular + Overtime + Additional + Deployment + Arbitration,
    total_compensation = total_earnings + Benefits
  )
```




### Summary Statistics

```{r}
#| label: summary-statistics

# =============================================================================
# SUMMARY STATISTICS
# =============================================================================

# Summary by year
yearly_summary <- final_df |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    total_compensation = base::sum(amount, na.rm = TRUE),
    total_employees = dplyr::n_distinct(employee_key),
    avg_compensation = total_compensation / total_employees,
    .groups = "drop"
  ) |>
  dplyr::arrange(fiscal_year)

yearly_summary |>
  dplyr::mutate(
    total_compensation = scales::dollar(total_compensation, accuracy = 1),
    avg_compensation = scales::dollar(avg_compensation, accuracy = 1)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Yearly Summary: Total Compensation and Headcount") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```



## Year-over-Year Comparison

```{r}
#| label: yoy-comparison

# =============================================================================
# APPENDIX C: YEAR-OVER-YEAR COMPARISON OF EE TOTALS, COMPENSATION, PCT, AVG COMP
# =============================================================================

yoy_summary <- yearly_summary |>
  dplyr::mutate(
    prev_total = dplyr::lag(total_compensation),
    prev_employees = dplyr::lag(total_employees),
    total_change = total_compensation - prev_total,
    total_pct_change = (total_change / prev_total) * 100,
    employee_change = total_employees - prev_employees
  )

yoy_summary |>
  dplyr::mutate(
    total_compensation = scales::dollar(total_compensation, accuracy = 1),
    avg_compensation = scales::dollar(avg_compensation, accuracy = 1),
    total_change = scales::dollar(total_change, accuracy = 1),
    total_pct_change = base::sprintf("%.1f%%", total_pct_change)
  ) |>
  dplyr::select(
    fiscal_year, total_employees, employee_change,
    total_compensation, total_change, total_pct_change, avg_compensation
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Year-over-Year Comparison") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```


------------------------------------------------------------------------


## Top 10 Earners by Year

```{r}
#| label: top-10-earners-per-year

# =============================================================================
# APPENDIX D: TOP 10 EARNERS BY YEAR
# =============================================================================

top_10_earners <- wide_df |>
  dplyr::group_by(fiscal_year) |>
  dplyr::slice_max(order_by = total_compensation, n = 10) |>
  dplyr::select(
    fiscal_year, name, department, job_title, total_compensation
  ) |>
  dplyr::mutate(
    rank = dplyr::row_number()
  ) |>
  dplyr::ungroup()


top_10_earners |>
  dplyr::mutate(
    total_compensation = scales::dollar(total_compensation, accuracy = 1)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Top 10 Earners by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```


------------------------------------------------------------------------


## Appendix E: Department Summary

```{r}
#| label: department-summary

# =============================================================================
# APPENDIX E: DEPARTMENT SUMMARY
# =============================================================================

dept_summary <- wide_df |>
  dplyr::group_by(fiscal_year, department) |>
  dplyr::summarise(
    employees = dplyr::n(),
    total_comp = base::sum(total_compensation, na.rm = TRUE),
    avg_comp = base::mean(total_compensation, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(fiscal_year, dplyr::desc(total_comp))

# Show FY25 as example
dept_summary |>
  dplyr::filter(fiscal_year == "FY25") |>
  dplyr::mutate(
    total_comp = scales::dollar(total_comp, accuracy = 1),
    avg_comp = scales::dollar(avg_comp, accuracy = 1)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Department Summary - FY25") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```


------------------------------------------------------------------------


## Overtime Analysis by Fiscal Year

```{r}
#| label: overtime-analysis

# =============================================================================
# APPENDIX F: OVERTIME ANALYSIS
# =============================================================================

overtime_summary <- wide_df |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    total_overtime = base::sum(overtime_earnings, na.rm = TRUE),
    total_regular = base::sum(regular_earnings, na.rm = TRUE),
    overtime_pct = (total_overtime / total_regular) * 100,
    employees_with_ot = base::sum(overtime_earnings > 0),
    total_employees = dplyr::n(),
    pct_with_ot = (employees_with_ot / total_employees) * 100,
    .groups = "drop"
  )

overtime_summary |>
  dplyr::mutate(
    total_overtime = scales::dollar(total_overtime, accuracy = 1),
    total_regular = scales::dollar(total_regular, accuracy = 1),
    overtime_pct = base::sprintf("%.1f%%", overtime_pct),
    pct_with_ot = base::sprintf("%.1f%%", pct_with_ot)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Overtime Analysis by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```



# OVERTIME BY DEPARTMENT - TOP 3 

```{r}
#| label: fig-overtime-by-dept
#| fig-cap: "Top 3 Departments by Overtime Usage"

# ======================================================================
# OVERTIME BY DEPARTMENT - TOP 3 LINE CHART
# ======================================================================

# STEP 1: Identify the top 3 departments by total overtime across all years
# We need to know which departments to include before we can chart them

top_3_depts <- all_data |>
  # Group by department to aggregate across all fiscal years
  dplyr::group_by(department) |>
  # Calculate total overtime for each department (all years combined)
  # na.rm = TRUE ignores missing values in the sum
  dplyr::summarise(
    total_ot = base::sum(overtime_earnings, na.rm = TRUE),
    .groups = "drop"
  ) |>
  # Keep only the 3 departments with highest total overtime
  # slice_max automatically sorts descending and takes top n rows
  dplyr::slice_max(order_by = total_ot, n = 3) |>
  # Extract just the department names as a character vector
  # This vector will be used to filter the main data
dplyr::pull(department)

# STEP 2: Summarize overtime by fiscal year for only those top 3 departments

dept_overtime <- all_data |>
  # Keep only rows where department is in our top 5 list
  # %in% checks if each value exists in the vector
  dplyr::filter(department %in% top_3_depts) |>
  # Group by both fiscal_year and department
  # This gives us one row per year-department combination
  dplyr::group_by(fiscal_year, department) |>
  # Sum overtime for each year-department combination
  dplyr::summarise(
    overtime_dollars = base::sum(overtime_earnings, na.rm = TRUE),
    .groups = "drop"
  ) |>
  # Create the label text that will appear above each point
  # paste0 concatenates strings with no separator
  # scales::dollar formats numbers as currency
  # scale = 1e-3 divides by 1000, suffix = "K" adds K for thousands
  # Result looks like "($150K)"
  dplyr::mutate(
    label_text = base::paste0(scales::dollar(overtime_dollars, 
                                                   accuracy = 1, 
                                                   scale = 1e-3, 
                                                   suffix = "K"))
  )

# STEP 3: Build the line chart using ggplot2

# Initialize the plot with data and aesthetic mappings
# aes() maps data columns to visual properties
# x = fiscal_year puts years on horizontal axis
# y = overtime_dollars puts dollar amounts on vertical axis
# color = department assigns a different color to each department
# group = department tells geom_line to connect points within each department
ggplot2::ggplot(dept_overtime, 
                ggplot2::aes(x = fiscal_year, 
                             y = overtime_dollars, 
                             color = department, 
                             group = department)) +

  # Add lines connecting points for each department
# linewidth controls thickness (1 is moderate)
ggplot2::geom_line(linewidth = 2) +

  # Add points at each data value
# size = 3 makes points visible but not overwhelming
ggplot2::geom_point(size = 4) +

  # Add text labels above each point
# aes(label = label_text) specifies which column contains the label
# vjust = -1.2 moves labels up (negative = above, positive = below)
# size = 3 controls text size
# show.legend = FALSE prevents "a" appearing in legend
ggplot2::geom_text(ggplot2::aes(label = label_text),
                   vjust = -1.2,
                   size = 5,
                   color = "black",
                   show.legend = FALSE) +

 # Format the y-axis labels as dollars in thousands
# expand = expansion(mult = c(0.05, 0.15)) adds padding:
#   - 0.05 = 5% padding at bottom
#   - 0.15 = 15% padding at top (makes room for labels)
ggplot2::scale_y_continuous(
  labels = scales::dollar_format(scale = 1e-3, suffix = "K"),
  expand = ggplot2::expansion(mult = c(0.05, 0.15))
) +

  # Add titles and axis labels
# title = main chart title
# subtitle = secondary descriptive text
# x = x-axis label
# y = y-axis label
# color = legend title (matches the aes color mapping)
ggplot2::labs(
  title = "Overtime Usage by Department",
  subtitle = "Top 3 Departments by Total Overtime Used",
  x = "Fiscal Year",
  y = "Overtime ($)",
  color = "Department"
) 

  # Apply a clean, minimal theme
# Removes gray background, uses light gridlines
# ggplot2::theme_minimal() +

  # Customize specific theme elements
# legend.position = "bottom" moves legend below chart (saves horizontal space)
# element_text(face = "bold") makes the title bold
# ggplot2::theme(
#   legend.position = "bottom",
#   plot.title = ggplot2::element_text(face = "bold")
# )

```
------------------------------------------------------------------------


## Appendix G: Department Compenstation Ranking Over Time (Bump Chart)
```{r}
#| label: dept-rank-bump-chart
#| fig-width: 13
#| fig-height: 8

# =============================================================================
# APPENDIX G: DEPARTMENT RANKING BUMP CHART
# =============================================================================

# Calculate ranks by year for all departments
dept_ranks_all <- final_df |>
  dplyr::group_by(department, fiscal_year) |>
  dplyr::summarise(
    total_compensation = base::sum(amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::group_by(fiscal_year) |>
  dplyr::mutate(
    rank = dplyr::min_rank(dplyr::desc(total_compensation))
  ) |>
  dplyr::ungroup()

# Get departments that were ever in top 10
top_10_ever <- dept_ranks_all |>
  dplyr::filter(rank <= 10) |>
  dplyr::distinct(department) |>
  dplyr::pull(department)

# Filter to only those departments
dept_ranks_top_10 <- dept_ranks_all |>
  dplyr::filter(department %in% top_10_ever)

okabe_ito <- c(
  "#000000", # black
  "#E69F00", # orange
  "#56B4E9", # sky blue
  "#009E73", # bluish green
  "#F0E442", # yellow
  "#0072B2", # blue
  "#D55E00", # vermilion
  "#CC79A7",  # reddish purple
  "firebrick",
  "navy",
  "darkgreen",
  "brown"
  
)


# Create bump chart
ggplot2::ggplot(dept_ranks_top_10, ggplot2::aes(x = fiscal_year, y = rank, 
                                          group = department, color = department)) +
  ggplot2::geom_line(linewidth = 1.5) +
  ggplot2::geom_point(size = 4) +
  ggplot2::geom_text(
    data = dplyr::filter(dept_ranks_top_10, fiscal_year == "FY24"),
    ggplot2::aes(label = department),
    hjust = -0.1,
    size = 3
  ) +
  ggplot2::geom_text(
    data = dplyr::filter(dept_ranks_top_10, fiscal_year == "FY20"),
    ggplot2::aes(label = department),
    hjust = 1.1,
    size = 3
  ) +
  ggplot2::scale_color_manual(values = okabe_ito) +
  ggplot2::scale_y_reverse(breaks = 1:15) +
  ggplot2::scale_x_discrete(expand = ggplot2::expansion(mult = c(0.25, 0.35))) +
  ggplot2::labs(
    title = "Department Compensation Rankings Over Time",
    subtitle = "Rank 1 = highest total compensation. Lines show how departments moved in rankings relative to each other.",
    x = "Fiscal Year",
    y = "Rank (1 = Highest)"
  ) +
  ggplot2::theme(legend.position = "none")
```


# Calculate compensation rankings by year for all departments (in table form)

```{r}
#| label: dept-rank-table

# Show the underlying data
dept_ranks_top_10 |>
  tidyr::pivot_wider(
    id_cols = department,
    names_from = fiscal_year,
    values_from = rank
  ) |>
  dplyr::arrange(FY24) |>
  flextable::flextable() |>
  flextable::add_header_lines("Department Rankings by Fiscal Year (1 = Highest TOTAL Compensation)") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```


------------------------------------------------------------------------


## Appendix H: City Manager Total Compensation Over Time

```{r}
#| label: plot-city-manager-total-comp
#| fig-width: 13
#| fig-height: 8

# =============================================================================
# APPENDIX H: CITY MANAGER COMPENSATION OVER TIME
# =============================================================================


# City Manager TOTAL COMPENSATION by Fiscal Year
# Total compensation = Leave + Regular + Overtime + Additional + Deployment + Arbitration + Benefits

# If you've already recoded earnings_type to pretty labels (e.g., "Regular"),
# these are the labels to include:
comp_types <- c(
  "Leave Payout", "Regular", "Overtime", "Additional",
  "Deployment", "Arbitration", "Benefits"
)

city_manager_data <- final_df |>
  dplyr::filter(
    job_title == "City Manager",
    earnings_type %in% comp_types
  ) |>
  dplyr::group_by(fiscal_year, name) |>
  dplyr::summarise(
    total_compensation = sum(amount, na.rm = TRUE),
    .groups = "drop"
  )

if (base::nrow(city_manager_data) > 0) {

  ggplot2::ggplot(
    city_manager_data,
    ggplot2::aes(x = fiscal_year, y = total_compensation, group = name, color = name)
  ) +
    ggplot2::geom_line(linewidth = 1) +
    ggplot2::geom_point(size = 3) +

    # Value labels ABOVE points
    ggplot2::geom_text(
      ggplot2::aes(label = scales::dollar(total_compensation, accuracy = 1)),
      vjust = -1.5,
      size = 4.5,
      color = "black",
      show.legend = FALSE
    ) +

    # Keep headroom for labels
    ggplot2::scale_y_continuous(
      expand = ggplot2::expansion(mult = c(0, 0.15))
    ) +

    ggplot2::labs(
      title = "City Manager - Total Compensation by Fiscal Year",
      subtitle = "City of Schertz",
      x = "Fiscal Year",
      y = NULL,
      color = "Employee"
    ) +

    # Remove y-axis clutter
    ggplot2::theme(
      legend.position = "bottom",
      axis.text.y  = ggplot2::element_blank(),
      axis.ticks.y = ggplot2::element_blank(),
      panel.grid.major.y = ggplot2::element_blank()
    )

} else {
  base::cat("No City Manager records found in the data.\n")
}

```




------------------------------------------------------------------------


## Appendix I: Total Compensation Growth by Year
```{r}
#| label: plot-total-compensation-growth-by-year
#| fig-width: 10
#| fig-height: 6

# =============================================================================
# APPENDIX I: TOTAL COMPENSATION GROWTH BY FISCAL YEAR - ALL DEPTS
# =============================================================================

ggplot2::ggplot(yearly_summary, 
                ggplot2::aes(x = fiscal_year, y = total_compensation, group = 1)) +
  ggplot2::geom_line(linewidth = 1.5, color = "steelblue") +
  ggplot2::geom_point(size = 4, color = "steelblue") +
  ggplot2::geom_text(
    ggplot2::aes(label = scales::dollar(total_compensation, scale = 1e-6, suffix = "M")),
    vjust = -2
  ) +
  
  ggplot2::theme(
    axis.text.y  = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    panel.grid.major.y = ggplot2::element_blank()
  ) +

  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = c(0.05, 0.15)
                                )
    ) +

  ggplot2::labs(
    title = "Total Employee Compensation by Fiscal Year - All Depts",
    subtitle = "City of Schertz",
    x = "Fiscal Year",
    y = "Total Compensation"
  )

```


------------------------------------------------------------------------


## Appendix J: Year-over-Year Growth Rate Bar Chart


```{r}
#| label: plot-growth-rate
#| fig-width: 10
#| fig-height: 6

# =============================================================================
# APPENDIX J: YEAR-OVER-YEAR GROWTH RATE BAR CHART
# =============================================================================

growth_data <- yearly_summary |>
  dplyr::arrange(fiscal_year) |>
  dplyr::mutate(
    prior_total = dplyr::lag(total_compensation),
    growth_rate = (total_compensation - prior_total) / prior_total,
    growth_pct = growth_rate * 100
  )

growth_data |>
  dplyr::filter(!base::is.na(growth_rate)) |>
  ggplot2::ggplot(ggplot2::aes(x = fiscal_year, y = growth_pct)) +
  ggplot2::geom_col(fill = "steelblue") +
  ggplot2::geom_text(
    ggplot2::aes(label = base::paste0(base::round(growth_pct, 1), "%")),
    vjust = -0.5
  ) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed") +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.15))) +
  ggplot2::labs(
    title = "Year-over-Year Compensation Growth",
    subtitle = "City of Schertz",
    x = "Fiscal Year",
    y = "Growth Rate (%)"
  )
```


```{r}
#| label: growth-rate-table

# APPENDIX J: YEAR-OVER-YEAR GROWTH RATE TABLE

growth_data |>
  dplyr::mutate(
    change_num = total_compensation - prior_total,
    total_compensation_disp = scales::dollar(total_compensation, accuracy = 1),
    prior_total_disp        = scales::dollar(prior_total, accuracy = 1),
    change_disp             = scales::dollar(change_num, accuracy = 1),
    growth_pct_disp         = sprintf("%.1f%%", growth_pct)
  ) |>
  dplyr::select(
    fiscal_year, total_employees,
    total_compensation = total_compensation_disp,
    prior_total        = prior_total_disp,
    change             = change_disp,
    growth_pct         = growth_pct_disp
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Annual Compensation Growth Details") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()

```







