---
title: "Schertz Compensation Data Pipeline - Explicit Column Version"
subtitle: "Compensation Review Project"
description: "Each earnings component explicitly captured and validated"
authors: 
  - name: "Dan Swart, CPA (ret)"
  - name: "Claude Opus 4.5"
date: today
date-format: long
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: false
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0

editor: source

quarto:
  render:
    cache-refresh: true

execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8

knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false
---

```{r}
#| label: setup
#| include: false

# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(DT.options = list(dom = 'pBlfrti'))
options(shiny.maxRequestSize = 50 * 1024^2)
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 

# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)

# Set global ggplot2 theme (NO theme_minimal)
ggplot2::theme_set(
  ggplot2::theme_bw(base_size = 16) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 16,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 14,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 14,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 14),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 12,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)

# Set seed for reproducibility
base::set.seed(123)
```

## Introduction

This document implements **explicit column handling** for the Schertz compensation data pipeline. Every earnings component is captured distinctly, validated independently, and then summed to verify totals match.

**Key Philosophy**: Each column name is specified **exactly** as it appears after `janitor::clean_names()`. No regex patterns, no guessing.

**Explicit Compensation Components** (FY23 structure):

| Component               | Excel Column Name        | Goes Into          |
|-------------------------|--------------------------|--------------------|
| Leave Payout            | FY23 LEAVE PAYOUT        | Total Earnings     |
| Regular Earnings        | FY23 REGULAR EARNINGS    | Total Earnings     |
| Overtime Earnings       | FY23 OVERTIME EARNINGS   | Total Earnings     |
| Additional Earnings     | FY23 ADDITIONAL EARNINGS | Total Earnings     |
| Deployment Earnings     | FY23 DEPLOYMENT EARNINGS | Total Earnings     |
| Arbitration/Settlements | ARBITRATION/ SETTLEMENTS | Total Earnings     |
| **Total Earnings**      | FY23 TOTAL EARNINGS      | (calculated)       |
| Additional Benefits     | FY23 ADDITIONAL BENEFITS | Total Compensation |
| **Total Compensation**  | FY23 TOTAL COMPENSATION  | (final)            |

**Validation Formula**:

```         
TOTAL_EARNINGS = LEAVE_PAYOUT + REGULAR_EARNINGS + OVERTIME_EARNINGS + 
                 ADDITIONAL_EARNINGS + DEPLOYMENT_EARNINGS + ARBITRATION_SETTLEMENTS

TOTAL_COMPENSATION = TOTAL_EARNINGS + ADDITIONAL_BENEFITS
```

------------------------------------------------------------------------

# PHASE 1: DISCOVERY

Run this section first to see exact column names after `clean_names()`.

## Step 1: Define Your Files

**Edit these filenames to match your actual files:**

```{r}
#| label: define-files

# =============================================================================
# EDIT THESE FILENAMES TO MATCH YOUR ACTUAL FILES
# =============================================================================

files <- base::list(
  FY20 = "FY20 Earnings.xlsx",
  FY21 = "FY21 Earnings.xlsx",
  FY22 = "FY22 Earnings.xlsx",
  FY23 = "FY23Employee_Comp_FY23.xlsx",
  FY24 = "FY24Employee_Comp_FY24_All.xlsx"
)

data_folder <- here::here("data_raw")

# Configuration paths for export
paths <- base::list(
  data_raw = here::here("data_raw"),
  data_clean = here::here("data_clean"),
  reports = here::here("reports"),
  output_test = here::here("output_test")
)

# Report settings
report_settings <- base::list(
  test_mode = FALSE
)
```

## Step 2: Read Files and Show Column Names

This reads each file, applies `clean_names()`, and shows you the exact column names to use.

```{r}
#| label: read-and-show-columns

# Read each file and capture column names
file_info <- purrr::map_dfr(base::names(files), function(fy) {
  filepath <- base::file.path(data_folder, files[[fy]])
  
  if (!base::file.exists(filepath)) {
    return(dplyr::tibble(
      fiscal_year = fy,
      filename = files[[fy]],
      status = "FILE NOT FOUND",
      row_count = NA_integer_,
      columns = NA_character_
    ))
  }
  
  df <- readxl::read_excel(filepath) |>
    janitor::clean_names()
  
  dplyr::tibble(
    fiscal_year = fy,
    filename = files[[fy]],
    status = "OK",
    row_count = base::nrow(df),
    columns = base::paste(base::names(df), collapse = ", ")
  )
})

# Display summary
file_info |>
  dplyr::select(fiscal_year, filename, status, row_count) |>
  flextable::flextable() |>
  flextable::add_header_lines("File Import Summary") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

## Step 3: Exact Column Names by File

**Copy these exact column names into Phase 2 below.**

```{r}
#| label: show-exact-columns

# Show columns for each file in a usable format
for (fy in base::names(files)) {
  filepath <- base::file.path(data_folder, files[[fy]])
  
  if (base::file.exists(filepath)) {
    df <- readxl::read_excel(filepath) |>
      janitor::clean_names()
    
    base::cat("\n")
    base::cat("# =========================================================\n")
    base::cat("#", fy, "- Columns after clean_names():\n")
    base::cat("# =========================================================\n")
    
    cols <- base::names(df)
    for (i in base::seq_along(cols)) {
      base::cat("#", base::sprintf("%2d", i), ":", cols[i], "\n")
    }
  }
}
```

------------------------------------------------------------------------

# ⛔ STOP HERE FIRST TIME

Run Phase 1 above. Then use the column names shown to fill in the exact mappings in Phase 2 below.

------------------------------------------------------------------------

# PHASE 2: PROCESSING

**After reviewing Phase 1 output, edit the column mappings below.**

## Step 4: Define Exact Column Mappings

**Edit these to match the exact column names from Phase 1:**

```{r}
#| label: define-column-mappings

# =============================================================================
# COLUMN MAPPINGS - EDIT THESE TO MATCH YOUR FILES EXACTLY
# =============================================================================
# Use the exact column names shown in Phase 1 (after clean_names)
# Set to NA if that column doesn't exist in a given year

column_maps <- base::list(
  
  # ----- FY20 -----
  FY20 = base::list(
    # Identity columns
    name = "name",                    # Single name column (or NA if split)
    last_name = NA,                   # NA if using single name column
    first_name = NA,                  # NA if using single name column
    department = "department",
    job_title = "job_title",
    hire_date = "hire_date",
    termination_date = "termination_date",
    
    # Earnings columns (set to NA if not present)
    leave_payout = NA,
    regular_earnings = "regular_earnings",
    overtime_earnings = "overtime_earnings",
    additional_earnings = "additional_earnings1",
    deployment_earnings = NA,
    arbitration_settlements = NA,
    
    # Benefits (set to NA if not present)
    additional_benefits = NA,
    
    # Totals for validation
    total_earnings = "total_earnings",
    total_compensation = NA           # NA if no total_comp column
  ),
  
  # ----- FY21 -----
  FY21 = base::list(
    status = "status",
    name = "name",
    last_name = NA,
    first_name = NA,
    department = "home_department",
    job_title = "job_title",
    hire_date = "hire_date",
    termination_date = "termination_date",
    
    leave_payout = NA,
    regular_earnings = "regular_earnings",
    overtime_earnings = "overtime_earnings",
    additional_earnings = "additional_earnings1",
    deployment_earnings = NA,
    arbitration_settlements = NA,
    
    additional_benefits = NA,
    
    total_earnings = "total_earnings",
    total_compensation = NA
  ),
  
  # ----- FY22 -----
  FY22 = base::list(
    status = "status",
    name = "name",
    last_name = NA,
    first_name = NA,
    department = "department",
    job_title = "job_title",
    hire_date = "hire_date",
    termination_date = "termination_date",
    
    leave_payout = NA,
    regular_earnings = "regular_earnings",
    overtime_earnings = "overtime_earnings",
    additional_earnings = "additional_earnings1",
    deployment_earnings = NA,
    arbitration_settlements = NA,
    
    additional_benefits = NA,
    
    total_earnings = "total_earnings",
    total_compensation = NA
  ),
  
  # ----- FY23 -----
  FY23 = base::list(
    name = NA,                                    # NA - using split names
    last_name = "last_name",
    first_name = "first_name",
    department = "department",
    job_title = "job_title",
    employee_category = "employee_category",
    hire_date = "hire_date",
    separation_date = "separation_date",
    fy23_annual_salary1 = "fy23_annual_salary1",
    
    leave_payout = "fy23_leave_payout2",
    regular_earnings = "fy23_regular_earnings3",
    overtime_earnings = "fy23_overtime_earnings4",
    additional_earnings = "fy23_additional_earnings5",
    deployment_earnings = "fy23_deployment_earnings6",
    arbitration_settlements = "arbitration_settlements",
    additional_benefits = "fy23_additional_benefits8",
    total_earnings = "fy23_total_earnings7",
    total_compensation = "fy23_total_compensation9"
  ),
  
  # ----- FY24 -----
  FY24 = base::list(
    name = NA,
    last_name = "last_name",
    first_name = "first_name",
    department = "home_department",
    job_title = "job_title",
    employee_category = "employee_category",
    hire_date = "hire_date",
    separation_date = "separation_date",
    fy24_annual_salary1 = "fy24_annual_salary1",
    
    leave_payout = "fy24_leave_payout2",
    regular_earnings = "fy24_regular_earnings3",
    overtime_earnings = "fy24_overtime_earnings4",
    additional_earnings = "fy24_additional",
    deployment_earnings = "fy24_deployment",
    arbitration_settlements = "arbitration_settlements",
    additional_benefits = "fy24_additional_1",
    total_earnings = "fy24_total_earnings7",
    total_compensation = "fy24_total_compensation9"
  )
)

base::cat("Defined column mappings for", base::length(column_maps), "fiscal years\n")
```

------------------------------------------------------------------------

## Step 5: Utility Functions

```{r}
#| label: utility-functions

# =============================================================================
# UTILITY FUNCTIONS
# =============================================================================

# Parse currency to numeric
parse_currency <- function(x) {
  if (base::is.numeric(x)) return(x)
  x <- stringr::str_remove_all(x, "[$,\\s]")
  base::suppressWarnings(base::as.numeric(x))
}

# Fix invisible spaces
fix_spaces <- function(x) {
  if (!base::is.character(x)) return(x)
  stringr::str_squish(x)
}

# Standardize name for matching
standardize_name <- function(x) {
  x <- fix_spaces(x)
  x <- stringr::str_to_upper(x)
  stringr::str_remove_all(x, "[^A-Z ]")
}

# Safe column extract - returns 0 if column is NA or doesn't exist
safe_extract <- function(df, col_name) {
  if (base::is.na(col_name) || !col_name %in% base::names(df)) {
    return(base::rep(0, base::nrow(df)))
  }
  result <- parse_currency(df[[col_name]])
  dplyr::coalesce(result, 0)
}
```

------------------------------------------------------------------------

## Step 6: Process Each File

```{r}
#| label: process-files

# =============================================================================
# PROCESS EACH FILE INTO WIDE FORMAT
# =============================================================================

process_file <- function(fy, filename, col_map) {
  
  filepath <- base::file.path(data_folder, filename)
  
  if (!base::file.exists(filepath)) {
    base::warning("File not found: ", filepath)
    return(NULL)
  }
  
  df <- readxl::read_excel(filepath) |>
    janitor::clean_names()
  
  # Handle name - either single column or split
  if (!base::is.na(col_map$name)) {
    # Single name column
    name_values <- fix_spaces(df[[col_map$name]])
    last_name <- NA_character_
    first_name <- NA_character_
  } else {
    # Split name columns
    last_name <- fix_spaces(df[[col_map$last_name]])
    first_name <- fix_spaces(df[[col_map$first_name]])
    name_values <- base::paste(last_name, first_name, sep = ", ")
  }
  
  # Build result
  result <- dplyr::tibble(
    fiscal_year = fy,
    name = name_values,
    name_std = standardize_name(name_values),
    last_name = last_name,
    first_name = first_name,
    department = fix_spaces(df[[col_map$department]]),
    job_title = fix_spaces(df[[col_map$job_title]]),
    
    # Earnings
    leave_payout = safe_extract(df, col_map$leave_payout),
    regular_earnings = safe_extract(df, col_map$regular_earnings),
    overtime_earnings = safe_extract(df, col_map$overtime_earnings),
    additional_earnings = safe_extract(df, col_map$additional_earnings),
    deployment_earnings = safe_extract(df, col_map$deployment_earnings),
    arbitration_settlements = safe_extract(df, col_map$arbitration_settlements),
    additional_benefits = safe_extract(df, col_map$additional_benefits),
    
    # Source totals for validation
    source_total_earnings = safe_extract(df, col_map$total_earnings),
    source_total_compensation = safe_extract(df, col_map$total_compensation)
  )
  
  # Calculate totals
  result <- result |>
    dplyr::mutate(
      calc_total_earnings = leave_payout + regular_earnings + overtime_earnings +
                            additional_earnings + deployment_earnings + arbitration_settlements,
      calc_total_compensation = calc_total_earnings + additional_benefits,
      employee_key = base::paste(name_std, department, sep = "_")
    )
  
  # For files without total_compensation, use total_earnings
  result <- result |>
    dplyr::mutate(
      source_total_compensation = dplyr::if_else(
        source_total_compensation == 0 & source_total_earnings > 0,
        source_total_earnings,
        source_total_compensation
      )
    )
  
  return(result)
}

# Process all files
all_data <- purrr::map2_dfr(
  base::names(files),
  files,
  function(fy, filename) {
    process_file(fy, filename, column_maps[[fy]])
  }
)

base::cat("Processed", base::nrow(all_data), "total rows\n")
base::cat("Fiscal years:", base::paste(base::unique(all_data$fiscal_year), collapse = ", "), "\n")
```

------------------------------------------------------------------------

## Step 7: Validation - Compare Calculated vs Source Totals

```{r}
#| label: validation

# =============================================================================
# VALIDATION: COMPARE CALCULATED VS SOURCE TOTALS
# =============================================================================

validation <- all_data |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    rows = dplyr::n(),
    calc_total = base::sum(calc_total_compensation, na.rm = TRUE),
    source_total = base::sum(source_total_compensation, na.rm = TRUE),
    difference = calc_total - source_total,
    .groups = "drop"
  ) |>
  dplyr::mutate(
    status = dplyr::if_else(base::abs(difference) <= 1, "PASS", "FAIL")
  )

validation |>
  dplyr::mutate(
    calc_total = scales::dollar(calc_total, accuracy = 1),
    source_total = scales::dollar(source_total, accuracy = 1),
    difference = scales::dollar(difference, accuracy = 0.01)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Validation: Calculated vs Source Totals") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::bg(i = ~ status == "FAIL", bg = "#ffcccc") |>
  flextable::bg(i = ~ status == "PASS", bg = "#ccffcc") |>
  flextable::autofit()
```

```{r}
#| label: validation-interpretation

# Report on validation status
validation_passed <- base::all(validation$status == "PASS")

if (!validation_passed) {
  base::cat("\n=== VALIDATION FAILED ===\n\n")
  
  failures <- validation |>
    dplyr::filter(status == "FAIL")
  
  for (i in base::seq_len(base::nrow(failures))) {
    row <- failures[i, ]
    base::cat("MISMATCH in", row$fiscal_year, ":\n")
    base::cat("  - Calculated:", scales::dollar(row$calc_total), "\n")
    base::cat("  - Source:", scales::dollar(row$source_total), "\n")
    base::cat("  - Difference:", scales::dollar(row$difference), "\n\n")
  }
  
  base::cat("ACTION REQUIRED:\n")
  base::cat("1. Check the column_maps for missing or incorrect column names\n")
  base::cat("2. Review the column names shown in Phase 1\n")
  base::cat("3. Update the column mappings to match the exact column names\n")
  
} else {
  base::cat("Validation: PASSED\n")
  base::cat("All calculated totals match source file totals within $1.00\n")
}
```

------------------------------------------------------------------------

## Step 8: Component Breakdown by Year

```{r}
#| label: component-breakdown

# =============================================================================
# COMPONENT BREAKDOWN BY FISCAL YEAR
# =============================================================================

component_summary <- all_data |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    `Leave Payout` = base::sum(leave_payout, na.rm = TRUE),
    `Regular` = base::sum(regular_earnings, na.rm = TRUE),
    `Overtime` = base::sum(overtime_earnings, na.rm = TRUE),
    `Additional` = base::sum(additional_earnings, na.rm = TRUE),
    `Deployment` = base::sum(deployment_earnings, na.rm = TRUE),
    `Arbitration` = base::sum(arbitration_settlements, na.rm = TRUE),
    `Benefits` = base::sum(additional_benefits, na.rm = TRUE),
    `TOTAL` = base::sum(calc_total_compensation, na.rm = TRUE),
    .groups = "drop"
  )

component_summary |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.numeric), ~ scales::dollar(.x, accuracy = 1))
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Component Breakdown by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

------------------------------------------------------------------------

## Step 9: Convert to Long Format

```{r}
#| label: convert-to-long

# =============================================================================
# CONVERT TO LONG FORMAT FOR ANALYSIS
# =============================================================================

final_df <- all_data |>
  dplyr::select(
    fiscal_year, name, name_std, last_name, first_name,
    department, job_title, employee_key,
    leave_payout, regular_earnings, overtime_earnings,
    additional_earnings, deployment_earnings, arbitration_settlements,
    additional_benefits
  ) |>
  tidyr::pivot_longer(
    cols = c(leave_payout, regular_earnings, overtime_earnings,
             additional_earnings, deployment_earnings, arbitration_settlements,
             additional_benefits),
    names_to = "earnings_type",
    values_to = "amount"
  ) |>
  dplyr::mutate(
    earnings_type = dplyr::case_when(
      earnings_type == "leave_payout" ~ "Leave Payout",
      earnings_type == "regular_earnings" ~ "Regular",
      earnings_type == "overtime_earnings" ~ "Overtime",
      earnings_type == "additional_earnings" ~ "Additional",
      earnings_type == "deployment_earnings" ~ "Deployment",
      earnings_type == "arbitration_settlements" ~ "Arbitration",
      earnings_type == "additional_benefits" ~ "Benefits",
      TRUE ~ earnings_type
    )
  )

base::cat("Long format data created:", base::nrow(final_df), "rows\n")
base::cat("Unique employees:", dplyr::n_distinct(final_df$employee_key), "\n")
```

------------------------------------------------------------------------

## Step 10: Checkpoint H - Explicit Component Breakdown

This checkpoint shows exactly how much of each component was captured, proving nothing was missed.

```{r}
#| label: checkpoint-h-component-breakdown

# =============================================================================
# CHECKPOINT H: EXPLICIT COMPONENT BREAKDOWN
# =============================================================================

# Create a detailed breakdown showing each component's contribution
component_breakdown <- final_df |>
  dplyr::group_by(fiscal_year, earnings_type) |>
  dplyr::summarise(
    amount = base::sum(amount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(
    names_from = earnings_type,
    values_from = amount,
    values_fill = 0
  )

# Add totals row
component_breakdown <- component_breakdown |>
  dplyr::mutate(
    `Total Earnings` = `Leave Payout` + Regular + Overtime + Additional + Deployment + Arbitration,
    `Total Compensation` = `Total Earnings` + Benefits
  )

# Display
component_breakdown |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.numeric), ~ scales::dollar(.x, accuracy = 1))
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Checkpoint H: Explicit Component Breakdown by Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

------------------------------------------------------------------------

## Step 11: Summary Statistics

```{r}
#| label: summary-statistics

# =============================================================================
# SUMMARY STATISTICS
# =============================================================================

# Summary by year
yearly_summary <- final_df |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    total_compensation = base::sum(amount, na.rm = TRUE),
    total_employees = dplyr::n_distinct(employee_key),
    avg_compensation = total_compensation / total_employees,
    .groups = "drop"
  ) |>
  dplyr::arrange(fiscal_year)

yearly_summary |>
  dplyr::mutate(
    total_compensation = scales::dollar(total_compensation, accuracy = 1),
    avg_compensation = scales::dollar(avg_compensation, accuracy = 1)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Yearly Summary: Total Compensation and Headcount") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

------------------------------------------------------------------------

## Step 12: Export Data

```{r}
#| label: export-data

# =============================================================================
# EXPORT DATA
# =============================================================================

# Ensure output directory exists
if (!base::dir.exists(paths$data_clean)) {
  base::dir.create(paths$data_clean, recursive = TRUE)
}

# Export
test_suffix <- base::ifelse(report_settings$test_mode, "_TEST", "")

rds_path <- base::file.path(paths$data_clean, base::paste0("compensation_long", test_suffix, ".rds"))
base::saveRDS(final_df, file = rds_path)

csv_path <- base::file.path(paths$data_clean, base::paste0("compensation_long", test_suffix, ".csv"))
readr::write_csv(final_df, file = csv_path)

# Also export the wide format for convenience
wide_rds_path <- base::file.path(paths$data_clean, base::paste0("compensation_wide", test_suffix, ".rds"))
base::saveRDS(all_data, file = wide_rds_path)

wide_csv_path <- base::file.path(paths$data_clean, base::paste0("compensation_wide", test_suffix, ".csv"))
readr::write_csv(all_data, file = wide_csv_path)

base::cat("Data exported:\n")
base::cat("  Long format:\n")
base::cat("    -", rds_path, "\n")
base::cat("    -", csv_path, "\n")
base::cat("  Wide format:\n")
base::cat("    -", wide_rds_path, "\n")
base::cat("    -", wide_csv_path, "\n")
```

------------------------------------------------------------------------

## Appendix A: Column Structure Comparison

This shows exactly what columns exist in each year's file, helping identify when the spreadsheet structure changes.

```{r}
#| label: column-structure-comparison

# =============================================================================
# APPENDIX A: COLUMN STRUCTURE COMPARISON
# =============================================================================

# Build a comparison of actual column names across years
col_comparison <- purrr::map_dfr(
  base::names(files),
  function(fy) {
    filepath <- base::file.path(data_folder, files[[fy]])
    
    if (base::file.exists(filepath)) {
      df <- readxl::read_excel(filepath) |>
        janitor::clean_names()
      
      dplyr::tibble(
        fiscal_year = fy,
        column_name = base::names(df)
      )
    } else {
      NULL
    }
  }
) |>
  dplyr::mutate(present = TRUE) |>
  tidyr::pivot_wider(
    names_from = fiscal_year,
    values_from = present,
    values_fill = FALSE
  )

col_comparison |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.logical), ~ dplyr::if_else(.x, "✓", ""))
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Column Presence by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

------------------------------------------------------------------------

## Appendix B: Data Quality Checks

```{r}
#| label: data-quality-checks

# =============================================================================
# APPENDIX B: DATA QUALITY CHECKS
# =============================================================================

# Check for missing names
missing_names <- all_data |>
  dplyr::filter(base::is.na(name) | name == "" | name == "NA, NA") |>
  dplyr::select(fiscal_year, name, department, job_title)

if (base::nrow(missing_names) > 0) {
  base::cat("WARNING: Found", base::nrow(missing_names), "records with missing names:\n")
  base::print(missing_names)
} else {
  base::cat("No missing names found.\n")
}

# Check for duplicate employee keys within a year
duplicates <- all_data |>
  dplyr::group_by(fiscal_year, employee_key) |>
  dplyr::filter(dplyr::n() > 1) |>
  dplyr::arrange(fiscal_year, employee_key)

if (base::nrow(duplicates) > 0) {
  base::cat("\nWARNING: Found", base::nrow(duplicates), "duplicate employee keys:\n")
  duplicates |>
    dplyr::select(fiscal_year, name, department, job_title) |>
    base::print()
} else {
  base::cat("No duplicate employee keys found within fiscal years.\n")
}

# Check for negative amounts
negative_amounts <- all_data |>
  dplyr::filter(
    leave_payout < 0 | regular_earnings < 0 | overtime_earnings < 0 |
    additional_earnings < 0 | deployment_earnings < 0 | arbitration_settlements < 0 |
    additional_benefits < 0
  )

if (base::nrow(negative_amounts) > 0) {
  base::cat("\nNOTE: Found", base::nrow(negative_amounts), "records with negative amounts (may be corrections/adjustments)\n")
} else {
  base::cat("No negative amounts found.\n")
}
```

------------------------------------------------------------------------

## Appendix C: Year-over-Year Comparison

```{r}
#| label: yoy-comparison

# =============================================================================
# APPENDIX C: YEAR-OVER-YEAR COMPARISON
# =============================================================================

yoy_summary <- yearly_summary |>
  dplyr::mutate(
    prev_total = dplyr::lag(total_compensation),
    prev_employees = dplyr::lag(total_employees),
    total_change = total_compensation - prev_total,
    total_pct_change = (total_change / prev_total) * 100,
    employee_change = total_employees - prev_employees
  )

yoy_summary |>
  dplyr::mutate(
    total_compensation = scales::dollar(total_compensation, accuracy = 1),
    avg_compensation = scales::dollar(avg_compensation, accuracy = 1),
    total_change = scales::dollar(total_change, accuracy = 1),
    total_pct_change = base::sprintf("%.1f%%", total_pct_change)
  ) |>
  dplyr::select(
    fiscal_year, total_employees, employee_change,
    total_compensation, total_change, total_pct_change, avg_compensation
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Year-over-Year Comparison") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

------------------------------------------------------------------------

## Appendix D: Top Earners by Year

```{r}
#| label: top-earners

# =============================================================================
# APPENDIX D: TOP EARNERS BY YEAR
# =============================================================================

top_earners <- all_data |>
  dplyr::group_by(fiscal_year) |>
  dplyr::slice_max(order_by = calc_total_compensation, n = 10) |>
  dplyr::select(
    fiscal_year, name, department, job_title, calc_total_compensation
  ) |>
  dplyr::mutate(
    rank = dplyr::row_number()
  ) |>
  dplyr::ungroup()

top_earners |>
  dplyr::mutate(
    calc_total_compensation = scales::dollar(calc_total_compensation, accuracy = 1)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Top 10 Earners by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

------------------------------------------------------------------------

## Appendix E: Department Summary

```{r}
#| label: department-summary

# =============================================================================
# APPENDIX E: DEPARTMENT SUMMARY
# =============================================================================

dept_summary <- all_data |>
  dplyr::group_by(fiscal_year, department) |>
  dplyr::summarise(
    employees = dplyr::n(),
    total_comp = base::sum(calc_total_compensation, na.rm = TRUE),
    avg_comp = base::mean(calc_total_compensation, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(fiscal_year, dplyr::desc(total_comp))

# Show FY24 as example
dept_summary |>
  dplyr::filter(fiscal_year == "FY24") |>
  dplyr::mutate(
    total_comp = scales::dollar(total_comp, accuracy = 1),
    avg_comp = scales::dollar(avg_comp, accuracy = 1)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Department Summary - FY24") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

------------------------------------------------------------------------

## Appendix F: Overtime Analysis

```{r}
#| label: overtime-analysis

# =============================================================================
# APPENDIX F: OVERTIME ANALYSIS
# =============================================================================

overtime_summary <- all_data |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    total_overtime = base::sum(overtime_earnings, na.rm = TRUE),
    total_regular = base::sum(regular_earnings, na.rm = TRUE),
    overtime_pct = (total_overtime / total_regular) * 100,
    employees_with_ot = base::sum(overtime_earnings > 0),
    total_employees = dplyr::n(),
    pct_with_ot = (employees_with_ot / total_employees) * 100,
    .groups = "drop"
  )

overtime_summary |>
  dplyr::mutate(
    total_overtime = scales::dollar(total_overtime, accuracy = 1),
    total_regular = scales::dollar(total_regular, accuracy = 1),
    overtime_pct = base::sprintf("%.1f%%", overtime_pct),
    pct_with_ot = base::sprintf("%.1f%%", pct_with_ot)
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Overtime Analysis by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

------------------------------------------------------------------------

## How to Add a New Fiscal Year

When you receive a new fiscal year's file:

1.  **Add the filename** to the `files` list in Step 1

2.  **Run Phase 1** to see the exact column names:

``` r
new_file <- readxl::read_excel(here::here("data_raw", "NEW_FILE.xlsx"))
base::names(janitor::clean_names(new_file))
```

3.  **Add a new entry to `column_maps`** in Step 4 with the exact column names

4.  **Run the pipeline** - Validation will tell you if any columns are missing

5.  **Update column names** if the spreadsheet structure has changed

This explicit approach means you'll never silently lose data when the spreadsheet structure changes.

------------------------------------------------------------------------

## Summary

| Year    | Structure                              | Notes                    |
|---------|----------------------------------------|--------------------------|
| FY20-22 | Single `name` column, no benefits      | Total = Total Earnings   |
| FY23-24 | Split `last_name`/`first_name`, has benefits | Full Total Compensation |

If validation shows **FAIL**, check the column mappings in Step 4.
