---
title: "US Census Data"
authors: 
  - name: "Dan Swart, CPA (ret)"
  - name: "Claude Sonnet 4.5"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: false     # Fast for drafts. Override for sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    # fig-width: 13
    E fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false


---


```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED


# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)


#
# # Sample Code:
# flextable::flextable(violations) |>
 #  flextable::set_header_labels(
 #    Variable = "Variable",
 #    Measurement = "Measurement",
 #    Likely_Impact = "Likely Impact"
 #  ) |>
#    flextable::add_header_lines(values = "Frequent Violations of Scientific Method in Current So-Called 'Equity' Research") |>
#   flextable::color(i = 1, color = "blue", part = "header") |>
#   flextable::italic(i = 1, part = "header") |>
#   flextable::align(i = 1, align = "center", part = "header") |>
#   flextable::fontsize(i = 1, size = 14, part = "header") |>
#   flextable::bg(i = 1, bg = "white", part = "header") |>
#   flextable::bg(i = 2, bg = "palegreen", part = "header") |>
#   flextable::bold(i = 1:2, part = "header") |>
#   flextable::bold(i = 1:7, j = 1, part = "body") |>
#   ftExtra::colformat_md() |> 
#   flextable::autofit()
  

# Flextable built-in themes:
  # flextable::theme_alafoli()	|>  # BLAH
  # flextable::theme_apa()  # THIS IS NICE
  # flextable::theme_booktabs() |>  # NICE, MORE COMPACT
  # flextable::theme_box() |>   # OK, INCLUDES CELL BORDERS
  # flextable::theme_tron() |>  # 'DARK MODE' BLUE TEXT
  # flextable::theme_tron_legacy() |>   # 'DARK MODE' YELLOW TEXT
  # flextable::theme_vader() |>    # 'DARK MODE' WHITE TEXT
  # flextable::theme_vanilla() |>   # NOT SPECIAL
  # flextable::theme_zebra()	|>
  #

# Flextable titles:
  # flextable::colformat_double(j = c("Mean", "SD", "N"), big.mark = ",", digits = 1) |>
  # flextable::flextable(variance_comparison) |>
  #  flextable::add_header_lines(values = "The Within-Group vs Between-Group Variance Problem") |>

# Flextable Title theming at table-level:
  #  flextable::color(i = 1, color = "blue", part = "header") |>
  #  flextable::italic(i = 1, part = "header") |>
  #  flextable::align(i = 1, align = "center", part = "header") |>
  #  flextable::fontsize(i = 1, size = 14, part = "header") |>
  #  flextable::bg(i = 1, bg = "white", part = "header") |>

# Flextable standard table background colors:
  #  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  

# Flextable reading markdown:
  #  ftExtra::colformat_md() |> 

# Flextable auto-sizing cell widths:
  #  flextable::autofit() 
 
 # Flextable background based on SPECIFIC cell contents:
  #
  # flextable::bg(i = ~ Impact_on_Validity == "High", j = "Impact_on_Validity", bg = "#ffcccc") |>
 # flextable::bg(i = ~ Impact_on_Validity == "Medium", j = "Impact_on_Validity", bg = "#ffffcc") |>
  #
  # 
  # Apply yellow background to any cell containing "Yes":

  # for (col in base::names(hypotheses_data)) {
  #   yes_rows <- base::which(hypotheses_data[[col]] == "Yes")
  #   if (base::length(yes_rows) > 0) {
  #     ft <- ft |>
  #       flextable::bg(i = yes_rows, j = col, bg = "yellow", part = "body")
  #   }
  # } 
  #
  # Apply to last row of table:
  #
  #



# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 16,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 14,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 14,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 14),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 12,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "bottom",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)



  
# Set seed for reproducibility
base::set.seed(123)

```


You can get a clean Schertz series for 2010–2024 with tidycensus, but you have to use two different mechanisms:

2010–2019 → time_series = TRUE against the 2019 vintage (Population Estimates API)

2020–2024 → flat-file approach with vintage = 2024 and variables = "POPESTIMATE"




## POPULATION GROWTH


### Get 2010–2019 via time_series = TRUE

2010–2019 using 'tidycensus::get_estimates() with argument 'time_series = TRUE'

This uses one call to get the full time series, then filters Schertz and maps the DATE code to calendar years using the Census mapping (DATE 3 = July 1, 2010; …; DATE 12 = July 1, 2019 → year = DATE + 2007). 


```{r}
library(tidycensus)
library(dplyr)

# 2010–2019 from Population Estimates API (2010-based series, vintage 2019)
schertz_pop_2010_2019 <- tidycensus::get_estimates(
  geography   = "place",
  state       = "TX",
  product     = "population",
  year        = 2019,      # 2010-based series, last vintage
  time_series = TRUE
) |>
  # keep Schertz only & total population
  filter(
    NAME == "Schertz city, Texas",
    variable == "POP"
  ) |>
  # Keep annual Oct 1 estimates. Census PEP estimates are as of July 1, not October 1—so there's no exact fiscal year match available. only (DATE 3–12)
  filter(DATE >= 3, DATE <= 12) |>
  mutate(
    year = DATE + 2007       # DATE 3 -> 2010, 4 -> 2011, ..., 12 -> 2019
  ) |>
  select(NAME, year, variable, value) |>
  arrange(year)

flextable::flextable(schertz_pop_2010_2019)
```




### Get 2020–2024 population data via flat file (post-2020)

```{r}
library(purrr)

years_2020_2024 <- 2020:2024

schertz_pop_2020_2024 <- map_dfr(
  years_2020_2024,
  \(yr) get_estimates(
    geography = "place",
    state     = "TX",
    place     = "Schertz",
    variables = "POPESTIMATE",  # post-2020 PEP
    vintage   = 2024,
    year      = yr
  ) |>
    filter(NAME == "Schertz city, Texas") |>
    mutate(
      year     = yr,
      variable = "POP"          # make name match pre-2020 chunk
    )
) |>
  select(NAME, year, variable, value) |>
  arrange(year)

flextable::flextable(schertz_pop_2020_2024)

```


### Combine both into a single 2010–2024 series

```{r}
schertz_pop_2010_2024 <- bind_rows(
  schertz_pop_2010_2019,
  schertz_pop_2020_2024
) |>
  arrange(year)

flextable::flextable(schertz_pop_2010_2024)

```


### Add change and percent change to tibble

```{r}
schertz_pop_2010_2024_with_changes <- schertz_pop_2010_2024 |>
  arrange(year) |>
  mutate(
    pop        = value,
    pop_lag    = lag(pop),
    change     = pop - pop_lag,
    pct_change = 100 * change / pop_lag
  )

flextable::flextable(schertz_pop_2010_2024_with_changes)

dplyr::glimpse(schertz_pop_2010_2024_with_changes)

```



### UNCOMMENT AND SAVE RESULTS AS EXCEL

```{r}
# writexl::write_xlsx(
#   schertz_pop_2010_2024_with_changes,
#   "RAW-schertz_pop_2010_2024_with_changes.xlsx"
# )
```





## HOUSING GROWTH

### Get Housing Units and Changes from ACS5

```{r}
#| label: get-housing-units_and_changes

years <- 2009:2022

schertz_housing_2009_2022_and_changes <- purrr::map(years, \(yr) {
  tidycensus::get_acs(
    geography = "place",
    state = "TX",
    variables = "B25001_001",
    year = yr,
    survey = "acs5"
  ) |> 
    dplyr::mutate(year = yr)
}) |> 
  purrr::list_rbind() |> 
  dplyr::filter(stringr::str_detect(NAME, "Schertz")) |> 
  dplyr::arrange(year) |> 
  dplyr::mutate(
    change = estimate - dplyr::lag(estimate),
    pct_change = change / dplyr::lag(estimate) * 100
  )

flextable::flextable(schertz_housing_2009_2022_and_changes)

```


### UNCOMMENT AND SAVE RESULTS AS EXCEL

```{r}
# writexl::write_xlsx(
#   schertz_housing_2009_2022_and_changes,
#   "schertz_housing_2009_2022.xlsx"
# )
```





## PLOT RESULTS

### Annual Change in Number of Housing Units

```{r}


# Plot 1: Annual change in number of housing units
ggplot2::ggplot(
  schertz_housing_2009_2022_and_changes,
  ggplot2::aes(x = year, y = change)
) +
  ggplot2::geom_col(fill = "steelblue") +
  ggplot2::labs(
    title = "Annual Change in Housing Units",
    subtitle = "Schertz, TX (ACS 5-Year Estimates)",
    x = NULL,
    y = "Units Added"
  ) +
  ggplot2::scale_x_continuous(breaks = years)
  # ggplot2::theme_minimal()

```

### Annual Percentage Change in Housing Units

```{r}

# Plot 2: Annual percentage change
ggplot2::ggplot(
  schertz_housing_2009_2022_and_changes,
  ggplot2::aes(x = year, y = pct_change / 100)
) +
  ggplot2::geom_col(fill = "darkgreen") +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  ggplot2::labs(
    title = "Annual Percentage Change in Housing Units",
    subtitle = "Schertz, TX (ACS 5-Year Estimates)",
    x = NULL,
    y = NULL
  ) +
  ggplot2::scale_x_continuous(breaks = years) 
  # ggplot2::theme_minimal()


```




