---
title: "FBI Violent and Property Crime Series for Schertz, TX"
subtitle: "2011-08 through 2026-01"
date: today
format:
  html:
    echo: true
    toc: true
    toc-depth: 2
    code-fold: true
  typst:
    echo: false
    toc: true
execute:
  warning: false
  message: false
  error: false
---

```{r}
#| label: setup
#| include: false

fmt_num <- function(x, digits = 0) {
  formatC(x, format = "f", digits = digits, big.mark = ",")
}

fmt_pct <- function(x, digits = 1) {
  paste0(formatC(x, format = "f", digits = digits), "%")
}

ds_style_flextable <- function(ft, title = NULL) {
  if (!base::is.null(title)) {
    ft <- flextable::add_header_lines(ft, values = title)
    ft <- flextable::color(ft, i = 1, part = "header", color = "blue")
    ft <- flextable::italic(ft, i = 1, part = "header")
    ft <- flextable::align(ft, i = 1, part = "header", align = "left")
    ft <- flextable::bold(ft, i = 1, part = "header")
    ft <- flextable::fontsize(ft, i = 1, part = "header", size = 16)
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "white")
    ft <- flextable::bg(ft, i = 2, part = "header", bg = "palegreen")
  } else {
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "palegreen")
  }
  flextable::autofit(ft)
}
```

## What this file can and cannot answer

The Excel export confirms that the uploaded file contains **one combined monthly crime series**, not separate violent and property series.

The metadata row shows the query included both:

- `offense : violent-crime`
- `offense : property-crime`

So this workbook supports a defensible analysis of the **combined monthly crime load** only.

It does **not** support separate longitudinal analyses of violent crime and property crime, because those two components are not split into separate rows or columns in the file.

### Practical implication

I can answer:

- Is the **combined** crime load now materially higher than in the past?

I cannot answer from this file alone:

- Is **violent crime** now materially higher than in the past?
- Is **property crime** now materially higher than in the past?

Those two questions require separate FBI exports, one for violent crime and one for property crime, or a workbook with those series split.

## Import the Excel series

```{r}
#| label: import-excel-series

raw <- readxl::read_excel(
  "schertz_violent_and_nonviolent_crime_2010-01-01_2026-01-01.xlsx"
)

crime_ts <- raw |>
  dplyr::transmute(
    date = as.Date(Date),
    crime_count = as.numeric(query1),
    ucl_excel = as.numeric(UCL),
    avg_excel = as.numeric(Average),
    lcl_excel = as.numeric(LCL)
  ) |>
  dplyr::filter(!is.na(date), !is.na(crime_count)) |>
  dplyr::arrange(date) |>
  dplyr::mutate(
    year = as.integer(format(date, "%Y")),
    month = as.integer(format(date, "%m"))
  )

metadata_df <- data.frame(
  Item = c(
    "Series type in workbook",
    "Date range in workbook",
    "First month",
    "Last month",
    "Rows in series"
  ),
  Value = c(
    "Single combined monthly series",
    paste0(min(crime_ts$date), " to ", max(crime_ts$date)),
    as.character(min(crime_ts$date)),
    as.character(max(crime_ts$date)),
    nrow(crime_ts)
  ),
  stringsAsFactors = FALSE
)

flextable::flextable(metadata_df) |>
  flextable::set_header_labels(Item = "Metadata Item", Value = "Value") |>
  ds_style_flextable(title = "What the Excel Workbook Actually Contains")
```

## Main finding

The combined monthly crime-load series does **not** show that Schertz PD is now under significantly increased crime load compared with the past.

The broader pattern is:

- the long-run trend is **down**
- recent years are **well below** the 2010s peak
- the Individuals/XmR chart shows **no sustained current high-load pattern**

There is one notable recent high-side point, but not a run of unusually high recent values.

```{r}
#| label: annual-summary-calc

crime_ts_full <- crime_ts |>
  dplyr::filter(year <= 2025)

annual_totals <- crime_ts_full |>
  dplyr::group_by(year) |>
  dplyr::summarise(
    annual_total = base::sum(crime_count, na.rm = TRUE),
    avg_monthly = base::mean(crime_count, na.rm = TRUE),
    median_monthly = stats::median(crime_count, na.rm = TRUE),
    max_month = base::max(crime_count, na.rm = TRUE),
    min_month = base::min(crime_count, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    yoy_change_pct = (annual_total / dplyr::lag(annual_total) - 1) * 100
  )

pre_2020_mean  <- base::mean(crime_ts_full$crime_count[crime_ts_full$year <= 2019])
post_2020_mean <- base::mean(crime_ts_full$crime_count[crime_ts_full$year >= 2020])
recent_2024_25 <- base::mean(crime_ts_full$crime_count[crime_ts_full$year >= 2024])

peak_row <- annual_totals |>
  dplyr::slice_max(order_by = annual_total, n = 1)
```

```{r}
#| label: summary-table

summary_df <- data.frame(
  Metric = c(
    "Average monthly count, 2011-2019",
    "Average monthly count, 2020-2025",
    "Average monthly count, 2024-2025",
    "Peak annual total",
    "2024 annual total",
    "2025 annual total"
  ),
  Value = c(
    fmt_num(pre_2020_mean, 1),
    fmt_num(post_2020_mean, 1),
    fmt_num(recent_2024_25, 1),
    paste0(fmt_num(peak_row$annual_total, 0), " in ", peak_row$year),
    fmt_num(annual_totals$annual_total[annual_totals$year == 2024], 0),
    fmt_num(annual_totals$annual_total[annual_totals$year == 2025], 0)
  ),
  stringsAsFactors = FALSE
)

flextable::flextable(summary_df) |>
  flextable::set_header_labels(Metric = "Metric", Value = "Value") |>
  ds_style_flextable(title = "Combined Crime Load Summary")
```

## Combined annual trend

```{r}
#| label: annual-plot
#| fig-width: 11
#| fig-height: 6

ggplot2::ggplot(
  annual_totals,
  ggplot2::aes(x = year, y = annual_total)
) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_point(size = 2.5) +
  ggplot2::geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  ggplot2::scale_x_continuous(breaks = annual_totals$year) +
  ggplot2::labs(
    title = "Combined Crime Load Has Trended Down, Not Up",
    subtitle = "Peak annual total is in the mid-2010s; recent years are much lower",
    x = "Year",
    y = "Annual combined count"
  ) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```

```{r}
#| label: annual-table

annual_display <- annual_totals |>
  dplyr::mutate(
    annual_total = fmt_num(annual_total, 0),
    avg_monthly = fmt_num(avg_monthly, 1),
    median_monthly = fmt_num(median_monthly, 1),
    max_month = fmt_num(max_month, 0),
    min_month = fmt_num(min_month, 0),
    yoy_change_pct = dplyr::if_else(
      base::is.na(yoy_change_pct),
      "",
      fmt_pct(yoy_change_pct, 1)
    )
  )

flextable::flextable(annual_display) |>
  flextable::set_header_labels(
    year = "Year",
    annual_total = "Annual Total",
    avg_monthly = "Avg Monthly",
    median_monthly = "Median Monthly",
    max_month = "Max Month",
    min_month = "Min Month",
    yoy_change_pct = "YoY % Change"
  ) |>
  ds_style_flextable(title = "Annual Combined Crime Load by Year")
```

## Individuals/XmR interpretation for the combined series

```{r}
#| label: xmr-calc

x <- crime_ts_full$crime_count
mr <- base::abs(x - dplyr::lag(x))
xbar <- base::mean(x, na.rm = TRUE)
mrbar <- base::mean(mr, na.rm = TRUE)
ucl <- xbar + 2.66 * mrbar
lcl <- base::max(0, xbar - 2.66 * mrbar)

crime_ts_xmr <- crime_ts_full |>
  dplyr::mutate(
    center_line = xbar,
    ucl = ucl,
    lcl = lcl,
    out_of_control = crime_count > ucl | crime_count < lcl
  )

ooc_points <- crime_ts_xmr |>
  dplyr::filter(out_of_control) |>
  dplyr::transmute(
    Month = format(date, "%Y-%m"),
    Count = crime_count,
    Direction = dplyr::if_else(crime_count > ucl, "Above UCL", "Below LCL")
  )
```

```{r}
#| label: xmr-plot
#| fig-width: 11
#| fig-height: 6

ggplot2::ggplot(crime_ts_xmr, ggplot2::aes(x = date, y = crime_count)) +
  ggplot2::geom_line(linewidth = 0.6) +
  ggplot2::geom_point(
    ggplot2::aes(shape = out_of_control),
    size = 2
  ) +
  ggplot2::geom_line(ggplot2::aes(y = center_line), linetype = "dashed") +
  ggplot2::geom_line(ggplot2::aes(y = ucl), linetype = "dotted") +
  ggplot2::geom_line(ggplot2::aes(y = lcl), linetype = "dotted") +
  ggplot2::scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), guide = "none") +
  ggplot2::labs(
    title = "Individuals/XmR View of Combined Monthly Crime Load",
    subtitle = paste0(
      "Center line = ", fmt_num(xbar, 1),
      " | UCL = ", fmt_num(ucl, 1),
      " | LCL = ", fmt_num(lcl, 1)
    ),
    x = "Month",
    y = "Monthly combined count"
  ) +
  ggplot2::theme_minimal(base_size = 14)
```

```{r}
#| label: xmr-summary-table

xmr_summary <- data.frame(
  Measure = c(
    "Center line (mean monthly count)",
    "Upper control limit",
    "Lower control limit",
    "Out-of-control points",
    "Interpretation"
  ),
  Value = c(
    fmt_num(xbar, 1),
    fmt_num(ucl, 1),
    fmt_num(lcl, 1),
    fmt_num(base::nrow(ooc_points), 0),
    "Recent period does not show a sustained unusually high pattern"
  ),
  stringsAsFactors = FALSE
)

flextable::flextable(xmr_summary) |>
  flextable::set_header_labels(Measure = "Measure", Value = "Value") |>
  ds_style_flextable(title = "XmR Summary for the Combined Series")
```

```{r}
#| label: ooc-table

flextable::flextable(ooc_points) |>
  ds_style_flextable(title = "Out-of-Control Months in the Combined Series")
```

## Direct answer to your research question

### Combined crime trend

Based on the Excel-confirmed combined series, the answer is **no**: the Schertz Police Department does **not** appear to be under significantly increased crime load now than in the past.

The strongest signals are:

1. The long-run history points **downward**, not upward.
2. Recent years are **well below** the earlier peak period.
3. The XmR chart does **not** show a sustained recent high-load pattern.

### Violent crime trend

This workbook does **not** contain a separate violent-crime time series, so that question cannot be answered from this file alone.

### Property crime trend

This workbook does **not** contain a separate property-crime time series, so that question cannot be answered from this file alone.

## What to export next if you want the full answer

To answer your full question cleanly, export three separate FBI series:

1. **Violent crime only**
2. **Property crime only**
3. **Combined violent + property crime**

With those three files, I can produce a complete side-by-side QMD covering:

- violent crime trend
- property crime trend
- combined trend
- separate annual summaries
- separate Individuals/XmR charts
- a final policy conclusion on whether present staffing claims are supported by the long-run data
