---
title: "ACS 5-Year Estimates of Per Capita Income & Earnings in Schertz, TX"
subtitle: "2012 - 2024"
author: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true     # Set to 'false' for drafts. Override if you're NOT sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: false
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: false
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: false
    error: false
    warning: false
    message: false
    cache: false

prefer-html: true

---


```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)




# # load libraries - NOT NEEDED

# Download button helpers live here; knitr::is_html_output() guard is inside
# the helpers so they silently no-op when rendering to typst
source(here::here("R", "download_button_helpers.R"))

# census helpers
source(here::here("R", "census_helpers.R"))


# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)



# Flextable built-in themes:
  # flextable::theme_alafoli()	|>  # BLAH
  # flextable::theme_apa()  # THIS IS NICE
  # flextable::theme_booktabs() |>  # NICE, MORE COMPACT
  # flextable::theme_box() |>   # OK, INCLUDES CELL BORDERS
  # flextable::theme_tron() |>  # 'DARK MODE' BLUE TEXT
  # flextable::theme_tron_legacy() |>   # 'DARK MODE' YELLOW TEXT
  # flextable::theme_vader() |>    # 'DARK MODE' WHITE TEXT
  # flextable::theme_vanilla() |>   # NOT SPECIAL
  # flextable::theme_zebra()	|>
  #


# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 16,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 14,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 14,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 14),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 12,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)


claude <- ellmer::chat_anthropic(model = "claude-opus-4-6-20260205")
  
# Set seed for reproducibility
base::set.seed(123)

```

\

#### **Filename:** 
acs5_estimates_schertz_per_capita_income_and_earnings_
analysis_2012_2024_with_buttons.qmd

\



```{r}
#| label: params
#| include: false

# =============================================================================
# PARAMETERS - CHANGE THESE FOR A DIFFERENT CITY/YEARS
# =============================================================================

params <- base::list(
  city       = "Schertz",
  state      = "TX",
  years      = 2012:2024,
  base_year  = 2024,
  cache_dir  = "cache/census"
)

# =============================================================================
```

```{r}
#| label: packages-and-options
#| include: false

# -----------------------------------------------------------------------------
# Package verification
# -----------------------------------------------------------------------------
required_pkgs <- c(
  "tidycensus", "tigris", "dplyr", "purrr", "tibble", "stringr", "tidyr",
  "digest", "lubridate", "httr2", "readxl", "ggplot2", "flextable",
  "readr", "writexl"
)

missing <- required_pkgs[
  !base::vapply(required_pkgs, requireNamespace, logical(1L), quietly = TRUE)
]

if (base::length(missing) > 0L) {
  base::stop("Install missing packages:\n", base::paste(missing, collapse = ", "))
}

# -----------------------------------------------------------------------------
# Options
# -----------------------------------------------------------------------------
base::options(tigris_use_cache = TRUE)
```


# Overview

This document retrieves per capita income and earnings data for **`r params$city`, `r params$state`** from the American Community Survey (ACS) 5-year estimates.

**Variables retrieved:**

| Variable | Description |
|----------|-------------|
| B19301_001 | Per Capita Income (all ages) |
| B20017_001 | Median Earnings (workers 16+ with earnings) |
| B19013_001 | Median Household Income |

**Note:** The Census Bureau does not publish a direct "per capita earnings for adults 20+" variable. The closest proxies are:

- **Per Capita Income (B19301)**: Total income divided by total population (all ages)
- **Median Earnings (B20017)**: Median of earnings for workers 16+ who had earnings

If you need earnings specifically for the 20+ population, you would need to use PUMS microdata.


# ACS Data Retrieval

```{r}
#| label: pull-data

# ACS variables for income/earnings
acs_vars <- c(
  per_capita_income   = "B19301_001",
  median_earnings     = "B20017_001",
  median_hh_income    = "B19013_001"
)

# Pull ACS data
acs_data <- ds_get_acs_place(
  city      = params$city,
  state     = params$state,
  years     = params$years,
  variables = base::unname(acs_vars),
  survey    = "acs5",
  cache_dir = params$cache_dir
)

# Add friendly labels
var_labels <- stats::setNames(base::names(acs_vars), acs_vars)
acs_data$variable_label <- var_labels[acs_data$series]   # was acs_data$variable
```

```{r}
#| label: cpi-data

# Get CPI for inflation adjustment
cpi_tbl <- ds_get_cpiu_rs_annual(
  start_year = base::min(params$years),
  end_year   = base::max(params$years),
  base_year  = params$base_year,
  cache_dir  = params$cache_dir
)

# -----------------------------------------------------------------------------
# Create inflation rate data frame (year-over-year % change in CPI)
# This object is available in the global environment for flextable/ggplot2
# -----------------------------------------------------------------------------
inflation_rate <- cpi_tbl |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    cpi_prior = dplyr::lag(cpi_annual_avg),
    inflation_pct = (cpi_annual_avg - cpi_prior) / cpi_prior * 100
  ) |>
  dplyr::filter(!base::is.na(inflation_pct)) |>
  dplyr::select(year, cpi_annual_avg, inflation_pct)

# Join CPI to ACS data and calculate real values
acs_data <- dplyr::left_join(acs_data, cpi_tbl[, c("year", "factor_to_base")], by = "year")
acs_data$estimate_real <- acs_data$estimate * acs_data$factor_to_base
acs_data$moe_real <- acs_data$moe * acs_data$factor_to_base
```


# Annual Inflation Rate

Year-over-year percent change in CPI-U-RS (Research Series Using Current Methods) — the BLS series that back-applies current methodology uniformly, recommended by the Census Bureau for deflating ACS income data.

```{r}
#| label: inflation-rate-table

inflation_display <- inflation_rate |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    Year = base::as.character(year),
    `Inflation Rate` = base::paste0(base::format(base::round(inflation_pct, 1), nsmall = 1), "%")
  ) |>
  dplyr::select(Year, `Inflation Rate`)

ft_inflation <- flextable::flextable(inflation_display)
ft_inflation <- ds_style_flextable(ft_inflation, title = "Annual Inflation Rate (CPI-U-RS)")
ft_inflation
```

::: {.content-visible when-format="html"}
```{r}
#| label: inflation-rate-download
#| echo: false
ds_download_data_btn(
  inflation_display,
  filename_base = "schertz_annual_inflation_rate",
  label = "Download Inflation Rate Data (CSV)",
  clean_df = inflation_rate
)
```
:::

```{r}
#| label: inflation-rate-chart
#| fig-width: 9
#| fig-height: 5

p_inflation <- ggplot2::ggplot(inflation_rate, ggplot2::aes(x = year, y = inflation_pct)) +
  ggplot2::geom_col(fill = "steelblue", width = 0.7) +
  ggplot2::geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  ggplot2::geom_text(
    ggplot2::aes(label = base::paste0(base::round(inflation_pct, 1), "%")),
    vjust = -0.5,
    size = 3
  ) +
  ggplot2::scale_x_continuous(breaks = inflation_rate$year) +
  ggplot2::scale_y_continuous(
    labels = function(x) base::paste0(x, "%"),
    expand = ggplot2::expansion(mult = c(0.05, 0.15))
  ) +
  ggplot2::labs(
    title = "Annual Inflation Rate",
    subtitle = "Year-over-year percent change in CPI-U-RS",
    x = "Year",
    y = "Inflation Rate",
    caption = "Source: BLS CPI-U-RS Research Series (r-cpi-u-rs-allitems.xlsx)"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    plot.title = ggplot2::element_text(face = "bold")
  )
print(p_inflation)
```

::: {.content-visible when-format="html"}
```{r}
#| label: inflation-rate-plot-download
#| echo: false
ds_download_plot_btn(
  p_inflation,
  filename_base = "schertz_annual_inflation_rate",
  label         = "Download Inflation Chart (PNG)"
)
```
:::


# Per Capita Income

Per capita income for all ages (B19301_001).

```{r}
#| label: per-capita-table

pci <- acs_data |>
  dplyr::filter(variable_label == "per_capita_income") |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    Year = base::as.character(year),
    Nominal = base::paste0("$", base::format(base::round(estimate, 0), big.mark = ",")),
    Real = base::paste0("$", base::format(base::round(estimate_real, 0), big.mark = ",")),
    MOE = base::paste0("\u00b1$", base::format(base::round(moe, 0), big.mark = ","))
  ) |>
  dplyr::select(Year, Nominal, Real, MOE)

ft_pci <- flextable::flextable(pci)
ft_pci <- ds_style_flextable(
  ft_pci,
  title = base::paste0(
    "Per Capita Income: ", params$city, ", ", params$state,
    " (Real values in ", params$base_year, " dollars)"
  )
)
ft_pci
```

::: {.content-visible when-format="html"}
```{r}
#| label: per-capita-download
#| echo: false
ds_download_data_btn(
  pci,
  filename_base = "schertz_per_capita_income",
  label = "Download Per Capita Income (CSV)",
  clean_df = acs_data |>
    dplyr::filter(variable_label == "per_capita_income") |>
    dplyr::select(year, estimate, moe, estimate_real, moe_real, factor_to_base) |>
    dplyr::arrange(year)
)
```
:::

```{r}
#| label: per-capita-chart
#| fig-width: 9
#| fig-height: 5

pci_chart <- acs_data |>
  dplyr::filter(variable_label == "per_capita_income") |>
  dplyr::arrange(year)

p_pci <- ggplot2::ggplot(pci_chart, ggplot2::aes(x = year)) +
  ggplot2::geom_ribbon(
    ggplot2::aes(
      ymin = estimate_real - moe_real,
      ymax = estimate_real + moe_real
    ),
    fill = "lightblue",
    alpha = 0.5
  ) +
  ggplot2::geom_line(ggplot2::aes(y = estimate_real), color = "blue", linewidth = 1) +
  ggplot2::geom_point(ggplot2::aes(y = estimate_real), color = "blue", size = 2) +
  ggplot2::scale_x_continuous(breaks = pci_chart$year) +
  ggplot2::scale_y_continuous(
    labels = function(x) base::paste0("$", base::format(x, big.mark = ","))
  ) +
  ggplot2::labs(
    title = base::paste0("Per Capita Income: ", params$city, ", ", params$state),
    subtitle = base::paste0("ACS 5-Year Estimates (", params$base_year, " dollars)"),
    x = "Year",
    y = "Per Capita Income",
    caption = "Shaded area represents margin of error. Source: ACS B19301_001"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    plot.title = ggplot2::element_text(face = "bold")
  )
print(p_pci)
```

::: {.content-visible when-format="html"}
```{r}
#| label: per-capita-plot-download
#| echo: false
ds_download_plot_btn(
  p_pci,
  filename_base = "schertz_per_capita_income",
  label         = "Download Per Capita Income Chart (PNG)",
  width = 9, height = 5
)
```
:::


# Median Earnings (Workers 16+ with Earnings)

Median earnings for the population 16 years and over with earnings (B20017_001).

```{r}
#| label: median-earnings-table

me <- acs_data |>
  dplyr::filter(variable_label == "median_earnings") |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    Year = base::as.character(year),
    Nominal = base::paste0("$", base::format(base::round(estimate, 0), big.mark = ",")),
    Real = base::paste0("$", base::format(base::round(estimate_real, 0), big.mark = ",")),
    MOE = base::paste0("\u00b1$", base::format(base::round(moe, 0), big.mark = ","))
  ) |>
  dplyr::select(Year, Nominal, Real, MOE)

ft_me <- flextable::flextable(me)
ft_me <- ds_style_flextable(
  ft_me,
  title = base::paste0(
    "Median Earnings (Workers 16+): ", params$city, ", ", params$state,
    " (Real values in ", params$base_year, " dollars)"
  )
)
ft_me
```

::: {.content-visible when-format="html"}
```{r}
#| label: median-earnings-download
#| echo: false
ds_download_data_btn(
  me,
  filename_base = "schertz_median_earnings",
  label = "Download Median Earnings (CSV)",
  clean_df = acs_data |>
    dplyr::filter(variable_label == "median_earnings") |>
    dplyr::select(year, estimate, moe, estimate_real, moe_real, factor_to_base) |>
    dplyr::arrange(year)
)
```
:::

```{r}
#| label: median-earnings-chart
#| fig-width: 9
#| fig-height: 5

me_chart <- acs_data |>
  dplyr::filter(variable_label == "median_earnings") |>
  dplyr::arrange(year)

p_me <- ggplot2::ggplot(me_chart, ggplot2::aes(x = year)) +
  ggplot2::geom_ribbon(
    ggplot2::aes(
      ymin = estimate_real - moe_real,
      ymax = estimate_real + moe_real
    ),
    fill = "lightgreen",
    alpha = 0.5
  ) +
  ggplot2::geom_line(ggplot2::aes(y = estimate_real), color = "darkgreen", linewidth = 1) +
  ggplot2::geom_point(ggplot2::aes(y = estimate_real), color = "darkgreen", size = 2) +
  ggplot2::scale_x_continuous(breaks = me_chart$year) +
  ggplot2::scale_y_continuous(
    labels = function(x) base::paste0("$", base::format(x, big.mark = ","))
  ) +
  ggplot2::labs(
    title = base::paste0("Median Earnings: ", params$city, ", ", params$state),
    subtitle = base::paste0("Workers 16+ with Earnings, ACS 5-Year (", params$base_year, " dollars)"),
    x = "Year",
    y = "Median Earnings",
    caption = "Shaded area represents margin of error. Source: ACS B20017_001"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    plot.title = ggplot2::element_text(face = "bold")
  )
print(p_me)
```

::: {.content-visible when-format="html"}
```{r}
#| label: median-earnings-plot-download
#| echo: false
ds_download_plot_btn(
  p_me,
  filename_base = "schertz_median_earnings",
  label         = "Download Median Earnings Chart (PNG)",
  width = 9, height = 5
)
```
:::


# Median Household Income

Median household income (B19013_001) for reference.

```{r}
#| label: median-hh-income-table

mhi <- acs_data |>
  dplyr::filter(variable_label == "median_hh_income") |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    Year = base::as.character(year),
    Nominal = base::paste0("$", base::format(base::round(estimate, 0), big.mark = ",")),
    Real = base::paste0("$", base::format(base::round(estimate_real, 0), big.mark = ",")),
    MOE = base::paste0("\u00b1$", base::format(base::round(moe, 0), big.mark = ","))
  ) |>
  dplyr::select(Year, Nominal, Real, MOE)

ft_mhi <- flextable::flextable(mhi)
ft_mhi <- ds_style_flextable(
  ft_mhi,
  title = base::paste0(
    "Median Household Income: ", params$city, ", ", params$state,
    " (Real values in ", params$base_year, " dollars)"
  )
)
ft_mhi
```

::: {.content-visible when-format="html"}
```{r}
#| label: median-hh-income-download
#| echo: false
ds_download_data_btn(
  mhi,
  filename_base = "schertz_median_hh_income",
  label = "Download Median HH Income (CSV)",
  clean_df = acs_data |>
    dplyr::filter(variable_label == "median_hh_income") |>
    dplyr::select(year, estimate, moe, estimate_real, moe_real, factor_to_base) |>
    dplyr::arrange(year)
)
```
:::


# Summary Comparison

All three measures side-by-side (inflation-adjusted to `r params$base_year` dollars).

```{r}
#| label: comparison-table

comparison <- acs_data |>
  dplyr::select(year, variable_label, estimate_real) |>
  tidyr::pivot_wider(
    names_from = variable_label,
    values_from = estimate_real
  ) |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    Year = base::as.character(year),
    `Per Capita Income` = base::paste0(
      "$", base::format(base::round(per_capita_income, 0), big.mark = ",")
    ),
    `Median Earnings` = base::paste0(
      "$", base::format(base::round(median_earnings, 0), big.mark = ",")
    ),
    `Median HH Income` = base::paste0(
      "$", base::format(base::round(median_hh_income, 0), big.mark = ",")
    )
  ) |>
  dplyr::select(Year, `Per Capita Income`, `Median Earnings`, `Median HH Income`)

ft_comp <- flextable::flextable(comparison)
ft_comp <- ds_style_flextable(
  ft_comp,
  title = base::paste0(
    "Income & Earnings Comparison: ", params$city, ", ", params$state,
    " (", params$base_year, " dollars)"
  )
)
ft_comp
```

::: {.content-visible when-format="html"}
```{r}
#| label: comparison-download
#| echo: false
ds_download_data_btn(
  comparison,
  filename_base = "schertz_income_earnings_comparison",
  label = "Download Comparison Table (CSV)",
  clean_df = acs_data |>
    dplyr::select(year, variable_label, estimate, moe, estimate_real, moe_real) |>
    tidyr::pivot_wider(
      id_cols = year,
      names_from = variable_label,
      values_from = c(estimate_real, moe_real)
    ) |>
    dplyr::arrange(year)
)
```
:::


# CPI Inflation Factors

Reference table showing CPI-W annual averages used for inflation adjustment.

```{r}
#| label: cpi-table

cpi_display <- cpi_tbl |>
  dplyr::mutate(
    Year = base::as.character(year),
    `CPI-U-RS Avg` = base::round(cpi_annual_avg, 2),
    `Base Year` = base::as.character(base_year),
    Factor = base::round(factor_to_base, 4)
  ) |>
  dplyr::select(Year, `CPI-U-RS Avg`, `Base Year`, Factor)

ft_cpi <- flextable::flextable(cpi_display)
ft_cpi <- ds_style_flextable(
  ft_cpi,
  title = base::paste0("CPI-U-RS Inflation Factors (Base: ", params$base_year, ")")
)
ft_cpi
```

::: {.content-visible when-format="html"}
```{r}
#| label: cpi-table-download
#| echo: false
ds_download_data_btn(
  cpi_display,
  filename_base = "schertz_cpi_inflation_factors",
  label = "Download CPI Factors (CSV)",
  clean_df = cpi_tbl
)
```
:::


# Export Data

```{r}
#| label: export-data

# Create output directory
base::dir.create("out", showWarnings = FALSE, recursive = TRUE)

# Prepare clean export data
export_data <- acs_data |>
  dplyr::select(
    year,
    series,           # census_helpers.R returns 'series', not 'variable'
    variable_label,
    estimate_nominal = estimate,
    moe_nominal = moe,
    estimate_real,
    moe_real,
    cpi_factor = factor_to_base
  ) |>
  dplyr::arrange(variable_label, year)

# File name base
file_base <- base::paste0(
  "out/",
  base::tolower(base::gsub(" ", "_", params$city)),
  "_",
  params$state,
  "_earnings_income_",
  base::min(params$years),
  "_",
  base::max(params$years)
)

# Save as RDS
base::saveRDS(export_data, base::paste0(file_base, ".rds"))

# Save inflation_rate as separate RDS for easy reuse
base::saveRDS(inflation_rate, base::paste0(file_base, "_inflation_rate.rds"))

# Save as CSV
readr::write_csv(export_data, base::paste0(file_base, ".csv"))
readr::write_csv(inflation_rate, base::paste0(file_base, "_inflation_rate.csv"))

# Save as Excel (3 sheets)
writexl::write_xlsx(
 base::list(
   earnings_income = export_data,
   inflation_rate = inflation_rate,
   cpi_factors = cpi_tbl
 ),
 base::paste0(file_base, ".xlsx")
)
```

::: {.content-visible when-format="html"}
```{r}
#| label: export-data-download
#| echo: false
ds_download_data_btn(
  export_data,
  filename_base = "schertz_earnings_income_export",
  label = "Download Full Export Data (CSV)"
)
```
:::

Data exported to `out/` directory:

- **RDS**: `r base::paste0(file_base, ".rds")` — Income/earnings data
- **RDS**: `r base::paste0(file_base, "_inflation_rate.rds")` — Inflation rate data
- **CSV**: `r base::paste0(file_base, ".csv")` — Income/earnings data
- **CSV**: `r base::paste0(file_base, "_inflation_rate.csv")` — Inflation rate data
- **Excel**: `r base::paste0(file_base, ".xlsx")` — Three sheets: earnings_income, inflation_rate, cpi_factors


# Using This Template

To analyze a different city, change these values in the `params` code chunk:

```r
params <- base::list(
  city       = "Austin",
  state      = "TX",
  years      = 2012:2023,
  base_year  = 2023,
  cache_dir  = "cache/census"
)
```

Clear cache when switching cities:

```r
unlink("cache", recursive = TRUE)
```


# Notes

- **Per Capita Income (B19301)**: Total income of all residents divided by total population. Includes all ages and all income types (wages, investments, transfers, etc.).

- **Median Earnings (B20017)**: Median of earnings for workers 16+ who had earnings in the past 12 months. Only includes wage/salary income and self-employment income.

- **ACS 5-Year Estimates**: End-year labels (e.g., 2023 = 2019-2023 data). More reliable for small geographies but reflects a 5-year period, not a single point in time.

- **For Adults 20+ Specifically**: You would need PUMS microdata to filter to exactly 20+ population. The B20017 variable covers 16+, and B19301 covers all ages.

- **Caching**: All API calls are cached to RDS files in the `cache/census` directory to minimize traffic. Clear with `unlink("cache", recursive = TRUE)`.


# Export Data

```{r}
#| label: export-data-2

# Create output directory
base::dir.create("out", showWarnings = FALSE, recursive = TRUE)

# Prepare export data
export_data <- acs_data |>
  dplyr::select(
    year,
    series,           # census_helpers.R returns 'series', not 'variable'
    variable_label,
    estimate_nominal = estimate,
    moe_nominal = moe,
    estimate_real,
    moe_real,
    cpi_factor = factor_to_base
  ) |>
  dplyr::arrange(variable_label, year)

# File paths
file_base <- base::paste0(
  "out/",
  base::tolower(base::gsub(" ", "_", params$city)),
  "_",
  params$state,
  "_income_earnings_",
  base::min(params$years),
  "_",
  base::max(params$years)
)

# Save as RDS
base::saveRDS(export_data, base::paste0(file_base, ".rds"))

# Save as CSV
readr::write_csv(export_data, base::paste0(file_base, ".csv"))

# Save as Excel
writexl::write_xlsx(
  base::list(
    income_earnings = export_data,
    cpi_factors = cpi_tbl
  ),
  base::paste0(file_base, ".xlsx")
)
```

::: {.content-visible when-format="html"}
```{r}
#| label: export-download
#| echo: false
ds_download_data_btn(
  export_data,
  filename_base = "schertz_income_earnings_export",
  label = "Download Income & Earnings Data (CSV)"
)
```
:::

Data exported to `out/` directory:

- **RDS**: `r base::paste0(file_base, ".rds")` — R native format, preserves all data types
- **CSV**: `r base::paste0(file_base, ".csv")` — Universal format
- **Excel**: `r base::paste0(file_base, ".xlsx")` — Two sheets: income_earnings + cpi_factors
