---
title: "Longitudinal Census Data Template"
subtitle: "PEP Population + ACS Socioeconomics + Decennial Anchors + Inflation Adjustment"
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    echo: true
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: false     # Fast for drafts. Override for sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    echo: false
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    error: false
    warning: false
    message: false
    cache: false


---

```{r}
#| label: params
#| include: false

# =============================================================================
# PARAMETERS - CHANGE THESE FOR A DIFFERENT CITY
# =============================================================================

params <- base::list(
  city                = "Schertz",
  state               = "TX",
  years_pep           = 2010:2024,
  years_acs           = 2011:2023,
  base_year           = 2024,
  cache_census_dir    = "cache/census",
  cache_inflation_dir = "cache/inflation"
)

# =============================================================================
```


```{r}
#| label: setup
#| include: false

# -----------------------------------------------------------------------------
# Package verification
# -----------------------------------------------------------------------------
required_pkgs <- c(
"tidycensus", "tigris", "dplyr", "purrr", "tibble", "stringr", "readr",
"digest", "lubridate", "quantmod", "zoo", "rlang", "ggplot2",
"flextable", "qicharts2", "downloadthis"
)

missing <- required_pkgs[
!base::vapply(required_pkgs, requireNamespace, logical(1L), quietly = TRUE)
]

if (base::length(missing) > 0L) {
base::stop(
  "Install missing packages:\n",
  base::paste(missing, collapse = ", ")
)
}

# -----------------------------------------------------------------------------
# Source helper functions
# -----------------------------------------------------------------------------
source("R/census_helpers.R")
# source("R/skew_check.R")
# source("R/quick_distrib.R")
# source("R/distrib_ridge.R")
# source("R/distrib_quantile_evolution.R")
# source("R/distrib_acceleration.R")
source("R/download_button_helpers.R")
# source("R/plot_city_pop.R")
# source("R/get_city_pep.R")
# source("R/get_city_pumas.R")
# source("R/get_pums_data.R")


# Standard table styler: blue italic title row + palegreen column header row.
# Defined here so all QMDs that source this setup get the same presentation
# without depending on census_helpers.R for a purely cosmetic function.
ds_style_flextable <- function(ft, title = NULL) {
  if (!base::is.null(title)) {
    ft <- flextable::add_header_lines(ft, values = title)
    ft <- flextable::color(ft, i = 1, part = "header", color = "blue")
    ft <- flextable::italic(ft, i = 1, part = "header")
    ft <- flextable::align(ft, i = 1, part = "header", align = "left")
    ft <- flextable::fontsize(ft, i = 1, part = "header", size = 12)
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "white")
    ft <- flextable::bg(ft, i = 2, part = "header", bg = "palegreen")
  } else {
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "palegreen")
  }
  ft <- flextable::autofit(ft)
  ft
}


# -----------------------------------------------------------------------------
# Options
# -----------------------------------------------------------------------------
# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)

# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 16,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 14,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 14,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 14),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 12,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)



  
# Set seed for reproducibility
base::set.seed(123)

```


\

#### **Filename:** 
city_census_template_with_buttons.qmd

\


# Overview

This template generates a unified longitudinal dataset for **`r params$city`, `r params$state`** combining:

- **PEP**: Annual population estimates (2010–2024)
- **ACS 5-Year**: Selected socioeconomic indicators
- **Decennial**: Census anchors (2000, 2010, 2020)
- **CPI-U**: Inflation adjustment factors for constant-dollar analysis

All API calls are cached to minimize traffic. Output uses a normalized schema: `source`, `series`, `year`, `estimate`, `moe`.


# PEP Annual Population

Population Estimates Program data for `r params$city`.

```{r}
#| label: pep-pull

pep <- ds_get_pep_population_place(
city      = params$city,
state     = params$state,
years     = params$years_pep,
cache_dir = params$cache_census_dir
)
```

```{r}
#| label: pep-table

pep_display <- pep |>
dplyr::arrange(series, year) |>
dplyr::mutate(
  Year = base::as.character(year),
  Series = series,
  Estimate = base::format(
    base::round(estimate, 1),
    big.mark = ",",
    scientific = FALSE
  )
) |>
dplyr::select(Year, Series, Estimate)

ft_pep <- flextable::flextable(pep_display)
ft_pep <- ds_style_flextable(
ft_pep,
title = base::paste0("PEP Population: ", params$city, ", ", params$state)
)
ft_pep
```

::: {.content-visible when-format="html"}
```{r}
#| label: pep-table-download
#| echo: false
ds_download_data_btn(
  pep_display,
  filename_base = base::paste0(base::tolower(params$city), "_pep_population"),
  label         = "Download PEP Data (CSV)"
)
```
:::


# ACS 5-Year Indicators

Selected American Community Survey variables. Modify `acs_vars` below to change which indicators are pulled.

```{r}
#| label: acs-vars

# -------------------------------------------------------------------------
# ACS VARIABLE CODES - MODIFY THIS SECTION FOR DIFFERENT INDICATORS
# -------------------------------------------------------------------------
# Format: friendly_name = "TABLE_VARIABLE"
# Common examples:
#   B19013_001 = Median household income
#   B25001_001 = Total housing units
#   B01003_001 = Total population (ACS estimate)
#   B25077_001 = Median home value
#   B23025_005 = Unemployed (civilian labor force)
# -------------------------------------------------------------------------

acs_vars <- c(
median_hh_income = "B19013_001",
housing_units    = "B25001_001",
population_acs   = "B01003_001"
)
```

```{r}
#| label: acs-pull

acs <- ds_get_acs_place(
city      = params$city,
state     = params$state,
years     = params$years_acs,
variables = base::unname(acs_vars),
survey    = "acs5",
cache_dir = params$cache_census_dir
)

# Map variable codes to friendly labels
var_labels <- stats::setNames(base::names(acs_vars), acs_vars)
acs$series_label <- var_labels[acs$series]
acs$series_label <- base::ifelse(
base::is.na(acs$series_label),
acs$series,
acs$series_label
)
```

```{r}
#| label: acs-table

acs_display <- acs |>
dplyr::arrange(series_label, year) |>
dplyr::mutate(
  Year = base::as.character(year),
  Estimate = base::format(
    base::round(estimate, 0),
    big.mark = ",",
    scientific = FALSE
  ),
  MOE = base::ifelse(
    base::is.na(moe),
    "—",
    base::format(base::round(moe, 0), big.mark = ",")
  )
) |>
dplyr::select(
  Year,
  Variable = series_label,
  Estimate,
  MOE
)

ft_acs <- flextable::flextable(acs_display)
ft_acs <- ds_style_flextable(
ft_acs,
title = base::paste0("ACS 5-Year Estimates: ", params$city, ", ", params$state)
)
ft_acs
```

::: {.content-visible when-format="html"}
```{r}
#| label: acs-table-download
#| echo: false
ds_download_data_btn(
  acs_display,
  filename_base = base::paste0(base::tolower(params$city), "_acs_estimates"),
  label         = "Download ACS Data (CSV)"
)
```
:::


# Decennial Census Anchors

Official Census counts from 2000, 2010, and 2020.

```{r}
#| label: decennial-pull

# Variable codes differ by census year
# 2000/2010 SF1: P001001 (total population)
# 2020 PL: P1_001N (total population)

dec_2000 <- ds_get_decennial_place(
params$city, params$state,
year = 2000L, variable = "P001001", sumfile = "sf1",
cache_dir = params$cache_census_dir
)

dec_2010 <- ds_get_decennial_place(
params$city, params$state,
year = 2010L, variable = "P001001", sumfile = "sf1",
cache_dir = params$cache_census_dir
)

dec_2020 <- ds_get_decennial_place(
params$city, params$state,
year = 2020L, variable = "P1_001N", sumfile = "pl",
cache_dir = params$cache_census_dir
)

decennial <- dplyr::bind_rows(dec_2000, dec_2010, dec_2020)
decennial$series <- "population_decennial"
```

```{r}
#| label: decennial-table

dec_display <- decennial |>
dplyr::arrange(year) |>
dplyr::mutate(
  Year = base::as.character(year),
  Population = base::format(estimate, big.mark = ",", scientific = FALSE)
) |>
dplyr::select(Year, Source = source, Population)

ft_dec <- flextable::flextable(dec_display)
ft_dec <- ds_style_flextable(
ft_dec,
title = base::paste0("Decennial Census: ", params$city, ", ", params$state)
)
ft_dec
```

::: {.content-visible when-format="html"}
```{r}
#| label: decennial-table-download
#| echo: false
ds_download_data_btn(
  dec_display,
  filename_base = base::paste0(base::tolower(params$city), "_decennial_census"),
  label         = "Download Decennial Data (CSV)"
)
```
:::


# Unified Longitudinal Dataset

Combined dataset with normalized schema across all sources.

```{r}
#| label: unify

# PEP data already has series column (population, density_per_sq_mile)
pep_long <- pep
pep_long$series_label <- base::paste0("pep_", pep_long$series)

acs_long <- acs
acs_long$series_label <- base::ifelse(
base::is.na(acs_long$series_label),
acs_long$series,
acs_long$series_label
)

dec_long <- decennial
dec_long$series_label <- "population_decennial"

longitudinal <- dplyr::bind_rows(pep_long, acs_long, dec_long)
longitudinal <- longitudinal |>
dplyr::select(source, series_label, year, estimate, moe) |>
dplyr::arrange(series_label, year)
```

```{r}
#| label: longitudinal-summary

summary_tbl <- longitudinal |>
dplyr::group_by(series_label) |>
dplyr::summarise(
  year_min = base::min(year, na.rm = TRUE),
  year_max = base::max(year, na.rm = TRUE),
  n_obs    = dplyr::n(),
  .groups  = "drop"
) |>
dplyr::mutate(
  year_min = base::as.character(year_min),
  year_max = base::as.character(year_max)
)

ft_summary <- flextable::flextable(summary_tbl)
ft_summary <- flextable::set_header_labels(
ft_summary,
series_label = "Series",
year_min     = "First Year",
year_max     = "Last Year",
n_obs        = "Observations"
)
ft_summary <- ds_style_flextable(ft_summary, title = "Longitudinal Dataset Summary")
ft_summary
```

::: {.content-visible when-format="html"}
```{r}
#| label: longitudinal-summary-download
#| echo: false
ds_download_data_btn(
  longitudinal,
  filename_base = base::paste0(base::tolower(params$city), "_longitudinal_all_series"),
  label         = "Download Longitudinal Data (CSV)"
)
```
:::


# Inflation Adjustment

CPI-U annual averages from FRED for converting nominal dollars to `r params$base_year` constant dollars.

```{r}
#| label: cpi-pull

year_min <- base::min(c(params$years_acs, params$years_pep), na.rm = TRUE)
year_max <- base::max(c(params$years_acs, params$years_pep), na.rm = TRUE)

cpi_tbl <- ds_get_cpiu_rs_annual(
start_year = year_min,
end_year   = year_max,
base_year  = params$base_year,
cache_dir  = params$cache_inflation_dir
)
```

```{r}
#| label: cpi-table

cpi_display <- cpi_tbl |>
dplyr::mutate(
  Year = base::as.character(year),
  `Base Year` = base::as.character(base_year),
  `CPI-U Avg` = base::round(cpi_annual_avg, 2),
  Factor = base::round(factor_to_base, 4)
) |>
dplyr::select(Year, `CPI-U Avg`, `Base Year`, Factor)

ft_cpi <- flextable::flextable(cpi_display)
ft_cpi <- ds_style_flextable(
ft_cpi,
title = base::paste0("CPI-U Inflation Factors (Base: ", params$base_year, ")")
)
ft_cpi
```

::: {.content-visible when-format="html"}
```{r}
#| label: cpi-table-download
#| echo: false
ds_download_data_btn(
  cpi_display,
  filename_base = base::paste0(base::tolower(params$city), "_cpi_factors"),
  label         = "Download CPI Data (CSV)"
)
```
:::


# Real Income Analysis

Median household income adjusted to `r params$base_year` dollars.
```{r}
#| label: income-real

income_nominal <- acs_long |>
dplyr::filter(series_label == "median_hh_income") |>
dplyr::select(source, series_label, year, estimate, moe)

if (base::nrow(income_nominal) > 0L) {
income_real <- ds_adjust_to_base_dollars(
  df        = income_nominal,
  value_col = "estimate",
  year_col  = "year",
  cpi_tbl   = cpi_tbl,
  out_col   = "estimate_real"
)
} else {
income_real <- tibble::tibble()
}
```

```{r}
#| label: income-table
#| eval: !expr nrow(income_real) > 0

income_display <- income_real |>
dplyr::arrange(year) |>
dplyr::mutate(
  Year          = base::as.character(year),
  Nominal       = base::paste0("$", base::format(base::round(estimate, 0), big.mark = ",")),
  Real          = base::paste0("$", base::format(base::round(estimate_real, 0), big.mark = ","))
) |>
dplyr::select(Year, Nominal, Real)

ft_income <- flextable::flextable(income_display)
ft_income <- ds_style_flextable(
ft_income,
title = base::paste0("Median Household Income (", params$base_year, " Dollars)")
)
ft_income
```

::: {.content-visible when-format="html"}
```{r}
#| label: income-table-download
#| echo: false
#| eval: !expr nrow(income_real) > 0
ds_download_data_btn(
  income_display,
  filename_base = base::paste0(base::tolower(params$city), "_income_real"),
  label         = "Download Income Data (CSV)"
)
```
:::


# Population Trend (XmR Chart)

Statistical process control chart for PEP population estimates.

## Version 1: qicharts2

```{r}
#| label: xmr-chart-v1
#| fig-width: 8
#| fig-height: 5

# Filter to population only (not density) for the chart
pop_chart <- pep |>
dplyr::filter(series == "population") |>
dplyr::select(year, estimate) |>
dplyr::arrange(year)

p_xmr1 <- ds_plot_xmr(
df        = pop_chart,
year_col  = "year",
value_col = "estimate",
title     = base::paste0("Population XmR: ", params$city, ", ", params$state),
subtitle  = "PEP Annual Population Estimates (qicharts2)",
y_label   = "Population"
)
print(p_xmr1)
```

::: {.content-visible when-format="html"}
```{r}
#| label: xmr-chart-v1-download
#| echo: false
ds_download_plot_btn(
  p_xmr1,
  filename_base = base::paste0(base::tolower(params$city), "_population_xmr_v1"),
  label         = "Download XmR Chart v1 (PNG)",
  width = 8, height = 5
)
```
:::

## Version 2: Custom ggplot2

```{r}
#| label: xmr-chart-v2
#| fig-width: 9
#| fig-height: 5

p_xmr2 <- ds_plot_xmr2(
df        = pop_chart,
year_col  = "year",
value_col = "estimate",
title     = base::paste0("Population XmR: ", params$city, ", ", params$state),
subtitle  = "PEP Annual Population Estimates (custom ggplot2)",
y_label   = "Population"
)
print(p_xmr2)
```

::: {.content-visible when-format="html"}
```{r}
#| label: xmr-chart-v2-download
#| echo: false
ds_download_plot_btn(
  p_xmr2,
  filename_base = base::paste0(base::tolower(params$city), "_population_xmr_v2"),
  label         = "Download XmR Chart v2 (PNG)",
  width = 9, height = 5
)
```
:::


# Export Data

```{r}
#| label: export

base::dir.create("out", showWarnings = FALSE, recursive = TRUE)

readr::write_csv(
longitudinal,
base::file.path("out", "longitudinal_place_series.csv")
)

readr::write_csv(
cpi_tbl,
base::file.path("out", "inflation_cpi_table.csv")
)

if (base::nrow(income_real) > 0L) {
readr::write_csv(
  income_real,
  base::file.path("out", "income_real_base_year.csv")
)
}
```

Data exported to `out/` directory:

- `longitudinal_place_series.csv` — All series combined
- `inflation_cpi_table.csv` — CPI factors
- `income_real_base_year.csv` — Inflation-adjusted income

::: {.content-visible when-format="html"}
```{r}
#| label: export-download
#| echo: false
ds_download_data_btn(
  longitudinal,
  filename_base = base::paste0(
    base::tolower(params$city), "_", params$state,
    "_all_series_longitudinal"
  ),
  label = "Download Complete Dataset (CSV)"
)
```
:::


# Using This Template for Other Cities

To analyze a different city, change only these values in the `params` code chunk at the top of the document:

```r
params <- base::list(
  city  = "Austin",
  state = "TX",
  ...
)
```

Everything else adapts automatically. Clear your cache directory if switching cities:

```r
unlink("cache", recursive = TRUE)
```


## Notes

- **1990 Decennial**: Not supported by tidycensus for place-level data. Use NHGIS for earlier years.
- **ACS Years**: ACS 5-year estimates are labeled by end-year. First available: 2009 (covering 2005–2009).
- **PEP 2020+**: Uses tidycensus which reads Census flat files directly.
