---
title: "Government Employee Earnings Analysis of PUMS Microdata By Level of Government - Local, State, Federal"
subtitle: "2023 - 2024"
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
format:
  html:
    default: true         
    code-copy: true
    code-link: true
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: false
    include-in-header:
      - text: |
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    linestretch: 1.0

   
    
  typst:
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

editor: source

execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8

knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false
---

```{r}
#| label: params
#| include: false

params <- list(
  city       = "Schertz",
  state      = "TX",
  years      = c(2019, 2021, 2023, 2024),
  min_age    = 20L,
  survey     = "acs1",
  base_year  = 2024,
  cache_dir  = "cache/acs1_pums"
)
```

```{r}
#| label: setup
#| include: false

required_pkgs <- c(
  "tidycensus", "tigris", "dplyr", "purrr", "tibble", "stringr", "tidyr",
  "digest", "lubridate", "quantmod", "zoo", "ggplot2", "flextable",
  "srvyr", "survey", "readr", "writexl", "sf", "mapview", "httr", "ggridges",
  "downloadthis"
)

missing <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) stop("Install: ", paste(missing, collapse = ", "))

suppressPackageStartupMessages({
  library(tidycensus)
  library(tigris)
})

source("R/skew_check.R")
source("R/quick_distrib.R")
source("R/distrib_ridge.R")
source("R/distrib_quantile_evolution.R")
source("R/distrib_acceleration.R")
source("R/download_button_helpers.R")

options(
  tigris_use_cache = TRUE,
  tidycensus.show_progress = FALSE,
  survey.lonely.psu = "adjust",
  scipen = 999
)

httr::set_config(httr::config(noprogress = TRUE))
mapview::mapviewOptions(verbose = FALSE)
mapview::mapviewOptions(basemaps = "OpenStreetMap")

year_colors <- c(
  "2019" = "#1b9e77",
  "2021" = "#d95f02", 
  "2023" = "#7570b3",
  "2024" = "#e7298a"
)

ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 16) +
    ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = 18),
      plot.subtitle = ggplot2::element_text(size = 14),
      legend.position = "right"
    )
)

flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  padding.top = 3, 
  padding.bottom = 3
)

set.seed(123)
```


\

#### **Filename:** 
govt_employee_earnings_analysis_of_acs1_pums_microdata_  
by_level_of_govt_local_state_fed_with_buttons_and_  
caching_HTML_ONLY.qmd

\






```{r}
#| label: helper-functions
#| include: false

# =============================================================================
# CACHING - data is fetched once and reused on subsequent renders
# =============================================================================

cache_path <- function(cache_dir, key) {
  dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
  file.path(cache_dir, paste0(key, ".rds"))
}

cache_key <- function(prefix, ...) {
  digest::digest(list(prefix = prefix, ...), algo = "xxhash64")
}

cache_read <- function(path) {
  if (file.exists(path)) readRDS(path) else NULL
}

cache_write <- function(x, path) {
  dir.create(dirname(path), showWarnings = FALSE, recursive = TRUE)
  saveRDS(x, path)
  invisible(x)
}

# =============================================================================
# GET PUMAS FOR A CITY
# =============================================================================

get_city_pumas <- function(city, state, cache_dir) {
  key <- cache_key("city_pumas", city, state)
  path <- cache_path(cache_dir, key)
  cached <- cache_read(path)
  if (!is.null(cached)) return(cached)
  
  pumas <- suppressMessages(tigris::pumas(state = state, year = 2020, progress_bar = FALSE))
  places <- suppressMessages(tigris::places(state = state, year = 2020, progress_bar = FALSE))
  
  city_sf <- places[grepl(paste0("^", city), places$NAME, ignore.case = TRUE), ]
  if (nrow(city_sf) == 0) stop("City '", city, "' not found in ", state)
  
  city_sf <- sf::st_transform(city_sf, sf::st_crs(pumas))
  hits <- suppressMessages(sf::st_intersects(pumas, city_sf, sparse = FALSE))
  city_pumas <- pumas[apply(hits, 1, any), ]
  
  puma_col <- grep("^PUMACE", names(city_pumas), value = TRUE)[1]
  
  result <- list(
    puma_codes = city_pumas[[puma_col]],
    puma_names = city_pumas$NAMELSAD20,
    city_name  = city_sf$NAME[1],
    n_pumas    = nrow(city_pumas)
  )
  
  cache_write(result, path)
}

# =============================================================================
# GET PUMS DATA - cached to avoid repeated API calls
# =============================================================================

get_pums_data <- function(state, puma_codes, year, min_age, survey, cache_dir) {
  key <- cache_key("pums", state, puma_codes, year, min_age, survey)
  path <- cache_path(cache_dir, key)
  cached <- cache_read(path)
  if (!is.null(cached)) return(cached)
  
  # Fetch from API (only happens once per year, then cached)
  raw <- suppressMessages(
    tidycensus::get_pums(
      variables        = c("AGEP", "WAGP", "SEMP", "COW", "PUMA"),
      state            = state,
      year             = year,
      survey           = survey,
      variables_filter = list(AGEP = min_age:99L),
      rep_weights      = "person",
      recode           = FALSE,
      show_call        = FALSE
    )
  )
  
  if (is.null(raw) || nrow(raw) == 0) return(NULL)
  
  # Filter to target PUMAs
  raw <- raw[raw$PUMA %in% puma_codes, ]
  if (nrow(raw) == 0) return(NULL)
  
  # Standardize column types for binding across years
  raw$SERIALNO <- as.character(raw$SERIALNO)
  raw$SPORDER  <- as.character(raw$SPORDER)
  raw$PUMA     <- as.character(raw$PUMA)
  raw$STATE    <- as.character(raw$STATE)
  raw$COW      <- as.character(raw$COW)
  raw$AGEP     <- as.integer(raw$AGEP)
  raw$WAGP     <- as.numeric(raw$WAGP)
  raw$SEMP     <- as.numeric(raw$SEMP)
  raw$WGTP     <- as.numeric(raw$WGTP)
  raw$PWGTP    <- as.numeric(raw$PWGTP)
  
  # Standardize replicate weights
 for (i in 1:80) {
    col <- paste0("PWGTP", i)
    if (col %in% names(raw)) raw[[col]] <- as.numeric(raw[[col]])
  }
  
  # Calculate earnings
  raw$earnings <- raw$WAGP + raw$SEMP
  raw$earnings[is.na(raw$earnings)] <- 0
  raw$earnings[raw$earnings < 0] <- 0
  
  # Classify government level: COW 3=Local, 4=State, 5=Federal
  raw$gov_level <- dplyr::case_when(
    raw$COW == "3" ~ "Local Government",
    raw$COW == "4" ~ "State Government",
    raw$COW == "5" ~ "Federal Government",
    TRUE ~ NA_character_
  )
  
  raw$year <- as.integer(year)
  
  cache_write(raw, path)
}

# =============================================================================
# GET CPI DATA
# =============================================================================

get_cpi <- function(start_year, end_year, base_year, cache_dir) {
  key <- cache_key("cpi", start_year, end_year, base_year)
  path <- cache_path(cache_dir, key)
  cached <- cache_read(path)
  if (!is.null(cached)) return(cached)
  
  xt <- suppressMessages(suppressWarnings(
    quantmod::getSymbols("CPIAUCSL", src = "FRED", auto.assign = FALSE)
  ))
  
  df <- tibble::tibble(
    date = as.Date(zoo::index(xt)),
    cpi  = as.numeric(xt[, 1])
  )
  df$year <- lubridate::year(df$date)
  df <- df[df$year >= start_year & df$year <= end_year, ]
  
  annual <- df |>
    dplyr::group_by(year) |>
    dplyr::summarise(cpi_avg = mean(cpi, na.rm = TRUE), .groups = "drop")
  
  base_cpi <- annual$cpi_avg[annual$year == base_year]
  annual$factor_to_base <- base_cpi / annual$cpi_avg
  annual$base_year <- base_year
  
  cache_write(annual, path)
}

# =============================================================================
# STYLE FLEXTABLE
# =============================================================================

style_ft <- function(ft, title = NULL) {
  if (!is.null(title)) {
    ft <- flextable::add_header_lines(ft, values = title)
    ft <- flextable::color(ft, i = 1, part = "header", color = "blue")
    ft <- flextable::italic(ft, i = 1, part = "header")
    ft <- flextable::fontsize(ft, i = 1, part = "header", size = 12)
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "white")
    ft <- flextable::bg(ft, i = 2, part = "header", bg = "palegreen")
  } else {
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "palegreen")
  }
  flextable::autofit(ft)
}
```


# Overview

This document analyzes earnings for **government employees** (adults `r params$min_age`+) in the **`r params$city`, `r params$state` area** using ACS `r params$survey` PUMS microdata.

**Years analyzed:** `r paste(params$years, collapse = ", ")`

**Excluded years:**

- **2020:** No 1-year ACS released (COVID response rate issues)
- **2022:** PUMA boundary inconsistencies in tidycensus


# PUMA Identification

```{r}
#| label: find-pumas

puma_info <- get_city_pumas(params$city, params$state, params$cache_dir)
puma_codes <- puma_info$puma_codes

puma_tbl <- tibble::tibble(
  `PUMA Code` = puma_info$puma_codes,
  `PUMA Name` = puma_info$puma_names
)
puma_tbl |>
  flextable::flextable() |>
  style_ft(title = paste0("PUMAs Containing ", params$city, ", ", params$state))
```

::: {.content-visible when-format="html"}
```{r}
#| label: find-pumas-download
#| echo: false
ds_download_data_btn(
  puma_tbl,
  filename_base = paste0(tolower(params$city), "_puma_codes"),
  label = "Download PUMA Table (CSV)"
)
```
:::

```{r}
#| label: puma-map
#| fig-height: 6

pumas_sf <- suppressMessages(tigris::pumas(state = params$state, year = 2020, progress_bar = FALSE))
places_sf <- suppressMessages(tigris::places(state = params$state, year = 2020, progress_bar = FALSE))

puma_col <- grep("^PUMACE", names(pumas_sf), value = TRUE)[1]
city_pumas_sf <- pumas_sf[pumas_sf[[puma_col]] %in% puma_codes, ]
city_sf <- places_sf[grepl(paste0("^", params$city), places_sf$NAME, ignore.case = TRUE), ]

city_pumas_sf$label <- paste0("PUMA ", city_pumas_sf[[puma_col]])
city_sf$label <- params$city

m <- mapview::mapview(city_pumas_sf, col.regions = "lightblue", alpha.regions = 0.3,
                      layer.name = "PUMA Coverage", label = city_pumas_sf$label)
m <- m + mapview::mapview(city_sf, col.regions = "tomato", alpha.regions = 0.5,
                          layer.name = params$city, label = city_sf$label)
m
```

::: {.content-visible when-format="html"}
```{r}
#| label: puma-map-download
#| echo: false
local({
  tmp <- tempfile(fileext = ".html")
  htmlwidgets::saveWidget(m@map, file = tmp, selfcontained = TRUE)
  downloadthis::download_file(
    path         = tmp,
    output_name  = paste0(tolower(params$city), "_puma_map"),
    button_label = "Download Map (HTML)",
    button_type  = "warning",
    has_icon     = TRUE,
    icon         = "fa fa-map"
  )
})
```
:::


# Data Retrieval

```{r}
#| label: pull-pums
#| output: false

# -----------------------------------------------------------------------------
# Two-level cache strategy:
#
#   Level 1 (per-year, inside get_pums_data): caches the full per-year PUMS
#   download from the Census API so we never re-hit the API.
#
#   Level 2 (combined, below): caches pums_list + pums_all + gov_employees as
#   a single RDS after all years are fetched, column-trimmed, and bound.
#   On repeat renders this level-2 cache is loaded in one readRDS() call,
#   skipping the lapply + bind_rows entirely — the main source of slowness.
#
# Cache key includes all params that affect the combined output so it
# automatically invalidates when years, city, age cutoff, or survey change.
# -----------------------------------------------------------------------------

keep_cols <- c("SERIALNO", "SPORDER", "PUMA", "STATE", "COW", "AGEP",
               "WAGP", "SEMP", "WGTP", "PWGTP",
               paste0("PWGTP", 1:80),
               "earnings", "gov_level", "year")

combined_key  <- cache_key("pums_combined",
                            params$state, puma_codes,
                            params$years, params$min_age,
                            params$survey)
combined_path <- cache_path(params$cache_dir, combined_key)
combined_cache <- cache_read(combined_path)

if (!is.null(combined_cache)) {
  # Fast path: everything already assembled
  pums_list    <- combined_cache$pums_list
  pums_all     <- combined_cache$pums_all
  gov_employees <- combined_cache$gov_employees
  message("pull-pums: loaded from combined cache (", combined_path, ")")
} else {
  # Slow path: fetch per-year (uses per-year cache if available)
  pums_list <- lapply(params$years, function(y) {
    tryCatch(
      get_pums_data(params$state, puma_codes, y, params$min_age,
                    params$survey, params$cache_dir),
      error = function(e) { message("Year ", y, " failed: ", e$message); NULL }
    )
  })
  names(pums_list) <- params$years
  pums_list <- pums_list[!vapply(pums_list, is.null, logical(1))]

  # Trim to needed columns before binding (reduces memory)
  pums_list <- lapply(pums_list, function(df) {
    df[, intersect(keep_cols, names(df)), drop = FALSE]
  })

  pums_all      <- dplyr::bind_rows(pums_list)
  gov_employees <- pums_all |> dplyr::filter(!is.na(gov_level), earnings > 0)

  # Save combined cache
  cache_write(
    list(pums_list = pums_list, pums_all = pums_all, gov_employees = gov_employees),
    combined_path
  )
  message("pull-pums: combined cache saved (", combined_path, ")")
}
```

**Data summary:**

- Years retrieved: `r paste(names(pums_list), collapse = ", ")`
- Total records: `r format(nrow(pums_all), big.mark = ",")`
- Government employees (earnings > 0): `r format(nrow(gov_employees), big.mark = ",")`

```{r}
#| label: data-summary-table

data_summary_tbl <- tibble::tibble(
  Year = as.character(names(pums_list)),
  `Total Records` = vapply(pums_list, nrow, integer(1)),
  `Gov Employees` = vapply(pums_list, function(df) sum(!is.na(df$gov_level) & df$earnings > 0), integer(1))
)
data_summary_tbl |>
  flextable::flextable() |>
  style_ft(title = "Records by Year")
```

::: {.content-visible when-format="html"}
```{r}
#| label: data-summary-table-download
#| echo: false
ds_download_data_btn(
  data_summary_tbl,
  filename_base = paste0(tolower(params$city), "_data_summary"),
  label = "Download Summary Table (CSV)"
)
```
:::

```{r}
#| label: gov-level-table

gov_level_tbl <- as.data.frame.matrix(table(gov_employees$gov_level, gov_employees$year)) |>
  tibble::rownames_to_column("Government Level")
gov_level_tbl |>
  flextable::flextable() |>
  style_ft(title = "Government Employees by Level and Year")
```

::: {.content-visible when-format="html"}
```{r}
#| label: gov-level-table-download
#| echo: false
ds_download_data_btn(
  gov_level_tbl,
  filename_base = paste0(tolower(params$city), "_gov_level_counts"),
  label = "Download Gov Level Table (CSV)"
)
```
:::

```{r}
#| label: cpi-data

cpi <- get_cpi(min(params$years), max(params$years), params$base_year, params$cache_dir)

cpi_display <- cpi |>
  dplyr::mutate(
    Year = as.character(year),
    `CPI (Annual Avg)` = round(cpi_avg, 1),
    `Adjustment Factor` = round(factor_to_base, 4)
  ) |>
  dplyr::select(Year, `CPI (Annual Avg)`, `Adjustment Factor`)
cpi_display |>
  flextable::flextable() |>
  style_ft(title = paste0("CPI Adjustment Factors (Base Year: ", params$base_year, ")"))
```

::: {.content-visible when-format="html"}
```{r}
#| label: cpi-data-download
#| echo: false
ds_download_data_btn(
  cpi_display,
  filename_base = paste0(tolower(params$city), "_cpi_factors"),
  label = "Download CPI Data (CSV)"
)
```
:::


# Distribution Analysis

## Ridge Plots

```{r}
#| label: ridge-all
#| fig-width: 11
#| fig-height: 7

p_ridge_all <- distrib_ridge(
  data = gov_employees,
  value_col = "earnings",
  year_col = "year",
  title = "All Government Employees: Earnings Distribution Shift (2019-2024)"
)
print(p_ridge_all)
```

::: {.content-visible when-format="html"}
```{r}
#| label: ridge-all-download
#| echo: false
ds_download_plot_btn(
  p_ridge_all,
  filename_base = paste0(tolower(params$city), "_ridge_all"),
  label = "Download Ridge Plot (PNG)",
  width = 11, height = 7
)
```
:::

```{r}
#| label: ridge-by-level
#| fig-width: 11
#| fig-height: 10

p_ridge_level <- distrib_ridge(
  data = gov_employees,
  value_col = "earnings",
  year_col = "year",
  group_col = "gov_level",
  title = "Earnings Distribution by Government Level (2019-2024)"
)
print(p_ridge_level)
```

::: {.content-visible when-format="html"}
```{r}
#| label: ridge-by-level-download
#| echo: false
ds_download_plot_btn(
  p_ridge_level,
  filename_base = paste0(tolower(params$city), "_ridge_by_level"),
  label = "Download Ridge Plot by Level (PNG)",
  width = 11, height = 10
)
```
:::


## Quantile Evolution

```{r}
#| label: quantile-all
#| fig-width: 10
#| fig-height: 6

p_quantile_all <- distrib_quantile_evolution(
  data = gov_employees,
  value_col = "earnings",
  year_col = "year",
  title = "All Government: Percentile Evolution (2019-2024)"
)
print(p_quantile_all)
```

::: {.content-visible when-format="html"}
```{r}
#| label: quantile-all-download
#| echo: false
ds_download_plot_btn(
  p_quantile_all,
  filename_base = paste0(tolower(params$city), "_quantile_all"),
  label = "Download Quantile Plot (PNG)",
  width = 10, height = 6
)
```
:::

```{r}
#| label: quantile-by-level
#| fig-width: 12
#| fig-height: 8

p_quantile_level <- distrib_quantile_evolution(
  data = gov_employees,
  value_col = "earnings",
  year_col = "year",
  group_col = "gov_level",
  title = "Percentile Evolution by Government Level (2019-2024)"
)
print(p_quantile_level)
```

::: {.content-visible when-format="html"}
```{r}
#| label: quantile-by-level-download
#| echo: false
ds_download_plot_btn(
  p_quantile_level,
  filename_base = paste0(tolower(params$city), "_quantile_by_level"),
  label = "Download Quantile Plot by Level (PNG)",
  width = 12, height = 8
)
```
:::


## Overlaid Density Plots

```{r}
#| label: overlay-local
#| fig-width: 10
#| fig-height: 6

local_data <- gov_employees |> dplyr::filter(gov_level == "Local Government")

p_overlay_local <- NULL
if (nrow(local_data) > 0) {
  p_overlay_local <- ggplot2::ggplot(local_data, ggplot2::aes(x = earnings, fill = factor(year),
                                            color = factor(year), weight = PWGTP)) +
    ggplot2::geom_density(alpha = 0.3, linewidth = 1) +
    ggplot2::scale_fill_manual(values = year_colors, name = "Year") +
    ggplot2::scale_color_manual(values = year_colors, name = "Year") +
    ggplot2::scale_x_continuous(
      labels = function(x) paste0("$", format(x/1000, big.mark = ","), "K"),
      limits = c(0, quantile(local_data$earnings, 0.98))
    ) +
    ggplot2::labs(
      title = "Local Government: Earnings Distribution (2019-2024)",
      x = "Earnings", y = "Density"
    )
  print(p_overlay_local)
}
```

::: {.content-visible when-format="html"}
```{r}
#| label: overlay-local-download
#| echo: false
if (!is.null(p_overlay_local)) {
  ds_download_plot_btn(
    p_overlay_local,
    filename_base = paste0(tolower(params$city), "_overlay_local"),
    label = "Download Local Density Plot (PNG)",
    width = 10, height = 6
  )
}
```
:::

```{r}
#| label: overlay-state
#| fig-width: 10
#| fig-height: 6

state_data <- gov_employees |> dplyr::filter(gov_level == "State Government")

p_overlay_state <- NULL
if (nrow(state_data) > 0) {
  p_overlay_state <- ggplot2::ggplot(state_data, ggplot2::aes(x = earnings, fill = factor(year),
                                            color = factor(year), weight = PWGTP)) +
    ggplot2::geom_density(alpha = 0.3, linewidth = 1) +
    ggplot2::scale_fill_manual(values = year_colors, name = "Year") +
    ggplot2::scale_color_manual(values = year_colors, name = "Year") +
    ggplot2::scale_x_continuous(
      labels = function(x) paste0("$", format(x/1000, big.mark = ","), "K"),
      limits = c(0, quantile(state_data$earnings, 0.98))
    ) +
    ggplot2::labs(
      title = "State Government: Earnings Distribution (2019-2024)",
      x = "Earnings", y = "Density"
    )
  print(p_overlay_state)
}
```

::: {.content-visible when-format="html"}
```{r}
#| label: overlay-state-download
#| echo: false
if (!is.null(p_overlay_state)) {
  ds_download_plot_btn(
    p_overlay_state,
    filename_base = paste0(tolower(params$city), "_overlay_state"),
    label = "Download State Density Plot (PNG)",
    width = 10, height = 6
  )
}
```
:::

```{r}
#| label: overlay-federal
#| fig-width: 10
#| fig-height: 6

federal_data <- gov_employees |> dplyr::filter(gov_level == "Federal Government")

p_overlay_federal <- NULL
if (nrow(federal_data) > 0) {
  p_overlay_federal <- ggplot2::ggplot(federal_data, ggplot2::aes(x = earnings, fill = factor(year),
                                              color = factor(year), weight = PWGTP)) +
    ggplot2::geom_density(alpha = 0.3, linewidth = 1) +
    ggplot2::scale_fill_manual(values = year_colors, name = "Year") +
    ggplot2::scale_color_manual(values = year_colors, name = "Year") +
    ggplot2::scale_x_continuous(
      labels = function(x) paste0("$", format(x/1000, big.mark = ","), "K"),
      limits = c(0, quantile(federal_data$earnings, 0.98))
    ) +
    ggplot2::labs(
      title = "Federal Government: Earnings Distribution (2019-2024)",
      x = "Earnings", y = "Density"
    )
  print(p_overlay_federal)
}
```

::: {.content-visible when-format="html"}
```{r}
#| label: overlay-federal-download
#| echo: false
if (!is.null(p_overlay_federal)) {
  ds_download_plot_btn(
    p_overlay_federal,
    filename_base = paste0(tolower(params$city), "_overlay_federal"),
    label = "Download Federal Density Plot (PNG)",
    width = 10, height = 6
  )
}
```
:::


## Acceleration Analysis

```{r}
#| label: acceleration

accel <- distrib_acceleration(
  data = gov_employees,
  value_col = "earnings",
  year_col = "year",
  group_col = "gov_level",
  title = "Earnings Change & Acceleration by Government Level (2019-2024)"
)

accel$table
```

::: {.content-visible when-format="html"}
```{r}
#| label: acceleration-download
#| echo: false
ds_download_data_btn(
  accel$data,
  filename_base = paste0(tolower(params$city), "_acceleration_data"),
  label = "Download Acceleration Data (CSV)"
)
```
:::


# Weighted Estimates

```{r}
#| label: calculate-estimates

calc_estimates <- function(data, cpi_tbl, by_level = TRUE) {
  years <- sort(unique(data$year))
  
  results <- purrr::map(years, function(y) {
    year_data <- data[data$year == y, ]
    cpi_factor <- cpi_tbl$factor_to_base[cpi_tbl$year == y]
    if (length(cpi_factor) == 0) cpi_factor <- 1
    
    svy <- srvyr::as_survey_rep(
      year_data,
      weights = PWGTP,
      repweights = dplyr::matches("^PWGTP\\d+$"),
      type = "JK1",
      scale = 4/80,
      rscales = rep(1, 80),
      mse = TRUE
    )
    
    if (by_level) {
      stats <- svy |>
        srvyr::group_by(gov_level) |>
        srvyr::summarise(
          n_employees = srvyr::survey_total(1, vartype = "se"),
          mean_earnings = srvyr::survey_mean(earnings, vartype = "se"),
          median_earnings = srvyr::survey_median(earnings, vartype = "se"),
          .groups = "drop"
        )
    } else {
      stats <- svy |>
        srvyr::summarise(
          n_employees = srvyr::survey_total(1, vartype = "se"),
          mean_earnings = srvyr::survey_mean(earnings, vartype = "se"),
          median_earnings = srvyr::survey_median(earnings, vartype = "se")
        )
      stats$gov_level <- "All Government"
    }
    
    stats$year <- y
    stats$mean_real <- stats$mean_earnings * cpi_factor
    stats$median_real <- stats$median_earnings * cpi_factor
    stats
  })
  
  dplyr::bind_rows(results)
}

gov_results <- calc_estimates(gov_employees, cpi, by_level = TRUE)
overall_results <- calc_estimates(gov_employees, cpi, by_level = FALSE)
```


## Employee Counts

```{r}
#| label: counts-table

counts_tbl <- gov_results |>
  dplyr::select(year, gov_level, n_employees) |>
  tidyr::pivot_wider(names_from = gov_level, values_from = n_employees) |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    dplyr::across(where(is.numeric), ~format(round(.), big.mark = ","))
  )
counts_tbl |>
  flextable::flextable() |>
  style_ft(title = paste0("Government Employees: ", params$city, " Area (2019-2024)"))
```

::: {.content-visible when-format="html"}
```{r}
#| label: counts-table-download
#| echo: false
ds_download_data_btn(
  counts_tbl,
  filename_base = paste0(tolower(params$city), "_employee_counts"),
  label = "Download Counts Table (CSV)"
)
```
:::


## Mean Earnings (Inflation-Adjusted)

```{r}
#| label: mean-table

mean_tbl <- gov_results |>
  dplyr::select(year, gov_level, mean_real) |>
  tidyr::pivot_wider(names_from = gov_level, values_from = mean_real) |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    dplyr::across(where(is.numeric), ~paste0("$", format(round(.), big.mark = ",")))
  )
mean_tbl |>
  flextable::flextable() |>
  style_ft(title = paste0("Mean Earnings (", params$base_year, " dollars)"))
```

::: {.content-visible when-format="html"}
```{r}
#| label: mean-table-download
#| echo: false
ds_download_data_btn(
  mean_tbl,
  filename_base = paste0(tolower(params$city), "_mean_earnings"),
  label = "Download Mean Earnings Table (CSV)"
)
```
:::

```{r}
#| label: mean-chart
#| fig-width: 9
#| fig-height: 5

p_mean_chart <- ggplot2::ggplot(gov_results, ggplot2::aes(x = year, y = mean_real, color = gov_level)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_point(size = 3) +
  ggplot2::scale_x_continuous(breaks = unique(gov_results$year)) +
  ggplot2::scale_y_continuous(labels = function(x) paste0("$", format(x, big.mark = ","))) +
  ggplot2::scale_color_manual(values = c(
    "Local Government" = "steelblue",
    "State Government" = "forestgreen",
    "Federal Government" = "darkorange"
  )) +
  ggplot2::labs(
    title = paste0("Mean Earnings: ", params$city, " Area (2019-2024)"),
    subtitle = paste0(params$base_year, " dollars"),
    x = "Year", y = "Mean Earnings", color = "Level"
  )
print(p_mean_chart)
```

::: {.content-visible when-format="html"}
```{r}
#| label: mean-chart-download
#| echo: false
ds_download_plot_btn(
  p_mean_chart,
  filename_base = paste0(tolower(params$city), "_mean_chart"),
  label = "Download Mean Earnings Chart (PNG)",
  width = 9, height = 5
)
```
:::


## Median Earnings (Inflation-Adjusted)

```{r}
#| label: median-table

median_tbl <- gov_results |>
  dplyr::select(year, gov_level, median_real) |>
  tidyr::pivot_wider(names_from = gov_level, values_from = median_real) |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    dplyr::across(where(is.numeric), ~paste0("$", format(round(.), big.mark = ",")))
  )
median_tbl |>
  flextable::flextable() |>
  style_ft(title = paste0("Median Earnings (", params$base_year, " dollars)"))
```

::: {.content-visible when-format="html"}
```{r}
#| label: median-table-download
#| echo: false
ds_download_data_btn(
  median_tbl,
  filename_base = paste0(tolower(params$city), "_median_earnings"),
  label = "Download Median Earnings Table (CSV)"
)
```
:::

```{r}
#| label: median-chart
#| fig-width: 9
#| fig-height: 5

p_median_chart <- ggplot2::ggplot(gov_results, ggplot2::aes(x = year, y = median_real, color = gov_level)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_point(size = 3) +
  ggplot2::scale_x_continuous(breaks = unique(gov_results$year)) +
  ggplot2::scale_y_continuous(labels = function(x) paste0("$", format(x, big.mark = ","))) +
  ggplot2::scale_color_manual(values = c(
    "Local Government" = "steelblue",
    "State Government" = "forestgreen",
    "Federal Government" = "darkorange"
  )) +
  ggplot2::labs(
    title = paste0("Median Earnings: ", params$city, " Area (2019-2024)"),
    subtitle = paste0(params$base_year, " dollars"),
    x = "Year", y = "Median Earnings", color = "Level"
  )
print(p_median_chart)
```

::: {.content-visible when-format="html"}
```{r}
#| label: median-chart-download
#| echo: false
ds_download_plot_btn(
  p_median_chart,
  filename_base = paste0(tolower(params$city), "_median_chart"),
  label = "Download Median Earnings Chart (PNG)",
  width = 9, height = 5
)
```
:::


## All Government Combined

```{r}
#| label: overall-table

overall_tbl <- overall_results |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    Year = as.character(year),
    Employees = format(round(n_employees), big.mark = ","),
    `Mean (Real)` = paste0("$", format(round(mean_real), big.mark = ",")),
    `Median (Real)` = paste0("$", format(round(median_real), big.mark = ","))
  ) |>
  dplyr::select(Year, Employees, `Mean (Real)`, `Median (Real)`)
overall_tbl |>
  flextable::flextable() |>
  style_ft(title = paste0("All Government: ", params$city, " Area (", params$base_year, " dollars)"))
```

::: {.content-visible when-format="html"}
```{r}
#| label: overall-table-download
#| echo: false
ds_download_data_btn(
  overall_tbl,
  filename_base = paste0(tolower(params$city), "_overall_results"),
  label = "Download Overall Table (CSV)"
)
```
:::


# Export

```{r}
#| label: export

dir.create("out", showWarnings = FALSE)

file_base <- paste0(
  "out/pums_gov_", tolower(gsub(" ", "_", params$city)), "_",
  params$state, "_", min(params$years), "_", max(params$years), "_", params$survey
)

saveRDS(gov_results, paste0(file_base, "_by_level.rds"))
saveRDS(overall_results, paste0(file_base, "_overall.rds"))
readr::write_csv(gov_results, paste0(file_base, "_by_level.csv"))
readr::write_csv(overall_results, paste0(file_base, "_overall.csv"))

writexl::write_xlsx(
  list(by_level = gov_results, overall = overall_results, cpi = cpi),
  paste0(file_base, ".xlsx")
)
```

Data exported to `r file_base`


# Notes

**Data:** ACS `r params$survey` PUMS for years `r paste(params$years, collapse = ", ")`

**Caching:** Data is cached in `r params$cache_dir`. First render fetches from Census API; subsequent renders load from cache. Delete the cache folder to force a refresh.

**COW codes:** 3 = Local, 4 = State, 5 = Federal government

**1-year limitations:** Larger standard errors than 5-year estimates; may be unreliable for small areas.
