---
title: "CPI-W (Urban Wage Earners) Analysis — Annual Data"
subtitle: "1970–2026"
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true     # Set to 'false' for drafts. Override if you're NOT sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false


params:
  data_updated: !expr format(Sys.Date(), "%B %Y")
---


```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED

# Download button helpers live here; knitr::is_html_output() guard is inside
# the helpers so they silently no-op when rendering to typst
source(here::here("R", "download_button_helpers.R"))

# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)



# Flextable built-in themes:
  # flextable::theme_alafoli()	|>  # BLAH
  # flextable::theme_apa()  # THIS IS NICE
  # flextable::theme_booktabs() |>  # NICE, MORE COMPACT
  # flextable::theme_box() |>   # OK, INCLUDES CELL BORDERS
  # flextable::theme_tron() |>  # 'DARK MODE' BLUE TEXT
  # flextable::theme_tron_legacy() |>   # 'DARK MODE' YELLOW TEXT
  # flextable::theme_vader() |>    # 'DARK MODE' WHITE TEXT
  # flextable::theme_vanilla() |>   # NOT SPECIAL
  # flextable::theme_zebra()	|>
  #


# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 28,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 22,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 18,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 18,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)


claude <- ellmer::chat_anthropic(model = "claude-opus-4-6-20260205")
  
# Set seed for reproducibility
base::set.seed(123)

```

\

#### **Filename:**
get_cpi_w_and_analyze_with_buttons.qmd

\

## Introduction

This document fetches **CPI-W** (Consumer Price Index for Urban Wage Earners and Clerical Workers, series `CWUR0000SA0`) from the Bureau of Labor Statistics public API v2, saves it locally to avoid redundant API calls, and produces charts and tables for municipal compensation benchmarking.

CPI-W is preferred over CPI-U for **employee compensation comparisons** because it tracks the price experience of wage-earning households specifically, and it is the index used for Social Security COLA adjustments.

---

## Step 1: Define Storage Paths

```{r}
#| label: define-paths

data_folder   <- here::here("data_clean")
cpi_file_path <- base::file.path(data_folder, "annual_cpi_data.rds")
csv_file_path <- base::file.path(data_folder, "annual_cpi_data.csv")

# Create data_clean/ on first run if it doesn't exist yet
if (!base::dir.exists(data_folder)) {
  base::dir.create(data_folder, recursive = TRUE)
}

base::cat("RDS path:", cpi_file_path, "\n")
base::cat("CSV path:", csv_file_path, "\n")
```

---

## Step 2: Define BLS API Fetch Functions

The BLS v2 API requires a **POST** request with a JSON body — a plain URL
query string (v1 style) returns only the most recent year and silently ignores
the year parameters. The 20-year chunk limit is enforced by BLS; requests
spanning more than 20 years must be split and re-combined.

```{r}
#| label: fetch-functions

# Single chunk: up to 20 years per BLS v2 limit
fetch_cpi_chunk <- function(start_year, end_year,
                             api_key = base::Sys.getenv("BLS_API_KEY")) {

  base_url  <- "https://api.bls.gov/publicAPI/v2/timeseries/data/"
  series_id <- "CWUR0000SA0"   # CPI-W — wage earners; most relevant for comp benchmarking

  base::cat("  Requesting", start_year, "-", end_year, "...\n")

  request_body <- base::list(
    seriesid  = base::list(series_id),
    startyear = base::as.character(start_year),
    endyear   = base::as.character(end_year)
  )

  # Key attached only when present; unregistered calls have tighter rate limits
  if (!base::is.null(api_key) && nchar(api_key) > 0) {
    request_body$registrationkey <- api_key
  }

  response <- httr2::request(base_url) |>
    httr2::req_headers("Content-Type" = "application/json") |>
    httr2::req_body_json(request_body) |>
    httr2::req_perform()

  json_data <- httr2::resp_body_json(response)

  if (json_data$status != "REQUEST_SUCCEEDED") {
    base::warning(
      "BLS API failed for ", start_year, "-", end_year,
      ": ", base::paste(json_data$message, collapse = " ")
    )
    return(NULL)
  }

  raw_rows <- json_data$Results$series[[1]]$data

  purrr::map_dfr(raw_rows, function(obs) {
    dplyr::tibble(
      year        = base::as.integer(obs$year),
      period      = obs$period,
      period_name = obs$periodName,
      value       = base::as.numeric(obs$value)
    )
  })
}


# Orchestrator: splits the full range into ≤20-year chunks and combines results
fetch_all_cpi <- function(start_year = 1970, end_year = NULL,
                           api_key = base::Sys.getenv("BLS_API_KEY")) {

  if (base::is.null(end_year)) {
    end_year <- base::as.integer(base::format(base::Sys.Date(), "%Y"))
  }

  all_raw <- dplyr::tibble()
  current_start <- start_year

  while (current_start <= end_year) {
    current_end <- base::min(current_start + 19, end_year)
    chunk <- fetch_cpi_chunk(current_start, current_end, api_key = api_key)
    if (!base::is.null(chunk)) {
      all_raw <- dplyr::bind_rows(all_raw, chunk)
    }
    current_start <- current_end + 1
    # Polite pause between chunks to stay within BLS rate limits
    if (current_start <= end_year) base::Sys.sleep(1)
  }

  # M13 = BLS pre-computed annual average; preferred over manual mean of M01-M12
  all_raw |>
    dplyr::filter(period == "M13" | base::grepl("^M[0-1][0-9]$", period)) |>
    dplyr::group_by(year) |>
    dplyr::summarise(
      annual_cpi = dplyr::if_else(
        base::any(period == "M13"),
        value[period == "M13"][1],
        base::mean(value, na.rm = TRUE)
      ),
      .groups = "drop"
    ) |>
    dplyr::arrange(year)
}
```

---

## Step 3: Load Cached Data or Fetch from BLS

```{r}
#| label: load-or-fetch

current_year <- base::as.integer(base::format(base::Sys.Date(), "%Y"))

if (base::file.exists(cpi_file_path)) {

  cpi_data <- base::readRDS(cpi_file_path)
  base::cat("Loaded cached data:", base::nrow(cpi_data), "years",
            "(", base::min(cpi_data$year), "-", base::max(cpi_data$year), ")\n")

  # Re-fetch if cache is more than one year stale
  if (base::max(cpi_data$year) < (current_year - 1)) {
    base::cat("Cache is stale — fetching fresh data from BLS...\n")
    cpi_data <- fetch_all_cpi(start_year = 1970)
  }

} else {

  base::cat("No cache found — fetching from BLS API...\n")
  cpi_data <- fetch_all_cpi(start_year = 1970)

}

base::cat("\nLoaded:", base::nrow(cpi_data), "years |",
          base::min(cpi_data$year), "-", base::max(cpi_data$year), "| CPI range:",
          base::round(base::min(cpi_data$annual_cpi), 1), "-",
          base::round(base::max(cpi_data$annual_cpi), 1), "\n")
```

---

## Step 4: Save Data Locally

```{r}
#| label: save-data

# RDS preserves types exactly; CSV provides human-readable backup
base::saveRDS(cpi_data, file = cpi_file_path)
readr::write_csv(cpi_data, csv_file_path)

base::cat("Saved RDS:", cpi_file_path, "\n")
base::cat("Saved CSV:", csv_file_path, "\n")
```

---

## Step 5: Full CPI Data Table

```{r}
#| label: display-full-data

cpi_data |>
  dplyr::mutate(annual_cpi = base::round(annual_cpi, 1)) |>
  dplyr::rename(Year = year, `Annual CPI-W` = annual_cpi) |>
  flextable::flextable() |>
  flextable::add_header_lines(
    base::paste0("CPI-W Annual Average (CWUR0000SA0) | 1970 - ",
                 base::max(cpi_data$year))
  ) |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: full-data-download
#| echo: false
ds_download_data_btn(
  cpi_data,
  filename_base = "cpi_w_annual_full",
  label         = "Download Full CPI-W Data (CSV)"
)
```
:::

---

## Step 6: Select Years to Analyze

Edit `selected_start_year` and `selected_end_year` below to change the analysis window.

```{r}
#| label: select-years

# =============================================================================
# EDIT THESE VALUES TO CHANGE THE ANALYSIS WINDOW
# =============================================================================
selected_start_year <- 1970
selected_end_year   <- 2024

# Uncomment to use specific years instead of a range:
# specific_years <- c(1970, 1980, 1990, 2000, 2010, 2020, 2024)

cpi_filtered <- cpi_data |>
  dplyr::filter(year >= selected_start_year & year <= selected_end_year)

# Verify the filter produced data before any downstream chunk tries to plot it
base::stopifnot(
  "cpi_filtered is empty — check selected_start_year / selected_end_year" =
    base::nrow(cpi_filtered) > 0
)

base::cat("Selected", base::nrow(cpi_filtered), "years:",
          base::min(cpi_filtered$year), "-", base::max(cpi_filtered$year), "\n")
```

---

## Step 7: Filtered Data Table

```{r}
#| label: display-filtered-data

cpi_filtered |>
  dplyr::mutate(annual_cpi = base::round(annual_cpi, 1)) |>
  dplyr::rename(Year = year, `Annual CPI-W` = annual_cpi) |>
  flextable::flextable() |>
  flextable::add_header_lines(
    base::paste0("CPI-W: ", selected_start_year, " - ", selected_end_year)
  ) |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: filtered-data-download
#| echo: false
ds_download_data_btn(
  cpi_filtered,
  filename_base = "cpi_w_filtered",
  label         = "Download Filtered CPI-W Data (CSV)"
)
```
:::

---

## Step 8: CPI Index Level Over Time

The index level shows the cumulative rise in prices — it is not the inflation
rate itself, but the base from which the rate is calculated.

```{r}
#| label: plot-cpi-level
#| fig-width: 13
#| fig-height: 8

plot_cpi_level <- ggplot2::ggplot(
  cpi_filtered,
  ggplot2::aes(x = year, y = annual_cpi)
) +
  ggplot2::geom_line(linewidth = 1.2, color = "steelblue") +
  ggplot2::geom_point(size = 2, color = "steelblue") +
  ggplot2::scale_x_continuous(
    breaks = base::seq(
      base::floor(base::min(cpi_filtered$year) / 5) * 5,
      base::max(cpi_filtered$year),
      by = 5
    )
  ) +
  ggplot2::scale_y_continuous(labels = scales::comma_format()) +
  ggplot2::labs(
    title    = "CPI-W Index Level Over Time",
    subtitle = base::paste0(
      "Annual Average | Urban Wage Earners & Clerical Workers | ",
      base::min(cpi_filtered$year), "\u2013", base::max(cpi_filtered$year)
    ),
    x       = "Year",
    y       = "CPI-W Index Value",
    caption = "Source: Bureau of Labor Statistics | Series CWUR0000SA0"
  ) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::theme(
    plot.title    = ggplot2::element_text(face = "bold", size = 16),
    plot.subtitle = ggplot2::element_text(color = "gray40"),
    axis.text.x   = ggplot2::element_text(angle = 45, hjust = 1),
    panel.grid.minor = ggplot2::element_blank()
  )

plot_cpi_level
```

::: {.content-visible when-format="html"}
```{r}
#| label: plot-cpi-level-download
#| echo: false
ds_download_plot_btn(
  plot_cpi_level,
  filename_base = "cpi_w_index_level",
  label         = "Download CPI Level Plot (PNG)"
)
```
:::

---

## Step 9: Year-over-Year Inflation Rate Table

```{r}
#| label: calculate-inflation

# Lag by one year to get year-over-year percentage change
cpi_with_inflation <- cpi_filtered |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    prior_cpi      = dplyr::lag(annual_cpi),
    inflation_rate = (annual_cpi - prior_cpi) / prior_cpi * 100
  )

# Separate object used by downstream plot chunks — filter once, reuse often
inflation_data <- cpi_with_inflation |>
  dplyr::filter(!base::is.na(inflation_rate))

inflation_data |>
  dplyr::mutate(
    annual_cpi     = base::round(annual_cpi, 1),
    prior_cpi      = base::round(prior_cpi, 1),
    inflation_rate = base::round(inflation_rate, 2)
  ) |>
  dplyr::select(year, prior_cpi, annual_cpi, inflation_rate) |>
  dplyr::rename(
    Year                = year,
    `Prior Year CPI-W`  = prior_cpi,
    `Current CPI-W`     = annual_cpi,
    `Inflation Rate (%)` = inflation_rate
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Year-over-Year CPI-W Inflation Rates") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: inflation-table-download
#| echo: false
ds_download_data_btn(
  inflation_data,
  filename_base = "cpi_w_inflation_rates",
  label         = "Download Inflation Rate Table (CSV)"
)
```
:::

---

## Step 10: Inflation Rate Over Time

```{r}
#| label: plot-inflation
#| fig-width: 13
#| fig-height: 8

plot_inflation <- ggplot2::ggplot(
  inflation_data,
  ggplot2::aes(x = year, y = inflation_rate)
) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  ggplot2::geom_area(alpha = 0.2, fill = "grey") +
  ggplot2::geom_line(linewidth = 2, color = "gray") +
  ggplot2::geom_point(size = 3, color = "black") +
  ggplot2::scale_x_continuous(
    breaks = base::seq(
      base::floor(base::min(inflation_data$year) / 5) * 5,
      base::max(inflation_data$year),
      by = 5
    )
  ) +
  ggplot2::scale_y_continuous(
    labels = function(x) base::paste0(x, "%")
  ) +
  ggplot2::labs(
    title    = "Annual CPI-W Inflation Rate",
    subtitle = base::paste0(
      base::min(inflation_data$year), "\u2013", base::max(inflation_data$year),
      " | Year-over-Year % Change"
    ),
    x       = "Year",
    y       = "Inflation Rate (%)",
    caption = "Source: Bureau of Labor Statistics | Series CWUR0000SA0"
  ) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::theme(
    plot.title    = ggplot2::element_text(face = "bold", size = 16),
    plot.subtitle = ggplot2::element_text(color = "gray40"),
    axis.text.x   = ggplot2::element_text(angle = 45, hjust = 1),
    panel.grid.minor = ggplot2::element_blank()
  )

plot_inflation
```

::: {.content-visible when-format="html"}
```{r}
#| label: plot-inflation-download
#| echo: false
ds_download_plot_btn(
  plot_inflation,
  filename_base = "cpi_w_inflation_rate",
  label         = "Download Inflation Rate Plot (PNG)"
)
```
:::

---

## Step 11: Inflation Rate with Historical Context

```{r}
#| label: plot-inflation-annotated
#| fig-width: 13
#| fig-height: 8

plot_inflation_annotated <- ggplot2::ggplot(
  inflation_data,
  ggplot2::aes(x = year, y = inflation_rate)
) +
  # Shade known high-inflation episodes so readers see structural context
  ggplot2::annotate("rect", xmin = 1973, xmax = 1975,
                    ymin = -Inf, ymax = Inf, alpha = 0.2, fill = "lightgreen") +
  ggplot2::annotate("rect", xmin = 1979, xmax = 1982,
                    ymin = -Inf, ymax = Inf, alpha = 0.2, fill = "lightgreen") +
  ggplot2::annotate("rect", xmin = 2021, xmax = 2023,
                    ymin = -Inf, ymax = Inf, alpha = 0.2, fill = "lightgreen") +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  # Fed 2% target is the policy anchor — always worth showing
  ggplot2::geom_hline(yintercept = 2, linetype = "dotted",
                      color = "darkgreen", linewidth = 1) +
  ggplot2::geom_line(linewidth = 2, color = "gray") +
  ggplot2::geom_point(size = 3, color = "black") +
  ggplot2::annotate("text", x = 1974,   y = 12,   size = 3, fontface = "italic",
                    label = "Oil Crisis\n1973\u201375") +
  ggplot2::annotate("text", x = 1980.5, y = 14.5, size = 3, fontface = "italic",
                    label = "Volcker Era\n1979\u201382") +
  ggplot2::annotate("text", x = 2022,   y = 9,    size = 3, fontface = "italic",
                    label = "Post-COVID\n2021\u201323") +
  ggplot2::annotate("text", x = 1975,   y = 2.6,  size = 3, fontface = "bold",
                    color = "darkgreen", label = "Fed Target: 2%") +
  ggplot2::scale_x_continuous(
    breaks = base::seq(
      base::floor(base::min(inflation_data$year) / 5) * 5,
      base::max(inflation_data$year),
      by = 5
    )
  ) +
  ggplot2::scale_y_continuous(
    labels = function(x) base::paste0(x, "%")
  ) +
  ggplot2::labs(
    title    = "Annual CPI-W Inflation Rate with Notable Periods Highlighted",
    subtitle = "Orange shading = high-inflation episodes | Green dotted line = Fed 2% target",
    x       = "Year",
    y       = "Inflation Rate (%)",
    caption = "Source: Bureau of Labor Statistics | Series CWUR0000SA0"
  ) +
  ggplot2::theme_minimal(base_size = 20) +
  ggplot2::theme(
    plot.title    = ggplot2::element_text(face = "bold", size = 28),
    plot.subtitle = ggplot2::element_text(color = "black"),
    axis.text.x   = ggplot2::element_text(angle = 45, hjust = 1),
    panel.grid.minor = ggplot2::element_blank()
  )

plot_inflation_annotated
```

::: {.content-visible when-format="html"}
```{r}
#| label: plot-inflation-annotated-download
#| echo: false
ds_download_plot_btn(
  plot_inflation_annotated,
  filename_base = "cpi_w_inflation_annotated",
  label         = "Download Annotated Plot (PNG)"
)
```
:::

---

## Step 12: Summary Statistics

```{r}
#| label: summary-stats

summary_stats <- inflation_data |>
  dplyr::summarise(
    `Average Inflation`  = base::paste0(base::round(base::mean(inflation_rate), 2), "%"),
    `Median Inflation`   = base::paste0(base::round(stats::median(inflation_rate), 2), "%"),
    `Std Deviation`      = base::paste0(base::round(stats::sd(inflation_rate), 2), "%"),
    `Maximum`            = base::paste0(
      base::round(base::max(inflation_rate), 2), "% (",
      year[inflation_rate == base::max(inflation_rate)], ")"
    ),
    `Minimum`            = base::paste0(
      base::round(base::min(inflation_rate), 2), "% (",
      year[inflation_rate == base::min(inflation_rate)], ")"
    ),
    `Years Above 5%`     = base::as.character(base::sum(inflation_rate > 5)),
    `Years Below 2%`     = base::as.character(base::sum(inflation_rate < 2 & inflation_rate > 0)),
    `Years of Deflation` = base::as.character(base::sum(inflation_rate < 0))
  ) |>
  tidyr::pivot_longer(
    cols      = dplyr::everything(),
    names_to  = "Statistic",
    values_to = "Value"
  )

summary_stats |>
  flextable::flextable() |>
  flextable::add_header_lines(
    base::paste0("CPI-W Inflation Summary | ",
                 base::min(inflation_data$year), "\u2013",
                 base::max(inflation_data$year))
  ) |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: summary-stats-download
#| echo: false
ds_download_data_btn(
  summary_stats,
  filename_base = "cpi_w_summary_stats",
  label         = "Download Summary Statistics (CSV)"
)
```
:::

---

## How to Customize

**Change the analysis window** — edit Step 6:

```r
selected_start_year <- 2000
selected_end_year   <- 2024
```

**Force a fresh fetch from BLS** — delete the cached file and re-render:

```r
base::file.remove(here::here("data_clean", "annual_cpi_data.rds"))
```

**Switch to a different CPI series** — change `series_id` in the
`fetch-functions` chunk:

| Series ID      | Description                                  |
|----------------|----------------------------------------------|
| `CWUR0000SA0`  | CPI-W — Urban Wage Earners (current)         |
| `CPIAUCSL`     | CPI-U — All Urban Consumers, seasonally adj  |
| `CPILFESL`     | Core CPI — excludes food and energy          |
| `CUURA316SA0`  | CPI-U — Dallas–Fort Worth metro area         |

**Store your BLS API key** (run once in the R console, not in this document):

```r
usethis::edit_r_environ()
# Add this line then save and restart R:
# BLS_API_KEY=your_key_here
```
