---
title: "***Personal and Confidential***"
subtitle: "Proof of Concept Proposal"
description: "Civic Fiscal Analysis Using the 5Q Decision Architecture Framework™" 
authors: 
  - name: "Dan Swart, CPA (ret)"
  - name: "Claude Opus 4.5"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: false     # Fast for drafts. Override for sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    # fig-width: 13
    E fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false


---


```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED


# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)


#
# # Sample Code:
# flextable::flextable(violations) |>
 #  flextable::set_header_labels(
 #    Variable = "Variable",
 #    Measurement = "Measurement",
 #    Likely_Impact = "Likely Impact"
 #  ) |>
#    flextable::add_header_lines(values = "Frequent Violations of Scientific Method in Current So-Called 'Equity' Research") |>
#   flextable::color(i = 1, color = "blue", part = "header") |>
#   flextable::italic(i = 1, part = "header") |>
#   flextable::align(i = 1, align = "center", part = "header") |>
#   flextable::fontsize(i = 1, size = 14, part = "header") |>
#   flextable::bg(i = 1, bg = "white", part = "header") |>
#   flextable::bg(i = 2, bg = "palegreen", part = "header") |>
#   flextable::bold(i = 1:2, part = "header") |>
#   flextable::bold(i = 1:7, j = 1, part = "body") |>
#   ftExtra::colformat_md() |> 
#   flextable::autofit()
  

# Flextable built-in themes:
  # flextable::theme_alafoli()	|>  # BLAH
  # flextable::theme_apa()  # THIS IS NICE
  # flextable::theme_booktabs() |>  # NICE, MORE COMPACT
  # flextable::theme_box() |>   # OK, INCLUDES CELL BORDERS
  # flextable::theme_tron() |>  # 'DARK MODE' BLUE TEXT
  # flextable::theme_tron_legacy() |>   # 'DARK MODE' YELLOW TEXT
  # flextable::theme_vader() |>    # 'DARK MODE' WHITE TEXT
  # flextable::theme_vanilla() |>   # NOT SPECIAL
  # flextable::theme_zebra()	|>
  #

# Flextable titles:
  # flextable::colformat_double(j = c("Mean", "SD", "N"), big.mark = ",", digits = 1) |>
  # flextable::flextable(variance_comparison) |>
  #  flextable::add_header_lines(values = "The Within-Group vs Between-Group Variance Problem") |>

# Flextable Title theming at table-level:
  #  flextable::color(i = 1, color = "blue", part = "header") |>
  #  flextable::italic(i = 1, part = "header") |>
  #  flextable::align(i = 1, align = "center", part = "header") |>
  #  flextable::fontsize(i = 1, size = 14, part = "header") |>
  #  flextable::bg(i = 1, bg = "white", part = "header") |>

# Flextable standard table background colors:
  #  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  

# Flextable reading markdown:
  #  ftExtra::colformat_md() |> 

# Flextable auto-sizing cell widths:
  #  flextable::autofit() 
 
 # Flextable background based on SPECIFIC cell contents:
  #
  # flextable::bg(i = ~ Impact_on_Validity == "High", j = "Impact_on_Validity", bg = "#ffcccc") |>
 # flextable::bg(i = ~ Impact_on_Validity == "Medium", j = "Impact_on_Validity", bg = "#ffffcc") |>
  #
  # 
  # Apply yellow background to any cell containing "Yes":

  # for (col in base::names(hypotheses_data)) {
  #   yes_rows <- base::which(hypotheses_data[[col]] == "Yes")
  #   if (base::length(yes_rows) > 0) {
  #     ft <- ft |>
  #       flextable::bg(i = yes_rows, j = col, bg = "yellow", part = "body")
  #   }
  # } 
  #
  # Apply to last row of table:
  #
  #



# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 16,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 14,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 14,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 14),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 12,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)



  
# Set seed for reproducibility
base::set.seed(123)

```

# Problem Statement

## What We Have Now

Governmental bodies currently use what can be called 'Compliance Disclosure' to satisfy the law.  

Compliance disclosure means records released in formats that satisfy legal requirements but obstruct analysis. PDFs instead of spreadsheets. Summaries without line items. Charts and graphs without making the underlying data accessible.  Replies that technically respond to requests but make evaluation nearly impossible.  'Transparency' is not the right word for 'Compliance Disclosure'.

**Compliance Disclosure:** The data exists. The meetings are open. Legally, all required data is available.  Obscures more than it reveals.



## What We Need Now

**Functional Transparency™** is what we need - information organized and readily accessible in ways that allow citizens to evaluate the ***quality*** of decisions, not just their existence.

**Functional Transparency™** is when a citizen can actually understand *why* a decision was made. What was considered. What was ignored. What assumptions drove the outcome.

Almost no government achieves **Functional Transparency™**.  Perhaps, not because they're hiding things—but because *no one has done the work* of connecting the numbers to the reasoning to the outcome in a form a citizen can actually use. DATA -> REASONING -> DECISIONS

Decisions are often made without examining all the pertinent areas. Boards and councils tend to make decisions based on rhetoric rather than empirical knowledge. Someone has to direct the search for relevant information—and that means knowing what *is* relevant information.


Determining what is 'relevant' is always a judgement based on incentives, context and available evidence.  
Many elected officials don't have a framework or experience to identify what may be essential, or the time for independent study of pending decisions. 

They receive an information packet and asked to vote. If the packet looks thorough, they may assume it *is* thorough. They may not know what's missing because they don't know what else *should* be there.

----------

# Business Proposal

***Phase I Goal***

To build a scalable model for civic engagement based on forming investigative teams to monitor and analyze local government decisions.  


***Purpose***

To improve access to public information from 'data-dump compliance' to Functional Transparency™.  Additionally, to provide incentives and methods for governing bodies to make well-considered decisions based on sufficient information that often is readily available.  The implicit promise of democratic governance is that *evidence matters*, and that *someone is paying attention*.


***Core Offering***

Translating 'compliance disclosures' into **Functional Transparency™** - reconstructing the decision logic behind public actions.  Making the "why" visible.

Our team will analyze the available data and **challenge decisions where warranted** — publicly, with documented evidence and reasoned argument. The goal is not to embarrass, but to establish a standard of decision development that makes certain all relevant aspects are considered.

**Decision competency.**  The ability to know what you need to know before you vote.


***Methodology***

The 5Q Decision Architecture Framework™ and TPIA/FOIA data request Template are the heart of the system.  The goal in Phase I is to stress-test and improve those instruments.  I expect many improvements will be made during Phase I.  If Phase I develops into a viable system further development will be considered.

The Schertz, TX compensation study will be the stress-test to refine the effectiveness and relevancy of the framework in action.

***The Challenge***

Using the 5Q Decision Architecture Framework™, can we reconstruct the decision logic and show what information and data were weighed, what wasn't, and whether the reasoning is consistent with stated policies and goals.


***Skills Needed***

*  Ability to fully describe and persuasively request information from any official, regardless of rank or notoriety. 
*  Ability to persuasively present the findings.
*  Investigative mindset.  
*  Data analysis skills that can create presentation-quality charts, graphs, and visualizations (coding, Excel, etc).  Basic statistical background helpful.
*  Expertise on the internal functioning of the governmental body involved, including the limitations and requirements under local policies.
*  Ability to create and improve data collection forms and techniques.

----------

# Example Phase I Workflow

A question is posed:  ***"Why has total compensation doubled in just the last five years?  It seems excessive."***

The team follows the 5Q Decision Architecture Framework™ which guides and focuses the information gathering needed to answer the question.  The framework will be customized for each specific engagement.

Information is gathered and analyzed numerically and contextually.

A report, presentation, or letter is prepared based on the empirical evidence gathered.   (duties remain the same?)



## Abbreviated Example of Using The 5Q Decision Architecture Framework™

Every significant fiscal decision can be evaluated against five essential questions:


1. Trajectory

What are the short-term and long-term consequences, financially and organizationally?

2. Proportionality

What share of the total budget does the category represent? 

3. Reversibility

How 'sticky' is the decision?  Should 'sunset' provisions be considered?

4. Assumption Testing

What has to be true for this proposal to be effective in achieving the stated goals provided? 

5. Opportunity Cost

Would there be greater civic benefit if the funds were directed elsewhere?




## Supporting Metrics

In addition to the five core questions, our team may employ supplementary analyses.  For example:

| Metric | Description |
|--------|-------------|
| **Per-Unit Analysis** | Cost per employee, cost per resident served, cost per student |
| **Variance History** | What did we budget vs. what did we actually spend on this decision in the past? | What results?
| **Commitment Creep** | What future obligations does this decision create? |
| **Sensitivity Testing** | What happens if revenue drops 10%? (e.g., new property tax exemptions) |
| **Internal Coherence** | Does this spending match our stated priorities? |

----------


# Internal Evaluation of Phase I - Proof of Concept" 

We must determine how much penetration can be achieved through:

- Texas Public Information Act (TPIA) requests
- Federal Of Information Act (FOIA) requests
- Publicly available budgets, financial documents, transcripts
- External data sources (U.S. Census Bureau, Texas Comptroller, state agencies)
- Meeting minutes, agendas, and recordings
- Other publicly accessible records
- Incidental correspondence (non-TIPA/FOIA requests)

Can ***anyone*** obtain the information needed for a detailed analytic analysis?

----------

# Phase I: Improvement Cycles

Phase I operates as a **continuous quality improvement cycle**: test, document, refine, repeat. Methods and materials remain in active development throughout.


## Location

City of Schertz, Texas

## Focus Question

Sharp increase in employee compensation found in FY2020–FY2024


## Methodology


### 1. Compensation Study

Complete 5Q analysis of workforce compensation trends, including:

- Trajectory projections (e.g., CAGR, 5-year and 10-year forecasts)
- Budget proportionality (e.g., personnel as share of total expenditures)
- Commitment analysis (e.g., locked-in costs, contractual obligations)
- Assumption testing (e.g., revenue projections vs. actuals)
- Opportunity costs of the decision


### 2. TIPA/FOIA Request Templates

Standardized information request specifications, including time requirements for response.  Templates may include:

- Document descriptions with specificity sufficient to avoid "we don't have that" responses, including any required request forms.
- Recommended fiscal year ranges
- Notes on likely availability
- Guidance on phased request strategies


### 3. Analytical Tools

Code-based data pipelines and workflows, documented for replication:

- Data import and cleaning pipelines
- Reconciliation and validation checkpoints
- Visualization templates (trajectory charts, proportionality analysis, etc.)
- Reusable coding functions (e.g., for CAGR calculation, variance analysis, and projection)


### 4. Lessons Learned Documentation

Systematic capture of what worked, what didn't, and improvements made:

- What documents were readily available?
- What required formal TIPA/FOIA requests?
- What was withheld, and on what grounds?
- What data gaps remained despite best efforts?
- What analytical approaches proved most effective?

### 5. Letters, Published Reports and Presentations

Findings are customized for the target audience, including media outlets, social media platforms, presentations, and formal reporting of results.  Any given transmittal may include:

- Key findings
- Methodology notes (reproducible)
- Data sources and limitations
- Recommendations (if warranted)
- Detailed, but not limited to, framework for ongoing monitoring


----------

# Future Development

## Scaling the Model

The templates, tools, and methodology developed in Phase I become the foundation for replication. Other analysts, whether independent researchers, citizen groups, or journalists can adopt the 5Q framework and associated materials to examine their own local governments.

Key assets for scaling:

- Standardized TIPA/FOIA request templates adaptable to any Texas municipality
- Software pipelines requiring only local data substitution (reusable code)
- Training documentation for framework application
- Quality standards for analysis and publication
- Possibly, legislative analysis


## Chapter Development

The long-term vision may include establishing like-minded analysis teams ("chapters") in other jurisdictions, operating under shared standards and methodology while maintaining local independence.  Similar to a franchise or Turning Point USA chapters?

Chapter model characteristics:

- Shared 5Q Decision Architecture™ methodology
- Common quality and documentation standards
- Local operational independence
- Network for knowledge sharing and mutual support
- Potential for coordinated multi-jurisdiction analysis


## Training and Education Opportunity

The 5Q Decision Architecture Framework™ can be taught. Potential audiences include:

| Audience | Application |
|----------|-------------|
| **Newly elected officials** | Analytical frameworks for evaluating staff recommendations |
| **Citizen oversight groups** | Structured methodology for fiscal monitoring |
| **Journalism programs** | Investigative techniques for local government coverage |
| **Civic education initiatives** | Curriculum for engaged citizenship |
| **Government finance professionals** | External perspective on decision quality |


Training formats may include:

- Workshops (half-day or full-day)
- Online modules
- Written guides and workbooks
- Mentored application to real decisions

----------


# What We Are Not

We are clear about what we are not:

- **Our independence requires separation from the entities we examine.**  We may accept government contracts only if independence in appearance and in fact are maintained.

- **We are not partisan operatives building cases for candidates.**  Our analysis applies the same standards regardless of who holds office.

- **We are not hobbyists who will lose interest after one cycle.**  We are building institutional capacity for sustained governmental scrutiny.

- **We are not collaborators helping officials feel better about their decisions.**  We are independent examiners who will report what we find.

We are building an institution; methodical, persistent, and serious—because local government accountability is the best antidote for government sprawl and administrative overreach (The Swamp).

----------


# Summary

One of the most important effects of public scrutiny is invisible. Some officials who know someone is watching behave differently than those who believe no one is looking at the books. It announces something: ***Someone is paying attention. Someone has the skills to find irregularities. Someone will notice.***  

That awareness changes behavior in ways that will never be directly observed. The effect of scrutiny is not nothing — it's just invisible.

There's a philosophical tradition — going back to the Romans and Greeks — that holds certain activities are intrinsically worthwhile because they express the kind of person you want to be and the kind of society you want to live in. 

Voting when a single vote won't determine the outcome, serving on a jury, documenting public spending: these acts matter not primarily because of their immediate instrumental effects but because they constitute citizenship itself. 

This approach embodies the engaged citizen that a democratic republic requires, whether or not others join. It is the antidote to disengagement, cynicism, and apathy.


```{r}
#| label: tbl-summary
#| echo: false
#| message: false

base::library(flextable)

summary_df <- data.frame(
  Element = c(
    "Stance",
    "Method",
    "Phase I Location",
    "Phase I Focus",
    "Key Outputs",
    "Future Vision"
  ),
  Description = c(
    "Investigative journalism model: independent, rigorous, willing to challenge publicly",
    "5Q Decision Architecture Framework™ (Trajectory, Proportionality, Reversibility, Opportunity Cost, Assumption Testing)",
    "City of Schertz, Texas",
    "Municipal employee compensation study, FY2015–FY2024",
    "Templates, analytical tools, methodology documentation, published analysis, lessons learned",
    "Scalable model for chapters nationwide; training for officials and citizens"
  )
)

summary_df |>
  flextable::flextable() |>
  flextable::set_header_labels(
    Element = "Element",
    Description = "Description"
  ) |>
  flextable::add_header_lines("Fifth Question Group: Summary") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::width(j = "Element", width = 1.5) |>
  flextable::width(j = "Description", width = 5.0) |>
  flextable::valign(valign = "top", part = "body") |>
  flextable::autofit()
```


### ENTITY NAMING DISCUSSION

Separate document

### TPIA/FOIA SAMPLE REQUEST LETTER

Separate document

### TPIA/FOIA SAMPLE TEMPLATE FOLLOWING 5Q DECISION ARCHITECTURE FRAMEWORK™

Separate document


---

# Contact

[Contact information to be added]

----------

# Version History

| Version | Date | Changes |
|---------|------|---------|
| 0.10 | `r Sys.Date()` | Initial proposal |

----------

*5Q Analytics Group*


----------












