---
title: "Texas Public Information Act Request"
subtitle: "City of Schertz Municipal Records"
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: false     # Fast for drafts. Override for sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    # fig-width: 13
    E fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false


---


```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED


# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)


#
# # Sample Code:
# flextable::flextable(violations) |>
 #  flextable::set_header_labels(
 #    Variable = "Variable",
 #    Measurement = "Measurement",
 #    Likely_Impact = "Likely Impact"
 #  ) |>
#    flextable::add_header_lines(values = "Frequent Violations of Scientific Method in Current So-Called 'Equity' Research") |>
#   flextable::color(i = 1, color = "blue", part = "header") |>
#   flextable::italic(i = 1, part = "header") |>
#   flextable::align(i = 1, align = "center", part = "header") |>
#   flextable::fontsize(i = 1, size = 14, part = "header") |>
#   flextable::bg(i = 1, bg = "white", part = "header") |>
#   flextable::bg(i = 2, bg = "palegreen", part = "header") |>
#   flextable::bold(i = 1:2, part = "header") |>
#   flextable::bold(i = 1:7, j = 1, part = "body") |>
#   ftExtra::colformat_md() |> 
#   flextable::autofit()
  

# Flextable built-in themes:
  # flextable::theme_alafoli()	|>  # BLAH
  # flextable::theme_apa()  # THIS IS NICE
  # flextable::theme_booktabs() |>  # NICE, MORE COMPACT
  # flextable::theme_box() |>   # OK, INCLUDES CELL BORDERS
  # flextable::theme_tron() |>  # 'DARK MODE' BLUE TEXT
  # flextable::theme_tron_legacy() |>   # 'DARK MODE' YELLOW TEXT
  # flextable::theme_vader() |>    # 'DARK MODE' WHITE TEXT
  # flextable::theme_vanilla() |>   # NOT SPECIAL
  # flextable::theme_zebra()	|>
  #

# Flextable titles:
  # flextable::colformat_double(j = c("Mean", "SD", "N"), big.mark = ",", digits = 1) |>
  # flextable::flextable(variance_comparison) |>
  #  flextable::add_header_lines(values = "The Within-Group vs Between-Group Variance Problem") |>

# Flextable Title theming at table-level:
  #  flextable::color(i = 1, color = "blue", part = "header") |>
  #  flextable::italic(i = 1, part = "header") |>
  #  flextable::align(i = 1, align = "center", part = "header") |>
  #  flextable::fontsize(i = 1, size = 14, part = "header") |>
  #  flextable::bg(i = 1, bg = "white", part = "header") |>

# Flextable standard table background colors:
  #  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  

# Flextable reading markdown:
  #  ftExtra::colformat_md() |> 

# Flextable auto-sizing cell widths:
  #  flextable::autofit() 
 
 # Flextable background based on SPECIFIC cell contents:
  #
  # flextable::bg(i = ~ Impact_on_Validity == "High", j = "Impact_on_Validity", bg = "#ffcccc") |>
 # flextable::bg(i = ~ Impact_on_Validity == "Medium", j = "Impact_on_Validity", bg = "#ffffcc") |>
  #
  # 
  # Apply yellow background to any cell containing "Yes":

  # for (col in base::names(hypotheses_data)) {
  #   yes_rows <- base::which(hypotheses_data[[col]] == "Yes")
  #   if (base::length(yes_rows) > 0) {
  #     ft <- ft |>
  #       flextable::bg(i = yes_rows, j = col, bg = "yellow", part = "body")
  #   }
  # } 
  #
  # Apply to last row of table:
  #
  #



# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 16,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 14,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 14,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 14),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 12,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)



  
# Set seed for reproducibility
base::set.seed(123)

```

```{r}
#| label: requestor-info
#| include: false

# =============================================================================
# CONFIGURATION - EDIT THESE VALUES FOR EACH REQUEST
# =============================================================================

# Requester Information
requester_name <- "Dan [Last Name]"
requester_organization <- "Fifth Question Group"  
requester_address <- "[Street Address]"
requester_city_state_zip <- "Live Oak, Texas [ZIP]"
requester_email <- "[email@domain.com]"
requester_phone <- "[Phone Number]"

# Target Government Entity
entity_name <- "City of Schertz"
entity_contact <- "City Secretary"
entity_address <- "1400 Schertz Parkway"
entity_city_state_zip <- "Schertz, Texas 78154"

# Request Parameters
fiscal_years_requested <- "FY2015 through FY2024"
fiscal_years_short <- "FY15-FY24"

# Date calculations
request_date <- base::format(base::Sys.Date(), "%B %d, %Y")
response_deadline <- base::format(base::Sys.Date() + 10, "%B %d, %Y")
```

`r request_date`
\vspace{12pt}

**VIA EMAIL AND U.S. MAIL**
\vspace{12pt}

`r entity_contact`
\newline
`r entity_name`
\newline
`r entity_address`
\newline
`r entity_city_state_zip`
\vspace{12pt}

**RE: Texas Public Information Act Request**
\vspace{12pt}

Dear `r entity_contact`:

Pursuant to the Texas Public Information Act, Texas Government Code Chapter 552, I am requesting access to and copies of the public records described below. This request is made on behalf of `r requester_organization`.

## Requested Records

I am requesting the following categories of records for `r fiscal_years_requested` (or the most recent available where historical records are not maintained):

### Category A: Compensation and Personnel Records

1. **Employee Compensation Reports** — Annual listing of all employees showing: name, department, job title, base salary, overtime pay, additional earnings/stipends, benefits costs, and total compensation.

2. **Position Classification and Pay Plans** — Official pay grade/step tables showing minimum, midpoint, and maximum salary for each job classification.

3. **Authorized Position Schedules** — List of all authorized (budgeted) positions by department, whether filled or vacant, for each fiscal year.

4. **Headcount Reports** — Monthly or annual employee counts by department (FTE and headcount).

5. **Pay Plan Amendments** — Resolutions, ordinances, or council actions amending pay scales, adding positions, or approving compensation changes.

6. **Cost of Living Adjustment (COLA) Documentation** — Records of COLA percentages approved by council, effective dates, and methodology used to determine adjustments.

7. **Merit/Step Increase Policies** — Written policies governing automatic step increases, merit increase pools, and eligibility criteria (current and any prior versions in effect during the request period).

### Category B: Budget and Financial Records

8. **Adopted Annual Budgets** — Complete adopted budget documents including all funds (General Fund, Enterprise Funds, Special Revenue Funds, Capital Projects Funds, Debt Service Funds, etc.).

9. **Budget Summaries by Expenditure Category** — Summary tables showing total expenditures by category: Personnel/Salaries, Benefits, Supplies, Contracted Services, Capital Outlay, Debt Service, Transfers.

10. **Budget Summaries by Department** — Summary tables showing total expenditures by department.

11. **Budget-to-Actual Reports** — Year-end reports comparing budgeted amounts to actual expenditures by category and department.

12. **Comprehensive Annual Financial Reports (CAFR/ACFR)** — Audited annual financial statements.

13. **Fund Balance Reports** — Year-end fund balance statements for all funds, showing reserved, unreserved, assigned, committed, and unassigned balances.

14. **Revenue Reports by Source** — Annual revenue summaries showing collections by source: property tax, sales tax, franchise fees, permits and licenses, fines and forfeitures, intergovernmental revenue, charges for services, interest income, and other revenues.

### Category C: Contractual Commitments and Long-Term Obligations

15. **Collective Bargaining Agreements / Memoranda of Understanding** — All current and recently expired (within past 10 years) agreements with police associations, fire associations, or other employee groups.

16. **Employment Contracts** — Individual employment contracts for City Manager, Assistant City Managers, department heads, or other contracted executive positions currently in effect.

17. **TMRS Actuarial Valuations** — Texas Municipal Retirement System actuarial valuation reports showing employer contribution rates, unfunded actuarial accrued liability, and funding ratio.

18. **OPEB Actuarial Valuations** — Other Post-Employment Benefits (retiree health insurance) actuarial valuation reports.

19. **Debt Service Schedules** — Amortization schedules for all outstanding bonds, certificates of obligation, notes, and other debt instruments showing annual principal and interest payments by fiscal year through maturity.

20. **Multi-Year Contracts** — List or copies of all contracts with terms exceeding one year, showing vendor/contractor name, annual cost, contract term, and expiration date.

21. **Lease Agreements** — All lease agreements for equipment, vehicles, facilities, or real property with terms exceeding one year.

22. **Pension Contribution Rate History** — Schedule of annual employer contribution rates to TMRS.

23. **Health Insurance Premium History** — Annual employer cost per employee for health insurance premiums by plan type.

### Category D: Budget Assumptions and Forecasts

24. **Revenue Projections/Forecasts** — Documents prepared during budget development projecting sales tax, property tax, and other revenues for upcoming fiscal years.

25. **Sales Tax Projections** — Specific forecasts of sales tax revenue used in budget preparation, including methodology and assumptions.

26. **Property Tax Calculations** — Certified tax roll values, effective tax rate calculations, rollback rate calculations, and property value growth assumptions used in budgeting.

27. **Budget Preparation Instructions/Guidelines** — Internal memoranda or guidelines issued to departments for budget preparation, including assumed inflation rates, salary increase assumptions, and growth projections.

28. **Five-Year Financial Forecasts** — Any long-range financial projections or forecasts prepared by staff or consultants (most recent three versions if available).

29. **Variance Analysis Reports** — Staff reports or memoranda explaining significant differences between budgeted and actual revenues or expenditures.

30. **Quarterly Financial Reports** — Interim financial reports presented to City Council showing year-to-date financial performance compared to budget.

### Category E: Capital Planning and Unfunded Needs

31. **Capital Improvement Plans (CIP)** — Multi-year capital project plans showing project descriptions, estimated costs, funding sources, and planned timing.

32. **Unfunded Project Lists** — Any lists or memoranda identifying projects or needs not included in adopted budgets or CIP due to funding constraints.

33. **Deferred Maintenance Assessments** — Assessments, inventories, or reports documenting maintenance backlogs for streets, facilities, utilities, parks, or equipment.

34. **Equipment Replacement Schedules** — Schedules showing planned replacement of vehicles, major equipment, or infrastructure components and associated costs.

35. **Reserve Fund Policies** — Written policies governing minimum fund balance levels, emergency reserves, stabilization funds, or contingency reserves.

36. **Department Budget Requests vs. Adopted** — Original department budget requests as submitted compared to final adopted budget amounts (showing reductions or additions made during budget process).

37. **Council Budget Workshop Materials** — Presentation materials, staff memoranda, and handouts from budget workshops showing options, alternatives, or trade-offs presented to City Council for consideration.

## Format and Delivery

I request that responsive records be provided in the following formats, where available:

- **Spreadsheets and data files:** Native electronic format (Excel, CSV, or similar)
- **Reports and documents:** PDF format
- **Large document sets:** Electronic delivery via email, file transfer, or portable media

If any requested records are available on the City's website, please provide the specific URLs rather than duplicating those records.

## Cost Estimate

If the estimated cost of fulfilling this request exceeds $100.00, please provide a written cost estimate before proceeding with document production. I am willing to pay reasonable charges for document search, retrieval, and copying as permitted under the Act.

## Response Deadline

Under the Texas Public Information Act, you are required to respond to this request promptly, and in any event within ten business days. If you determine that any portion of the requested records is exempt from disclosure, please identify the specific records withheld and cite the specific statutory exception(s) claimed for each.

If you intend to seek an Attorney General ruling on any portion of this request, please notify me in writing within the time period required by law and provide me with a copy of your request for ruling.

## Contact Information

Please direct all correspondence regarding this request to:

> `r requester_name`
> `r requester_organization`
> `r requester_address`
> `r requester_city_state_zip`
> Email: `r requester_email`
> Phone: `r requester_phone`

Thank you for your prompt attention to this request. Please do not hesitate to contact me if you have any questions or require clarification regarding the scope of this request.

\vspace{24pt}

Respectfully submitted,

\vspace{36pt}

`r requester_name`
\newline
`r requester_organization`

\vspace{24pt}

---

*This request is made pursuant to the Texas Public Information Act, Texas Government Code Chapter 552. The Act requires governmental bodies to make public information available for public inspection and copying. Failure to comply with the Act may result in civil penalties and attorney's fees.*
