---
title: "Schertz Population and Housing Growth per U.S. Census"
subtitle: "2010-2024"
description: ""
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: false     # Fast for drafts. Override for sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false


---


```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED


# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = base::list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)




# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 28,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 22,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 18,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 14),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 12,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "bottom",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)


  
# Set seed for reproducibility
base::set.seed(123)

```


```{r}
#| label: helper-functions
#| include: false

# ── TABLE DOWNLOAD BUTTON ────────────────────────────────────────────────────
# Creates an HTML download button that delivers a data frame as a CSV file.
# Only renders in HTML output; silently returns NULL in other formats.
#
# Arguments:
#   df        : data frame to download
#   filename  : suggested filename (default "data.csv")
#   label     : button label text (default "Download CSV")
#
create_table_download_button <- function(
    df,
    filename = "data.csv",
    label    = "Download CSV"
) {
  if (!knitr::is_html_output()) return(base::invisible(NULL))

  # Serialise the data frame to CSV in memory, then base64-encode
  csv_text <- utils::capture.output(
    utils::write.csv(df, stdout(), row.names = FALSE)
  )
  csv_raw  <- base::paste(csv_text, collapse = "\n")
  b64      <- jsonlite::base64_enc(base::charToRaw(csv_raw))

  href <- base::paste0("data:text/csv;base64,", b64)

  htmltools::browsable(
    htmltools::tagList(
      htmltools::tags$a(
        href     = href,
        download = filename,
        htmltools::tags$button(
          style = base::paste0(
            "background-color:#2196F3;color:white;border:none;",
            "padding:6px 14px;border-radius:4px;cursor:pointer;",
            "font-size:13px;margin:4px 0 8px 0;"
          ),
          label
        )
      )
    )
  )
}


# ── PLOT DOWNLOAD BUTTON ─────────────────────────────────────────────────────
# Creates an HTML download button that delivers a ggplot object as a PNG file.
# Only renders in HTML output; silently returns NULL in other formats.
#
# Arguments:
#   plot_obj  : a ggplot object
#   filename  : suggested filename (default "plot.png")
#   label     : button label text (default "Download PNG")
#   width     : image width in inches (default 13)
#   height    : image height in inches (default 8)
#   dpi       : resolution (default 300)
#
create_plot_download_button <- function(
    plot_obj,
    filename = "plot.png",
    label    = "Download PNG",
    width    = 13,
    height   = 8,
    dpi      = 300
) {
  if (!knitr::is_html_output()) return(base::invisible(NULL))

  tmp <- base::tempfile(fileext = ".png")
  ggplot2::ggsave(
    filename = tmp,
    plot     = plot_obj,
    width    = width,
    height   = height,
    dpi      = dpi,
    bg       = "white"
  )

  b64  <- jsonlite::base64_enc(base::readBin(tmp, "raw", base::file.info(tmp)$size))
  href <- base::paste0("data:image/png;base64,", b64)
  base::unlink(tmp)

  htmltools::browsable(
    htmltools::tagList(
      htmltools::tags$a(
        href     = href,
        download = filename,
        htmltools::tags$button(
          style = base::paste0(
            "background-color:#4CAF50;color:white;border:none;",
            "padding:6px 14px;border-radius:4px;cursor:pointer;",
            "font-size:13px;margin:4px 0 8px 0;"
          ),
          label
        )
      )
    )
  )
}

```

\

#### **Filename:** 
schertz_2010_2024_population_and_housing_growth_per_us_
census_with_buttons.qmd

\


You can get a clean Schertz series for 2010–2024 with tidycensus, but you have to use two different mechanisms:

2010–2019 → time_series = TRUE against the 2019 vintage (Population Estimates API)

2020–2024 → flat-file approach with vintage = 2024 and variables = "POPESTIMATE"




## POPULATION GROWTH


### Get 2010–2019 via time_series = TRUE

2010–2019 using 'tidycensus::get_estimates() with argument 'time_series = TRUE'

Use this one call to get the full time series, then filter Schertz and map the DATE code to calendar years using the Census mapping (DATE 3 = July 1, 2010; …; DATE 12 = July 1, 2019 → year = DATE + 2007). 


```{r}
#| label: get-pop-2010-2019

# 2010–2019 from Population Estimates API (2010-based series, vintage 2019)
schertz_pop_2010_2019 <- tidycensus::get_estimates(
  geography   = "place",
  state       = "TX",
  product     = "population",
  year        = 2019,      # 2010-based series, last vintage
  time_series = TRUE
) |>
  # keep Schertz only & total population
  dplyr::filter(
    NAME == "Schertz city, Texas",
    variable == "POP"
  ) |>
  # Keep annual July 1 estimates only (DATE 3–12)
  # Census PEP estimates are as of July 1, not October 1—so there's no exact fiscal year match available.
  dplyr::filter(DATE >= 3, DATE <= 12) |>
  dplyr::mutate(
    year = DATE + 2007       # DATE 3 -> 2010, 4 -> 2011, ..., 12 -> 2019
  ) |>
  dplyr::select(NAME, year, variable, value) |>
  dplyr::arrange(year)

flextable::flextable(schertz_pop_2010_2019) |>
  flextable::add_header_lines(values = "Schertz Population 2010–2019 (Population Estimates API)") |>
  flextable::color(i = 1, color = "white", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 13, part = "header") |>
  flextable::bg(i = 1, bg = "#2196F3", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::colformat_int(j = "year", big.mark = "") |>
  flextable::autofit()

create_table_download_button(
  schertz_pop_2010_2019,
  filename = "schertz_pop_2010_2019.csv",
  label    = "Download Table CSV"
)
```




### Get 2020–2024 population data via flat file (post-2020)

```{r}
#| label: get-pop-2020-2024

years_2020_2024 <- 2020:2024

schertz_pop_2020_2024 <- purrr::map_dfr(
  years_2020_2024,
  \(yr) tidycensus::get_estimates(
    geography = "place",
    state     = "TX",
    place     = "Schertz",
    variables = "POPESTIMATE",  # post-2020 PEP
    vintage   = 2024,
    year      = yr
  ) |>
    dplyr::filter(NAME == "Schertz city, Texas") |>
    dplyr::mutate(
      year     = yr,
      variable = "POP"          # make name match pre-2020 chunk
    )
) |>
  dplyr::select(NAME, year, variable, value) |>
  dplyr::arrange(year)

flextable::flextable(schertz_pop_2020_2024) |>
  flextable::add_header_lines(values = "Schertz Population 2020–2024 (Post-2020 Flat File)") |>
  flextable::color(i = 1, color = "white", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 13, part = "header") |>
  flextable::bg(i = 1, bg = "#2196F3", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::colformat_int(j = "year", big.mark = "") |>
  flextable::autofit()

create_table_download_button(
  schertz_pop_2020_2024,
  filename = "schertz_pop_2020_2024.csv",
  label    = "Download Table CSV"
)
```


### Combine both into a single 2010–2024 series

```{r}
#| label: combine-pop-series

schertz_pop_2010_2024 <- dplyr::bind_rows(
  schertz_pop_2010_2019,
  schertz_pop_2020_2024
) |>
  dplyr::arrange(year)

flextable::flextable(schertz_pop_2010_2024) |>
  flextable::add_header_lines(values = "Schertz Population 2010–2024 (Combined Series)") |>
  flextable::color(i = 1, color = "white", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 13, part = "header") |>
  flextable::bg(i = 1, bg = "#2196F3", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::colformat_int(j = "year", big.mark = "") |>
  flextable::autofit()

create_table_download_button(
  schertz_pop_2010_2024,
  filename = "schertz_pop_2010_2024.csv",
  label    = "Download Table CSV"
)
```


### Add change and percent change to tibble

```{r}
#| label: pop-with-changes

schertz_pop_2010_2024_with_changes <- schertz_pop_2010_2024 |>
  dplyr::mutate(year = base::as.character(year)) |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    pop        = value,
    pop_lag    = dplyr::lag(pop),
    change     = pop - pop_lag,
    pct_change = 100 * change / pop_lag
  )

flextable::flextable(schertz_pop_2010_2024_with_changes) |>
  flextable::add_header_lines(values = "Schertz Population 2010–2024 with Annual Changes") |>
  flextable::color(i = 1, color = "white", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 13, part = "header") |>
  flextable::bg(i = 1, bg = "#2196F3", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()

create_table_download_button(
  schertz_pop_2010_2024_with_changes,
  filename = "schertz_pop_2010_2024_with_changes.csv",
  label    = "Download Table CSV"
)
```



### UNCOMMENT AND SAVE RESULTS AS EXCEL

```{r}
#| label: save-pop-excel

# writexl::write_xlsx(
#  schertz_pop_2010_2024_with_changes,
#  "schertz_pop_and_housing/schertz_pop_2010_2024_with_changes.xlsx"
# )
```




## HOUSING GROWTH

### B25001_001 is "Total Housing Units" - it counts ALL types of housing units regardless of structure type.

This includes:

-  Single-family detached homes
-  Single-family attached (townhomes, row houses)
-  Apartments/condos in multi-unit buildings
-  Mobile homes/manufactured housing
-  RVs, boats, vans used as housing
-  Any other structure used as separate living quarters

Key point: It counts housing units, not structures. A 100-unit apartment building = 100 housing units.


It includes both:

-  Occupied units (people living there)
-  Vacant units (for sale, for rent, seasonal, abandoned, etc.)



If you want to break it down by structure type, you'd need table B25024 (Units in Structure), which shows:

-  B25024_002: 1, detached
-  B25024_003: 1, attached
-  B25024_004: 2 units
-  B25024_005: 3 or 4 units
-  B25024_006: 5 to 9 units
-  etc.

This analysis shows total housing stock growth in Schertz across all housing types.

### Get Housing Units and Changes from ACS5

```{r}
#| label: get-housing-units-2009-2022

years_2009_2022 <- 2009:2022

schertz_housing_2009_2022_with_changes <- purrr::map(years_2009_2022, \(yr) {
  tidycensus::get_acs(
    geography = "place",
    state     = "TX",
    variables = "B25001_001",
    year      = yr,
    survey    = "acs5"
  ) |>
    dplyr::mutate(year = yr)
}) |>
  purrr::list_rbind() |>
  dplyr::filter(stringr::str_detect(NAME, "Schertz")) |>
  dplyr::mutate(year = base::as.character(year)) |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    change     = estimate - dplyr::lag(estimate),
    pct_change = change / dplyr::lag(estimate) * 100
  )

flextable::flextable(schertz_housing_2009_2022_with_changes) |>
  flextable::add_header_lines(values = "Schertz Housing Units 2009–2022 (ACS5)") |>
  flextable::color(i = 1, color = "white", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 13, part = "header") |>
  flextable::bg(i = 1, bg = "#2196F3", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()

create_table_download_button(
  schertz_housing_2009_2022_with_changes,
  filename = "schertz_housing_2009_2022_with_changes.csv",
  label    = "Download Table CSV"
)
```


```{r}
#| label: get-housing-units-2009-2024

years_2009_2024 <- 2009:2024

schertz_housing_2009_2024_with_changes <- purrr::map(years_2009_2024, \(yr) {
  tidycensus::get_acs(
    geography = "place",
    state     = "TX",
    variables = "B25001_001",
    year      = yr,
    survey    = "acs5"
  ) |>
    dplyr::mutate(year = yr)
}) |>
  purrr::list_rbind() |>
  dplyr::filter(stringr::str_detect(NAME, "Schertz")) |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    change     = estimate - dplyr::lag(estimate),
    pct_change = change / dplyr::lag(estimate) * 100
  )

flextable::flextable(schertz_housing_2009_2024_with_changes) |>
  flextable::add_header_lines(values = "Schertz Housing Units 2009–2024 (ACS5)") |>
  flextable::color(i = 1, color = "white", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 13, part = "header") |>
  flextable::bg(i = 1, bg = "#2196F3", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()

create_table_download_button(
  schertz_housing_2009_2024_with_changes,
  filename = "schertz_housing_2009_2024_with_changes.csv",
  label    = "Download Table CSV"
)
```


### UNCOMMENT AND SAVE RESULTS AS EXCEL

```{r}
#| label: save-housing-excel

# writexl::write_xlsx(
#   schertz_housing_2009_2024_with_changes,
#   "schertz_pop_and_housing/schertz_housing_2009_2024_with_changes.xlsx"
# )
```




## PLOT RESULTS

### Annual Change in Number of Housing Units

```{r}
#| label: plot-housing-unit-change

# Plot 1: Annual change in number of housing units
p_housing_change <- ggplot2::ggplot(
  schertz_housing_2009_2022_with_changes,
  ggplot2::aes(x = year, y = change)
) +
  ggplot2::geom_col(fill = "steelblue") +
  ggplot2::labs(
    title    = "Annual Change in Housing Units",
    subtitle = "Schertz, TX (ACS 5-Year Estimates)",
    x        = NULL,
    y        = "Units Added"
  ) +
  ggplot2::scale_x_discrete(breaks = years_2009_2022)

p_housing_change

create_plot_download_button(
  p_housing_change,
  filename = "schertz_housing_annual_change.png",
  label    = "Download Plot PNG"
)
```

### Annual Percentage Change in Housing Units

```{r}
#| label: plot-housing-pct-change

p_housing_pct <- ggplot2::ggplot(
  schertz_housing_2009_2022_with_changes,
  ggplot2::aes(
    x = base::factor(year, levels = base::as.character(2009:2022)),
    y = pct_change / 100
  )
) +
  ggplot2::geom_col(fill = "darkgreen") +
  ggplot2::scale_y_continuous(
    labels = scales::percent_format(accuracy = 0.1)
  ) +
  ggplot2::scale_x_discrete(drop = FALSE) +
  ggplot2::labs(
    title    = "Annual Percent Change in Housing Units",
    subtitle = "Schertz, TX (ACS 5-Year Estimates)",
    x        = NULL,
    y        = "Pct Change"
  )

p_housing_pct

create_plot_download_button(
  p_housing_pct,
  filename = "schertz_housing_pct_change.png",
  label    = "Download Plot PNG"
)
```
