---
title: "Top Employee Total Compensation"
subtitle: "City of Schertz | FY20 - FY25"
description: "Directors, Managers, and Public Safety Leaders"
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: false     # Fast for drafts. Override for sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    # fig-width: 13
    E fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false


params:
  data_updated: !expr format(Sys.Date(), "%B %Y")
---



```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED


# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)

# -----------------------------------------------------------------------------
# Download button helper (HTML only)
# -----------------------------------------------------------------------------
ds_download_data_btn <- function(df, filename_base, label = "Download Data (CSV)",
                                  clean_df = NULL) {
  write_df <- if (!base::is.null(clean_df)) clean_df else df
  tmp <- base::tempfile(fileext = ".csv")
  readr::write_excel_csv(write_df, tmp)
  downloadthis::download_file(
    path         = tmp,
    output_name  = filename_base,
    button_label = label,
    button_type  = "primary",
    has_icon     = TRUE,
    icon         = "fa fa-download"
  )
}



# Flextable built-in themes:
  # flextable::theme_alafoli()	|>  # BLAH
  # flextable::theme_apa()  # THIS IS NICE
  # flextable::theme_booktabs() |>  # NICE, MORE COMPACT
  # flextable::theme_box() |>   # OK, INCLUDES CELL BORDERS
  # flextable::theme_tron() |>  # 'DARK MODE' BLUE TEXT
  # flextable::theme_tron_legacy() |>   # 'DARK MODE' YELLOW TEXT
  # flextable::theme_vader() |>    # 'DARK MODE' WHITE TEXT
  # flextable::theme_vanilla() |>   # NOT SPECIAL
  # flextable::theme_zebra()	|>
  #


# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 16,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 14,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 14,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 14),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 12,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)



  
# Set seed for reproducibility
base::set.seed(123)

```

## Introduction
This document presents **total compensation** for the City of Schertz's top-paid employees across fiscal years FY20 through FY25. Top employees include department directors, managers, and public safety leadership (Police, Fire, EMS).

::: {.callout-note}
## Interactive Tables
**Data last updated:** `r params$data_updated`
The tables below are fully interactive:

- **Sort**: Click any column header to sort
- **Search**: Use the search box to find specific employees
- **Filter**: Use column filters to narrow results
- **Export**: Use the Copy, CSV, or Excel buttons to export data
- **Print**: Use your browser's print function (Ctrl+P / Cmd+P)
:::

---

## Step 1: Load Cached Data
```{r}
#| label: load-cached-data
#| cache: true

# =============================================================================
# LOAD CACHED DATA FROM OMNIBUS ANALYSIS
# =============================================================================
# Purpose: Load the pre-processed all_data object from the main analysis
# Cache: TRUE - data loads once, subsequent renders use cached version
# 
# Expected columns in all_data:
#   - fiscal_year: FY20, FY21, FY22, FY23, FY24, FY25
#   - name: Employee name (Last, First format)
#   - name_std: Standardized name for matching
#   - department: Department name
#   - job_title: Job title
#   - regular_earnings: Base salary/regular pay
#   - additional_benefits: Healthcare, pension, etc.
#   - calc_total_compensation: Total compensation (earnings + benefits)
#   - leave_payout, overtime_earnings, additional_earnings, etc.

# Path to cleaned data (adjust if your path differs)
data_path <- here::here("data_clean", "compensation_wide_fy20_fy25.rds")

# Check if file exists; if not, provide instructions
if (base::file.exists(data_path)) {
  all_data <- base::readRDS(data_path)
  base::cat("Loaded data:", base::nrow(all_data), "rows\n")
  base::cat("Fiscal years:", base::paste(base::unique(all_data$fiscal_year), collapse = ", "), "\n")
} else {
  base::stop(
    "Data file not found at: ", data_path, "\n",
    "Please run the omnibus analysis first to generate the cleaned data,\n",
    "or update the data_path variable to point to your RDS file."
  )
}
```

---

## Step 2: Define Top Employee Filter Criteria
```{r}
#| label: define-top-employee-criteria
#| cache: true

# =============================================================================
# DEFINE CRITERIA FOR TOP EMPLOYEES
# =============================================================================
# Purpose: Create patterns to identify directors, managers, and public safety leaders
# 
# Categories captured:
#   1. Department Directors (e.g., "Director", "Assistant Director")
#   2. Department Managers (e.g., "Manager", "Superintendent")
#   3. City Leadership (City Manager, Assistant City Manager, etc.)
#   4. Public Safety Chiefs/Leaders (Police Chief, Fire Chief, EMS Director)
#
# Note: Uses case-insensitive regex matching on job_title

# Pattern for leadership job titles (case-insensitive)
leadership_pattern <- base::paste(
  # City executive leadership
  "city manager",
  "assistant city manager",
  "deputy city manager",
  

  # Department directors
  "director",
  "assistant director",
  "deputy director",
  
  # Department managers and supervisors
  "manager",
  "superintendent",
  
  # Public safety leadership
  "chief",              # Police Chief, Fire Chief, Assistant Chief
  "commander",          # Police Commander
  "captain",            # Fire Captain, Police Captain
  "battalion chief",    # Fire Battalion Chief
  "marshal",            # Fire Marshal
  
  sep = "|"
)

# Verify the pattern
base::cat("Leadership title pattern:\n")
base::cat(leadership_pattern, "\n")
```

---

## Step 3: Filter Top Employees
```{r}
#| label: filter-top-employees
#| cache: true

# =============================================================================
# FILTER DATA TO TOP EMPLOYEES ONLY
# =============================================================================
# Purpose: Apply leadership pattern to identify top-paid positions
# Method: Case-insensitive regex match on job_title column
#
# Output: top_employees tibble containing only leadership positions

top_employees <- all_data |>
  # Filter to leadership positions using case-insensitive regex
  dplyr::filter(
    stringr::str_detect(
      job_title, 
      stringr::regex(leadership_pattern, ignore_case = TRUE)
    )
  ) |>
  # Select relevant columns for analysis
  dplyr::select(
    fiscal_year,
    name,
    name_std,
    department,
    job_title,
    regular_earnings,
    additional_benefits,
    calc_total_compensation
  ) |>
  # Sort by fiscal year and total compensation (highest first)
  dplyr::arrange(fiscal_year, dplyr::desc(calc_total_compensation))

# Report filtering results
base::cat("\n=== TOP EMPLOYEES FILTER RESULTS ===\n")
base::cat("Total employees in dataset:", base::nrow(all_data), "\n")
base::cat("Top employees identified:", base::nrow(top_employees), "\n")
base::cat("Unique individuals:", dplyr::n_distinct(top_employees$name_std), "\n")
```

---

## Step 4: Verify Captured Job Titles
```{r}
#| label: verify-job-titles
#| cache: true

# =============================================================================
# VERIFY WHICH JOB TITLES WERE CAPTURED
# =============================================================================
# Purpose: Display all unique job titles captured to verify filtering accuracy
# This helps identify any missing titles or false positives

captured_titles <- top_employees |>
  dplyr::distinct(job_title) |>
  dplyr::arrange(job_title)

# Display captured titles
captured_titles |>
  flextable::flextable() |>
  flextable::add_header_lines("Job Titles Captured as Top Employees") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::autofit()
```

::: {.content-visible when-format="html"}
```{r}
#| label: captured-titles-download
#| echo: false
ds_download_data_btn(
  captured_titles,
  filename_base = "schertz_captured_leadership_titles",
  label = "Download Job Titles (CSV)"
)
```
:::

---

## Step 5: Create Wide Table - Total Compensation by FY
```{r}
#| label: total-comp-wide-table
#| cache: true

# =============================================================================
# CREATE WIDE FORMAT TABLE: TOTAL COMPENSATION BY FISCAL YEAR
# =============================================================================
# Purpose: Pivot data so each fiscal year is a column
# Shows total compensation for each top employee across all years
#
# Output columns:
#   - name: Employee name
#   - department: Most recent department
#   - job_title: Most recent job title
#   - FY20 through FY25: Total compensation for each year

# First, get the most recent job info for each employee
most_recent_info <- top_employees |>
  dplyr::group_by(name_std) |>
  dplyr::slice_max(order_by = fiscal_year, n = 1) |>
  dplyr::ungroup() |>
  dplyr::select(name_std, name, department, job_title)

# Create wide format with FY columns
total_comp_wide <- top_employees |>
  # Select key columns for pivot
  dplyr::select(name_std, fiscal_year, calc_total_compensation) |>
  # Pivot fiscal years to columns
  tidyr::pivot_wider(
    names_from = fiscal_year,
    values_from = calc_total_compensation,
    values_fill = NA_real_
  ) |>
  # Join back the most recent employee info
  dplyr::left_join(most_recent_info, by = "name_std") |>
  # Reorder columns: name info first, then FY columns
  dplyr::select(
    name, 
    department, 
    job_title,
    dplyr::any_of(c("FY20", "FY21", "FY22", "FY23", "FY24", "FY25"))
  ) |>
  # Sort by FY25 total compensation (or most recent available year)
  dplyr::arrange(
    dplyr::desc(
      dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20)
    )
  )

# Report table dimensions
base::cat("\n=== WIDE TABLE DIMENSIONS ===\n")
base::cat("Rows (unique employees):", base::nrow(total_comp_wide), "\n")
base::cat("Columns:", base::ncol(total_comp_wide), "\n")
```

---

## Table 1: Top Employee Total Compensation by Fiscal Year

::: {.content-visible when-format="html"}
```{r}
#| label: display-total-comp-dt
#| cache: false

# =============================================================================
# HTML OUTPUT: INTERACTIVE DT::DATATABLE
# =============================================================================
# Purpose: Display interactive table for HTML format
# Features: Sorting, searching, pagination, formatted currency
# Export buttons: Copy, CSV, Excel, Print

total_comp_wide |>
  # Format currency columns for display
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("FY"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  DT::datatable(
    caption = "Top Employee Total Compensation by Fiscal Year (FY20-FY25)",
    extensions = "Buttons",
    options = base::list(
      pageLength = 25,
      scrollX = TRUE,
      dom = "Bfrtip",
      buttons = base::list(
        "copy", 
        "csv", 
        "excel",
        base::list(
          extend = "print",
          title = "Top Employee Total Compensation - City of Schertz"
        )
      )
    ),
    rownames = FALSE,
    filter = "top"
  )
```
:::

::: {.content-visible when-format="typst"}
```{r}
#| label: display-total-comp-flextable
#| cache: false

# =============================================================================
# PDF/TYPST OUTPUT: FLEXTABLE
# =============================================================================
# Purpose: Display formatted table for PDF/typst rendering
# Styling: Dan's preferred flextable formatting

total_comp_wide |>
  # Format currency columns
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("FY"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  flextable::flextable() |>
  # Add title row

  flextable::add_header_lines("Top Employee Total Compensation by Fiscal Year") |>
  # Row 1 styling: blue text, white background, italic
 flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  # Row 2 styling: palegreen background (column headers)
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  # Footer note
  flextable::add_footer_lines(
    "Source: City of Schertz Employee Compensation Data | Blank cells indicate employee not in position that year"
  ) |>
  flextable::italic(part = "footer") |>
  flextable::fontsize(size = 10, part = "footer") |>
  # Column formatting
  ftExtra::colformat_md() |>
  flextable::autofit()
```
:::

::: {.content-visible when-format="html"}
```{r}
#| label: total-comp-raw-download
#| echo: false
ds_download_data_btn(
  total_comp_wide,
  filename_base = "schertz_top_employees_total_comp_fy20_fy25_numeric",
  label = "Download Raw Numeric Data (CSV)"
)
```
:::

---

## Object Reference

| Object Name | Description | Rows | Columns |
|-------------|-------------|------|---------|
| `all_data` | Full compensation dataset (cached from omnibus) | `r base::nrow(all_data)` | `r base::ncol(all_data)` |
| `top_employees` | Filtered to leadership positions | `r base::nrow(top_employees)` | `r base::ncol(top_employees)` |
| `total_comp_wide` | Wide format: FY columns for total comp | `r base::nrow(total_comp_wide)` | `r base::ncol(total_comp_wide)` |
| `captured_titles` | Unique job titles matched | `r base::nrow(captured_titles)` | 1 |

---

## Data Notes

- **Total Compensation** = Regular Earnings + Overtime + Additional Earnings + Leave Payout + Deployment + Arbitration + Additional Benefits
- **FY20-FY22**: Benefits were imputed based on FY23-FY25 ratios (benefits data not available in source files)
- **Blank cells**: Employee not in this position during that fiscal year

---

*Report generated: `r base::format(base::Sys.time(), "%B %d, %Y at %I:%M %p")`*
*Data source: City of Schertz Employee Compensation Records*
