---
title: "Schertz FY 2025–26: Compensation Growth vs. Tax Rate (Sustainability Review)"
subtitle: ""
authors:
    - Dan Swart, CPA (ret.)
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    echo: true
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true     # Set to 'false' for drafts. Override if you're NOT sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    echo: false
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false


params:
  data_updated: !expr format(Sys.Date(), "%B %Y")
---

```{r}
#| label: setup
#| include: false

# Capture current filename for use in document
file_name <- if (!base::is.null(knitr::current_input())) {
  paste0(tools::file_path_sans_ext(knitr::current_input()), ".qmd")
} else {
  "interactive_session.qmd"
}

file_stem <- base::paste0(tools::file_path_sans_ext(file_name), "_chat")



# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED

# Download button helpers live here; knitr::is_html_output() guard is inside
# the helpers so they silently no-op when rendering to typst

source(here::here("R", "report_format_helpers.R"))
source(here::here("R", "flextable_helpers.R"))
source(here::here("R", "download_button_helpers.R"))

# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  theme_fun = base::identity,          # <-- add this
  post_process_html = identity,
  post_process_docx = identity
)

# Standard table styler: blue italic title row + palegreen column header row.
# Defined here so all QMDs that source this setup get the same presentation
# without depending on census_helpers.R for a purely cosmetic function.
ds_style_flextable <- function(ft, title = NULL, title_size = 16) {
  # Fixed layout helps long text wrap instead of expanding columns
  ft <- flextable::set_table_properties(ft, layout = "fixed")

  if (!base::is.null(title)) {
    ft <- flextable::add_header_lines(ft, values = title)
    ft <- flextable::color(ft, i = 1, part = "header", color = "blue")
    ft <- flextable::italic(ft, i = 1, part = "header")
    ft <- flextable::align(ft, i = 1, part = "header", align = "left")
    ft <- flextable::bold(ft, i = 1, part = "header")
    ft <- flextable::fontsize(ft, i = 1, part = "header", size = title_size)
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "white")
    ft <- flextable::bg(ft, i = 2, part = "header", bg = "palegreen")

    # Optional: make wrapped titles read better
    ft <- flextable::line_spacing(ft, i = 1, part = "header", space = 1.2)
  } else {
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "palegreen")
  }

  ft <- flextable::autofit(ft)
  ft
}



# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 28,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 22,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 18,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 18,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)


# claude <- ellmer::chat_anthropic(model = "claude-opus-4-6-20260205")
  
# Set seed for reproducibility
base::set.seed(123)

```


\

#### **Filename:**

::: {.content-visible when-format="html"}
[`r file_name`]{style="word-break: break-all; font-size: 1.00em;"}
:::

```{=typst}
#let wrap_underscores(s) = s.replace("_", "_\u{200B}")
#text(size: 16pt)[#raw(wrap_underscores("`r file_name`"))]
```

\


## Purpose and scope

This memo reviews one issue only: **whether recurring compensation growth is being supported by a stable revenue base, or whether it increases pressure for future property tax rate increases**.

It is written as a policy critique focused on long-term sustainability and tax impact.

## What the budget says (direct quotes)

The budget states the merit/step program begins midyear:

> “These changes are planned to take effect midyear and cost $784,000.” (Budget Highlights, p. 7)

The budget shows the adopted/proposed tax rates:

- FY 2024–25 rate: **$0.4900**  

- No-New-Revenue rate: **$0.5064**  

- Budget proposed rate: **$0.5064** (Tax Rates table, p. 11)  


The budget also reports weakness in the existing tax base:

> “Without new value from annexations and development, the value of the existing properties decreased by $83,012,028.” (Property tax discussion, p. 11)

## Full-year correction for the merit/step program

Because the merit/step program starts midyear, the $784,000 amount is a **half-year** cost. For sustainability and “time-to-double” analysis, the recurring **full-year** amount should be used:

- Half-year shown in the budget: **$784,000**
- Full-year recurring amount: **$784,000 × 2 = $1,568,000**

This matters because recurring pay commitments do not end after one year. They become a permanent part of the payroll base.

## Key sustainability concern: short-term financing vs. long-term commitments

A fundamental financial principle is to match **short-term financing** with **short-term costs**, and **long-term revenue sources** with **long-term commitments**.

When a city supports a **long-term, recurring** payroll commitment using revenue gains that are not reliable (for example, gains that depend on a changing tax base or one-time conditions), the result is predictable: **future tax rate increases become not optional, but inevitable** unless services are reduced or hiring is frozen.

---

## Inputs and assumptions (transparent and editable)

```{r}
#| label: inputs-assumptions-table

assumptions <- data.frame(
  Item = c(
    "Merit/Step cost shown in budget (half-year)",
    "Full-year recurring Merit/Step cost",
    "FY 2024-25 tax rate",
    "FY 2025-26 No-New-Revenue / Proposed rate",
    "FY 2025-26 total taxable value"
  ),
  Value = c(
    "$784,000 (midyear start)",
    "$1,568,000 (784,000 × 2)",
    "$0.4900 per $100",
    "$0.5064 per $100",
    "$5,734,815,455"
  ),
  Source = c(
    "Budget Highlights (p. 7)",
    "Calculated from budget statement",
    "Tax Rates table (p. 11)",
    "Tax Rates table (p. 11)",
    "Property tax discussion (p. 11)"
  ),
  stringsAsFactors = FALSE
)

assumptions_spec <- list(
  title = "Inputs and Assumptions",
  labels = list(Item = "Item", Value = "Value", Source = "Source")
)

ds_ft(assumptions, assumptions_spec)

ds_download_data_btn(
  assumptions,
  filename_base = base::paste0(file_stem, "_inputs_assumptions"),
  label = "Download Inputs Table (CSV)"
)
```

---

## Time-to-double and structural pressure

This chart shows the relationship between ongoing compensation growth (as a percent per year) and how quickly costs can double. It is a simple planning tool: higher ongoing growth rates lead to much faster doubling-time.

```{r}
#| label: time-to-double-plot

growth_rates <- data.frame(
  annual_growth_rate = seq(0.02, 0.08, by = 0.01)
) |>
  dplyr::mutate(
    annual_growth_pct = annual_growth_rate * 100,
    doubling_time_years = log(2) / log(1 + annual_growth_rate)
  )

plot_time_to_double <- ggplot2::ggplot(growth_rates, ggplot2::aes(x = doubling_time_years, y = annual_growth_pct)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::labs(
    title = "Time to Double vs. Ongoing Compensation Growth",
    x = "Approximate years to double (% per year)",
    y = "Ongoing compensation growth (% per year)"
  ) +
  ggplot2::theme_minimal()

plot_time_to_double

ds_download_plot_btn(
  plot_time_to_double,
  filename_base = base::paste0(file_stem, "_time_to_double"),
  label = "Download Time-to-Double Plot (PNG)"
)
```

### Plain-language takeaway

As total compensation keeps rising every year, it eventually doubles, usually much sooner than you think.  

When payroll doubles, the city needs much more revenue just to keep service levels the same.

---

## 5-year projection model: payroll pressure vs. projected tax collected

### What this model does

This model compares:

1. **Payroll pressure** = starting payroll base + recurring annual increase (full-year $1.568M)  
2. **Projected tax collected** (simplified) = taxable value × assumed growth × tax rate

This is not a full revenue forecast. It is a focused stress test to keep the conversation on compensation growth vs. the property tax base.

### Growth rates used (shown in scenario names)

- **Optimistic:** 4.0% annual taxable value growth  
- **Baseline:** 2.0% annual taxable value growth  
- **Stress:** -0.5% annual taxable value decline  

### Model code (editable)



```{r}
#| label: five-year-model

# Inputs (edit if you want)
tax_rate_per_100 <- 0.5064
taxable_value_start <- 5734815455

# Merit/Step full-year recurring cost (from midyear $784k statement)
merit_step_full_year <- 784000 * 2

# Illustrative payroll base (replace with a confirmed payroll base if desired)
payroll_base_start <- 48623715

years <- 0:5

scenarios <- data.frame(
  scenario = c("Optimistic (4% Growth)",
               "Baseline (2% Growth)",
               "Stress (-0.5% Growth)"),
  taxable_value_growth = c(0.04, 0.02, -0.005)
)

proj <- tidyr::expand_grid(
  scenarios,
  year = years
) |>
  dplyr::mutate(
    taxable_value = taxable_value_start * (1 + taxable_value_growth) ^ year,
    projected_tax_collected = (taxable_value / 100) * tax_rate_per_100,
    payroll_pressure = payroll_base_start + (merit_step_full_year * year),
    gap = projected_tax_collected - payroll_pressure
  )
```

### Projection table (includes Gap)

Gap is defined as:

- **gap = projected_tax_collected − payroll_pressure**

So:
- **Positive gap** = more tax capacity than payroll pressure (less stress)
- **Negative gap** = payroll pressure is larger (more stress)

```{r}
#| label: projection-table-with-gap

proj_table <- proj |>
  dplyr::mutate(
    projected_tax_collected = round(projected_tax_collected, 0),
    payroll_pressure = round(payroll_pressure, 0),
    gap = round(gap, 0)
  ) |>
  dplyr::select(
    scenario,
    year,
    projected_tax_collected,
    payroll_pressure,
    gap
  )

proj_table_display <- proj_table |>
  dplyr::mutate(
    projected_tax_collected = scales::dollar(projected_tax_collected, accuracy = 1),
    payroll_pressure = scales::dollar(payroll_pressure, accuracy = 1),
    gap = scales::dollar(gap, accuracy = 1)
  )

proj_table_spec <- list(
  title = "5-Year Projection Table with Gap",
  labels = list(
    scenario = "Scenario",
    year = "Year",
    projected_tax_collected = "Projected Tax Collected",
    payroll_pressure = "Payroll Pressure",
    gap = "Gap"
  )
)

ds_ft(proj_table_display, proj_table_spec)

ds_download_data_btn(
  proj_table,
  filename_base = base::paste0(file_stem, "_projection_table_with_gap"),
  label = "Download Projection Table (CSV)"
)

```


---


## Bottom line (focused only on compensation vs tax rate)

- The merit/step program cost shown in the budget is **midyear** ($784,000), which becomes **$1.568M per year** when annualized. (p. 7)  

- The property tax rate is proposed at the **No-New-Revenue** rate ($0.5064), and the budget reports the **existing** tax base declined without new growth. (p. 11)  

- That combination creates a sustainability risk: **recurring payroll growth** can keep stacking up even when the tax base is not reliably growing.  


If the city continues to add recurring compensation costs without matching them to long-term, stable funding, **large tax rate increases are not optional, but inevitable**.

The unpleasant option:  staffing or services must be reduced.

