---
title: "Your Taxes Went Up. Here Is How That Decision Was Made."
subtitle: "A Plain-Language Review of the City of Schertz FY 2025-26 Budget Process for Staffing and Compensation"
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    echo: true
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true     # Set to 'false' for drafts. Override if you're NOT sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    echo: false
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    error: false
    warning: false
    message: false
    cache: false


params:
  data_updated: !expr format(Sys.Date(), "%B %Y")
---

```{r}
#| label: setup
#| include: false

# Capture current filename for use in document
file_name <- if (!base::is.null(knitr::current_input())) {
  paste0(tools::file_path_sans_ext(knitr::current_input()), ".qmd")
} else {
  "interactive_session.qmd"
}

file_stem <- base::paste0(tools::file_path_sans_ext(file_name), "_chat")



# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED

# Download button helpers live here; knitr::is_html_output() guard is inside
# the helpers so they silently no-op when rendering to typst

source(here::here("R", "report_format_helpers.R"))
source(here::here("R", "flextable_helpers.R"))
source(here::here("R", "download_button_helpers.R"))

# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  theme_fun = base::identity,          # <-- add this
  post_process_html = identity,
  post_process_docx = identity
)

# Standard table styler: blue italic title row + palegreen column header row.
# Defined here so all QMDs that source this setup get the same presentation
# without depending on census_helpers.R for a purely cosmetic function.
ds_style_flextable <- function(ft, title = NULL, title_size = 16) {
  # Fixed layout helps long text wrap instead of expanding columns
  ft <- flextable::set_table_properties(ft, layout = "fixed")

  if (!base::is.null(title)) {
    ft <- flextable::add_header_lines(ft, values = title)
    ft <- flextable::color(ft, i = 1, part = "header", color = "blue")
    ft <- flextable::italic(ft, i = 1, part = "header")
    ft <- flextable::align(ft, i = 1, part = "header", align = "left")
    ft <- flextable::bold(ft, i = 1, part = "header")
    ft <- flextable::fontsize(ft, i = 1, part = "header", size = title_size)
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "white")
    ft <- flextable::bg(ft, i = 2, part = "header", bg = "palegreen")

    # Optional: make wrapped titles read better
    ft <- flextable::line_spacing(ft, i = 1, part = "header", space = 1.2)
  } else {
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "palegreen")
  }

  ft <- flextable::autofit(ft)
  ft
}



# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 28,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 22,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 18,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 18,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)


# claude <- ellmer::chat_anthropic(model = "claude-opus-4-6-20260205")
  
# Set seed for reproducibility
base::set.seed(123)

```


\

#### **Filename:**

::: {.content-visible when-format="html"}
[`r file_name`]{style="word-break: break-all; font-size: 1.00em;"}
:::

```{=typst}
#let wrap_underscores(s) = s.replace("_", "_\u{200B}")
#text(size: 16pt)[#raw(wrap_underscores("`r file_name`"))]
```

\
\


---

## What the Budget Proposes

The City of Schertz is proposing a property tax rate of **\$0.5064 per \$100
of taxable value** for FY 2025-26, up from \$0.4900 last year. That is an
increase of \$0.0164 per \$100 — or roughly **\$49 per year** for a home with
a taxable value of \$300,000. That same home paid about \$1,470 in city
property taxes last year. Under the proposed rate, it would pay about \$1,519.

The proposed budget also increases utility fees. Water rates increase 2%.
Sewer rates increase 3%. The average monthly bill goes up approximately \$2.46.
The drainage fee nearly doubles — from \$5.20 to \$8.00 per month — an
increase of \$33.60 per year per household.

In total, the City is proposing to spend approximately **\$139.3 million**
across all funds in FY 2025-26. That is \$4.1 million more than the current
year — a 3% increase.

```{r}
#| label: impact-table

home_vals <- c(200000, 300000, 400000, 500000)
tax_increase <- 0.0164

tax_impact   <- home_vals / 100 * tax_increase
sewer_impact <- 2.46 * 12
drainage_imp <- (8.00 - 5.20) * 12

impact_tbl <- data.frame(
  Home_Value   = fmt_dollar(home_vals),
  Tax_Impact   = fmt_dollar(tax_impact, digits = 2),
  Util_Impact  = fmt_dollar(sewer_impact, digits = 2),
  Drain_Impact = fmt_dollar(drainage_imp, digits = 2),
  Total        = fmt_dollar(tax_impact + sewer_impact + drainage_imp, digits = 2),
  stringsAsFactors = FALSE
)

impact_spec <- list(
  title = "Estimated Annual Impact of Proposed Rate and Fee Increases — by Home Value",
  labels = list(
    Home_Value = "Taxable Home Value",
    Tax_Impact = "Property Tax Increase",
    Util_Impact = "Water/Sewer Increase",
    Drain_Impact = "Drainage Fee Increase",
    Total = "Total Annual Increase"
  )
)

ds_ft(impact_tbl, impact_spec)

ds_download_data_btn(
  impact_tbl,
  filename_base = base::paste0(file_stem, "_impact_table"),
  label = "Download Impact Table (CSV)"
)
```

\
\

Where is that money going? The two largest drivers are **employee compensation**
and **new positions**. This document examines how those decisions were made —
and what questions the process did not answer.

---

## The Two Biggest Cost Drivers

### Compensation Increases

The budget proposes a **merit program** for non-public-safety employees
(0–4% raise based on performance evaluation) and a **step increase** for
public safety employees (up to 4% based on tenure). The City states that
these increases will cost **\$784,000 citywide**.

That \$784,000 figure deserves a closer look. The raises take effect
*mid-year* — halfway through the fiscal year. So the budget only reflects
*half* the annual cost. Starting in FY 2026-27, those same raises cost
**\$1,568,000 per year** — before any additional raises are proposed.
That additional \$784,000 will appear in next year's budget as part of the
existing baseline, largely invisible as a discrete cost.

When you look at total employee compensation — wages, overtime, retirement
contributions, health insurance, workers' compensation, and all related costs —
the numbers are larger still. Reading through every department in the budget,
General Fund personnel costs are proposed at approximately **\$37.5 million**
for FY 2025-26, up from \$36.9 million last year. Across all funds combined
(including Water/Sewer, Drainage, and EMS), total personnel costs rise from
approximately \$50.3 million to \$52.6 million.



### New Positions

The budget adds **26 new positions** across all funds:

```{r}
#| label: positions-table

pos_tbl <- data.frame(
  Fund = c("General Fund","General Fund","General Fund","General Fund",
           "Water & Sewer","EMS","Drainage"),
  Count = c(6, 1, 2, 1, 6, 4, 6),
  Description = c(
    "Firefighters — to staff the new Fire Station 4",
    "Assistant City Secretary",
    "Athletics Coordinator and Recreation Specialist (Soccer Complex takeover)",
    "Senior Center Supervisor (Senior Center operations takeover)",
    "New water/wastewater crew plus one floater position",
    "EMTs to staff additional ambulances at peak hours",
    "New drainage maintenance crew plus a Floodplain Manager"
  ),
  stringsAsFactors = FALSE
)

pos_spec <- list(
  title = "Proposed New Positions — FY 2025-26",
  labels = list(
    Fund = "Fund",
    Count = "Positions",
    Description = "Description"
  )
)

ds_ft(pos_tbl, pos_spec)

ds_download_data_btn(
  pos_tbl,
  filename_base = base::paste0(file_stem, "_positions_table"),
  label = "Download Positions Table (CSV)"
)
```

\
\

Every one of these positions represents a permanent, recurring cost. Unlike a
road repair or an equipment purchase, a position does not depreciate. It comes
back every year — with a raise.

---

## How These Decisions Were Made

Here is the process, as it actually works:

The **City Manager's office** prepares the proposed budget and presents it to
City Council. Council reviews it, asks questions at public hearings, and votes.
Citizens can attend those hearings and comment.

In practice, the information Council receives comes almost entirely from the
City Manager's office. Council members are part-time elected officials; the
City Manager and department heads are full-time professionals. The professionals
know their departments in detail. That is appropriate — but it also means that
the information flowing to Council is shaped by the people whose budgets are
being reviewed.

This is not unusual. It is how most city governments work. The question is
whether the information provided is sufficient for Council to evaluate the
requests — or whether Council is being asked to trust the professionals
without the data needed to verify their claims.

---

\

## What the Justifications Actually Said

For each of the 26 proposed new positions, what case did the budget document
make?

For the **6 firefighters**, the justification is clear and concrete: a new
fire station is being built and cannot operate without staff. This is a
straightforward operational necessity.

For the **Soccer Complex and Senior Center positions**, the City is taking
over operations previously managed by other entities. The headcount follows
the operations transfer. Whether those positions are sized correctly for the
actual workload is not addressed.

For the **Assistant City Secretary**, the budget says the position is needed
to help "with existing operations." That is the complete justification. No
workload data. No description of what existing operations are currently
unmet. No alternative considered.

For the **Water/Sewer crew**, the justification is growth. Fair enough —
Schertz is a growing city. But how many service connections are currently
maintained per crew? How does that compare to what an efficiently run
water utility maintains? The budget does not say.

For the **4 EMTs**, the budget describes the goal of reducing response times
and staffing peak call hours. Have response times actually increased? By how
much? What is the call volume trend? What do the best-performing EMS
operations of comparable size look like? None of this appears.

For the **Drainage crew**, the need is described as maintaining existing
channels and planning for expansions. How many channel-miles are currently
maintained per crew? What is the backlog of deferred maintenance? The budget
does not say.

For the **Floodplain Manager**, a regulatory mandate is cited. That is a
legitimate driver — but the specific mandate and its requirements are not named.

**The pattern is consistent:** the City tells you *what* it wants to hire.
It does not show you *why that number* is the right answer, or whether the
existing workforce is being used as efficiently as possible before more is
added.

---

\

## What Frugal Governance Looks Like

Frugality in government is not the same as cutting services. It means getting
the maximum value from every dollar before asking taxpayers for more. A
frugal governing body asks different questions than one that simply reviews
what staff proposes.

Here is the difference in practice.

**What a deferential Council asks:**
*"Does this request seem reasonable given the department's description of its needs?"*

**What a frugal Council asks:**
*"Show us the trend in calls per unit, work orders per employee, and response
times over the last three years. Show us how the most efficient comparable
city handles this function. Explain what you tried before concluding that
adding headcount was the answer."*

The first approach depends on trust. The second approach verifies. Both
trust and verification have their place — but for permanent spending
commitments that compound over time, verification is the governing body's
obligation to the people who pay the bills.

There is a principle that applies directly here: **governing bodies should
look to outside organizations not for examples of how to spend as much as
others, but for examples of how to deliver the same — or better — service
for less.** The benchmark is efficiency, not average practice. If another
Texas city of similar size maintains its water infrastructure with fewer
crew members per service connection, the question is not whether Schertz
is close to the average. The question is how to get to where that efficient
city is.

---

## The Compounding Reality

Every compensation increase and every new position adds to a base that grows
from year to year. Consider what that means over time.

General Fund personnel costs have grown at a compound annual rate of
approximately **12%** over the three years for which data is available in
this budget — from roughly \$26.6 million in FY 2022-23 to a proposed
\$37.5 million in FY 2025-26. At that rate, personnel costs would double
in approximately **6 years**.

```{r}
#| label: doubling-table

base_gf  <- 37500000
base_all <- 52600000
cagr     <- 0.12
years    <- c(0, 3, 6, 9, 12)

proj_tbl <- data.frame(
  Years_From_Now = paste0("FY ", 2026 + years),
  GF_Personnel   = fmt_dollar(base_gf  * (1 + cagr)^years),
  All_Personnel  = fmt_dollar(base_all * (1 + cagr)^years),
  stringsAsFactors = FALSE
)

proj_spec <- list(
  title = "Illustrative Doubling Projections at Current Growth Rate",
  labels = list(
    Years_From_Now = "Year",
    GF_Personnel = "General Fund Personnel Costs",
    All_Personnel = "All Funds Personnel Costs"
  )
)

ds_ft(proj_tbl, proj_spec)

ds_download_data_btn(
  proj_tbl,
  filename_base = base::paste0(file_stem, "_doubling_projection_table"),
  label = "Download Doubling Projection Table (CSV)"
)
```

\
\

Property tax revenue, constrained by state law and the voter-approval rate
cap, grows at roughly 7% per year under good conditions — a doubling time
of about 10 years. Personnel costs growing at 12% per year outpace revenue
growing at 7% per year by a compounding margin. The gap between what the
City costs and what taxpayers can fund through current revenue sources does
not stay flat — it widens, year after year, until something breaks.

That something is usually: a larger tax increase than anyone expected, a
sudden reduction in services, or a draw-down of reserves. None of these
outcomes announce themselves in advance. They accumulate quietly, then arrive
all at once.

---

## The Constituency Dynamic

When a City Manager's office proposes adding 26 positions, it is worth
understanding what those positions mean organizationally — beyond their
direct cost.

Each new employee becomes part of the organization. The organization has
an interest in retaining and supporting its employees. That interest
generates future budget pressure — for training, for equipment, for
compensation increases, for additional support positions. Each new position
is not just a line item; it is the beginning of an organizational commitment.

This is not a criticism of the employees. It is a description of how
organizations behave. A growing workforce generates internal political
support for further growth. Department heads who have grown their teams
advocate for their teams. The City Manager who approved the growth defends
it. Over time, the employees of a city government become its most
reliable constituency — present at every budget meeting, professionally
invested in the outcome, and far more organized than the general public.

The remedy is not hostility toward city employees. The remedy is a
governing body that maintains independent standards for what evidence
it requires — and holds those standards consistently, regardless of
how requests are framed.

---


## What Citizens Can Ask

Public comment at City Council meetings is most effective when it is
specific. The following questions are grounded in the budget document and
ask for information the City should be able to provide.

**On staffing:**

1. *"For each of the 26 new positions proposed, what is the three-year trend
   in the primary workload indicator for that department — calls, service
   connections, work orders, or equivalent — and what is the current output
   per existing employee?"*

2. *"What alternatives to adding headcount were evaluated before each new
   position was proposed, and what was the result of that evaluation?"*

3. *"What are the City's adopted service-level standards for Police, Fire,
   EMS, Water/Sewer, and Drainage — and how does current performance compare
   to those standards?"*

**On compensation:**

4. *"The proposed merit/step program costs \$784,000 in FY 2025-26 because
   it takes effect mid-year. What is the full annualized cost beginning in
   FY 2026-27, and is that amount shown in the five-year financial forecast?"*

5. *"General Fund personnel costs have grown at approximately 12% per year
   for three years. What is the long-run property tax revenue growth rate,
   and what is the plan for closing the gap between compensation growth and
   revenue growth?"*

**On frugality:**

6. *"Which Texas city of comparable size delivers comparable services at the
   lowest cost per unit of output — and what practices make that possible?"*

7. *"Before adding positions in any department, what is the productivity
   trend for existing employees in that department over the last three years?"*

These are not hostile questions. They are the questions a governing body
should ask before voting — and that citizens are entitled to ask when
their governing body does not.

---

\

## Three Recommendations

**1. Require data before approving staffing requests.**
Every proposed position should come with three years of workload trend data,
current unit productivity figures, a description of alternatives considered,
and the full annualized cost — not just the budget-year cost. Requests that
lack this information should be deferred, not approved.

**2. Set service-level standards before the next budget cycle.**
Without adopted standards, there is no objective way to evaluate whether
current service is adequate or whether more resources are genuinely needed.
The City should establish measurable standards for each major function, set
by Council — not by the departments being measured.

**3. Demand the full-year cost, every time.**
When the City presents a compensation increase as taking effect mid-year,
the full annualized run-rate — what it costs starting the following year —
must be shown alongside the budget-year figure. Budget documents that
present only the discounted cost are not giving Council or the public an
accurate picture of the commitment being made.

---

\

*This document is based on a reading of the City of Schertz FY 2025-26
Proposed Budget. All figures come directly from that document. The analysis
is intended to inform public understanding of the budget process and to
identify questions that merit public attention.*
