---
title: "Texas Public Information Act (TPIA) Request"
subtitle: "City of Schertz Municipal Records"
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: false     # Fast for drafts. Override for sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    # fig-width: 13
    E fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: false
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false


---


```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED


# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)




# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 16,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 14,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 14,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 14),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 12,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)



  
# Set seed for reproducibility
base::set.seed(123)

```

```{r}
#| label: requestor-info
#| include: false

# =============================================================================
# CONFIGURATION - EDIT THESE VALUES FOR EACH REQUEST
# =============================================================================

# Requester Information
requester_name <- "Dan Swart"
requester_organization <- "A private citizen"  
requester_address <- "1509 Gwendolyn Way"
requester_city_state_zip <- "Schertz, Texas 78154"
requester_email <- "danswartcpa@gmail.com"
requester_phone <- "(916) 802-1040"

# Target Government Entity
entity_name <- "City of Schertz"
entity_contact <- "City Secretary"
entity_address <- "1400 Schertz Parkway"
entity_city_state_zip <- "Schertz, Texas 78154"

# Requester Time Coverage Parameters
fiscal_years_requested <- "FY23, FY24 (if available)"
fiscal_years_short <- ""

# Date calculations
request_date <- "February 09, 2026"
response_deadline <- base::format(base::Sys.Date() + 10, "%B %d, %Y")
```

\

`r request_date`  

\ 


**HAND DELIVERED**


`r entity_contact`  
`r entity_name`  
`r entity_address`  
`r entity_city_state_zip`  

**RE: Texas Public Information Act Request**

Dear `r entity_contact`:

Pursuant to the Texas Public Information Act, Texas Government Code Chapter 552, I am requesting access to and copies of the public records described below. This request is made by a private citizen.

## Requested Records

I am requesting the following information. For all numeric information, please provide the original, FINAL machine-readable electronic files (such as Excel spreadsheets or comma-separated values (CSV) files), including all detailed line-item data used to create totals, summaries, and published figures.

IF POSSIBLE, ALL NUMERICAL VALUES SHOULD BE IN NUMERICAL FORMAT (NOT TEXT) WITHIN EACH SPREADSHEET.

This request includes the final datasets and working spreadsheets actually used to prepare the published official reports for the City, not solely rendered documents or summary tables. If multiple versions exist, please provide the version used for publication or final decision-making. They must match the figures published by the City in final form.

SOURCE DOCUMENT:   CITY OF SCHERTZ ACFR FY23
                   AS PRESENTED IN PDF FILE FORMAT
                   AT URL:   https://laserfiche.schertzweb.com/WebLink/browse.aspx?id=55963&dbid=1&repo=SCHERTZ&cr=1


INFORMATION REQUESTED: 

ALL TABLES AND CHARTS PRESENTED IN THE "STATISTICAL SECTION" (pp 111 - 150 of the pdf version of the document) OF THE ACFR FY23, INCLUDING THE FOLLOWING:
 
01. Net Position by Component
02. Change in Net Position - Expenses
03. Change in Net Position - Revenues
04. Fund Balances of Governmental Funds
05. Changes in Fund Balances
06. Tax Revenues by Source
07. Water & Sewer Consumption/Rates
08. Principal Water & Sewer Consumers
09. Property Tax Rates
10. Principal Property Tax Payers
11. Tax Levies and Collections
12. Taxable Assessed Value
13. Outstanding Debt Ratios
14. General Bonded Debt Ratios
15. Direct & Overlapping Debt
16. Debt Margin Information
17. Pledged Revenue Coverage
18. Demographic & Economic Stats
19. Top 10 Employers
20. FTE Employees by Function
21. Operating Indicators - Part 1
22. Operating Indicators - Part 2
23. Capital Asset Statistics

The CITY OF SCHERTZ ACFR FY24 is not published on the website at this time.  If it is in final, but unpublished form please provide the above Excel files for FY24 as well.


To reduce administrative burden, I am not requesting the creation of new records or analyses. Existing electronic records may be provided in the format in which they are ordinarily maintained. If responsive records exist across multiple files or systems, production of those files is acceptable.


If responsive records are available on a rolling basis, please provide them as they are identified and ready, rather than waiting for the entire request to be completed.


If any portion of this request is unclear, overly broad, or believed to be exempt from disclosure, please identify the specific concern and the applicable statutory citation, so the request may be clarified or narrowed if appropriate.


## Format and Delivery

I request that responsive records be provided in the following formats, where available:

- **FINAL Spreadsheets and data files:**  Native electronic format (Excel, CSV, or similar) - these must match exactly the figures published in the published City of Schertz ACFR fy23  

- **Large document sets:**  Electronic delivery via email, file transfer, or portable media

- **Numerical values should be formated as numbers (not text) just as used by staff for calculating values published on the City website.

If any requested records are available on the City's website, please provide the specific URLs rather than duplicating those records.  PDF file format is NOT an acceptable substitute for spreadsheet requests.


## Cost Estimate

If the estimated cost of fulfilling this request exceeds $20.00, please provide a written cost estimate before proceeding with document production. I am willing to pay reasonable charges for document search, retrieval, and copying as permitted under the Act.


## Response Deadline

Under the Texas Public Information Act, you are required to respond to this request promptly, and in any event within ten business days. If you determine that any portion of the requested records is exempt from disclosure, please identify the specific records withheld and cite the specific statutory exception(s) claimed for each.

If you intend to seek an Attorney General ruling on any portion of this request, please notify me in writing within the time period required by law and provide me with a copy of your request for ruling.


## Contact Information

Please direct all correspondence regarding this request to:

> `r requester_name`  
<!-- > `r requester_organization`    -->
> `r requester_address`  
> `r requester_city_state_zip`  
> Email: `r requester_email`  
> Phone: `r requester_phone`  
<br>

Thank you for your prompt attention to this request. Please do not hesitate to contact me if you have any questions or require clarification regarding the scope of this request.
\

Respectfully submitted,  
\
\

`r requester_name` 

\
\

---

*This request is made pursuant to the Texas Public Information Act, Texas Government Code Chapter 552. The Act requires governmental bodies to make public information available for public inspection and copying. Failure to comply with the Act may result in civil penalties and attorney's fees.*
