---
title: "TPIA/FOIA Request Template for Municipal Fiscal Analysis"
subtitle: "Following the 5Q Decision Architecture Framework™"
description:  ""
authors: 
  - name: "Dan Swart, CPA (ret)"
  - name: "Claude Opus 4.5"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: false     # Fast for drafts. Override for sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    # fig-width: 13
    E fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false


---


```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED


# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)


#
# # Sample Code:
# flextable::flextable(violations) |>
 #  flextable::set_header_labels(
 #    Variable = "Variable",
 #    Measurement = "Measurement",
 #    Likely_Impact = "Likely Impact"
 #  ) |>
#    flextable::add_header_lines(values = "Frequent Violations of Scientific Method in Current So-Called 'Equity' Research") |>
#   flextable::color(i = 1, color = "blue", part = "header") |>
#   flextable::italic(i = 1, part = "header") |>
#   flextable::align(i = 1, align = "center", part = "header") |>
#   flextable::fontsize(i = 1, size = 14, part = "header") |>
#   flextable::bg(i = 1, bg = "white", part = "header") |>
#   flextable::bg(i = 2, bg = "palegreen", part = "header") |>
#   flextable::bold(i = 1:2, part = "header") |>
#   flextable::bold(i = 1:7, j = 1, part = "body") |>
#   ftExtra::colformat_md() |> 
#   flextable::autofit()
  

# Flextable built-in themes:
  # flextable::theme_alafoli()	|>  # BLAH
  # flextable::theme_apa()  # THIS IS NICE
  # flextable::theme_booktabs() |>  # NICE, MORE COMPACT
  # flextable::theme_box() |>   # OK, INCLUDES CELL BORDERS
  # flextable::theme_tron() |>  # 'DARK MODE' BLUE TEXT
  # flextable::theme_tron_legacy() |>   # 'DARK MODE' YELLOW TEXT
  # flextable::theme_vader() |>    # 'DARK MODE' WHITE TEXT
  # flextable::theme_vanilla() |>   # NOT SPECIAL
  # flextable::theme_zebra()	|>
  #

# Flextable titles:
  # flextable::colformat_double(j = c("Mean", "SD", "N"), big.mark = ",", digits = 1) |>
  # flextable::flextable(variance_comparison) |>
  #  flextable::add_header_lines(values = "The Within-Group vs Between-Group Variance Problem") |>

# Flextable Title theming at table-level:
  #  flextable::color(i = 1, color = "blue", part = "header") |>
  #  flextable::italic(i = 1, part = "header") |>
  #  flextable::align(i = 1, align = "center", part = "header") |>
  #  flextable::fontsize(i = 1, size = 14, part = "header") |>
  #  flextable::bg(i = 1, bg = "white", part = "header") |>

# Flextable standard table background colors:
  #  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  

# Flextable reading markdown:
  #  ftExtra::colformat_md() |> 

# Flextable auto-sizing cell widths:
  #  flextable::autofit() 
 
 # Flextable background based on SPECIFIC cell contents:
  #
  # flextable::bg(i = ~ Impact_on_Validity == "High", j = "Impact_on_Validity", bg = "#ffcccc") |>
 # flextable::bg(i = ~ Impact_on_Validity == "Medium", j = "Impact_on_Validity", bg = "#ffffcc") |>
  #
  # 
  # Apply yellow background to any cell containing "Yes":

  # for (col in base::names(hypotheses_data)) {
  #   yes_rows <- base::which(hypotheses_data[[col]] == "Yes")
  #   if (base::length(yes_rows) > 0) {
  #     ft <- ft |>
  #       flextable::bg(i = yes_rows, j = col, bg = "yellow", part = "body")
  #   }
  # } 
  #
  # Apply to last row of table:
  #
  #



# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 16,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 14,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 14,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 14),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 12,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)



  
# Set seed for reproducibility
base::set.seed(123)

```


```{r}
#| label: load-libraries
#| include: false

# =============================================================================
# LOAD LIBRARIES
# =============================================================================

# Required packages
base::library(dplyr)
base::library(flextable)
base::library(ftExtra)

# Set flextable defaults
flextable::set_flextable_defaults(

  font.size = 10,
  padding = 3
)
```

# Overview

This document provides detailed specifications for public records requests supporting the **Five Questions Framework** for municipal fiscal analysis. Each section corresponds to one of the five core analytical questions and lists the specific documents needed to complete that analysis.

**Framework Questions:**

1. **Trajectory** — Where does this decision lead?
2. **Proportionality** — What share of the budget is this?
3. **Reversibility** — How locked in are we?
4. **Assumption Testing** — What has to be true for this to work?
5. **Opportunity Cost** — What else could this money buy?

---

# Question 1: Trajectory

**Purpose:** Calculate compound annual growth rates for compensation, headcount, and revenue. Project future obligations and identify unsustainable trends.

```{r}
#| label: tbl-trajectory
#| tbl-cap: "Documents Required for Trajectory Analysis"

# =============================================================================
# QUESTION 1: TRAJECTORY - WHERE DOES THIS LEAD?
# =============================================================================

trajectory_docs <- dplyr::tibble(
  item_no = 1:7,
  document = c(
    "Employee Compensation Reports",
    "Position Classification and Pay Plans",
    "Authorized Position Schedules",
    "Headcount Reports",
    "Pay Plan Amendments",
    "COLA Documentation",
    "Merit/Step Increase Policies"
  ),
  description = c(
    "Annual listing of all employees showing: name, department, job title, base salary, overtime pay, additional earnings, benefits, total compensation",
    "Official pay grade/step tables showing minimum, midpoint, and maximum salary for each classification",
    "List of all authorized (budgeted) positions by department, whether filled or vacant",
    "Monthly or annual employee counts by department (FTE and headcount)",
    "Resolutions, ordinances, or council actions amending pay scales, adding positions, or approving compensation changes",
    "Records of COLA percentages approved by council, effective dates, and methodology",
    "Written policies governing automatic step increases, merit increase pools, and eligibility criteria"
  ),
  years_needed = c(
    "FY2015–FY2024 (10 years)",
    "FY2015–FY2024",
    "FY2015–FY2024",
    "FY2015–FY2024",
    "FY2015–FY2024",
    "FY2015–FY2024",
    "Current + prior versions"
  ),
  likely_availability = c(
    "May require compilation",
    "Usually public/online",
    "Usually in budget documents",
    "May require compilation",
    "Council minutes/ordinances",
    "Council minutes",
    "HR policy manual"
  ),
  analysis_use = c(
    "Calculate CAGR for total compensation",
    "Identify structural pay increases",
    "Track position growth over time",
    "Normalize compensation by headcount",
    "Identify discrete council decisions",
    "Separate COLA from merit increases",
    "Project automatic future increases"
  )
)

trajectory_docs |>
  flextable::flextable() |>
  flextable::set_header_labels(
    item_no = "#",
    document = "Document",
    description = "Specific Description",
    years_needed = "Years Needed",
    likely_availability = "Likely Availability",
    analysis_use = "Analysis Use"
  ) |>
  flextable::add_header_lines("Question 1: Trajectory — Where Does This Lead?") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::width(j = "item_no", width = 0.3) |>
  flextable::width(j = "document", width = 1.2) |>
  flextable::width(j = "description", width = 2.0) |>
  flextable::width(j = "years_needed", width = 1.0) |>
  flextable::width(j = "likely_availability", width = 1.0) |>
  flextable::width(j = "analysis_use", width = 1.2) |>
  flextable::valign(valign = "top", part = "body") |>
  flextable::autofit()
```

---

# Question 2: Proportionality

**Purpose:** Calculate personnel costs as a percentage of total budget. Identify what categories are being crowded out over time.

```{r}
#| label: tbl-proportionality
#| tbl-cap: "Documents Required for Proportionality Analysis"

# =============================================================================
# QUESTION 2: PROPORTIONALITY - WHAT SHARE OF THE BUDGET IS THIS?
# =============================================================================

proportionality_docs <- dplyr::tibble(
  item_no = 8:14,
  document = c(
    "Adopted Annual Budgets",
    "Budget Summaries by Expenditure Category",
    "Budget Summaries by Department",
    "Budget-to-Actual Reports",
    "Comprehensive Annual Financial Reports (CAFR/ACFR)",
    "Fund Balance Reports",
    "Revenue Reports by Source"
  ),
  description = c(
    "Complete adopted budget documents including all funds (General Fund, Enterprise Funds, Special Revenue Funds, Capital Projects Funds, Debt Service Funds)",
    "Summary tables showing total expenditures by category: Personnel/Salaries, Benefits, Supplies, Contracted Services, Capital Outlay, Debt Service, Transfers",
    "Summary tables showing total expenditures by department",
    "Year-end reports comparing budgeted amounts to actual expenditures by category and department",
    "Audited annual financial statements including all required schedules and notes",
    "Year-end fund balance statements for all funds, showing reserved, unreserved, assigned, committed, and unassigned balances",
    "Annual revenue summaries by source: property tax, sales tax, franchise fees, permits, fines, intergovernmental, charges for services, interest, other"
  ),
  years_needed = c(
    "FY2015–FY2025",
    "FY2015–FY2024",
    "FY2015–FY2024",
    "FY2015–FY2024",
    "FY2015–FY2024",
    "FY2015–FY2024",
    "FY2015–FY2024"
  ),
  likely_availability = c(
    "Usually online",
    "In budget documents",
    "In budget documents",
    "May require request",
    "Usually online",
    "In CAFR",
    "In CAFR/budget"
  ),
  analysis_use = c(
    "Total budget baseline for ratios",
    "Personnel as % of total spending",
    "Department growth comparison",
    "Variance analysis; assumption testing",
    "Audited actuals; fund balances",
    "Reserve adequacy over time",
    "Revenue growth vs. expense growth"
  )
)

proportionality_docs |>
  flextable::flextable() |>
  flextable::set_header_labels(
    item_no = "#",
    document = "Document",
    description = "Specific Description",
    years_needed = "Years Needed",
    likely_availability = "Likely Availability",
    analysis_use = "Analysis Use"
  ) |>
  flextable::add_header_lines("Question 2: Proportionality — What Share of the Budget Is This?") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::width(j = "item_no", width = 0.3) |>
  flextable::width(j = "document", width = 1.2) |>
  flextable::width(j = "description", width = 2.0) |>
  flextable::width(j = "years_needed", width = 1.0) |>
  flextable::width(j = "likely_availability", width = 1.0) |>
  flextable::width(j = "analysis_use", width = 1.2) |>
  flextable::valign(valign = "top", part = "body") |>
  flextable::autofit()
```

---

# Question 3: Reversibility

**Purpose:** Determine what percentage of future budgets is already committed by contracts, pensions, debt service, and policy. Assess fiscal flexibility.

```{r}
#| label: tbl-reversibility
#| tbl-cap: "Documents Required for Reversibility Analysis"

# =============================================================================
# QUESTION 3: REVERSIBILITY - HOW LOCKED IN ARE WE?
# =============================================================================

reversibility_docs <- dplyr::tibble(
  item_no = 15:24,
  document = c(
    "Collective Bargaining Agreements / MOUs",
    "Employment Contracts",
    "TMRS Actuarial Valuations",
    "OPEB Actuarial Valuations",
    "Debt Service Schedules",
    "Multi-Year Contracts",
    "Lease Agreements",
    "Pension Contribution Rate History",
    "Health Insurance Premium History",
    "Step Increase Cost Projections"
  ),
  description = c(
    "All current and recently expired agreements with police associations, fire associations, or other employee groups",
    "Individual employment contracts for City Manager, department heads, or other contracted executive positions",
    "Texas Municipal Retirement System actuarial reports showing contribution rates, unfunded liability, and funding ratio",
    "Other Post-Employment Benefits (retiree health insurance) actuarial reports showing liability and funding status",
    "Amortization schedules for all outstanding bonds, certificates of obligation, and other debt showing annual P&I by year",
    "List of all contracts with terms exceeding one year, showing vendor, annual cost, and expiration date",
    "All lease agreements for equipment, vehicles, or facilities with terms exceeding one year",
    "Annual employer contribution rates to TMRS",
    "Annual employer cost per employee for health insurance premiums by plan type",
    "Any documents projecting the cost of automatic step increases for upcoming fiscal years"
  ),
  years_needed = c(
    "Current + prior 2 agreements",
    "All currently in effect",
    "FY2015–FY2024",
    "FY2015–FY2024",
    "Current schedules through maturity",
    "Currently in effect",
    "Currently in effect",
    "FY2015–FY2024",
    "FY2015–FY2024",
    "Most recent available"
  ),
  likely_availability = c(
    "Should be public",
    "Usually public in TX",
    "Annual reports available",
    "Less common; may need request",
    "In CAFR or bond documents",
    "May require compilation",
    "May require compilation",
    "TMRS provides",
    "HR/Finance records",
    "May not exist"
  ),
  analysis_use = c(
    "Identify contractual salary obligations",
    "Executive compensation commitments",
    "Future pension cost trajectory",
    "Future retiree benefit obligations",
    "Fixed debt payments by year",
    "Committed expenditures",
    "Fixed lease obligations",
    "Pension cost trend analysis",
    "Benefit cost trend analysis",
    "Automatic future personnel costs"
  )
)

reversibility_docs |>
  flextable::flextable() |>
  flextable::set_header_labels(
    item_no = "#",
    document = "Document",
    description = "Specific Description",
    years_needed = "Years Needed",
    likely_availability = "Likely Availability",
    analysis_use = "Analysis Use"
  ) |>
  flextable::add_header_lines("Question 3: Reversibility — How Locked In Are We?") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::width(j = "item_no", width = 0.3) |>
  flextable::width(j = "document", width = 1.2) |>
  flextable::width(j = "description", width = 2.0) |>
  flextable::width(j = "years_needed", width = 1.0) |>
  flextable::width(j = "likely_availability", width = 1.0) |>
  flextable::width(j = "analysis_use", width = 1.2) |>
  flextable::valign(valign = "top", part = "body") |>
  flextable::autofit()
```

---

# Question 4: Assumption Testing

**Purpose:** Identify the assumptions embedded in budget projections and test whether past assumptions proved accurate. Expose hidden risks and incompatible assumptions.

```{r}
#| label: tbl-assumptions
#| tbl-cap: "Documents Required for Assumption Testing Analysis"

# =============================================================================
# QUESTION 4: ASSUMPTION Testing - WHAT HAS TO BE TRUE?
# =============================================================================

assumption_docs <- dplyr::tibble(
  item_no = 25:34,
  document = c(
    "Revenue Projections/Forecasts",
    "Sales Tax Projections",
    "Property Tax Calculations",
    "Budget Preparation Instructions/Guidelines",
    "Five-Year Financial Forecasts",
    "Variance Analysis Reports",
    "Quarterly Financial Reports",
    "Economic Development Projections",
    "Actuarial Assumption Documentation",
    "Health Insurance Renewal Documents"
  ),
  description = c(
    "Documents prepared during budget development projecting sales tax, property tax, and other revenues for upcoming fiscal years",
    "Specific forecasts of sales tax revenue used in budget preparation, including methodology and data sources",
    "Certified tax rolls, effective tax rate calculations, rollback rate calculations, and property value growth assumptions",
    "Internal memos or guidelines issued to departments for budget preparation, including inflation assumptions, salary increase assumptions, and growth projections",
    "Any long-range financial projections or forecasts prepared by staff or consultants",
    "Staff reports or memos explaining significant differences between budgeted and actual revenues or expenditures",
    "Interim financial reports presented to council showing year-to-date performance vs. budget",
    "Any projections of new development, population growth, or commercial growth used in financial planning",
    "Documents showing assumed rates of return, mortality tables, and other assumptions used in pension/OPEB valuations",
    "Annual renewal documents showing projected premium increases and plan design changes"
  ),
  years_needed = c(
    "FY2018–FY2025",
    "FY2018–FY2025",
    "FY2018–FY2025",
    "FY2020–FY2025",
    "Most recent 3 versions",
    "FY2018–FY2024",
    "FY2020–FY2024",
    "FY2018–FY2025",
    "Most recent valuations",
    "FY2020–FY2025"
  ),
  likely_availability = c(
    "Internal working papers",
    "May be in budget message",
    "Tax office records",
    "Internal documents",
    "May not exist",
    "May require request",
    "Council packet materials",
    "Economic development dept",
    "In actuarial reports",
    "HR/Finance records"
  ),
  analysis_use = c(
    "Compare projected vs. actual revenue",
    "Test sales tax assumptions",
    "Test property value assumptions",
    "Identify embedded assumptions",
    "Compare forecasts to outcomes",
    "Understand past forecast accuracy",
    "Track in-year performance",
    "Test growth assumptions",
    "Assess actuarial risk exposure",
    "Test benefit cost assumptions"
  )
)

assumption_docs |>
  flextable::flextable() |>
  flextable::set_header_labels(
    item_no = "#",
    document = "Document",
    description = "Specific Description",
    years_needed = "Years Needed",
    likely_availability = "Likely Availability",
    analysis_use = "Analysis Use"
  ) |>
  flextable::add_header_lines("Question 4: Assumption Testing — What Has to Be True?") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::width(j = "item_no", width = 0.3) |>
  flextable::width(j = "document", width = 1.2) |>
  flextable::width(j = "description", width = 2.0) |>
  flextable::width(j = "years_needed", width = 1.0) |>
  flextable::width(j = "likely_availability", width = 1.0) |>
  flextable::width(j = "analysis_use", width = 1.2) |>
  flextable::valign(valign = "top", part = "body") |>
  flextable::autofit()
```

---

# Question 5: Opportunity Costs

**Purpose:** Identify deferred projects, unfunded needs, and alternative uses of funds. Quantify trade-offs in concrete terms.

```{r}
#| label: tbl-opportunity
#| tbl-cap: "Documents Required for Opportunity Cost Analysis"

# =============================================================================
# QUESTION 5: OPPORTUNITY COSTS - WHAT ELSE COULD THIS MONEY BUY?
# =============================================================================

opportunity_docs <- dplyr::tibble(
  item_no = 35:44,
  document = c(
    "Capital Improvement Plans (CIP)",
    "Unfunded Project Lists",
    "Deferred Maintenance Assessments",
    "Equipment Replacement Schedules",
    "Reserve Fund Policies",
    "Strategic Plans",
    "Citizen Survey Results",
    "Department Budget Requests vs. Adopted",
    "Council Budget Workshop Materials",
    "Grant Applications (Unfunded)"
  ),
  description = c(
    "Multi-year capital project plans showing project descriptions, costs, funding sources, and timing",
    "Any lists of projects identified as needs but not included in adopted budgets or CIP due to funding constraints",
    "Assessments or reports documenting maintenance backlogs for streets, facilities, utilities, parks, or equipment",
    "Schedules showing planned replacement of vehicles, equipment, or infrastructure and associated costs",
    "Written policies governing minimum fund balance levels, emergency reserves, stabilization funds, or rainy day funds",
    "Adopted strategic plans identifying council priorities, goals, or performance measures",
    "Results of any citizen satisfaction surveys identifying service priorities, concerns, or unmet needs",
    "Original department budget requests as submitted compared to final adopted amounts (showing what was cut)",
    "Presentation materials from budget workshops showing options, alternatives, or trade-offs presented to council",
    "Applications for grants that were submitted but not awarded, showing unfunded project needs"
  ),
  years_needed = c(
    "FY2018–FY2025",
    "FY2020–FY2025",
    "Most recent available",
    "Current schedule",
    "Current policy",
    "Most recent version",
    "FY2018–FY2024",
    "FY2020–FY2025",
    "FY2020–FY2025",
    "FY2020–FY2024"
  ),
  likely_availability = c(
    "Usually public/online",
    "Internal working papers",
    "Public works/facilities dept",
    "Fleet/facilities records",
    "Finance policy manual",
    "Usually public/online",
    "City manager/comm. dept",
    "May not be maintained",
    "Council packet materials",
    "Grant administration records"
  ),
  analysis_use = c(
    "Identify planned vs. deferred projects",
    "Quantify unfunded needs",
    "Calculate infrastructure gap",
    "Assess equipment condition/risk",
    "Evaluate reserve adequacy",
    "Compare spending to stated priorities",
    "Identify citizen-perceived gaps",
    "Quantify budget reductions by area",
    "Understand alternatives considered",
    "Identify unfunded opportunities"
  )
)

opportunity_docs |>
  flextable::flextable() |>
  flextable::set_header_labels(
    item_no = "#",
    document = "Document",
    description = "Specific Description",
    years_needed = "Years Needed",
    likely_availability = "Likely Availability",
    analysis_use = "Analysis Use"
  ) |>
  flextable::add_header_lines("Question 5: Opportunity Cost — What Else Could This Money Buy?") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::width(j = "item_no", width = 0.3) |>
  flextable::width(j = "document", width = 1.2) |>
  flextable::width(j = "description", width = 2.0) |>
  flextable::width(j = "years_needed", width = 1.0) |>
  flextable::width(j = "likely_availability", width = 1.0) |>
  flextable::width(j = "analysis_use", width = 1.2) |>
  flextable::valign(valign = "top", part = "body") |>
  flextable::autofit()
```

---

# Master Document Checklist

```{r}
#| label: tbl-master
#| tbl-cap: "Complete Document Request Checklist"

# =============================================================================
# MASTER CHECKLIST - ALL DOCUMENTS
# =============================================================================

master_checklist <- dplyr::bind_rows(
  trajectory_docs |> dplyr::mutate(framework_question = "1. Trajectory"),
  proportionality_docs |> dplyr::mutate(framework_question = "2. Proportionality"),
  reversibility_docs |> dplyr::mutate(framework_question = "3. Reversibility"),
  assumption_docs |> dplyr::mutate(framework_question = "4. Assumptions"),
  opportunity_docs |> dplyr::mutate(framework_question = "5. Opportunity Cost")
) |>
  dplyr::select(
    item_no,
    framework_question,
    document,
    years_needed,
    likely_availability
  ) |>
  dplyr::mutate(
    received = "",
    notes = ""
  )

master_checklist |>
  flextable::flextable() |>
  flextable::set_header_labels(
    item_no = "#",
    framework_question = "Framework Question",
    document = "Document",
    years_needed = "Years Needed",
    likely_availability = "Likely Availability",
    received = "Received?",
    notes = "Notes"
  ) |>
  flextable::add_header_lines("Master Document Request Checklist") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::width(j = "item_no", width = 0.3) |>
  flextable::width(j = "framework_question", width = 1.0) |>
  flextable::width(j = "document", width = 1.5) |>
  flextable::width(j = "years_needed", width = 1.0) |>
  flextable::width(j = "likely_availability", width = 1.0) |>
  flextable::width(j = "received", width = 0.6) |>
  flextable::width(j = "notes", width = 1.0) |>
  flextable::valign(valign = "top", part = "body") |>
  flextable::autofit()
```

---
 
# Request Strategy Notes

## Phased Approach

Consider submitting requests in phases to manage response time and costs:
 
| Phase | Priority | Focus | Documents |
|-------|----------|-------|-----------|
| 1 | High | Core Financial | Budgets, CAFRs, Compensation Reports (#1, 8-14) |
| 2 | High | Commitments | Contracts, actuarial reports, debt schedules (#15-23) |
| 3 | Medium | Assumptions | Forecasts, variance reports, internal memos (#24-34) |
| 4 | Lower | Opportunity Cost | CIP, deferred maintenance, workshop materials (#35-44) |

## Documents Likely Available Online

Many Texas cities post these documents publicly:

- Adopted annual budgets
- Comprehensive Annual Financial Reports (CAFR/ACFR)
- Capital Improvement Plans
- Strategic plans
- Pay plans / salary schedules
- Council meeting agendas and minutes

**Always check the city website first** before submitting a formal TPIA request.

## Documents Requiring Formal Request

These typically require a written request:

- Individual employee compensation data
- Internal budget preparation memos
- Department budget requests (pre-adoption)
- Variance analysis reports
- Multi-year contract inventories
- Deferred maintenance assessments

## Potential Challenges

| Document Type | Potential Issue | Response Strategy |
|--------------|-----------------|-------------------|
| Internal memos | May claim "working papers" exception | Request final versions only; narrow scope |
| Pending negotiations | May claim deliberative process | Request completed/expired agreements only |
| Individual salaries | Rarely contested in TX | Cite Tex. Gov't Code § 552.022 |
| Cost estimates | High compilation costs | Request phased delivery; negotiate scope |

---

# Version History

| Version | Date | Changes |
|---------|------|---------|
| 0.10 | `r Sys.Date()` | Initial document |
