---
title: "Schertz Total Compensation - Directors, Managers, and Public Safety Leaders"
subtitle: "FY20 - FY25"
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    echo: true
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true     # Set to 'false' for drafts. Override if you're NOT sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    echo: false
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    error: false
    warning: false
    message: false
    cache: false


params:
  data_updated: !expr format(Sys.Date(), "%B %Y")
---

```{r}
#| label: setup
#| include: false

# Capture current filename for use in document
file_name <- if (!base::is.null(knitr::current_input())) {
  paste0(tools::file_path_sans_ext(knitr::current_input()), ".qmd")
} else {
  "interactive_session.qmd"
}

file_stem <- tools::file_path_sans_ext(file_name)



# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED

# Download button helpers live here; knitr::is_html_output() guard is inside
# the helpers so they silently no-op when rendering to typst

source(here::here("R", "download_button_helpers.R"))
source(here::here("R", "report_format_helpers.R"))
source(here::here("R", "flextable_helpers.R"))
source(here::here("R", "download_button_helpers.R"))

# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  theme_fun = base::identity,          # <-- add this
  post_process_html = identity,
  post_process_docx = identity
)

# Standard table styler: blue italic title row + palegreen column header row.
# Defined here so all QMDs that source this setup get the same presentation
# without depending on census_helpers.R for a purely cosmetic function.
ds_style_flextable <- function(ft, title = NULL, title_size = 16) {
  # Fixed layout helps long text wrap instead of expanding columns
  ft <- flextable::set_table_properties(ft, layout = "fixed")

  if (!base::is.null(title)) {
    ft <- flextable::add_header_lines(ft, values = title)
    ft <- flextable::color(ft, i = 1, part = "header", color = "blue")
    ft <- flextable::italic(ft, i = 1, part = "header")
    ft <- flextable::align(ft, i = 1, part = "header", align = "left")
    ft <- flextable::bold(ft, i = 1, part = "header")
    ft <- flextable::fontsize(ft, i = 1, part = "header", size = title_size)
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "white")
    ft <- flextable::bg(ft, i = 2, part = "header", bg = "palegreen")

    # Optional: make wrapped titles read better
    ft <- flextable::line_spacing(ft, i = 1, part = "header", space = 1.2)
  } else {
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "palegreen")
  }

  ft <- flextable::autofit(ft)
  ft
}



# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 28,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 22,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 18,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 18,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)


# claude <- ellmer::chat_anthropic(model = "claude-opus-4-6-20260205")
  
# Set seed for reproducibility
base::set.seed(123)

```


\

#### **Filename:**

::: {.content-visible when-format="html"}
[`r file_name`]{style="word-break: break-all; font-size: 1.00em;"}
:::

```{=typst}
#let wrap_underscores(s) = s.replace("_", "_\u{200B}")
#text(size: 16pt)[#raw(wrap_underscores("`r file_name`"))]
```

\

\


## Introduction
This document presents **total compensation** for the City of Schertz's top-paid employees across fiscal years FY20 through FY25. Top employees include department directors, managers, and public safety leadership (Police, Fire, EMS).

::: {.callout-note}
## Interactive Tables
**Data last updated:** `r params$data_updated`
The tables below are fully interactive:

- **Sort**: Click any column header to sort
- **Search**: Use the search box to find specific employees
- **Filter**: Use column filters to narrow results
- **Export**: Use the Copy, CSV, or Excel buttons to export data
- **Print**: Use your browser's print function (Ctrl+P / Cmd+P)
:::

---

## Step 1: Load Cached Data

```{r}
#| label: load-cached-data
#| cache: true

# =============================================================================
# LOAD CACHED DATA FROM OMNIBUS ANALYSIS
# =============================================================================

data_path <- here::here("data_clean", "compensation_wide_fy20_fy25.rds")

if (base::file.exists(data_path)) {
  all_data <- base::readRDS(data_path)

  data_load_df <- data.frame(
    Item = c("Data file", "Rows loaded", "Fiscal years found"),
    Value = c(
      data_path,
      base::format(base::nrow(all_data), big.mark = ","),
      base::paste(base::sort(base::unique(all_data$fiscal_year)), collapse = ", ")
    ),
    stringsAsFactors = FALSE
  )

  data_load_spec <- list(
    title = "Loaded Compensation Data Summary",
    labels = list(Item = "Item", Value = "Value")
  )

  ds_ft(data_load_df, data_load_spec)
} else {
  base::stop(
    "Data file not found at: ", data_path, "
",
    "Please run the omnibus analysis first to generate the cleaned data,
",
    "or update the data_path variable to point to your RDS file."
  )
}
```

---

## Step 2: Define Top Employee Filter Criteria
```{r}
#| label: define-top-employee-criteria
#| cache: true

# =============================================================================
# DEFINE CRITERIA FOR TOP EMPLOYEES
# =============================================================================

leadership_pattern <- base::paste(
  "city manager",
  "assistant city manager",
  "deputy city manager",
  "director",
  "assistant director",
  "deputy director",
  "manager",
  "superintendent",
  "chief",
  "commander",
  "captain",
  "battalion chief",
  "marshal",
  sep = "|"
)

leadership_criteria_df <- data.frame(
  Category = c(
    "City executive leadership",
    "Department directors",
    "Department managers and supervisors",
    "Public safety leadership",
    "Regex pattern used"
  ),
  Detail = c(
    "City Manager, Assistant City Manager, Deputy City Manager",
    "Director, Assistant Director, Deputy Director",
    "Manager, Superintendent",
    "Chief, Commander, Captain, Battalion Chief, Marshal",
    leadership_pattern
  ),
  stringsAsFactors = FALSE
)

leadership_criteria_spec <- list(
  title = "Leadership Position Filter Criteria",
  labels = list(Category = "Category", Detail = "Detail")
)

ds_ft(leadership_criteria_df, leadership_criteria_spec)
```

---

## Step 3: Filter Top Employees
```{r}
#| label: filter-top-employees
#| cache: true

# =============================================================================
# FILTER DATA TO TOP EMPLOYEES ONLY
# =============================================================================

top_employees <- all_data |>
  dplyr::filter(
    stringr::str_detect(
      job_title,
      stringr::regex(leadership_pattern, ignore_case = TRUE)
    )
  ) |>
  dplyr::select(
    fiscal_year,
    name,
    name_std,
    department,
    job_title,
    regular_earnings,
    additional_benefits,
    calc_total_compensation
  ) |>
  dplyr::arrange(fiscal_year, dplyr::desc(calc_total_compensation))

filter_results_df <- data.frame(
  Item = c("Total employees in dataset", "Top employees identified", "Unique individuals"),
  Value = c(
    base::format(base::nrow(all_data), big.mark = ","),
    base::format(base::nrow(top_employees), big.mark = ","),
    base::format(dplyr::n_distinct(top_employees$name_std), big.mark = ",")
  ),
  stringsAsFactors = FALSE
)

filter_results_spec <- list(
  title = "Top Employees Filter Results",
  labels = list(Item = "Item", Value = "Value"),
  align_right = "Value"
)

ds_ft(filter_results_df, filter_results_spec)
```

---

## Step 4: Verify Captured Job Titles
```{r}
#| label: verify-job-titles
#| cache: true

# =============================================================================
# VERIFY WHICH JOB TITLES WERE CAPTURED
# =============================================================================

captured_titles <- top_employees |>
  dplyr::distinct(job_title) |>
  dplyr::arrange(job_title)

captured_titles_spec <- list(
  title = "Job Titles Captured for Leadership Filtering",
  labels = list(job_title = "Job Title")
)

ds_ft(captured_titles, captured_titles_spec)

ds_download_data_btn(
  captured_titles,
  filename_base = "schertz_top_employee_job_titles",
  label = "Download Captured Job Titles (CSV)"
)
```

::: {.content-visible when-format="html"}
:::

---

## Step 5: Create Wide Table - Total Compensation by FY
```{r}
#| label: total-comp-wide-table
#| cache: true

# =============================================================================
# CREATE WIDE FORMAT TABLE: TOTAL COMPENSATION BY FISCAL YEAR
# =============================================================================

most_recent_info <- top_employees |>
  dplyr::group_by(name_std) |>
  dplyr::slice_max(order_by = fiscal_year, n = 1) |>
  dplyr::ungroup() |>
  dplyr::select(name_std, name, department, job_title)

total_comp_wide <- top_employees |>
  dplyr::select(name_std, fiscal_year, calc_total_compensation) |>
  tidyr::pivot_wider(
    names_from = fiscal_year,
    values_from = calc_total_compensation,
    values_fill = NA_real_
  ) |>
  dplyr::left_join(most_recent_info, by = "name_std") |>
  dplyr::select(
    name,
    department,
    job_title,
    dplyr::any_of(c("FY20", "FY21", "FY22", "FY23", "FY24", "FY25"))
  ) |>
  dplyr::arrange(
    dplyr::desc(
      dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20)
    )
  )

wide_table_summary_df <- data.frame(
  Item = c("Rows (unique employees)", "Columns"),
  Value = c(
    base::format(base::nrow(total_comp_wide), big.mark = ","),
    base::format(base::ncol(total_comp_wide), big.mark = ",")
  ),
  stringsAsFactors = FALSE
)

wide_table_summary_spec <- list(
  title = "Wide Table Dimensions",
  labels = list(Item = "Item", Value = "Value"),
  align_right = "Value"
)

ds_ft(wide_table_summary_df, wide_table_summary_spec)
```

---

## Table 1: Top Employee Total Compensation by Fiscal Year

::: {.content-visible when-format="html"}
:::

::: {.content-visible when-format="typst"}
:::

::: {.content-visible when-format="html"}
:::

---

## Object Reference

| Object Name | Description | Rows | Columns |
|-------------|-------------|------|---------|
| `all_data` | Full compensation dataset (cached from omnibus) | `r base::nrow(all_data)` | `r base::ncol(all_data)` |
| `top_employees` | Filtered to leadership positions | `r base::nrow(top_employees)` | `r base::ncol(top_employees)` |
| `total_comp_wide` | Wide format: FY columns for total comp | `r base::nrow(total_comp_wide)` | `r base::ncol(total_comp_wide)` |
| `captured_titles` | Unique job titles matched | `r base::nrow(captured_titles)` | 1 |

---

## Data Notes

- **Total Compensation** = Regular Earnings + Overtime + Additional Earnings + Leave Payout + Deployment + Arbitration + Additional Benefits
- **FY20-FY22**: Benefits were imputed based on FY23-FY25 ratios (benefits data not available in source files)
- **Blank cells**: Employee not in this position during that fiscal year

---

*Report generated: `r base::format(base::Sys.time(), "%B %d, %Y at %I:%M %p")`*
*Data source: City of Schertz Employee Compensation Records*


```{r}
#| label: display-total-comp-flextable
#| cache: false

# =============================================================================
# DISPLAY TOTAL COMPENSATION TABLE
# =============================================================================

total_comp_display <- total_comp_wide |>
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("FY"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  )

total_comp_spec <- list(
  title = "Top Employee Total Compensation by Fiscal Year",
  footer_lines = "Source: City of Schertz Employee Compensation Data | Blank cells indicate employee not in position that year",
  footer_fontsize = 10,
  footer_italic = TRUE
)

ds_ft(total_comp_display, total_comp_spec)

ds_download_data_btn(
  total_comp_wide,
  filename_base = "schertz_top_employees_total_comp_fy20_fy25",
  label = "Download Raw Numeric Data (CSV)"
)
```
