---
title: "City of Schertz Year-over-Year Dollar and Pct Change for Directors, Managers, and Public Safety Leaders"
subtitle: "FY20 - FY25"
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    echo: true
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true     # Set to 'false' for drafts. Override if you're NOT sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    echo: false
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    error: false
    warning: false
    message: false
    cache: false


params:
  data_updated: !expr format(Sys.Date(), "%B %Y")
---

```{r}
#| label: setup
#| include: false

# Capture current filename for use in document
file_name <- if (!base::is.null(knitr::current_input())) {
  paste0(tools::file_path_sans_ext(knitr::current_input()), ".qmd")
} else {
  "interactive_session.qmd"
}

file_stem <- base::paste0(tools::file_path_sans_ext(file_name), "_chat")



# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED

# Download button helpers live here; knitr::is_html_output() guard is inside
# the helpers so they silently no-op when rendering to typst

source(here::here("R", "report_format_helpers.R"))
source(here::here("R", "flextable_helpers.R"))
source(here::here("R", "download_button_helpers.R"))

# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  theme_fun = base::identity,          # <-- add this
  post_process_html = identity,
  post_process_docx = identity
)

# Standard table styler: blue italic title row + palegreen column header row.
# Defined here so all QMDs that source this setup get the same presentation
# without depending on census_helpers.R for a purely cosmetic function.
ds_style_flextable <- function(ft, title = NULL, title_size = 16) {
  # Fixed layout helps long text wrap instead of expanding columns
  ft <- flextable::set_table_properties(ft, layout = "fixed")

  if (!base::is.null(title)) {
    ft <- flextable::add_header_lines(ft, values = title)
    ft <- flextable::color(ft, i = 1, part = "header", color = "blue")
    ft <- flextable::italic(ft, i = 1, part = "header")
    ft <- flextable::align(ft, i = 1, part = "header", align = "left")
    ft <- flextable::bold(ft, i = 1, part = "header")
    ft <- flextable::fontsize(ft, i = 1, part = "header", size = title_size)
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "white")
    ft <- flextable::bg(ft, i = 2, part = "header", bg = "palegreen")

    # Optional: make wrapped titles read better
    ft <- flextable::line_spacing(ft, i = 1, part = "header", space = 1.2)
  } else {
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "palegreen")
  }

  ft <- flextable::autofit(ft)
  ft
}



# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 28,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 22,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 18,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 18,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)


# claude <- ellmer::chat_anthropic(model = "claude-opus-4-6-20260205")
  
# Set seed for reproducibility
base::set.seed(123)

```

\
\

#### **Filename:**

::: {.content-visible when-format="html"}
[`r file_name`]{style="word-break: break-all; font-size: 1.00em;"}
:::

```{=typst}
#let wrap_underscores(s) = s.replace("_", "_\u{200B}")
#text(size: 16pt)[#raw(wrap_underscores("`r file_name`"))]
```

\
\

## Introduction
This document presents **year-over-year compensation changes** (both dollar amounts and percentages) for the City of Schertz's top-paid employees. This analysis reveals the trajectory of pay increases for leadership positions from FY20 through FY25.

::: {.callout-note}
## Interactive Tables
**Data last updated:** `r params$data_updated`

- **Sort**: Click any column header to sort (e.g., find largest increases)
- **Search**: Use the search box to find specific employees
- **Filter**: Use column filters to narrow results
- **Export**: Use the Copy, CSV, Excel, or Print buttons
:::

---

## Step 1: Load Cached Data
```{r}
#| label: load-cached-data
#| cache: true

# =============================================================================
# LOAD CACHED DATA FROM OMNIBUS ANALYSIS
# =============================================================================

data_path <- here::here("data_clean", "compensation_wide_fy20_fy25.rds")

if (base::file.exists(data_path)) {
  all_data <- base::readRDS(data_path)

  data_load_df <- data.frame(
    Item = c("Data file", "Rows loaded", "Fiscal years found"),
    Value = c(
      data_path,
      base::format(base::nrow(all_data), big.mark = ","),
      base::paste(base::sort(base::unique(all_data$fiscal_year)), collapse = ", ")
    ),
    stringsAsFactors = FALSE
  )

  data_load_spec <- list(
    title = "Loaded Compensation Data Summary",
    labels = list(Item = "Item", Value = "Value")
  )

  ds_ft(data_load_df, data_load_spec)
} else {
  base::stop(
    "Data file not found at: ", data_path, "
",
    "Please run the omnibus analysis first to generate the cleaned data."
  )
}
```

---

## Step 2: Filter to Top Employees
```{r}
#| label: filter-top-employees
#| cache: true

# =============================================================================
# FILTER DATA TO TOP EMPLOYEES ONLY
# =============================================================================

leadership_pattern <- base::paste(
  "city manager", "assistant city manager", "deputy city manager",
  "director", "assistant director", "deputy director",
  "manager", "superintendent",
  "chief", "commander", "captain", "battalion chief", "marshal",
  sep = "|"
)

top_employees <- all_data |>
  dplyr::filter(
    stringr::str_detect(
      job_title,
      stringr::regex(leadership_pattern, ignore_case = TRUE)
    )
  ) |>
  dplyr::select(
    fiscal_year, name, name_std, department, job_title,
    calc_total_compensation
  ) |>
  dplyr::arrange(fiscal_year, dplyr::desc(calc_total_compensation))

filter_results_df <- data.frame(
  Item = c("Top employees identified", "Unique individuals"),
  Value = c(
    base::format(base::nrow(top_employees), big.mark = ","),
    base::format(dplyr::n_distinct(top_employees$name_std), big.mark = ",")
  ),
  stringsAsFactors = FALSE
)

filter_results_spec <- list(
  title = "Top Employees Filter Results",
  labels = list(Item = "Item", Value = "Value"),
  align_right = "Value"
)

ds_ft(filter_results_df, filter_results_spec)
```

---

## Step 3: Calculate Year-over-Year Changes
```{r}
#| label: calculate-yoy-changes
#| cache: true

# =============================================================================
# CALCULATE YEAR-OVER-YEAR CHANGES ($ AND %)
# =============================================================================

yoy_changes <- top_employees |>
  dplyr::arrange(name_std, fiscal_year) |>
  dplyr::group_by(name_std) |>
  dplyr::mutate(
    prior_total_comp = dplyr::lag(calc_total_compensation),
    change_dollars = calc_total_compensation - prior_total_comp,
    change_pct = dplyr::if_else(
      prior_total_comp > 0,
      (calc_total_compensation - prior_total_comp) / prior_total_comp * 100,
      NA_real_
    ),
    fy_transition = dplyr::if_else(
      !base::is.na(prior_total_comp),
      base::paste0(dplyr::lag(fiscal_year), "→", fiscal_year),
      NA_character_
    )
  ) |>
  dplyr::ungroup()

calc_results_df <- data.frame(
  Item = c("Total records", "Records with changes calculated"),
  Value = c(
    base::format(base::nrow(yoy_changes), big.mark = ","),
    base::format(base::sum(!base::is.na(yoy_changes$change_dollars)), big.mark = ",")
  ),
  stringsAsFactors = FALSE
)

calc_results_spec <- list(
  title = "Year-over-Year Change Calculations",
  labels = list(Item = "Item", Value = "Value"),
  align_right = "Value"
)

ds_ft(calc_results_df, calc_results_spec)
```

---

## Step 4: Create Wide Format - Dollar Changes
```{r}
#| label: create-dollar-changes-wide
#| cache: true

# =============================================================================
# CREATE WIDE FORMAT: DOLLAR CHANGES BY FISCAL YEAR TRANSITION
# =============================================================================

most_recent_info <- top_employees |>
  dplyr::group_by(name_std) |>
  dplyr::slice_max(order_by = fiscal_year, n = 1) |>
  dplyr::ungroup() |>
  dplyr::select(name_std, name, department, job_title)

dollar_changes_wide <- yoy_changes |>
  dplyr::filter(!base::is.na(fy_transition)) |>
  dplyr::select(name_std, fy_transition, change_dollars) |>
  tidyr::pivot_wider(
    names_from = fy_transition,
    values_from = change_dollars,
    values_fill = NA_real_
  ) |>
  dplyr::left_join(most_recent_info, by = "name_std") |>
  dplyr::select(
    name, department, job_title,
    dplyr::any_of(c("FY20→FY21", "FY21→FY22", "FY22→FY23", "FY23→FY24", "FY24→FY25"))
  ) |>
  dplyr::arrange(
    dplyr::desc(
      dplyr::coalesce(`FY24→FY25`, `FY23→FY24`, `FY22→FY23`, `FY21→FY22`, `FY20→FY21`)
    )
  )

dollar_changes_df <- data.frame(
  Item = "Dollar changes table employees",
  Value = base::format(base::nrow(dollar_changes_wide), big.mark = ","),
  stringsAsFactors = FALSE
)

dollar_changes_meta_spec <- list(
  title = "Dollar Changes Table Summary",
  labels = list(Item = "Item", Value = "Value"),
  align_right = "Value"
)

ds_ft(dollar_changes_df, dollar_changes_meta_spec)
```

---

## Step 5: Create Wide Format - Percent Changes
```{r}
#| label: create-pct-changes-wide
#| cache: true

# =============================================================================
# CREATE WIDE FORMAT: PERCENT CHANGES BY FISCAL YEAR TRANSITION
# =============================================================================

pct_changes_wide <- yoy_changes |>
  dplyr::filter(!base::is.na(fy_transition)) |>
  dplyr::select(name_std, fy_transition, change_pct) |>
  tidyr::pivot_wider(
    names_from = fy_transition,
    values_from = change_pct,
    values_fill = NA_real_
  ) |>
  dplyr::left_join(most_recent_info, by = "name_std") |>
  dplyr::select(
    name, department, job_title,
    dplyr::any_of(c("FY20→FY21", "FY21→FY22", "FY22→FY23", "FY23→FY24", "FY24→FY25"))
  ) |>
  dplyr::arrange(
    dplyr::desc(
      dplyr::coalesce(`FY24→FY25`, `FY23→FY24`, `FY22→FY23`, `FY21→FY22`, `FY20→FY21`)
    )
  )

pct_changes_df <- data.frame(
  Item = "Percent changes table employees",
  Value = base::format(base::nrow(pct_changes_wide), big.mark = ","),
  stringsAsFactors = FALSE
)

pct_changes_meta_spec <- list(
  title = "Percent Changes Table Summary",
  labels = list(Item = "Item", Value = "Value"),
  align_right = "Value"
)

ds_ft(pct_changes_df, pct_changes_meta_spec)
```

---

## Table 2A: Year-over-Year Dollar Changes by Management Employee

::: {.content-visible when-format="html"}
```{r}
#| label: display-dollar-changes-dt
#| cache: false

# =============================================================================
# HTML OUTPUT: DOLLAR CHANGES - INTERACTIVE DT::DATATABLE
# =============================================================================

dollar_changes_wide |>
  dplyr::mutate(
    dplyr::across(
      dplyr::contains("→"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  DT::datatable(
    caption = "Year-over-Year Dollar Change in Total Compensation",
    extensions = "Buttons",
    options = base::list(
      pageLength = 25,
      scrollX = TRUE,
      dom = "Bfrtip",
      buttons = base::list(
        "copy", 
        "csv", 
        "excel",
        base::list(
          extend = "print",
          title = "YOY Dollar Changes - City of Schertz Top Employees"
        )
      )
    ),
    rownames = FALSE,
    filter = "top"
  )
```
:::

::: {.content-visible when-format="typst"}
```{r}
#| label: display-dollar-changes-flextable
#| cache: false

# =============================================================================
# PDF/TYPST OUTPUT: DOLLAR CHANGES - FLEXTABLE
# =============================================================================

dollar_changes_display <- dollar_changes_wide |>
  dplyr::mutate(
    dplyr::across(
      dplyr::contains("→"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  )

dollar_changes_spec <- list(
  title = "Year-over-Year Dollar Change in Total Compensation",
  footer_lines = "Positive = increase | Negative = decrease | Blank = not in position",
  footer_fontsize = 10,
  footer_italic = TRUE
)

ds_ft(dollar_changes_display, dollar_changes_spec)

ds_download_data_btn(
  dollar_changes_wide,
  filename_base = base::paste0(file_stem, "_schertz_yoy_dollar_changes"),
  label = "Download Table CSV"
)
```
:::

---

## Table 2B: Year-over-Year Percent Change by Management Employee

::: {.content-visible when-format="html"}
```{r}
#| label: display-pct-changes-dt
#| cache: false

# =============================================================================
# HTML OUTPUT: PERCENT CHANGES - INTERACTIVE DT::DATATABLE
# =============================================================================

pct_changes_wide |>
  dplyr::mutate(
    dplyr::across(
      dplyr::contains("→"),
      ~ base::sprintf("%.1f%%", .x)
    )
  ) |>
  DT::datatable(
    caption = "Year-over-Year Percent Change in Total Compensation",
    extensions = "Buttons",
    options = base::list(
      pageLength = 25,
      scrollX = TRUE,
      dom = "Bfrtip",
      buttons = base::list(
        "copy", 
        "csv", 
        "excel",
        base::list(
          extend = "print",
          title = "YOY Percent Changes - City of Schertz Top Employees"
        )
      )
    ),
    rownames = FALSE,
    filter = "top"
  )
```
:::

::: {.content-visible when-format="typst"}
```{r}
#| label: display-pct-changes-flextable
#| cache: false

# =============================================================================
# PDF/TYPST OUTPUT: PERCENT CHANGES - FLEXTABLE
# =============================================================================

pct_changes_display <- pct_changes_wide |>
  dplyr::mutate(
    dplyr::across(
      dplyr::contains("→"),
      ~ base::sprintf("%.1f%%", .x)
    )
  )

pct_changes_spec <- list(
  title = "Year-over-Year Percent Change in Total Compensation by Employee",
  footer_lines = "Percent change = (Current - Prior) / Prior × 100 | Blank = not in position",
  footer_fontsize = 10,
  footer_italic = TRUE
)

ds_ft(pct_changes_display, pct_changes_spec)

ds_download_data_btn(
  pct_changes_wide,
  filename_base = base::paste0(file_stem, "_schertz_yoy_pct_changes"),
  label = "Download Table CSV"
)
```
:::

---

## Summary Statistics
```{r}
#| label: summary-statistics
#| cache: true

# =============================================================================
# SUMMARY STATISTICS FOR YEAR-OVER-YEAR CHANGES
# =============================================================================

change_summary <- yoy_changes |>
  dplyr::filter(!base::is.na(change_dollars)) |>
  dplyr::group_by(fy_transition) |>
  dplyr::summarise(
    n_employees = dplyr::n(),
    avg_dollar_change = base::mean(change_dollars, na.rm = TRUE),
    median_dollar_change = stats::median(change_dollars, na.rm = TRUE),
    min_change = base::min(change_dollars, na.rm = TRUE),
    max_change = base::max(change_dollars, na.rm = TRUE),
    avg_pct_change = base::mean(change_pct, na.rm = TRUE),
    .groups = "drop"
  )

change_summary_display <- change_summary |>
  dplyr::mutate(
    avg_dollar_change = scales::dollar(avg_dollar_change, accuracy = 1),
    median_dollar_change = scales::dollar(median_dollar_change, accuracy = 1),
    min_change = scales::dollar(min_change, accuracy = 1),
    max_change = scales::dollar(max_change, accuracy = 1),
    avg_pct_change = base::sprintf("%.1f%%", avg_pct_change)
  ) |>
  dplyr::rename(
    `FY Transition` = fy_transition,
    `N Employees` = n_employees,
    `Avg $ Change` = avg_dollar_change,
    `Median $ Change` = median_dollar_change,
    `Min Change` = min_change,
    `Max Change` = max_change,
    `Avg % Change` = avg_pct_change
  )

change_summary_spec <- list(
  title = "Summary: Year-over-Year Changes Across All Top Employees"
)

ds_ft(change_summary_display, change_summary_spec)

ds_download_data_btn(
  change_summary,
  filename_base = base::paste0(file_stem, "_schertz_yoy_change_summary"),
  label = "Download Table CSV"
)
```

---

## Object Reference

| Object Name | Description | Rows | Columns |
|-------------|-------------|------|---------|
| `all_data` | Full compensation dataset (cached) | `r base::nrow(all_data)` | `r base::ncol(all_data)` |
| `top_employees` | Filtered to leadership positions | `r base::nrow(top_employees)` | `r base::ncol(top_employees)` |
| `yoy_changes` | With calculated YOY changes | `r base::nrow(yoy_changes)` | `r base::ncol(yoy_changes)` |
| `dollar_changes_wide` | Wide format: $ changes by FY transition | `r base::nrow(dollar_changes_wide)` | `r base::ncol(dollar_changes_wide)` |
| `pct_changes_wide` | Wide format: % changes by FY transition | `r base::nrow(pct_changes_wide)` | `r base::ncol(pct_changes_wide)` |
| `change_summary` | Aggregate statistics by transition | `r base::nrow(change_summary)` | `r base::ncol(change_summary)` |

---

## Data Notes

- **Dollar Change** = Current Year Total Comp − Prior Year Total Comp
- **Percent Change** = (Current − Prior) / Prior × 100
- **Positive values** = compensation increase; **Negative** = decrease
- **Blank cells** = employee not in position during one or both years of transition

---

*Report generated: `r base::format(base::Sys.time(), "%B %d, %Y at %I:%M %p")`*
*Data source: City of Schertz Employee Compensation Records*
