---
title: "City of Schertz — FY 2025-26 Proposed Budget: Compensation Increase Analysis"
subtitle: "Questions for the Board of Trustees / City Council Candidate Review"
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true     # Set to 'false' for drafts. Override if you're NOT sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false

prefer-html: true

params:
  data_updated: !expr format(Sys.Date(), "%B %Y")
---


```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED

# Download button helpers live here; knitr::is_html_output() guard is inside
# the helpers so they silently no-op when rendering to typst
source(here::here("R", "download_button_helpers.R"))

# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  theme_fun = base::identity,          # <-- add this
  post_process_html = identity,
  post_process_docx = identity
)

# Standard table styler: blue italic title row + palegreen column header row.
# Defined here so all QMDs that source this setup get the same presentation
# without depending on census_helpers.R for a purely cosmetic function.
ds_style_flextable <- function(ft, title = NULL) {
  if (!base::is.null(title)) {
    ft <- flextable::add_header_lines(ft, values = title)
    ft <- flextable::color(ft, i = 1, part = "header", color = "blue")
    ft <- flextable::italic(ft, i = 1, part = "header")
    ft <- flextable::align(ft, i = 1, part = "header", align = "left")
    ft <- flextable::fontsize(ft, i = 1, part = "header", size = 12)
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "white")
    ft <- flextable::bg(ft, i = 2, part = "header", bg = "palegreen")
  } else {
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "palegreen")
  }
  ft <- flextable::autofit(ft)
  ft
}


# Flextable built-in themes:
  # flextable::theme_alafoli()	|>  # BLAH
  # flextable::theme_apa()  # THIS IS NICE
  # flextable::theme_booktabs() |>  # NICE, MORE COMPACT
  # flextable::theme_box() |>   # OK, INCLUDES CELL BORDERS
  # flextable::theme_tron() |>  # 'DARK MODE' BLUE TEXT
  # flextable::theme_tron_legacy() |>   # 'DARK MODE' YELLOW TEXT
  # flextable::theme_vader() |>    # 'DARK MODE' WHITE TEXT
  # flextable::theme_vanilla() |>   # NOT SPECIAL
  # flextable::theme_zebra()	|>
  #


# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 28,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 22,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 18,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 18,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)


# claude <- ellmer::chat_anthropic(model = "claude-opus-4-6-20260205")
  
# Set seed for reproducibility
base::set.seed(123)

```

\

#### **Filename:** 
city_of_schertz_base_salary_vs_additional_earnings_fy20_fy25_for_
directors_managers_and_public_safety_leaders_searchable_dt_tables_with_
buttons.qmd

\




```{r}
#| label: helper-functions
#| include: false

# No external packages required for this document — all computations are
# in base R.  Formatting helpers are inline below.

fmt_dollar <- function(x, digits = 0) {
  paste0("$", formatC(x, format = "f", digits = digits, big.mark = ","))
}

fmt_pct <- function(x, digits = 1) {
  paste0(formatC(x * 100, format = "f", digits = digits), "%")
}
```

---

## Background and Source

This analysis is drawn from a careful reading of the **City of Schertz FY 2025-26
Proposed Budget** document.  All figures come directly from the budget narratives,
department summary tables, and the detailed account-level budget worksheets
(ExpCategory 511 — Personnel Services totals).  No automated parsing routines
were used; every number below was read from the document.

The budget presents expenditure data across four columns most relevant to this
analysis:

| Column | Description |
|---|---|
| 2024-25 Budget | Original adopted budget for the current year |
| 2024-25 YTD Activity | Actual spending through 8/31/2025 |
| **2024-25 Projected** | Staff's year-end estimate — the baseline for comparison |
| **2025-26 Budget (AB)** | Proposed budget — the subject of this analysis |

The comparison used throughout is **2024-25 Projected vs. 2025-26 Proposed
Budget**, which is the standard "apples-to-apples" measure of the incremental
cost of the new budget.

---

## Question 1: What is the Total Increase in Compensation Requested?

### Explicit Statement from the Budget Document

The budget message contains one explicit, composite dollar figure for the
compensation policy changes:

> *"Citywide the budget funds the Merit program for non-uniformed public safety
> personnel and the Step program for uniformed public safety. These programs move
> employees within their pay range but does not affect the starting wages of new
> employees. Non-uniformed employees will be eligible for a 0–4% increase based
> on performance and uniformed public safety employees could receive a 4% increase
> based on the tenure schedule. **These changes are planned to take effect
> midyear and cost \$784,000.**"*

This **\$784,000** is the mid-year (six-month) cost of merit and step increases
for **existing employees, across all funds**.

The budget also states that the medical insurance plan was changed to a new
provider in a way that **prevents any increase in premiums for 2026**, so the
benefits cost impact for health insurance is essentially flat in the proposed
budget.

### New Positions Added (FY 2025-26)

Beyond the merit/step program, 26 new positions were authorized:

| Fund | Positions | Detail |
|---|---|---|
| General Fund | 10 | 6 Firefighters (Fire Station 4 staffing); 1 Asst. City Secretary; 1 Athletics Coordinator; 1 Recreation Specialist; 1 Senior Center Supervisor |
| Water & Sewer Fund | 6 | 3 Worker I's, 1 Worker II, 1 Crew Leader (new crew); 1 Worker I (floater) |
| EMS Fund | 4 | 4 EMTs (Medic 8 and Medic 9 peak-hour coverage) |
| Drainage Fund | 6 | 3 Worker I's, 1 Worker II, 1 Crew Leader (new crew); 1 Floodplain Manager |
| **Total** | **26** | |

### Compiled Personnel Services Totals by Department

The table below was compiled manually from the **ExpCategory: 511 —
Personnel Services Total** rows in the budget worksheets.  "Personnel
Services" in Schertz's chart of accounts (Category 511) encompasses:
base wages, overtime, longevity pay, clothing/certification allowances,
employer FICA, employer TMRS (retirement), employer health insurance,
workers' compensation insurance, uniforms, training and travel, and
miscellaneous employee-related costs.  **Benefits are therefore
included.**

```{r dept-table}
#| label: dept-table
#| tbl-cap: "Personnel Services (ExpCategory 511) by Department — FY 2024-25 Projected vs. FY 2025-26 Proposed Budget"

# --------------------------------------------------------------------------
# All figures read directly from the Schertz FY 2025-26 Proposed Budget PDF.
# Columns: department name, fund, 2024-25 projected, 2025-26 budget.
# --------------------------------------------------------------------------

dept_data <- data.frame(
  Fund = c(
    # General Fund departments (Fund 101)
    rep("General Fund (101)", 24),
    # Water & Sewer Fund (Fund 202)
    "Water & Sewer (202)",
    "Water & Sewer (202)",
    # Drainage Fund (Fund 202/579)
    "Drainage (202)",
    # EMS Fund (Fund 203)
    "EMS (203)",
    # Special Revenue
    "Special Revenue (615)"
  ),
  Dept_ID = c(
    150, 151, 152, 153, 154, 156, 164, 170, 173,
    256, 257, 258, 259, 359, 460, 463, 464,
    665, 853, 866, 868, 872, 877, 878,
    176, 575,
    579,
    280,
    609
  ),
  Department = c(
    "City Council",
    "City Manager",
    "Municipal Court",
    "311 / Customer Relations",
    "Planning & Zoning",
    "Economic Development",
    "City Secretary",
    "Public Affairs",
    "Engineering",
    "Police",
    "Fire Rescue",
    "Inspections",
    "Neighborhood Services",
    "Streets",
    "Parks, Recreation & Community Svcs",
    "Swim Pool",
    "Event Facilities",
    "Library",
    "Information Technology",
    "Human Resources",
    "Finance",
    "Purchasing & Asset Management",
    "Fleet Services",
    "Facility Services",
    # W&S
    "Utility Billing",
    "Water & Sewer Administration",
    # Drainage
    "Drainage Operations",
    # EMS
    "Emergency Medical Services",
    # Special Revenue
    "Historical Committee"
  ),
  FY2425_Proj = c(
    # General Fund — read from ExpCategory 511 Total, column 7
     69525.00,    # City Council (150)
   1067488.00,    # City Manager (151)
    383922.00,    # Municipal Court (152)
    137060.00,    # 311/Customer Relations (153)
    463444.00,    # Planning & Zoning (154)
    543724.00,    # Economic Development (156)
    251676.00,    # City Secretary (164)
    355153.00,    # Public Affairs (170)
   1230173.00,    # Engineering (173)
  12506800.00,    # Police (256)
   8329400.00,    # Fire Rescue (257)
   1388161.00,    # Inspections (258)
    523406.00,    # Neighborhood Services (259)
   1237425.00,    # Streets (359)
   1411968.30,    # Parks (460)
     22300.00,    # Swim Pool (463)
    536887.00,    # Event Facilities (464)
   1204714.30,    # Library (665)
   1394110.00,    # IT (853)
    759136.60,    # Human Resources (866)
    819292.76,    # Finance (868)
    300511.08,    # Purchasing & Asset Mgt (872)
    862355.00,    # Fleet Services (877)
   1149335.00,    # Facility Services (878)
    # Water & Sewer
    774837.00,    # Utility Billing (176)
   2801020.00,    # W&S Administration (575)
    # Drainage
    597942.00,    # Drainage (579)
    # EMS
   9218278.73,    # EMS (203-280)
    # Special Revenue
      6350.00     # Historical Committee (615-609)
  ),
  FY2526_Budget = c(
    # General Fund — read from ExpCategory 511 Total, column 8
    109839.00,    # City Council (150)
   1017376.00,    # City Manager (151)
    373139.00,    # Municipal Court (152)
    150518.00,    # 311/Customer Relations (153)
    456245.00,    # Planning & Zoning (154)
    554647.00,    # Economic Development (156)
    336046.00,    # City Secretary (164)
    348964.00,    # Public Affairs (170)
   1272482.00,    # Engineering (173)
  12249146.00,    # Police (256)
   8829785.00,    # Fire Rescue (257)
   1390202.00,    # Inspections (258)
    521791.00,    # Neighborhood Services (259)
   1323243.00,    # Streets (359)
   1584494.30,    # Parks (460)
     24575.00,    # Swim Pool (463)
    521117.00,    # Event Facilities (464)
   1198296.00,    # Library (665)
   1367412.00,    # IT (853)
    771430.00,    # Human Resources (866)
    820727.00,    # Finance (868)
    302455.00,    # Purchasing & Asset Mgt (872)
    864312.00,    # Fleet Services (877)
   1103336.00,    # Facility Services (878)
    # Water & Sewer
    788936.00,    # Utility Billing (176)
   3462953.00,    # W&S Administration (575)
    # Drainage
   1018261.00,    # Drainage (579)
    # EMS
   9814819.00,    # EMS (203-280)
    # Special Revenue
      6350.00     # Historical Committee (615-609)
  ),
  stringsAsFactors = FALSE
)

dept_data$Change    <- dept_data$FY2526_Budget - dept_data$FY2425_Proj
dept_data$Pct_Chg   <- dept_data$Change / dept_data$FY2425_Proj

# Display table
display_df <- dept_data
display_df$FY2425_Proj   <- fmt_dollar(display_df$FY2425_Proj)
display_df$FY2526_Budget <- fmt_dollar(display_df$FY2526_Budget)
display_df$Change        <- ifelse(
  dept_data$Change >= 0,
  paste0("+", fmt_dollar(dept_data$Change)),
  paste0("-", fmt_dollar(abs(dept_data$Change)))
)
display_df$Pct_Chg <- ifelse(
  dept_data$Pct_Chg >= 0,
  paste0("+", fmt_pct(dept_data$Pct_Chg)),
  paste0("-", fmt_pct(abs(dept_data$Pct_Chg)))
)

knitr::kable(
  display_df[, c("Fund", "Department", "FY2425_Proj", "FY2526_Budget",
                 "Change", "Pct_Chg")],
  col.names = c("Fund", "Department", "FY 2024-25 Projected",
                "FY 2025-26 Budget", "Change ($)", "Change (%)"),
  align = c("l", "l", "r", "r", "r", "r")
)
```

```{r subtotals}
#| label: subtotals
#| tbl-cap: "Personnel Services Subtotals by Fund"

# Subtotals by fund
library_funds <- c("General Fund (101)", "Water & Sewer (202)",
                   "Drainage (202)", "EMS (203)", "Special Revenue (615)")

fund_summary <- do.call(rbind, lapply(unique(dept_data$Fund), function(f) {
  sub <- dept_data[dept_data$Fund == f, ]
  data.frame(
    Fund          = f,
    FY2425_Proj   = sum(sub$FY2425_Proj),
    FY2526_Budget = sum(sub$FY2526_Budget),
    Change        = sum(sub$Change),
    stringsAsFactors = FALSE
  )
}))

# Grand total row
grand <- data.frame(
  Fund          = "**GRAND TOTAL (All Funds)**",
  FY2425_Proj   = sum(dept_data$FY2425_Proj),
  FY2526_Budget = sum(dept_data$FY2526_Budget),
  Change        = sum(dept_data$Change),
  stringsAsFactors = FALSE
)

fund_summary <- rbind(fund_summary, grand)
fund_summary$Pct_Chg <- fund_summary$Change / fund_summary$FY2425_Proj

fs_display <- fund_summary
fs_display$FY2425_Proj   <- fmt_dollar(fund_summary$FY2425_Proj)
fs_display$FY2526_Budget <- fmt_dollar(fund_summary$FY2526_Budget)
fs_display$Change        <- ifelse(
  fund_summary$Change >= 0,
  paste0("+", fmt_dollar(fund_summary$Change)),
  paste0("-", fmt_dollar(abs(fund_summary$Change)))
)
fs_display$Pct_Chg <- ifelse(
  fund_summary$Pct_Chg >= 0,
  paste0("+", fmt_pct(fund_summary$Pct_Chg)),
  paste0("-", fmt_pct(abs(fund_summary$Pct_Chg)))
)

knitr::kable(
  fs_display,
  col.names = c("Fund", "FY 2024-25 Projected", "FY 2025-26 Budget",
                "Change ($)", "Change (%)"),
  align = c("l", "r", "r", "r", "r")
)
```

### Key Findings — Total Compensation Increase

```{r summary-findings}
gf_proj   <- sum(dept_data$FY2425_Proj[dept_data$Fund == "General Fund (101)"])
gf_budget <- sum(dept_data$FY2526_Budget[dept_data$Fund == "General Fund (101)"])
gf_change <- gf_budget - gf_proj

all_proj   <- sum(dept_data$FY2425_Proj)
all_budget <- sum(dept_data$FY2526_Budget)
all_change <- all_budget - all_proj

# Non-GF increases (fee-funded)
non_gf_change <- all_change - gf_change
```

- **All-funds Personnel Services increase: `r fmt_dollar(all_change)`** (from
  `r fmt_dollar(all_proj)` to `r fmt_dollar(all_budget)`, a
  `r fmt_pct(all_change/all_proj)` increase).

- **General Fund (tax-supported) net increase: `r fmt_dollar(gf_change)`** — the
  portion relevant to property tax funding.

- **Fee-funded Personnel Services increase: `r fmt_dollar(non_gf_change)`**
  (Water/Sewer, Drainage, EMS — covered by utility fees and service contracts,
  not property taxes).

- The **explicitly stated merit/step program cost is \$784,000 citywide**
  (mid-year implementation, covering approximately the second half of FY 2025-26).
  The annualized cost of the merit/step program would be approximately
  **\$1,568,000**.

- The difference between the merit/step policy cost and the net GF personnel
  change (`r fmt_dollar(gf_change)`) reflects offsetting factors in individual
  departments — notably a **`r fmt_dollar(abs(dept_data$Change[dept_data$Department == "Police"]))` decrease in Police Personnel Services**
  and smaller decreases in several other departments (City Manager, Facility
  Services, IT), partially offset by the 6 new firefighters
  (`r fmt_dollar(dept_data$Change[dept_data$Department == "Fire Rescue"])` Fire
  Rescue increase) and other new positions.

#### Notable Departmental Changes — General Fund

```{r notable-changes}
gf <- dept_data[dept_data$Fund == "General Fund (101)", ]
gf_sorted <- gf[order(gf$Change, decreasing = TRUE), ]

gf_sorted$FY2425_Proj   <- fmt_dollar(gf_sorted$FY2425_Proj)
gf_sorted$FY2526_Budget <- fmt_dollar(gf_sorted$FY2526_Budget)
gf_sorted$Change        <- ifelse(
  as.numeric(gsub("[^-0-9.]", "", gf_sorted$Change)) >= 0,
  paste0("+", fmt_dollar(gf[order(gf$Change, decreasing = TRUE), "Change"])),
  paste0("-", fmt_dollar(abs(gf[order(gf$Change, decreasing = TRUE), "Change"])))
)
```

The five largest increases and five largest decreases in the General Fund:

```{r gf-top-changes}
#| tbl-cap: "Largest Personnel Services Changes — General Fund"

gf_num <- dept_data[dept_data$Fund == "General Fund (101)", ]
gf_num <- gf_num[order(gf_num$Change, decreasing = TRUE), ]

top5    <- head(gf_num, 5)
bottom5 <- tail(gf_num, 5)
highlight <- rbind(top5, bottom5)

hl_display <- highlight[, c("Department", "FY2425_Proj", "FY2526_Budget",
                             "Change", "Pct_Chg")]
hl_display$FY2425_Proj   <- fmt_dollar(highlight$FY2425_Proj)
hl_display$FY2526_Budget <- fmt_dollar(highlight$FY2526_Budget)
hl_display$Change        <- ifelse(
  highlight$Change >= 0,
  paste0("+", fmt_dollar(highlight$Change)),
  paste0("-", fmt_dollar(abs(highlight$Change)))
)
hl_display$Pct_Chg <- ifelse(
  highlight$Pct_Chg >= 0,
  paste0("+", fmt_pct(highlight$Pct_Chg)),
  paste0("-", fmt_pct(abs(highlight$Pct_Chg)))
)

knitr::kable(
  hl_display,
  col.names = c("Department", "FY 2024-25 Proj.", "FY 2025-26 Budget",
                "Change ($)", "Change (%)"),
  align = c("l", "r", "r", "r", "r")
)
```

**Police decrease explanation:** The Police department shows a decrease of
`r fmt_dollar(abs(dept_data$Change[dept_data$Department == "Police"]))` in
Personnel Services.  Examining the individual account lines reveals that the
FY 2024-25 projected year-end included elevated overtime actuals and medical
services that will not recur at that level.  Regular pay (account 511110)
actually *increases* from the FY 2024-25 budget (\$7,484,154) to the FY 2025-26
budget (\$7,773,694), consistent with the merit/step program.  The net dollar
decrease is primarily an artifact of the 2024-25 projection being inflated
above budget due to mid-year actuals.

---

## Question 2: Does the Increase Include Benefits?

**Yes.**  Expenditure Category 511 — "Personnel Services" — is the
account category that Schertz uses for all compensation-related costs.
It includes the following account types (verified from the Police and EMS
detailed account listings in the budget worksheet):

| Account Type | Example Accounts | Included? |
|---|---|---|
| Regular wages | 511110 | ✓ |
| Overtime | 511120 | ✓ |
| Longevity pay | 511210 | ✓ |
| Clothing / uniform allowance | 511220, 511606 | ✓ |
| Certification allowance | 511230 | ✓ |
| Employer FICA (Social Security + Medicare) | 511310 | ✓ |
| Employer TMRS (retirement — Texas Municipal) | 511350 | ✓ |
| Employer health insurance | 511410 | ✓ |
| Workers' compensation insurance | 511500 | ✓ |
| Unemployment compensation | 511600 | ✓ |
| Training and travel | 511612 | ✓ |
| Medical services (pre-employment, physicals) | 511617 | ✓ |
| Employee recognition | 511609 | ✓ |

The budget explicitly states that the medical insurance provider change was
structured to **prevent any premium increase**, meaning that health insurance
costs are held approximately flat in the new budget year.  The overall
Personnel Services increases therefore reflect primarily the wage/salary and
associated payroll-tax components.

In addition to ExpCategory 511, the Non-Departmental budget (Dept 101) carries
a **Contingencies-Personnel** appropriation (account 101-101-534013) of
**\$300,000** in both FY 2024-25 and FY 2025-26 — no change.  This is a
reserve for unplanned personnel-related needs and is not part of the
ExpCategory 511 totals compiled above.

---

## Question 3: Property Tax Rate Required to Cover the Compensation Increases

### Overview of the Proposed Tax Rate

The budget is funded at the **No-New-Revenue (NNR) rate** of **\$0.5064** per
\$100 of taxable value — an increase of \$0.0164 over the FY 2024-25 rate of
\$0.4900.

```{r tax-rates}
#| label: tax-rates
#| tbl-cap: "City of Schertz Property Tax Rate — FY 2024-25 vs. FY 2025-26 Proposed"

tax_tbl <- data.frame(
  Component = c("Maintenance & Operations (M&O)", "Interest & Sinking (I&S)",
                "Total Rate"),
  FY2425 = c(0.3326, 0.1574, 0.4900),
  FY2526 = c(0.3383, 0.1681, 0.5064),
  stringsAsFactors = FALSE
)
tax_tbl$Change <- tax_tbl$FY2526 - tax_tbl$FY2425

knitr::kable(
  tax_tbl,
  col.names = c("Component", "FY 2024-25 ($/100)", "FY 2025-26 Proposed ($/100)",
                "Change ($/100)"),
  digits = 4,
  align = c("l", "r", "r", "r")
)
```

The total taxable value as certified by the Guadalupe County Tax Assessor is
**\$5,734,815,455**.

### Revenue per Cent of Tax Rate

```{r revenue-per-cent}
# Derived from budgeted Ad Valorem Tax revenue and the proposed rate.
# FY 2025-26 Ad Valorem Tax Current budgeted: $20,668,000 at $0.5064 rate.
# Revenue per $0.01 per $100 = $20,668,000 / 50.64 (rate in cents)

rate_2526_cents <- 50.64          # $0.5064 expressed as cents per $100
av_revenue_2526 <- 20668000       # From budget line 101-101-411100
rev_per_cent    <- av_revenue_2526 / rate_2526_cents

cat(sprintf(
  "Budgeted Ad Valorem Tax Revenue (FY 2025-26): %s\n",
  fmt_dollar(av_revenue_2526)
))
cat(sprintf(
  "Proposed tax rate: $%.4f per $100 valuation (%.2f cents per $100)\n",
  0.5064, rate_2526_cents
))
cat(sprintf(
  "Estimated revenue per $0.01 per $100 of tax rate: %s\n",
  fmt_dollar(rev_per_cent)
))
```

Note: This per-cent figure implicitly incorporates the actual collection rate
(delinquencies, exemptions).  It is a more reliable basis for rate calculations
than applying the statutory taxable value directly.

### Tax Rate Required to Cover General Fund Personnel Increase

The property tax (M&O rate) supports only the **General Fund**.  Water/Sewer,
Drainage, and EMS personnel costs are funded by fees and service contracts,
not property taxes.

```{r gf-tax-calc}
# General Fund Personnel Services net change
gf_pers_increase <- gf_change    # computed above

# Tax rate increase needed (in cents per $100)
cents_needed <- gf_pers_increase / rev_per_cent

# Dollar rate
rate_needed <- cents_needed / 100    # convert cents to $/100

cat(sprintf(
  "General Fund Personnel Services net increase:   %s\n",
  fmt_dollar(gf_pers_increase)
))
cat(sprintf(
  "Revenue per $0.01 per $100 of rate:             %s\n",
  fmt_dollar(rev_per_cent)
))
cat(sprintf(
  "Tax rate increase needed (cents per $100):       %.4f cents\n",
  cents_needed
))
cat(sprintf(
  "Tax rate increase needed ($/100):               $%.4f per $100\n",
  rate_needed
))
cat(sprintf(
  "This equals %.1f%% of the current M&O rate ($0.3326)\n",
  rate_needed / 0.3326 * 100
))
```

### Dollar Impact on Taxpayers

To translate the tax rate increase into a dollar cost per taxpayer, we need a
typical taxable home value.  The budget does not publish the median taxable
value directly, but we can derive an approximation.

```{r taxpayer-impact}
# Total taxable value from budget: $5,734,815,455
# Total homesteads approximated from the document:
#   2,616 properties use Disabled Veteran exemption (~23% of homesteads)
#   => total homesteads ≈ 2,616 / 0.23 ≈ 11,374
# Total taxable value includes commercial properties too.
# For illustrative purposes, compute per typical home values.

total_taxable <- 5734815455

# Illustrative taxable home values
home_values <- c(200000, 300000, 400000, 500000)

# Rate increase needed solely for GF personnel increase
rate_increase <- rate_needed    # $/100

impact_df <- data.frame(
  Home_Value = home_values,
  Annual_Impact = home_values / 100 * rate_increase,
  stringsAsFactors = FALSE
)

# Also compute impact of the full proposed rate increase ($0.0164)
full_rate_increase <- 0.5064 - 0.4900   # total proposed increase
impact_df$Full_Rate_Impact <- home_values / 100 * full_rate_increase

knitr::kable(
  data.frame(
    Taxable_Value  = fmt_dollar(impact_df$Home_Value),
    Personnel_Only = paste0(
      fmt_dollar(impact_df$Annual_Impact, digits = 2), " / yr"
    ),
    Full_Rate_Increase = paste0(
      fmt_dollar(impact_df$Full_Rate_Impact, digits = 2), " / yr"
    )
  ),
  col.names = c("Taxable Home Value",
                "Annual Cost: GF Personnel Increase Only",
                "Annual Cost: Full Proposed Rate Increase ($0.0164)"),
  align = c("r", "r", "r"),
  caption = "Estimated Annual Property Tax Impact by Home Value"
)
```

### Citywide Merit/Step Program — Standalone Rate Calculation

The budget explicitly states the merit/step program costs **\$784,000 citywide**
for the half-year (mid-year implementation).  If we isolate only this policy
cost and ask how much of it falls on property taxpayers, we need the General
Fund share.

The document does not publish a GF-only merit/step figure, but we can estimate
it proportionally using the ratio of GF Personnel Services to total personnel
services:

```{r merit-step-calc}
gf_share <- gf_proj / all_proj
gf_merit_step <- 784000 * gf_share

cat(sprintf(
  "Total merit/step program cost (half-year, all funds): %s\n",
  fmt_dollar(784000)
))
cat(sprintf(
  "General Fund share of total personnel services:        %s (%.1f%%)\n",
  fmt_dollar(gf_proj), gf_share * 100
))
cat(sprintf(
  "Estimated GF portion of merit/step cost:               %s\n",
  fmt_dollar(gf_merit_step)
))

rate_merit_step <- (gf_merit_step / rev_per_cent) / 100
cat(sprintf(
  "Tax rate to fund GF merit/step cost alone:             $%.4f per $100\n",
  rate_merit_step
))
```

---

## Question 4: Is the Proposed Tax Rate Sufficient to Cover the Compensation Increases?

```{r sufficiency-check}
# Proposed M&O rate increase revenue
mo_rate_increase <- 0.3383 - 0.3326          # $0.0057
mo_revenue_increase <- mo_rate_increase * 100 * rev_per_cent  # rate in cents

# Total proposed General Fund revenue increase (all sources)
gf_rev_2425 <- 52632269   # 2024-25 year-end estimate from summary
gf_rev_2526 <- 52499128   # 2025-26 proposed budget from summary
gf_rev_change <- gf_rev_2526 - gf_rev_2425

cat(sprintf(
  "GF Revenue FY 2024-25 (year-end est.):    %s\n", fmt_dollar(gf_rev_2425)
))
cat(sprintf(
  "GF Revenue FY 2025-26 (proposed):         %s\n", fmt_dollar(gf_rev_2526)
))
cat(sprintf(
  "Net GF Revenue change:                     %s\n", fmt_dollar(gf_rev_change)
))
cat(sprintf(
  "\nM&O rate increase ($0.0057) additional revenue est.: %s\n",
  fmt_dollar(mo_revenue_increase)
))
cat(sprintf(
  "GF Personnel Services net increase (full year):      %s\n",
  fmt_dollar(gf_change)
))
cat(sprintf(
  "Rate increase revenue MINUS personnel increase:      %s\n",
  fmt_dollar(mo_revenue_increase - gf_change)
))
cat(sprintf(
  "  --> Rate increase DOES%s cover the full-year GF personnel increase.\n",
  ifelse(mo_revenue_increase >= gf_change, "", " NOT")
))
cat(sprintf(
  "\nMemo: merit/step cost as stated in budget = $784,000 (HALF-YEAR only)\n"
))
cat(sprintf(
  "      Annualized merit/step cost (full year estimate) = %s\n",
  fmt_dollar(784000 * 2)
))
cat(sprintf(
  "      Rate increase revenue vs. annualized merit/step: %s\n",
  fmt_dollar(mo_revenue_increase - 784000 * 2)
))
```

### Deficit Budget Context

Despite the No-New-Revenue rate increase, the General Fund remains in
**deficit**:

| | Amount |
|---|---|
| FY 2025-26 GF Revenue | $52,499,128 |
| FY 2025-26 GF Expenditures | $53,754,934 |
| **Surplus / (Deficit)** | **(\$1,255,806)** |
| Resulting Reserve as % of Expenditures | 26.8% |
| City Reserve Policy Minimum | ≤ 26% |

The proposed budget would draw down reserves by \$1,255,806, leaving the
reserve balance at \$13,936,617, which is *above* the City's own 26% maximum
policy target.  This means the budget proposes to overspend relative to
current revenue while still technically meeting the reserve floor — but the
reserve direction is downward.

Other funds are also in deficit:

| Fund | Revenue | Expenditures | Surplus/(Deficit) |
|---|---|---|---|
| Water & Sewer | \$35,391,478 | \$36,695,153 | (\$1,303,675) |
| Drainage | \$2,316,283 | \$2,818,831 | (\$502,548) |
| SEDC | \$8,928,608 | \$16,701,463 | (\$7,772,855) |
| EMS | \$17,504,469 | \$17,149,568 | \$354,901 |

The EMS fund is the only major operating fund with a projected surplus, made
possible by a 20% increase in contracted service revenues and \$100/transport
fee increase.

---

## Question 5: Annual Growth Rate and Doubling Time for Total Compensation

### Data Available

The budget worksheets present four fiscal years of Personnel Services data for every department:

| Year | Description |
|---|---|
| FY 2022-23 | Actual expenditures |
| FY 2023-24 | Actual expenditures |
| FY 2024-25 | Year-end projected (staff estimate) |
| FY 2025-26 | Proposed budget |

This yields a **3-year span** for CAGR calculation (FY 2022-23 → FY 2025-26 proposed), with a 2-year cross-check using FY 2024-25 projected as the endpoint to avoid dependence on the proposed budget.

### Complete Four-Year Dataset

```{r cagr-data}
#| label: cagr-data
#| tbl-cap: "Personnel Services — Four-Year History, All Departments"

# Values read from ExpCategory: 511 - Personnel Services Total rows,
# column 2 (FY22-23 Actual) and column 4 (FY23-24 Actual).

dept_data$FY2223_Act <- c(
  # General Fund (24 departments, same order as dept_data)
    65991.00,    # City Council (150)
  1268219.37,    # City Manager (151)
   349992.25,    # Municipal Court (152)
   130167.92,    # 311/Customer Relations (153)
   332998.93,    # Planning & Zoning (154)
         0.00,   # Economic Development (156) — dept created FY23-24
   195204.99,    # City Secretary (164)
   338315.64,    # Public Affairs (170)
   995485.25,    # Engineering (173)
  9667983.16,    # Police (256)
  6825077.73,    # Fire Rescue (257)
  1056532.12,    # Inspections (258)
   423553.18,    # Neighborhood Services (259)
   996132.16,    # Streets (359)
  1035692.62,    # Parks (460)
    17692.84,    # Swim Pool (463)
   440489.69,    # Event Facilities (464)
  1001807.06,    # Library (665)
  1147821.16,    # IT (853)
   643658.58,    # Human Resources (866)
   696231.81,    # Finance (868)
   267811.27,    # Purchasing & Asset Mgt (872)
   694009.96,    # Fleet Services (877)
   897424.16,    # Facility Services (878)
  # Water & Sewer
   601713.12,    # Utility Billing (176)
  1942709.33,    # W&S Administration (575)
  # Drainage
   634080.66,    # Drainage (579)
  # EMS
  7602824.08,    # EMS (203-280)
  # Special Revenue
      3558.32    # Historical Committee (615-609)
)

dept_data$FY2324_Act <- c(
  # General Fund
    70269.22,    # City Council (150)
  1133902.71,    # City Manager (151)
   365992.45,    # Municipal Court (152)
   150130.32,    # 311/Customer Relations (153)
   415440.81,    # Planning & Zoning (154)
   373804.15,    # Economic Development (156)
   218750.26,    # City Secretary (164)
   361689.77,    # Public Affairs (170)
  1050157.62,    # Engineering (173)
 11191198.57,    # Police (256)
  8283649.91,    # Fire Rescue (257)
  1261620.00,    # Inspections (258)
   506751.71,    # Neighborhood Services (259)
  1091718.73,    # Streets (359)
  1322151.35,    # Parks (460)
    15723.23,    # Swim Pool (463)
   543478.03,    # Event Facilities (464)
  1154459.87,    # Library (665)
  1343640.33,    # IT (853)
   713034.39,    # Human Resources (866)
   727058.93,    # Finance (868)
   234729.59,    # Purchasing & Asset Mgt (872)
   897046.05,    # Fleet Services (877)
  1115683.58,    # Facility Services (878)
  # Water & Sewer
   653083.93,    # Utility Billing (176)
  2210979.68,    # W&S Administration (575)
  # Drainage
   613853.27,    # Drainage (579)
  # EMS
  8986816.34,    # EMS (203-280)
  # Special Revenue
        0.00     # Historical Committee (615-609) — no activity FY23-24
)

# Four-year totals by scope
calc_totals <- function(fund_filter = NULL) {
  sub <- if (is.null(fund_filter)) dept_data else dept_data[dept_data$Fund %in% fund_filter, ]
  c(FY2223 = sum(sub$FY2223_Act),
    FY2324 = sum(sub$FY2324_Act),
    FY2425 = sum(sub$FY2425_Proj),
    FY2526 = sum(sub$FY2526_Budget))
}

gf_funds <- "General Fund (101)"
gf_totals  <- calc_totals(gf_funds)
all_totals <- calc_totals()

summary_tbl <- rbind(
  data.frame(Scope = "General Fund",        t(gf_totals)),
  data.frame(Scope = "All Funds Combined",  t(all_totals))
)

disp <- summary_tbl
for (col in c("FY2223","FY2324","FY2425","FY2526")) disp[[col]] <- fmt_dollar(as.numeric(disp[[col]]))

knitr::kable(disp,
  col.names = c("Scope","FY 2022-23 Actual","FY 2023-24 Actual",
                "FY 2024-25 Projected","FY 2025-26 Proposed"),
  align = c("l","r","r","r","r"))
```

### Annualized Adjustment for the Merit/Step Program

The FY 2025-26 budget **deliberately understates** the full cost of the
compensation increases because the merit/step program is implemented
**mid-year** — only half the annual cost appears in the budget.  The other
half hits the City in FY 2026-27 before any new raises are negotiated.

To project where compensation actually lands on a full-year run-rate basis,
the CAGR calculation below uses an **annualized FY 2025-26 endpoint**: the
budgeted Personnel Services totals with the second half of the merit/step cost
added back.  This is the correct basis for doubling-time projections because
doubling time is a forward-looking statement — it answers "how long until
compensation is twice what it is today?" — and "today" should mean the full
run-rate cost, not the discounted budget figure.

```{r annualize-adj}
#| label: annualize-adj

# The budget states the merit/step program costs $784,000 citywide for the
# half-year.  The full annualized incremental cost is $784,000 × 2 = $1,568,000.
# The second half ($784,000) is not in the proposed budget at all.
# We allocate the adjustment proportionally by fund share of total personnel.

merit_half   <- 784000          # as stated in budget document
merit_full   <- merit_half * 2  # annualized
merit_addback <- merit_half     # the half NOT in the budget

# GF share of total personnel services (FY2526 budget)
gf_pers_budget  <- sum(dept_data$FY2526_Budget[dept_data$Fund == "General Fund (101)"])
all_pers_budget <- sum(dept_data$FY2526_Budget)
gf_share_2526   <- gf_pers_budget / all_pers_budget

# Annualized endpoints
gf_2526_ann  <- gf_totals["FY2526"]  + merit_addback * gf_share_2526
all_2526_ann <- all_totals["FY2526"] + merit_addback

cat(sprintf("Merit/step half-year cost (in budget):        %s\n", fmt_dollar(merit_half)))
cat(sprintf("Merit/step addback for annualization:         %s\n", fmt_dollar(merit_addback)))
cat(sprintf("Annualized GF Personnel Services (FY25-26):   %s\n", fmt_dollar(gf_2526_ann)))
cat(sprintf("Annualized All-Funds Personnel Services:      %s\n", fmt_dollar(all_2526_ann)))
cat(sprintf("\nDifference from budgeted figure:\n"))
cat(sprintf("  General Fund:  %s\n", fmt_dollar(gf_2526_ann  - gf_totals["FY2526"])))
cat(sprintf("  All Funds:     %s\n", fmt_dollar(all_2526_ann - all_totals["FY2526"])))
```

### Compound Annual Growth Rates (CAGR)

$$\text{CAGR} = \left(\frac{V_n}{V_0}\right)^{1/n} - 1 \qquad t_{\text{double}} = \frac{\ln 2}{\ln(1 + \text{CAGR})}$$

Two sets of CAGR are presented: one using the **as-budgeted** FY 2025-26
figures (half-year merit/step), and one using the **annualized** figures
(full-year run-rate).  The annualized CAGR is the appropriate basis for
doubling-time projections.

```{r cagr-calc}
#| label: cagr-calc
#| tbl-cap: "CAGR and Doubling Time — As-Budgeted vs. Annualized"

cagr          <- function(v0, vn, n) (vn / v0)^(1/n) - 1
doubling_time <- function(r) log(2) / log(1 + r)

scopes <- list(
  list(label = "General Fund — as budgeted",   v0 = gf_totals["FY2223"],
       v2425 = gf_totals["FY2425"],  v2526 = gf_totals["FY2526"],
       v2526a = gf_2526_ann,  vy1324 = gf_totals["FY2324"]),
  list(label = "General Fund — annualized",    v0 = gf_totals["FY2223"],
       v2425 = gf_totals["FY2425"],  v2526 = gf_2526_ann,
       v2526a = gf_2526_ann, vy1324 = gf_totals["FY2324"]),
  list(label = "All Funds — as budgeted",      v0 = all_totals["FY2223"],
       v2425 = all_totals["FY2425"], v2526 = all_totals["FY2526"],
       v2526a = all_2526_ann, vy1324 = all_totals["FY2324"]),
  list(label = "All Funds — annualized",       v0 = all_totals["FY2223"],
       v2425 = all_totals["FY2425"], v2526 = all_2526_ann,
       v2526a = all_2526_ann, vy1324 = all_totals["FY2324"])
)

results <- do.call(rbind, lapply(scopes, function(s) {
  c3 <- cagr(s$v0,     s$v2526,  3)
  c2 <- cagr(s$v0,     s$v2425,  2)
  y1 <- cagr(s$vy1324, s$v2425,  1)
  y2 <- cagr(s$v2425,  s$v2526,  1)
  data.frame(
    Scope    = s$label,
    CAGR_3yr = fmt_pct(c3),
    DT_3yr   = sprintf("%.1f yrs", doubling_time(c3)),
    CAGR_2yr = fmt_pct(c2),
    DT_2yr   = sprintf("%.1f yrs", doubling_time(c2)),
    YoY_2425 = fmt_pct(y1),
    YoY_2526 = fmt_pct(y2),
    stringsAsFactors = FALSE
  )
}))

knitr::kable(results,
  col.names = c("Scope",
                "3-yr CAGR\n(FY22-23→FY25-26)", "Doubling Time",
                "2-yr CAGR\n(FY22-23→FY24-25)", "Doubling Time",
                "YoY\nFY24-25", "YoY\nFY25-26"),
  align = c("l","r","r","r","r","r","r"))
```

```{r cagr-print}
gf_c3_ann  <- cagr(gf_totals["FY2223"],  gf_2526_ann,  3)
gf_dt3_ann <- doubling_time(gf_c3_ann)
all_c3_ann  <- cagr(all_totals["FY2223"], all_2526_ann, 3)
all_dt3_ann <- doubling_time(all_c3_ann)

cat(sprintf("General Fund (annualized) — 3-yr CAGR: %s  →  doubles every %.1f years\n",
            fmt_pct(gf_c3_ann),  gf_dt3_ann))
cat(sprintf("All Funds    (annualized) — 3-yr CAGR: %s  →  doubles every %.1f years\n",
            fmt_pct(all_c3_ann), all_dt3_ann))
```

### Interpretation

At the annualized 3-year CAGR, compare the compensation doubling time against:

- **Property tax revenue** — Schertz's long-run growth has averaged ~7% per year
  (doubling time ~10 years), constrained by the voter-approval rate cap.
- **Population growth** — approximately 2–3% per year in the Schertz area
  (doubling time 23–35 years).

If compensation costs compound faster than either revenue or population, the
gap grows exponentially and must eventually be closed by rate increases,
service cuts, or reserve drawdowns.  The year-over-year figures show whether
the growth rate is accelerating or decelerating.

---

## Supplemental: Full-Year (Annualized) Compensation Cost Impact

The proposed FY 2025-26 budget includes only **half a year** of merit/step
increases.  The City will face the **full annualized run-rate** beginning in
FY 2026-27, before any new compensation adjustments are negotiated.  This
section presents those forward-looking figures.

```{r fullyr-impact}
#| label: fullyr-impact
#| tbl-cap: "Full-Year Run-Rate Compensation — FY 2026-27 Baseline Projection"

# Full-year GF and all-funds personnel cost at the annualized rate
cat("=== FULL-YEAR ANNUALIZED COMPENSATION RUN-RATE ===\n\n")

cat(sprintf("As-budgeted GF Personnel Services (FY25-26):       %s\n",
            fmt_dollar(gf_totals["FY2526"])))
cat(sprintf("Annualized GF Personnel Services:                  %s\n",
            fmt_dollar(gf_2526_ann)))
cat(sprintf("Additional GF cost hitting FY26-27 (addback):      %s\n\n",
            fmt_dollar(gf_2526_ann - gf_totals["FY2526"])))

cat(sprintf("As-budgeted All-Funds Personnel Services (FY25-26):%s\n",
            fmt_dollar(all_totals["FY2526"])))
cat(sprintf("Annualized All-Funds Personnel Services:           %s\n",
            fmt_dollar(all_2526_ann)))
cat(sprintf("Additional all-funds cost hitting FY26-27:         %s\n\n",
            fmt_dollar(all_2526_ann - all_totals["FY2526"])))

# Tax rate equivalent of the addback (GF only)
addback_gf     <- gf_2526_ann - gf_totals["FY2526"]
rev_per_cent_v <- av_revenue_2526 / rate_2526_cents
rate_addback   <- (addback_gf / rev_per_cent_v) / 100

cat(sprintf("Tax rate equivalent of the GF addback:\n"))
cat(sprintf("  Additional revenue needed:   %s\n", fmt_dollar(addback_gf)))
cat(sprintf("  Revenue per $0.01 of rate:   %s\n", fmt_dollar(rev_per_cent_v)))
cat(sprintf("  Rate increase required:      $%.4f per $100\n\n", rate_addback))

cat("=== TAXPAYER IMPACT — FULL-YEAR MERIT/STEP COST ===\n\n")
home_vals <- c(200000, 300000, 400000, 500000)
impact_tbl <- data.frame(
  Home_Value   = fmt_dollar(home_vals),
  HalfYr_Cost  = paste0(fmt_dollar(home_vals / 100 * (gf_2526_ann - gf_totals["FY2526"]) /
                                     (rev_per_cent_v * 100) * 100, digits = 2), " / yr"),
  FullYr_Total = paste0(fmt_dollar(home_vals / 100 *
                                     (gf_2526_ann - sum(dept_data$FY2425_Proj[
                                       dept_data$Fund == "General Fund (101)"])) /
                                     (rev_per_cent_v * 100) * 100, digits = 2), " / yr")
)
knitr::kable(impact_tbl,
  col.names = c("Taxable Home Value",
                "Additional Cost: Merit/Step Addback\n(2nd-half hitting FY26-27)",
                "Total GF Personnel Increase vs FY24-25\n(Annualized, at full-year rate)"),
  align = c("r","r","r"))
```

**Bottom line:** The FY 2025-26 budget presents the merit/step program as a
\$784,000 cost.  That figure is accurate for the current budget year only.
By FY 2026-27, the identical compensation structure costs **\$1,568,000
per year** before any additional raises — simply because employees will
receive the increase for a full 12 months instead of 6.  This is a structural
cost commitment, not a one-time event.

---

## Summary of Findings

```{r final-summary}
cat("=== COMPENSATION INCREASES ===\n\n")
cat(sprintf(
  "Merit/Step program (stated, half-year, all funds):      %s\n",
  fmt_dollar(784000)
))
cat(sprintf(
  "Net Personnel Services increase, all funds:             %s\n",
  fmt_dollar(all_change)
))
cat(sprintf(
  "  -- General Fund (property-tax supported):             %s\n",
  fmt_dollar(gf_change)
))
cat(sprintf(
  "  -- Fee-funded funds (W&S, Drainage, EMS):             %s\n",
  fmt_dollar(non_gf_change)
))
cat(sprintf(
  "Benefits included in above figures?                     YES\n"
))

cat("\n=== PROPERTY TAX RATE ANALYSIS ===\n\n")
cat(sprintf(
  "Current M&O rate:                                       $0.3326 per $100\n"
))
cat(sprintf(
  "Proposed M&O rate (No-New-Revenue):                     $0.3383 per $100\n"
))
cat(sprintf(
  "M&O rate increase proposed:                             $0.0057 per $100\n"
))
cat(sprintf(
  "Rate needed solely for GF personnel increase:           $%.4f per $100\n",
  rate_needed
))
cat(sprintf(
  "Rate needed solely for GF merit/step (est.):            $%.4f per $100\n",
  rate_merit_step
))
cat(sprintf(
  "Revenue per $0.01 of rate (at 2025-26 taxable base):   %s\n",
  fmt_dollar(rev_per_cent)
))

cat("\n=== DEFICIT CONTEXT ===\n\n")
cat(sprintf(
  "General Fund deficit:                                   (%s)\n",
  fmt_dollar(abs(-1255806))
))
cat(sprintf(
  "Proposed rate (NNR) covers full-year GF personnel increase: %s\n",
  ifelse(mo_revenue_increase >= gf_change,
         "YES",
         paste0("NO — shortfall of ", fmt_dollar(gf_change - mo_revenue_increase)))
))
cat(sprintf(
  "Budget is structurally balanced (revenue >= expenditures):         NO\n"
))
```

### Interpretation

1. **The explicit merit/step compensation policy costs \$784,000** across all
   funds for a half-year.  The annualized run-rate is approximately
   **\$1.57 million** citywide.

2. **The total net increase in Personnel Services across all funds is
   `r fmt_dollar(all_change)`**, inclusive of all wages, benefits, and
   employee-related costs.  Benefits are included in this figure.

3. **Only the General Fund increase (`r fmt_dollar(gf_change)`) is
   property-tax-supported.**  The fee-funded funds (Water/Sewer, Drainage, EMS)
   cover their own personnel increases through utility fees and service contract
   revenues.

4. **The proposed M&O rate increase of \$0.0057** (from \$0.3326 to \$0.3383)
   generates approximately \$`r format(round(mo_revenue_increase), big.mark=",")`
   in additional revenue — **not sufficient** to cover the General Fund
   personnel net increase of `r fmt_dollar(gf_change)` on its own, let alone
   the full General Fund deficit of \$1,255,806. The rate increase covers only
   a portion of the compensation increase; the remainder is absorbed by
   drawing down reserves.

5. **Important note on the \$784,000 merit/step figure vs. the budget totals.**
   The \$784,000 stated in the budget message is explicitly a **half-year cost**
   (the program takes effect mid-fiscal-year). All Personnel Services totals
   compiled in this analysis are **full fiscal-year** figures from the budget
   worksheets. These two figures cannot be directly compared without adjustment.
   The annualized (full-year) cost of the merit/step program is approximately
   **\$1,568,000 citywide**. Conversely, the full-year net General Fund
   Personnel Services change of `r fmt_dollar(gf_change)` is suppressed below
   the annualized merit/step cost because some departments show decreases
   (notably Police, whose FY 2024-25 projected year-end was inflated by
   mid-year overtime actuals that will not recur).

5. **A taxpayer with a \$300,000 taxable home value** would pay approximately
   **`r fmt_dollar(300000/100*rate_needed, digits=2)` per year** attributable
   to the GF personnel services increase, and
   **`r fmt_dollar(300000/100*full_rate_increase, digits=2)` per year**
   from the full proposed rate increase (which also covers debt service increases
   and other operating cost growth).

---

*Analysis prepared by Dan Swart, CPA (ret.)| 
Sources: City of Schertz FY 2025-26 Proposed Budget
         (Excel budget document dated August 2025) | All figures read directly from budget document; no automated parsing routines were used.*
