---
title: "City of Schertz Base Salary vs Additional Earnings FY20-FY25 for Directors, Managers, and Public Safety Leaders-Searchable DT Tables"
subtitle: "Filename:  city_of_schertz_base_salary_vs_additional_earnings_fy20_fy25_for_directors_managers_and_public_safety_leaders_searchable_dt_tables.qmd"
description: ""
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true     # Set to 'false' for drafts. Override if you're NOT sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false


params:
  data_updated: !expr format(Sys.Date(), "%B %Y")
---


```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED


# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  post_process_html = identity,
  post_process_docx = identity
)



# Flextable built-in themes:
  # flextable::theme_alafoli()	|>  # BLAH
  # flextable::theme_apa()  # THIS IS NICE
  # flextable::theme_booktabs() |>  # NICE, MORE COMPACT
  # flextable::theme_box() |>   # OK, INCLUDES CELL BORDERS
  # flextable::theme_tron() |>  # 'DARK MODE' BLUE TEXT
  # flextable::theme_tron_legacy() |>   # 'DARK MODE' YELLOW TEXT
  # flextable::theme_vader() |>    # 'DARK MODE' WHITE TEXT
  # flextable::theme_vanilla() |>   # NOT SPECIAL
  # flextable::theme_zebra()	|>
  #


# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 28,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 22,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 18,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 18,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)



  
# Set seed for reproducibility
base::set.seed(123)

```






## Introduction
This document breaks down **total compensation** into its components for the City of Schertz's top-paid employees:

1. **Base Salary** (Regular Earnings) - Core salary/wages
2. **Additional Earnings** - Overtime, leave payout, deployment, arbitration, etc.
3. **Additional Benefits** - Healthcare, pension contributions, etc.

This breakdown helps understand:
- What portion of pay increases go to base salary vs. benefits
- How overtime and other earnings affect total compensation
- The true cost of employment beyond base salary

::: {.callout-note}
## Interactive Tables
**Data last updated:** `r params$data_updated`

- **Sort**: Click any column header to sort
- **Search**: Use the search box to find specific employees  
- **Filter**: Use column filters to narrow results
- **Export**: Use the Copy, CSV, Excel, or Print buttons
:::

---

## Step 1: Load Cached Data
```{r}
#| label: load-cached-data
#| cache: true

# =============================================================================
# LOAD CACHED DATA FROM OMNIBUS ANALYSIS
# =============================================================================
# Purpose: Load the pre-processed all_data object from the main analysis
# Cache: TRUE - data loads once, subsequent renders use cached version

data_path <- here::here("data_clean", "compensation_wide_fy20_fy25.rds")

if (base::file.exists(data_path)) {
  all_data <- base::readRDS(data_path)
  base::cat("Loaded data:", base::nrow(all_data), "rows\n")
  base::cat("Fiscal years:", base::paste(base::unique(all_data$fiscal_year), collapse = ", "), "\n")
} else {
  base::stop(
    "Data file not found at: ", data_path, "\n",
    "Please run the omnibus analysis first to generate the cleaned data."
  )
}
```

---

## Step 2: Filter to Top Employees
```{r}
#| label: filter-top-employees
#| cache: true

# =============================================================================
# FILTER DATA TO TOP EMPLOYEES ONLY
# =============================================================================
# Purpose: Apply leadership pattern to identify top-paid positions

# Pattern for leadership job titles (case-insensitive)
leadership_pattern <- base::paste(
  "city manager", "assistant city manager", "deputy city manager",
  "director", "assistant director", "deputy director",
  "manager", "superintendent",
  "chief", "commander", "captain", "battalion chief", "marshal",
  sep = "|"
)

# Filter to leadership positions with full earnings breakdown
top_employees_full <- all_data |>
  dplyr::filter(
    stringr::str_detect(
      job_title, 
      stringr::regex(leadership_pattern, ignore_case = TRUE)
    )
  ) |>
  dplyr::select(
    fiscal_year,
    name,
    name_std,
    department,
    job_title,
    # Base salary component
    regular_earnings,
    # Additional earnings components
    leave_payout,
    overtime_earnings,
    additional_earnings,
    deployment_earnings,
    arbitration_settlements,
    # Benefits component
    additional_benefits,
    # Total (for validation)
    calc_total_compensation
  ) |>
  dplyr::arrange(fiscal_year, dplyr::desc(calc_total_compensation))

base::cat("Top employees identified:", base::nrow(top_employees_full), "\n")
base::cat("Unique individuals:", dplyr::n_distinct(top_employees_full$name_std), "\n")
```

---

## Step 3: Calculate Component Totals
```{r}
#| label: calculate-components
#| cache: true

# =============================================================================
# CALCULATE COMPENSATION COMPONENTS
# =============================================================================
# Purpose: Aggregate individual earnings items into three main categories
#
# Component Structure:
#   BASE SALARY = regular_earnings
#   
#   ADDITIONAL EARNINGS = leave_payout + overtime_earnings + additional_earnings
#                        + deployment_earnings + arbitration_settlements
#   
#   ADDITIONAL BENEFITS = additional_benefits (healthcare, pension, etc.)
#   
#   TOTAL COMPENSATION = BASE + ADDITIONAL EARNINGS + ADDITIONAL BENEFITS

top_employees_components <- top_employees_full |>
  dplyr::mutate(
    # Base Salary (Regular Earnings only)
    base_salary = regular_earnings,
    
    # Additional Earnings (everything else that's not base or benefits)
    add_earnings = dplyr::coalesce(leave_payout, 0) +
                   dplyr::coalesce(overtime_earnings, 0) +
                   dplyr::coalesce(additional_earnings, 0) +
                   dplyr::coalesce(deployment_earnings, 0) +
                   dplyr::coalesce(arbitration_settlements, 0),
    
    # Additional Benefits (healthcare, pension, etc.)
    add_benefits = dplyr::coalesce(additional_benefits, 0),
    
    # Verify total matches
    calc_total = base_salary + add_earnings + add_benefits,
    
    # Check for any discrepancy (should be zero or very small)
    discrepancy = base::abs(calc_total - calc_total_compensation)
  )

# Validate that components sum correctly
max_discrepancy <- base::max(top_employees_components$discrepancy, na.rm = TRUE)
base::cat("\nValidation: Max discrepancy between components and total:", 
          scales::dollar(max_discrepancy), "\n")

if (max_discrepancy > 1) {
  base::warning("Discrepancy exceeds $1 - check component calculations")
}
```

---

## Step 4: Create Wide Tables for Each Component
```{r}
#| label: create-component-tables
#| cache: true

# =============================================================================
# CREATE WIDE FORMAT TABLES FOR EACH COMPONENT
# =============================================================================
# Purpose: Create separate tables showing Base, Additional Earnings, and Benefits
#          across fiscal years for each employee

# Get most recent employee info
most_recent_info <- top_employees_components |>
  dplyr::group_by(name_std) |>
  dplyr::slice_max(order_by = fiscal_year, n = 1) |>
  dplyr::ungroup() |>
  dplyr::select(name_std, name, department, job_title)

# --- TABLE 3A: Total Compensation (for reference) ---
total_comp_wide <- top_employees_components |>
  dplyr::select(name_std, fiscal_year, calc_total_compensation) |>
  tidyr::pivot_wider(
    names_from = fiscal_year,
    values_from = calc_total_compensation,
    names_prefix = "Tot_",
    values_fill = NA_real_
  ) |>
  dplyr::left_join(most_recent_info, by = "name_std")

# --- TABLE 3B: Base Salary (Regular Earnings) ---
base_salary_wide <- top_employees_components |>
  dplyr::select(name_std, fiscal_year, base_salary) |>
  tidyr::pivot_wider(
    names_from = fiscal_year,
    values_from = base_salary,
    names_prefix = "Base_",
    values_fill = NA_real_
  ) |>
  dplyr::left_join(most_recent_info, by = "name_std")

# --- TABLE 3C: Additional Earnings ---
add_earnings_wide <- top_employees_components |>
  dplyr::select(name_std, fiscal_year, add_earnings) |>
  tidyr::pivot_wider(
    names_from = fiscal_year,
    values_from = add_earnings,
    names_prefix = "AddE_",
    values_fill = NA_real_
  ) |>
  dplyr::left_join(most_recent_info, by = "name_std")

# --- TABLE 3D: Additional Benefits ---
add_benefits_wide <- top_employees_components |>
  dplyr::select(name_std, fiscal_year, add_benefits) |>
  tidyr::pivot_wider(
    names_from = fiscal_year,
    values_from = add_benefits,
    names_prefix = "Ben_",
    values_fill = NA_real_
  ) |>
  dplyr::left_join(most_recent_info, by = "name_std")

base::cat("\nTables created:\n")
base::cat("  Total Comp:", base::nrow(total_comp_wide), "employees\n")
base::cat("  Base Salary:", base::nrow(base_salary_wide), "employees\n")
base::cat("  Add Earnings:", base::nrow(add_earnings_wide), "employees\n")
base::cat("  Add Benefits:", base::nrow(add_benefits_wide), "employees\n")
```

---

## Step 5: Create Combined Breakdown Table
```{r}
#| label: create-combined-breakdown
#| cache: true

# =============================================================================
# CREATE COMBINED BREAKDOWN TABLE
# =============================================================================
# Purpose: Show all three components side-by-side for each fiscal year
# Format: Employee | FY | Base | Add Earnings | Benefits | Total

combined_breakdown <- top_employees_components |>
  dplyr::select(
    name_std, name, department, job_title,
    fiscal_year,
    base_salary, add_earnings, add_benefits, calc_total_compensation
  ) |>
  dplyr::rename(
    `Base Salary` = base_salary,
    `Add'l Earnings` = add_earnings,
    `Add'l Benefits` = add_benefits,
    `Total Comp` = calc_total_compensation
  ) |>
  dplyr::arrange(name, fiscal_year)

base::cat("Combined breakdown table:", base::nrow(combined_breakdown), "rows\n")
```

---

## Table 3A: Total Compensation by Fiscal Year

::: {.content-visible when-format="html"}
```{r}
#| label: display-total-comp-dt
#| cache: false

# =============================================================================
# HTML OUTPUT: TOTAL COMPENSATION - INTERACTIVE DT::DATATABLE
# =============================================================================

total_comp_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("Tot_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "Tot_"),
    dplyr::starts_with("Tot_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("FY"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  DT::datatable(
    caption = "Table 3A: Total Compensation by Fiscal Year",
    extensions = "Buttons",
    options = base::list(
      pageLength = 25,
      scrollX = TRUE,
      dom = "Bfrtip",
      buttons = base::list(
        "copy", "csv", "excel",
        base::list(extend = "print", title = "Total Compensation - City of Schertz")
      )
    ),
    rownames = FALSE,
    filter = "top"
  )
```
:::

::: {.content-visible when-format="typst"}
```{r}
#| label: display-total-comp-flextable
#| cache: false

# =============================================================================
# PDF/TYPST OUTPUT: TOTAL COMPENSATION - FLEXTABLE
# =============================================================================

total_comp_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("Tot_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "Tot_"),
    dplyr::starts_with("Tot_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("FY"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Table 3A: Total Compensation by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::add_footer_lines(
    "Total Comp = Base Salary + Additional Earnings + Additional Benefits"
  ) |>
  flextable::italic(part = "footer") |>
  flextable::fontsize(size = 10, part = "footer") |>
  ftExtra::colformat_md() |>
  flextable::autofit()
```
:::

---

## Table 3B: Base Salary (Regular Earnings) by Fiscal Year

::: {.content-visible when-format="html"}
```{r}
#| label: display-base-salary-dt
#| cache: false

# =============================================================================
# HTML OUTPUT: BASE SALARY - INTERACTIVE DT::DATATABLE
# =============================================================================

base_salary_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("Base_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "Base_"),
    dplyr::starts_with("Base_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("FY"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  DT::datatable(
    caption = "Table 3B: Base Salary (Regular Earnings) by Fiscal Year",
    extensions = "Buttons",
    options = base::list(
      pageLength = 25,
      scrollX = TRUE,
      dom = "Bfrtip",
      buttons = base::list(
        "copy", "csv", "excel",
        base::list(extend = "print", title = "Base Salary - City of Schertz")
      )
    ),
    rownames = FALSE,
    filter = "top"
  )
```
:::

::: {.content-visible when-format="typst"}
```{r}
#| label: display-base-salary-flextable
#| cache: false

# =============================================================================
# PDF/TYPST OUTPUT: BASE SALARY - FLEXTABLE
# =============================================================================

base_salary_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("Base_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "Base_"),
    dplyr::starts_with("Base_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("FY"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Table 3B: Base Salary (Regular Earnings) by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::add_footer_lines(
    "Base Salary = Regular Earnings only (excludes OT, bonuses, benefits)"
  ) |>
  flextable::italic(part = "footer") |>
  flextable::fontsize(size = 10, part = "footer") |>
  ftExtra::colformat_md() |>
  flextable::autofit()
```
:::

---

## Table 3C: Additional Earnings by Fiscal Year

::: {.content-visible when-format="html"}
```{r}
#| label: display-add-earnings-dt
#| cache: false

# =============================================================================
# HTML OUTPUT: ADDITIONAL EARNINGS - INTERACTIVE DT::DATATABLE
# =============================================================================

add_earnings_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("AddE_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "AddE_"),
    dplyr::starts_with("AddE_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("FY"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  DT::datatable(
    caption = "Table 3C: Additional Earnings (OT, Leave, Deployment, etc.) by Fiscal Year",
    extensions = "Buttons",
    options = base::list(
      pageLength = 25,
      scrollX = TRUE,
      dom = "Bfrtip",
      buttons = base::list(
        "copy", "csv", "excel",
        base::list(extend = "print", title = "Additional Earnings - City of Schertz")
      )
    ),
    rownames = FALSE,
    filter = "top"
  )
```
:::

::: {.content-visible when-format="typst"}
```{r}
#| label: display-add-earnings-flextable
#| cache: false

# =============================================================================
# PDF/TYPST OUTPUT: ADDITIONAL EARNINGS - FLEXTABLE
# =============================================================================

add_earnings_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("AddE_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "AddE_"),
    dplyr::starts_with("AddE_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("FY"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Table 3C: Additional Earnings by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::add_footer_lines(
    "Additional Earnings = Leave Payout + Overtime + Additional + Deployment + Arbitration"
  ) |>
  flextable::italic(part = "footer") |>
  flextable::fontsize(size = 10, part = "footer") |>
  ftExtra::colformat_md() |>
  flextable::autofit()
```
:::

---

## Table 3D: Additional Benefits by Fiscal Year

::: {.content-visible when-format="html"}
```{r}
#| label: display-add-benefits-dt
#| cache: false

# =============================================================================
# HTML OUTPUT: ADDITIONAL BENEFITS - INTERACTIVE DT::DATATABLE
# =============================================================================

add_benefits_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("Ben_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "Ben_"),
    dplyr::starts_with("Ben_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("FY"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  DT::datatable(
    caption = "Table 3D: Additional Benefits (Healthcare, Pension, etc.) by Fiscal Year",
    extensions = "Buttons",
    options = base::list(
      pageLength = 25,
      scrollX = TRUE,
      dom = "Bfrtip",
      buttons = base::list(
        "copy", "csv", "excel",
        base::list(extend = "print", title = "Additional Benefits - City of Schertz")
      )
    ),
    rownames = FALSE,
    filter = "top"
  )
```
:::

::: {.content-visible when-format="typst"}
```{r}
#| label: display-add-benefits-flextable
#| cache: false

# =============================================================================
# PDF/TYPST OUTPUT: ADDITIONAL BENEFITS - FLEXTABLE
# =============================================================================

add_benefits_wide |>
  dplyr::select(
    name, department, job_title,
    dplyr::starts_with("Ben_")
  ) |>
  dplyr::rename_with(
    ~ stringr::str_remove(.x, "Ben_"),
    dplyr::starts_with("Ben_")
  ) |>
  dplyr::arrange(dplyr::desc(dplyr::coalesce(FY25, FY24, FY23, FY22, FY21, FY20))) |>
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("FY"),
      ~ scales::dollar(.x, accuracy = 1)
    )
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Table 3D: Additional Benefits by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::add_footer_lines(
    "Additional Benefits = Healthcare, pension contributions, etc. | FY20-FY22 benefits are imputed"
  ) |>
  flextable::italic(part = "footer") |>
  flextable::fontsize(size = 10, part = "footer") |>
  ftExtra::colformat_md() |>
  flextable::autofit()
```
:::

---

## Component Summary by Fiscal Year
```{r}
#| label: component-summary
#| cache: true

# =============================================================================
# AGGREGATE COMPONENT SUMMARY BY FISCAL YEAR
# =============================================================================
# Purpose: Show how much of total compensation goes to each component

component_summary <- top_employees_components |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    n_employees = dplyr::n(),
    total_base = base::sum(base_salary, na.rm = TRUE),
    total_add_earnings = base::sum(add_earnings, na.rm = TRUE),
    total_benefits = base::sum(add_benefits, na.rm = TRUE),
    total_comp = base::sum(calc_total_compensation, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    # Calculate percentages
    pct_base = total_base / total_comp * 100,
    pct_add_earnings = total_add_earnings / total_comp * 100,
    pct_benefits = total_benefits / total_comp * 100
  )

component_summary |>
  dplyr::mutate(
    total_base = scales::dollar(total_base, accuracy = 1),
    total_add_earnings = scales::dollar(total_add_earnings, accuracy = 1),
    total_benefits = scales::dollar(total_benefits, accuracy = 1),
    total_comp = scales::dollar(total_comp, accuracy = 1),
    pct_base = base::sprintf("%.1f%%", pct_base),
    pct_add_earnings = base::sprintf("%.1f%%", pct_add_earnings),
    pct_benefits = base::sprintf("%.1f%%", pct_benefits)
  ) |>
  dplyr::select(
    fiscal_year, n_employees,
    total_base, pct_base,
    total_add_earnings, pct_add_earnings,
    total_benefits, pct_benefits,
    total_comp
  ) |>
  dplyr::rename(
    `FY` = fiscal_year,
    `N` = n_employees,
    `Base Salary` = total_base,
    `% Base` = pct_base,
    `Add'l Earnings` = total_add_earnings,
    `% Add'l` = pct_add_earnings,
    `Benefits` = total_benefits,
    `% Ben` = pct_benefits,
    `Total Comp` = total_comp
  ) |>
  flextable::flextable() |>
  flextable::add_header_lines("Component Summary: All Top Employees by Fiscal Year") |>
  flextable::color(i = 1, color = "darkblue", part = "header") |>
  flextable::italic(i = 1, part = "header") |>
  flextable::align(i = 1, align = "left", part = "header") |>
  flextable::fontsize(i = 1, size = 14, part = "header") |>
  flextable::bg(i = 1, bg = "white", part = "header") |>
  flextable::bg(i = 2, bg = "palegreen", part = "header") |>
  flextable::add_footer_lines(
    "Percentages show each component's share of total compensation"
  ) |>
  flextable::italic(part = "footer") |>
  flextable::fontsize(size = 10, part = "footer") |>
  ftExtra::colformat_md() |>
  flextable::autofit()
```

---

## Object Reference

| Object Name | Description | Rows | Columns |
|-------------|-------------|------|---------|
| `all_data` | Full compensation dataset (cached) | `r base::nrow(all_data)` | `r base::ncol(all_data)` |
| `top_employees_full` | Filtered with all earnings columns | `r base::nrow(top_employees_full)` | `r base::ncol(top_employees_full)` |
| `top_employees_components` | With calculated component totals | `r base::nrow(top_employees_components)` | `r base::ncol(top_employees_components)` |
| `total_comp_wide` | Wide format: Total comp by FY | `r base::nrow(total_comp_wide)` | `r base::ncol(total_comp_wide)` |
| `base_salary_wide` | Wide format: Base salary by FY | `r base::nrow(base_salary_wide)` | `r base::ncol(base_salary_wide)` |
| `add_earnings_wide` | Wide format: Add'l earnings by FY | `r base::nrow(add_earnings_wide)` | `r base::ncol(add_earnings_wide)` |
| `add_benefits_wide` | Wide format: Benefits by FY | `r base::nrow(add_benefits_wide)` | `r base::ncol(add_benefits_wide)` |
| `combined_breakdown` | Long format: All components | `r base::nrow(combined_breakdown)` | `r base::ncol(combined_breakdown)` |
| `component_summary` | Aggregate by FY | `r base::nrow(component_summary)` | `r base::ncol(component_summary)` |

---

## Data Notes

| Component | Includes |
|-----------|----------|
| **Base Salary** | Regular Earnings only |
| **Add'l Earnings** | Leave Payout + Overtime + Additional + Deployment + Arbitration |
| **Add'l Benefits** | Healthcare, pension, other employer contributions |
| **Total Comp** | Base + Add'l Earnings + Add'l Benefits |

- **FY20-FY22**: Benefits were imputed based on FY23-FY25 ratios (not in source files)
- **Blank cells**: Employee not in position during that fiscal year

---

*Report generated: `r base::format(base::Sys.time(), "%B %d, %Y at %I:%M %p")`*
*Data source: City of Schertz Employee Compensation Records*
