---
title: "Schertz Compensation Pipeline - Simple Explicit Version"
subtitle: "Manual Column Specification Workflow"
author: "Dan Swart, CPA (ret)"
date: today
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

options(scipen = 999)

flextable::set_flextable_defaults(

font.size = 12, 
  font.family = "Cabin",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4
)
```

# PHASE 1: DISCOVERY

Run this section first to see exact column names after `clean_names()`.

## Step 1: Define Your Files

**Edit these filenames to match your actual files:**

```{r}
#| label: define-files

# =============================================================================
# EDIT THESE FILENAMES TO MATCH YOUR ACTUAL FILES
# =============================================================================

files <- base::list(
  FY20 = "FY20 Earnings.xlsx",
  FY21 = "FY21 Earnings.xlsx",
  FY22 = "FY22 Earnings.xlsx",
  FY23 = "FY23Employee_Comp_FY23.xlsx",
  FY24 = "FY24Employee_Comp_FY24_All.xlsx"
)

data_folder <- here::here("data_raw")
```

## Step 2: Read Files and Show Column Names

This reads each file, applies `clean_names()`, and shows you the exact column names to use.

```{r}
#| label: read-and-show-columns

# Read each file and capture column names
file_info <- purrr::map_dfr(base::names(files), function(fy) {
  filepath <- base::file.path(data_folder, files[[fy]])
  
  if (!base::file.exists(filepath)) {
    return(dplyr::tibble(
      fiscal_year = fy,
      filename = files[[fy]],
      status = "FILE NOT FOUND",
      row_count = NA_integer_,
      columns = NA_character_
    ))
  }
  
  df <- readxl::read_excel(filepath) |>
    janitor::clean_names()
  
  dplyr::tibble(
    fiscal_year = fy,
    filename = files[[fy]],
    status = "OK",
    row_count = base::nrow(df),
    columns = base::paste(base::names(df), collapse = ", ")
  )
})

# Display summary
file_info |>
  dplyr::select(fiscal_year, filename, status, row_count) |>
  flextable::flextable() |>
  flextable::autofit() |>
  flextable::set_caption("File Import Summary")
```

## Step 3: Exact Column Names by File

**Copy these exact column names into Phase 2 below.**

```{r}
#| label: show-exact-columns

# Show columns for each file in a usable format
for (fy in base::names(files)) {
  filepath <- base::file.path(data_folder, files[[fy]])
  
  if (base::file.exists(filepath)) {
    df <- readxl::read_excel(filepath) |>
      janitor::clean_names()
    
    base::cat("\n")
    base::cat("# =========================================================\n")
    base::cat("#", fy, "- Columns after clean_names():\n")
    base::cat("# =========================================================\n")
    
    cols <- base::names(df)
    for (i in base::seq_along(cols)) {
      base::cat("#", base::sprintf("%2d", i), ":", cols[i], "\n")
    }
  }
}
```

------------------------------------------------------------------------

# ⛔ STOP HERE FIRST TIME

Run Phase 1 above. Then use the column names shown to fill in the exact mappings in Phase 2 below.

------------------------------------------------------------------------

# PHASE 2: PROCESSING

**After reviewing Phase 1 output, edit the column mappings below.**

## Step 4: Define Exact Column Mappings

**Edit these to match the exact column names from Phase 1:**

```{r}
#| label: define-column-mappings

# =============================================================================
# COLUMN MAPPINGS - EDIT THESE TO MATCH YOUR FILES EXACTLY
# =============================================================================
# Use the exact column names shown in Phase 1 (after clean_names)
# Set to NA if that column doesn't exist in a given year

column_maps <- base::list(
  
  # ----- FY20 -----
  FY20 = base::list(
    # Identity columns
    name = "name",                    # Single name column (or NA if split)
    last_name = NA,                   # NA if using single name column
    first_name = NA,                  # NA if using single name column
    department = "department",
    job_title = "job_title",
    hire_date = "hire_date",
    termination_date = "termination_date",
    
    # Earnings columns (set to NA if not present)
    leave_payout = NA,
    regular_earnings = "regular_earnings",
    overtime_earnings = "overtime_earnings",
    additional_earnings = "additional_earnings1",
    deployment_earnings = NA,
    arbitration_settlements = NA,
    
    # Benefits (set to NA if not present)
    additional_benefits = NA,
    
    # Totals for validation
    total_earnings = "total_earnings",
    total_compensation = NA           # NA if no total_comp column
  ),
  
  # ----- FY21 -----
  FY21 = base::list(
    status = "status",
    name = "name",
    last_name = NA,
    first_name = NA,
    department = "home_department",
    job_title = "job_title",
    hire_date = "hire_date",
    termination_date = "termination_date",
    
    
    leave_payout = NA,
    regular_earnings = "regular_earnings",
    overtime_earnings = "overtime_earnings",
    additional_earnings = "additional_earnings1",
    deployment_earnings = NA,
    arbitration_settlements = NA,
    
    additional_benefits = NA,
    
    total_earnings = "total_earnings",
    total_compensation = NA
  ),
  
  # ----- FY22 -----
  FY22 = base::list(
    status = "status",
    name = "name",
    last_name = NA,
    first_name = NA,
    department = "department",
    job_title = "job_title",
    hire_date = "hire_date",
    termination_date = "termination_date",
    
    leave_payout = NA,
    regular_earnings = "regular_earnings",
    overtime_earnings = "overtime_earnings",
    additional_earnings = "additional_earnings1",
    deployment_earnings = NA,
    arbitration_settlements = NA,
    
    additional_benefits = NA,
    
    total_earnings = "total_earnings",
    total_compensation = NA
  ),
  
  # ----- FY23 -----
  FY23 = base::list(
    name = NA,                                    # NA - using split names
    last_name = "last_name",
    first_name = "first_name",
    department = "department",
    job_title = "job_title",
    employee_category = "employee_category",
    hire_date = "hire_date",
    separation_date = "separation_date",
    fy23_annual_salary1 = "fy23_annual_salary1",
    
    leave_payout = "fy23_leave_payout2",
    regular_earnings = "fy23_regular_earnings3",
    overtime_earnings = "fy23_overtime_earnings4",
    additional_earnings = "fy23_additional_earnings5",
    deployment_earnings = "fy23_deployment_earnings6",
    arbitration_settlements = "arbitration_settlements",
    additional_benefits = "fy23_additional_benefits8",
    total_earnings = "fy23_total_earnings7",
    total_compensation = "fy23_total_compensation9"
  ),
  
  # ----- FY24 -----
  FY24 = base::list(
    name = NA,
    last_name = "last_name",
    first_name = "first_name",
    department = "home_department",
    job_title = "job_title",
    employee_category = "employee_category",
    hire_date = "hire_date",
    separation_date = "separation_date",
    fy24_annual_salary1 = "fy23_annual_salary1",
    
    leave_payout = "fy24_leave_payout2",
    regular_earnings = "fy24_regular_earnings3",
    overtime_earnings = "fy24_overtime_earnings4",
    additional_earnings = "fy24_additional",
    deployment_earnings = "fy24_deployment",
    arbitration_settlements = "arbitration_settlements",
    additional_benefits = "fy24_additional_1",
    total_earnings = "fy24_total_earnings7",
    total_compensation = "fy24_total_compensation9"
  )
)
```

## Step 5: Utility Functions

```{r}
#| label: utility-functions

# Parse currency to numeric
parse_currency <- function(x) {
  if (base::is.numeric(x)) return(x)
  x <- stringr::str_remove_all(x, "[$,\\s]")
  base::suppressWarnings(base::as.numeric(x))
}

# Fix invisible spaces
fix_spaces <- function(x) {
  if (!base::is.character(x)) return(x)
  stringr::str_squish(x)
}

# Standardize name for matching
standardize_name <- function(x) {
  x <- fix_spaces(x)
  x <- stringr::str_to_upper(x)
  stringr::str_remove_all(x, "[^A-Z ]")
}

# Safe column extract - returns 0 if column is NA or doesn't exist
safe_extract <- function(df, col_name) {
  if (base::is.na(col_name) || !col_name %in% base::names(df)) {
    return(base::rep(0, base::nrow(df)))
  }
  result <- parse_currency(df[[col_name]])
  dplyr::coalesce(result, 0)
}
```

## Step 6: Process Each File

```{r}
#| label: process-files

process_file <- function(fy, filename, col_map) {
  
  filepath <- base::file.path(data_folder, filename)
  
  if (!base::file.exists(filepath)) {
    base::warning("File not found: ", filepath)
    return(NULL)
  }
  
  df <- readxl::read_excel(filepath) |>
    janitor::clean_names()
  
  # Handle name - either single column or split
  if (!base::is.na(col_map$name)) {
    # Single name column
    name_values <- fix_spaces(df[[col_map$name]])
    last_name <- NA_character_
    first_name <- NA_character_
  } else {
    # Split name columns
    last_name <- fix_spaces(df[[col_map$last_name]])
    first_name <- fix_spaces(df[[col_map$first_name]])
    name_values <- base::paste(last_name, first_name, sep = ", ")
  }
  
  # Build result
  result <- dplyr::tibble(
    fiscal_year = fy,
    name = name_values,
    name_std = standardize_name(name_values),
    last_name = last_name,
    first_name = first_name,
    department = fix_spaces(df[[col_map$department]]),
    job_title = fix_spaces(df[[col_map$job_title]]),
    
    # Earnings
    leave_payout = safe_extract(df, col_map$leave_payout),
    regular_earnings = safe_extract(df, col_map$regular_earnings),
    overtime_earnings = safe_extract(df, col_map$overtime_earnings),
    additional_earnings = safe_extract(df, col_map$additional_earnings),
    deployment_earnings = safe_extract(df, col_map$deployment_earnings),
    arbitration_settlements = safe_extract(df, col_map$arbitration_settlements),
    additional_benefits = safe_extract(df, col_map$additional_benefits),
    
    # Source totals for validation
    source_total_earnings = safe_extract(df, col_map$total_earnings),
    source_total_compensation = safe_extract(df, col_map$total_compensation)
  )
  
  # Calculate totals
  result <- result |>
    dplyr::mutate(
      calc_total_earnings = leave_payout + regular_earnings + overtime_earnings +
                            additional_earnings + deployment_earnings + arbitration_settlements,
      calc_total_compensation = calc_total_earnings + additional_benefits,
      employee_key = base::paste(name_std, department, sep = "_")
    )
  
  # For files without total_compensation, use total_earnings
  result <- result |>
    dplyr::mutate(
      source_total_compensation = dplyr::if_else(
        source_total_compensation == 0 & source_total_earnings > 0,
        source_total_earnings,
        source_total_compensation
      )
    )
  
  return(result)
}

# Process all files
all_data <- purrr::map2_dfr(
  base::names(files),
  files,
  function(fy, filename) {
    process_file(fy, filename, column_maps[[fy]])
  }
)

base::cat("Processed", base::nrow(all_data), "total rows\n")
base::cat("Fiscal years:", base::paste(base::unique(all_data$fiscal_year), collapse = ", "), "\n")
```

## Step 7: Validation - Compare Calculated vs Source Totals

```{r}
#| label: validation

validation <- all_data |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    rows = dplyr::n(),
    calc_total = base::sum(calc_total_compensation, na.rm = TRUE),
    source_total = base::sum(source_total_compensation, na.rm = TRUE),
    difference = calc_total - source_total,
    .groups = "drop"
  ) |>
  dplyr::mutate(
    status = dplyr::if_else(base::abs(difference) <= 1, "PASS", "FAIL")
  )

validation |>
  dplyr::mutate(
    calc_total = scales::dollar(calc_total, accuracy = 1),
    source_total = scales::dollar(source_total, accuracy = 1),
    difference = scales::dollar(difference, accuracy = 0.01)
  ) |>
  flextable::flextable() |>
  flextable::bg(i = ~ status == "FAIL", bg = "#ffcccc") |>
  flextable::bg(i = ~ status == "PASS", bg = "#ccffcc") |>
  flextable::autofit() |>
  flextable::set_caption("Validation: Calculated vs Source Totals")
```

## Step 8: Component Breakdown by Year

```{r}
#| label: component-breakdown

component_summary <- all_data |>
  dplyr::group_by(fiscal_year) |>
  dplyr::summarise(
    `Leave Payout` = base::sum(leave_payout, na.rm = TRUE),
    `Regular` = base::sum(regular_earnings, na.rm = TRUE),
    `Overtime` = base::sum(overtime_earnings, na.rm = TRUE),
    `Additional` = base::sum(additional_earnings, na.rm = TRUE),
    `Deployment` = base::sum(deployment_earnings, na.rm = TRUE),
    `Arbitration` = base::sum(arbitration_settlements, na.rm = TRUE),
    `Benefits` = base::sum(additional_benefits, na.rm = TRUE),
    `TOTAL` = base::sum(calc_total_compensation, na.rm = TRUE),
    .groups = "drop"
  )

component_summary |>
  dplyr::mutate(
    dplyr::across(dplyr::where(is.numeric), ~ scales::dollar(.x, accuracy = 1))
  ) |>
  flextable::flextable() |>
  flextable::autofit() |>
  flextable::set_caption("Component Breakdown by Fiscal Year")
```

## Step 9: Convert to Long Format and Export

```{r}
#| label: export

# Pivot to long format
final_long <- all_data |>
  dplyr::select(
    fiscal_year, name, name_std, last_name, first_name,
    department, job_title, employee_key,
    leave_payout, regular_earnings, overtime_earnings,
    additional_earnings, deployment_earnings, arbitration_settlements,
    additional_benefits
  ) |>
  tidyr::pivot_longer(
    cols = c(leave_payout, regular_earnings, overtime_earnings,
             additional_earnings, deployment_earnings, arbitration_settlements,
             additional_benefits),
    names_to = "earnings_type",
    values_to = "amount"
  ) |>
  dplyr::mutate(
    earnings_type = dplyr::case_when(
      earnings_type == "leave_payout" ~ "Leave Payout",
      earnings_type == "regular_earnings" ~ "Regular",
      earnings_type == "overtime_earnings" ~ "Overtime",
      earnings_type == "additional_earnings" ~ "Additional",
      earnings_type == "deployment_earnings" ~ "Deployment",
      earnings_type == "arbitration_settlements" ~ "Arbitration",
      earnings_type == "additional_benefits" ~ "Benefits",
      TRUE ~ earnings_type
    )
  )

# Export
output_dir <- here::here("data-clean")
if (!base::dir.exists(output_dir)) base::dir.create(output_dir, recursive = TRUE)

base::saveRDS(final_long, base::file.path(output_dir, "compensation_long.rds"))
readr::write_csv(final_long, base::file.path(output_dir, "compensation_long.csv"))

base::cat("Exported to:", output_dir, "\n")
base::cat("- compensation_long.rds\n")
base::cat("- compensation_long.csv\n")
```

------------------------------------------------------------------------

# Summary

| Year | Status | Notes |
|------|--------|-------|
| FY20-22 | Single `name` column, no benefits | |
| FY23-24 | Split `last_name`/`first_name`, has benefits | |

If validation shows FAIL, check the column mappings in Step 4.
