---
title: "City of Schertz — FY 2025-26 Proposed Budget: Compensation Increase Analysis"
subtitle: ""
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true     # Set to 'false' for drafts. Override if you're NOT sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false

prefer-html: true

params:
  data_updated: !expr format(Sys.Date(), "%B %Y")
---


```{r}
#| label: setup
#| include: false

# Capture current filename for use in document
file_name <- if (!base::is.null(knitr::current_input())) {
  paste0(tools::file_path_sans_ext(knitr::current_input()), ".qmd")
} else {
  "interactive_session.qmd"
}

file_stem <- tools::file_path_sans_ext(file_name)



# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED

# Download button helpers live here; knitr::is_html_output() guard is inside
# the helpers so they silently no-op when rendering to typst
source(here::here("R", "download_button_helpers.R"))

# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  theme_fun = base::identity,          # <-- add this
  post_process_html = identity,
  post_process_docx = identity
)

# Standard table styler: blue italic title row + palegreen column header row.
# Defined here so all QMDs that source this setup get the same presentation
# without depending on census_helpers.R for a purely cosmetic function.
ds_style_flextable <- function(ft, title = NULL) {
  if (!base::is.null(title)) {
    ft <- flextable::add_header_lines(ft, values = title)
    ft <- flextable::color(ft, i = 1, part = "header", color = "blue")
    ft <- flextable::italic(ft, i = 1, part = "header")
    ft <- flextable::align(ft, i = 1, part = "header", align = "left")
    ft <- flextable::fontsize(ft, i = 1, part = "header", size = 12)
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "white")
    ft <- flextable::bg(ft, i = 2, part = "header", bg = "palegreen")
  } else {
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "palegreen")
  }
  ft <- flextable::autofit(ft)
  ft
}


# Flextable built-in themes:
  # flextable::theme_alafoli()	|>  # BLAH
  # flextable::theme_apa()  # THIS IS NICE
  # flextable::theme_booktabs() |>  # NICE, MORE COMPACT
  # flextable::theme_box() |>   # OK, INCLUDES CELL BORDERS
  # flextable::theme_tron() |>  # 'DARK MODE' BLUE TEXT
  # flextable::theme_tron_legacy() |>   # 'DARK MODE' YELLOW TEXT
  # flextable::theme_vader() |>    # 'DARK MODE' WHITE TEXT
  # flextable::theme_vanilla() |>   # NOT SPECIAL
  # flextable::theme_zebra()	|>
  #


# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 28,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 22,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 18,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 18,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)


# claude <- ellmer::chat_anthropic(model = "claude-opus-4-6-20260205")
  
# Set seed for reproducibility
base::set.seed(123)

```




```{r}
#| label: helper-functions
#| include: false

# No external packages required for this document — all computations are
# in base R.  Formatting helpers are inline below.

fmt_dollar <- function(x, digits = 0) {
  paste0("$", formatC(x, format = "f", digits = digits, big.mark = ","))
}

fmt_pct <- function(x, digits = 1) {
  paste0(formatC(x * 100, format = "f", digits = digits), "%")
}



```


\

#### **Filename:**

::: {.content-visible when-format="html"}
[`r file_name`]{style="word-break: break-all; font-size: 1.00em;"}
:::

```{=typst}
#let wrap_underscores(s) = s.replace("_", "_\u{200B}")
#text(size: 16pt)[#raw(wrap_underscores("`r file_name`"))]
```

\
\

---

## Background and Source

This analysis is drawn from a careful reading of the **City of Schertz FY 2025-26
Proposed Budget** document.  All figures come directly from the budget narratives,
department summary tables, and the detailed account-level budget worksheets
(ExpCategory 511 — Personnel Services totals).  No automated parsing routines
were used; every number below was read from the document.

The budget presents expenditure data across four columns most relevant to this
analysis:

```{r}
#| label: budget-columns

cols_df <- data.frame(
  Column = c("2024-25 Budget", "2024-25 YTD Activity",
             "2024-25 Projected", "2025-26 Budget (AB)"),
  Description = c(
    "Original adopted budget for the current year",
    "Actual spending through 8/31/2025",
    "Staff's year-end estimate — the baseline for comparison",
    "Proposed budget — the subject of this analysis"
  ),
  stringsAsFactors = FALSE
)

flextable::flextable(cols_df) |>
  flextable::set_header_labels(Column = "Column", Description = "Description") |>
  flextable::bold(i = 3:4, part = "body") |>
  flextable::bg(i = 3:4, bg = "#FFFACD", part = "body") |>
  ds_style_flextable(title = "Budget Document — Expenditure Columns Used in This Analysis")
```

```{r}
#| label: download-budget-col-names

ds_download_data_btn(cols_df, "25_26_budget_cols_described")

```

\
\

The comparison used throughout is **2024-25 Projected vs. 2025-26 Proposed
Budget**, which is the standard "apples-to-apples" measure of the incremental cost of the new budget.

---

## Question 1: What is the Total Increase in Compensation Requested?

### Explicit Statement from the Budget Document

The budget message contains one explicit, composite dollar figure for the
compensation policy changes:

> *"Citywide the budget funds the Merit program for non-uniformed public safety
> personnel and the Step program for uniformed public safety. These programs move
> employees within their pay range but does not affect the starting wages of new
> employees. Non-uniformed employees will be eligible for a 0–4% increase based
> on performance and uniformed public safety employees could receive a 4% increase
> based on the tenure schedule. **These changes are planned to take effect
> midyear and cost \$784,000.**

This **\$784,000** is the mid-year (six-month) cost of merit and step increases
for **existing employees, across all funds**.

The budget also states that the medical insurance plan was changed to a new
provider in a way that **prevents any increase in premiums for 2026**, so the
benefits cost impact for health insurance is essentially flat in the proposed
budget.

### New Positions Added (FY 2025-26)

Beyond the merit/step program, 26 new positions were authorized:

```{r}
#| label: new-positions

new_pos <- data.frame(
  Fund      = c("General Fund", "Water & Sewer Fund", "EMS Fund",
                "Drainage Fund", "TOTAL"),
  Positions = c(10, 6, 4, 6, 26),
  Detail    = c(
    "6 Firefighters (Fire Station 4 staffing); 1 Asst. City Secretary; 1 Athletics Coordinator; 1 Recreation Specialist; 1 Senior Center Supervisor",
    "3 Worker I's, 1 Worker II, 1 Crew Leader (new crew); 1 Worker I (floater)",
    "4 EMTs (Medic 8 and Medic 9 peak-hour coverage)",
    "3 Worker I's, 1 Worker II, 1 Crew Leader (new crew); 1 Floodplain Manager",
    ""
  ),
  stringsAsFactors = FALSE
)

n_total <- base::nrow(new_pos)

flextable::flextable(new_pos) |>
  flextable::set_header_labels(
    Fund = "Fund", Positions = "Positions", Detail = "Detail"
  ) |>
  flextable::align(j = "Positions", align = "center", part = "all") |>
  flextable::bold(i = n_total, part = "body") |>
  flextable::bg(i = n_total, bg = "#DDEEFF", part = "body") |>
  flextable::hline(i = n_total - 1,
                   border = officer::fp_border(color = "navy", width = 2),
                   part = "body") |>
  ds_style_flextable(title = "New Positions Proposed — FY 2025-26 (All Funds)")
```



```{r}
#| label: download-new-positions-added

ds_download_data_btn(new_pos, "25_26_budget_new_positions_added")

```

\
\

### Compiled Personnel Services Totals by Department

The table below was compiled manually from the **ExpCategory: 511 —
Personnel Services Total** rows in the budget worksheets.  "Personnel
Services" in Schertz's chart of accounts (Category 511) encompasses:
base wages, overtime, longevity pay, clothing/certification allowances,
employer FICA, employer TMRS (retirement), employer health insurance,
workers' compensation insurance, uniforms, training and travel, and
miscellaneous employee-related costs.  **Benefits are therefore
included.**

```{r}
#| tbl-cap: "Personnel Services (ExpCategory 511) by Department — FY 2024-25 Projected vs. FY 2025-26 Proposed Budget"

# --------------------------------------------------------------------------
# All figures read directly from the Schertz FY 2025-26 Proposed Budget PDF.
# Columns: department name, fund, 2024-25 projected, 2025-26 budget.
# --------------------------------------------------------------------------

dept_data <- data.frame(
  Fund = c(
    # General Fund departments (Fund 101)
    rep("General Fund (101)", 24),
    # Water & Sewer Fund (Fund 202)
    "Water & Sewer (202)",
    "Water & Sewer (202)",
    # Drainage Fund (Fund 202/579)
    "Drainage (202)",
    # EMS Fund (Fund 203)
    "EMS (203)",
    # Special Revenue
    "Special Revenue (615)"
  ),
  Dept_ID = c(
    150, 151, 152, 153, 154, 156, 164, 170, 173,
    256, 257, 258, 259, 359, 460, 463, 464,
    665, 853, 866, 868, 872, 877, 878,
    176, 575,
    579,
    280,
    609
  ),
  Department = c(
    "City Council",
    "City Manager",
    "Municipal Court",
    "311 / Customer Relations",
    "Planning & Zoning",
    "Economic Development",
    "City Secretary",
    "Public Affairs",
    "Engineering",
    "Police",
    "Fire Rescue",
    "Inspections",
    "Neighborhood Services",
    "Streets",
    "Parks, Recreation & Community Svcs",
    "Swim Pool",
    "Event Facilities",
    "Library",
    "Information Technology",
    "Human Resources",
    "Finance",
    "Purchasing & Asset Management",
    "Fleet Services",
    "Facility Services",
    # W&S
    "Utility Billing",
    "Water & Sewer Administration",
    # Drainage
    "Drainage Operations",
    # EMS
    "Emergency Medical Services",
    # Special Revenue
    "Historical Committee"
  ),
  FY2425_Proj = c(
    # General Fund — read from ExpCategory 511 Total, column 7
     69525.00,    # City Council (150)
   1067488.00,    # City Manager (151)
    383922.00,    # Municipal Court (152)
    137060.00,    # 311/Customer Relations (153)
    463444.00,    # Planning & Zoning (154)
    543724.00,    # Economic Development (156)
    251676.00,    # City Secretary (164)
    355153.00,    # Public Affairs (170)
   1230173.00,    # Engineering (173)
  12506800.00,    # Police (256)
   8329400.00,    # Fire Rescue (257)
   1388161.00,    # Inspections (258)
    523406.00,    # Neighborhood Services (259)
   1237425.00,    # Streets (359)
   1411968.30,    # Parks (460)
     22300.00,    # Swim Pool (463)
    536887.00,    # Event Facilities (464)
   1204714.30,    # Library (665)
   1394110.00,    # IT (853)
    759136.60,    # Human Resources (866)
    819292.76,    # Finance (868)
    300511.08,    # Purchasing & Asset Mgt (872)
    862355.00,    # Fleet Services (877)
   1149335.00,    # Facility Services (878)
    # Water & Sewer
    774837.00,    # Utility Billing (176)
   2801020.00,    # W&S Administration (575)
    # Drainage
    597942.00,    # Drainage (579)
    # EMS
   9218278.73,    # EMS (203-280)
    # Special Revenue
      6350.00     # Historical Committee (615-609)
  ),
  FY2526_Budget = c(
    # General Fund — read from ExpCategory 511 Total, column 8
    109839.00,    # City Council (150)
   1017376.00,    # City Manager (151)
    373139.00,    # Municipal Court (152)
    150518.00,    # 311/Customer Relations (153)
    456245.00,    # Planning & Zoning (154)
    554647.00,    # Economic Development (156)
    336046.00,    # City Secretary (164)
    348964.00,    # Public Affairs (170)
   1272482.00,    # Engineering (173)
  12249146.00,    # Police (256)
   8829785.00,    # Fire Rescue (257)
   1390202.00,    # Inspections (258)
    521791.00,    # Neighborhood Services (259)
   1323243.00,    # Streets (359)
   1584494.30,    # Parks (460)
     24575.00,    # Swim Pool (463)
    521117.00,    # Event Facilities (464)
   1198296.00,    # Library (665)
   1367412.00,    # IT (853)
    771430.00,    # Human Resources (866)
    820727.00,    # Finance (868)
    302455.00,    # Purchasing & Asset Mgt (872)
    864312.00,    # Fleet Services (877)
   1103336.00,    # Facility Services (878)
    # Water & Sewer
    788936.00,    # Utility Billing (176)
   3462953.00,    # W&S Administration (575)
    # Drainage
   1018261.00,    # Drainage (579)
    # EMS
   9814819.00,    # EMS (203-280)
    # Special Revenue
      6350.00     # Historical Committee (615-609)
  ),
  stringsAsFactors = FALSE
)

dept_data$Change  <- dept_data$FY2526_Budget - dept_data$FY2425_Proj
dept_data$Pct_Chg <- dept_data$Change / dept_data$FY2425_Proj

disp_dept <- data.frame(
  Fund          = dept_data$Fund,
  Department    = dept_data$Department,
  FY2425_Proj   = fmt_dollar(dept_data$FY2425_Proj),
  FY2526_Budget = fmt_dollar(dept_data$FY2526_Budget),
  Change        = ifelse(dept_data$Change >= 0,
                         paste0("+", fmt_dollar(dept_data$Change)),
                         paste0("(", fmt_dollar(base::abs(dept_data$Change)), ")")),
  Pct_Chg       = ifelse(dept_data$Pct_Chg >= 0,
                         paste0("+", fmt_pct(dept_data$Pct_Chg)),
                         paste0("(", fmt_pct(base::abs(dept_data$Pct_Chg)), ")")),
  stringsAsFactors = FALSE
)

dept_rows_neg <- base::which(dept_data$Change < 0)
dept_rows_pos <- base::which(dept_data$Change > 0)

flextable::flextable(disp_dept) |>
  flextable::set_header_labels(
    Fund = "Fund", Department = "Department",
    FY2425_Proj = "FY 2024-25 Projected", FY2526_Budget = "FY 2025-26 Budget",
    Change = "Change ($)", Pct_Chg = "Change (%)"
  ) |>
  flextable::align(j = c("FY2425_Proj","FY2526_Budget","Change","Pct_Chg"),
                   align = "right", part = "all") |>
  flextable::color(i = dept_rows_neg, j = "Change", color = "firebrick", part = "body") |>
  flextable::color(i = dept_rows_pos, j = "Change", color = "darkgreen", part = "body") |>
  ds_style_flextable(title = "Personnel Services (ExpCategory 511) by Department — FY 2024-25 Projected vs. FY 2025-26 Proposed")
```




```{r}
#| label: download-personnel_services_by_dept

ds_download_data_btn(disp_dept, "25_26_budget_personnel_services_by_dept")

```

\
\


```{r}
#| tbl-cap: "Personnel Services Subtotals by Fund"

# Subtotals by fund
fund_summary <- do.call(rbind, lapply(unique(dept_data$Fund), function(f) {
  sub <- dept_data[dept_data$Fund == f, ]
  data.frame(Fund = f,
             FY2425_Proj   = base::sum(sub$FY2425_Proj),
             FY2526_Budget = base::sum(sub$FY2526_Budget),
             Change        = base::sum(sub$Change),
             stringsAsFactors = FALSE)
}))

grand <- data.frame(
  Fund          = "GRAND TOTAL — All Funds",
  FY2425_Proj   = base::sum(dept_data$FY2425_Proj),
  FY2526_Budget = base::sum(dept_data$FY2526_Budget),
  Change        = base::sum(dept_data$Change),
  stringsAsFactors = FALSE
)
fund_summary          <- rbind(fund_summary, grand)
fund_summary$Pct_Chg  <- fund_summary$Change / fund_summary$FY2425_Proj
n_rows                <- base::nrow(fund_summary)

fs_disp <- data.frame(
  Fund          = fund_summary$Fund,
  FY2425_Proj   = fmt_dollar(fund_summary$FY2425_Proj),
  FY2526_Budget = fmt_dollar(fund_summary$FY2526_Budget),
  Change        = ifelse(fund_summary$Change >= 0,
                         paste0("+", fmt_dollar(fund_summary$Change)),
                         paste0("(", fmt_dollar(base::abs(fund_summary$Change)), ")")),
  Pct_Chg       = ifelse(fund_summary$Pct_Chg >= 0,
                         paste0("+", fmt_pct(fund_summary$Pct_Chg)),
                         paste0("(", fmt_pct(base::abs(fund_summary$Pct_Chg)), ")")),
  stringsAsFactors = FALSE
)

flextable::flextable(fs_disp) |>
  flextable::set_header_labels(
    Fund = "Fund", FY2425_Proj = "FY 2024-25 Projected",
    FY2526_Budget = "FY 2025-26 Budget", Change = "Change ($)", Pct_Chg = "Change (%)"
  ) |>
  flextable::align(j = c("FY2425_Proj","FY2526_Budget","Change","Pct_Chg"),
                   align = "right", part = "all") |>
  flextable::bold(i = n_rows, part = "body") |>
  flextable::bg(i = n_rows, bg = "#DDEEFF", part = "body") |>
  flextable::hline(i = n_rows - 1,
                   border = officer::fp_border(color = "navy", width = 2),
                   part = "body") |>
  ds_style_flextable(title = "Personnel Services Subtotals by Fund — FY 2024-25 Projected vs. FY 2025-26 Proposed")
```



```{r}
#| label: download-personnel-services-subtotals-by-fund

ds_download_data_btn(fs_disp, "25_26_budget_personnel_services_subtotals_by_fund")

```

\
\

### Key Findings — Total Compensation Increase

```{r summary-findings}
gf_proj   <- sum(dept_data$FY2425_Proj[dept_data$Fund == "General Fund (101)"])
gf_budget <- sum(dept_data$FY2526_Budget[dept_data$Fund == "General Fund (101)"])
gf_change <- gf_budget - gf_proj

all_proj   <- sum(dept_data$FY2425_Proj)
all_budget <- sum(dept_data$FY2526_Budget)
all_change <- all_budget - all_proj

# Non-GF increases (fee-funded)
non_gf_change <- all_change - gf_change
```

- **All-funds Personnel Services increase:  
  `r fmt_dollar(all_change)`** (from
  `r fmt_dollar(all_proj)` to `r fmt_dollar(all_budget)`, a
  `r fmt_pct(all_change/all_proj)` increase).

- **General Fund (tax-supported) net increase:  
  `r fmt_dollar(gf_change)`** — portion relevant to property tax funding.

- **Fee-funded Personnel Services increase:  
  `r fmt_dollar(non_gf_change)`**
  (Water/Sewer, Drainage, EMS — presumably covered by utility fee increases and service contracts, not property taxes).

- The **explicitly stated merit/step program cost is \$784,000 citywide:**    
  (mid-year implementation, covering approximately the second half of FY 2025-26).  
  
  The annualized cost of the merit/step program would be approximately
  **\$1,568,000**.

- The difference between the merit/step policy cost and the net GF personnel
  change (`r fmt_dollar(gf_change)`) reflects offsetting factors in individual
  departments:  
  — notably a **`r fmt_dollar(abs(dept_data$Change[dept_data$Department == "Police"]))` decrease in Police Personnel Services**  
  
  — and smaller decreases in several other departments (City Manager, Facility
  Services, IT),  
  
  — partially offset by the 6 new firefighters
  (`r fmt_dollar(dept_data$Change[dept_data$Department == "Fire Rescue"])` Fire Rescue increase) and other new positions.



#### Notable Departmental Changes — General Fund

```{r notable-changes}
gf_num    <- dept_data[dept_data$Fund == "General Fund (101)", ]
gf_num    <- gf_num[base::order(gf_num$Change, decreasing = TRUE), ]
highlight <- rbind(head(gf_num, 5), tail(gf_num, 5))

# Pre-compute row indices for coloring (formulae evaluate against hl_disp
# which has character Change — use numeric indices from the source data instead)
rows_neg  <- base::which(highlight$Change < 0)
rows_pos  <- base::which(highlight$Change > 0)

hl_disp <- data.frame(
  Department    = highlight$Department,
  FY2425_Proj   = fmt_dollar(highlight$FY2425_Proj),
  FY2526_Budget = fmt_dollar(highlight$FY2526_Budget),
  Change        = ifelse(highlight$Change >= 0,
                         paste0("+", fmt_dollar(highlight$Change)),
                         paste0("(", fmt_dollar(base::abs(highlight$Change)), ")")),
  Pct_Chg       = ifelse(highlight$Pct_Chg >= 0,
                         paste0("+", fmt_pct(highlight$Pct_Chg)),
                         paste0("(", fmt_pct(base::abs(highlight$Pct_Chg)), ")")),
  stringsAsFactors = FALSE
)

flextable::flextable(hl_disp) |>
  flextable::set_header_labels(
    Department = "Department", FY2425_Proj = "FY 2024-25 Proj.",
    FY2526_Budget = "FY 2025-26 Budget", Change = "Change ($)", Pct_Chg = "Change (%)"
  ) |>
  flextable::align(j = c("FY2425_Proj","FY2526_Budget","Change","Pct_Chg"),
                   align = "right", part = "all") |>
  flextable::color(i = rows_neg, j = "Change", color = "firebrick", part = "body") |>
  flextable::color(i = rows_pos, j = "Change", color = "darkgreen", part = "body") |>
  flextable::hline(i = 5, border = officer::fp_border(color = "navy", width = 2),
                   part = "body") |>
  ds_style_flextable(title = "Top 5 Increases / Top 5 Decreases — General Fund Personnel Services")
```


```{r}
#| label: download-notable-dept-changes-gen-fund

ds_download_data_btn(hl_disp, "25_26_budget_notable_dept_changes_gen_fund")

```

\
\


**Police decrease explanation:** The Police department shows a decrease of
`r fmt_dollar(abs(dept_data$Change[dept_data$Department == "Police"]))` in
Personnel Services.  

Examining the individual account lines reveals that the
FY 2024-25 projected year-end included elevated overtime actuals and medical services that will not recur at that level.  

Regular pay (account 511110)
actually *increases* from the FY 2024-25 budget (\$7,484,154) to the FY 2025-26 budget (\$7,773,694), consistent with the merit/step program.  

The net dollar decrease is primarily an artifact of the 2024-25 projection being inflated above budget due to mid-year actuals.

---

## Question 2: Does the Increase Include Benefits?

**Yes.**  Expenditure Category 511 — "Personnel Services" — is the
account category that Schertz uses for all compensation-related costs.
It includes the following account types (verified from the Police and EMS
detailed account listings in the budget worksheet):

```{r}
#| label: account-types

acct_df <- data.frame(
  Account_Type = c("Regular wages", "Overtime", "Longevity pay",
                   "Clothing / uniform allowance", "Certification allowance",
                   "Employer FICA (Social Security + Medicare)",
                   "Employer TMRS (retirement — Texas Municipal)",
                   "Employer health insurance", "Workers' compensation insurance",
                   "Unemployment compensation", "Training and travel",
                   "Medical services (pre-employment, physicals)",
                   "Employee recognition"),
  Example_Accounts = c("511110","511120","511210","511220, 511606","511230",
                        "511310","511350","511410","511500","511600",
                        "511612","511617","511609"),
  Included = rep("\u2713", 13),
  stringsAsFactors = FALSE
)

flextable::flextable(acct_df) |>
  flextable::set_header_labels(
    Account_Type     = "Account Type",
    Example_Accounts = "Example Accounts",
    Included         = "Included?"
  ) |>
  flextable::align(j = "Included", align = "center", part = "all") |>
  flextable::color(j = "Included", color = "darkgreen", part = "body") |>
  flextable::bold(j = "Included", part = "body") |>
  ds_style_flextable(title = "ExpCategory 511 — Personnel Services Account Types")
```


```{r}
#| label: download-personnel-svs-acct-types

ds_download_data_btn(acct_df, "25_26_budget_personnel_svs_acct_types")

```

\
\


The budget explicitly states that the medical insurance provider change was
structured to **prevent any premium increase**, meaning that health insurance
costs are held approximately flat in the new budget year.  The overall
Personnel Services increases therefore reflect primarily the wage/salary and
associated payroll-tax components.

In addition to ExpCategory 511, the Non-Departmental budget (Dept 101) carries
a **Contingencies-Personnel** appropriation (account 101-101-534013) of
**\$300,000** in both FY 2024-25 and FY 2025-26 — no change.  This is a
reserve for unplanned personnel-related needs and is not part of the
ExpCategory 511 totals compiled above.

---

## Question 3: Property Tax Rate Required to Cover the Compensation Increases

### Overview of the Proposed Tax Rate

The budget is funded at the **No-New-Revenue (NNR) rate** of **\$0.5064** per
\$100 of taxable value — an increase of \$0.0164 over the FY 2024-25 rate of
\$0.4900.

```{r}
#| tbl-cap: "City of Schertz Property Tax Rate — FY 2024-25 vs. FY 2025-26 Proposed"

tax_tbl <- data.frame(
  Component = c("Maintenance & Operations (M&O)", "Interest & Sinking (I&S)", "Total Rate"),
  FY2425    = c(0.3326, 0.1574, 0.4900),
  FY2526    = c(0.3383, 0.1681, 0.5064),
  stringsAsFactors = FALSE
)
tax_tbl$Change <- tax_tbl$FY2526 - tax_tbl$FY2425

tax_disp <- data.frame(
  Component = tax_tbl$Component,
  FY2425    = base::sprintf("$%.4f", tax_tbl$FY2425),
  FY2526    = base::sprintf("$%.4f", tax_tbl$FY2526),
  Change    = base::sprintf("+$%.4f", tax_tbl$Change),
  stringsAsFactors = FALSE
)

flextable::flextable(tax_disp) |>
  flextable::set_header_labels(
    Component = "Component",
    FY2425 = "FY 2024-25 ($/100)",
    FY2526 = "FY 2025-26 Proposed ($/100)",
    Change = "Change ($/100)"
  ) |>
  flextable::align(j = c("FY2425","FY2526","Change"), align = "right", part = "all") |>
  flextable::bold(i = 3, part = "body") |>
  flextable::hline(i = 2, border = officer::fp_border(color = "navy", width = 1),
                   part = "body") |>
  ds_style_flextable(title = "City of Schertz Property Tax Rate — FY 2024-25 vs. FY 2025-26 Proposed")
```



```{r}
#| label: download-changes-in-prop-tax-rates

ds_download_data_btn(tax_disp, "25_26_budget_changes_in_prop_tax_rates")

```

\
\



The total taxable value as certified by the Guadalupe County Tax Assessor is
**\$5,734,815,455**.

### Revenue per Cent of Tax Rate

```{r revenue-per-cent}
# Derived from budgeted Ad Valorem Tax revenue and the proposed rate.
# FY 2025-26 Ad Valorem Tax Current budgeted: $20,668,000 at $0.5064 rate.
# Revenue per $0.01 per $100 = $20,668,000 / 50.64 (rate in cents)

rate_2526_cents <- 50.64
av_revenue_2526 <- 20668000
rev_per_cent    <- av_revenue_2526 / rate_2526_cents

rpc_df <- data.frame(
  Item  = c("Budgeted Ad Valorem Tax Revenue (FY 2025-26)",
            "Proposed tax rate",
            "Revenue per $0.01 per $100 of rate"),
  Value = c(fmt_dollar(av_revenue_2526),
            base::sprintf("$%.4f per $100 (%.2f cents per $100)", 0.5064, rate_2526_cents),
            fmt_dollar(rev_per_cent)),
  stringsAsFactors = FALSE
)

flextable::flextable(rpc_df) |>
  flextable::set_header_labels(Item = "Item", Value = "Value") |>
  flextable::align(j = "Value", align = "right", part = "all") |>
  flextable::bold(i = 3, part = "body") |>
  ds_style_flextable(title = "Ad Valorem Revenue Baseline — FY 2025-26")
```

Note: This per-cent figure implicitly incorporates the actual collection rate (delinquencies, exemptions).  It is a more reliable basis for rate calculations than applying the statutory taxable value directly.


```{r}
#| label: download-ad-velorem-revenue-baseline

ds_download_data_btn(rpc_df, "25_26_budget_add_valorem_rev_baseline")

```

\
\





### Tax Rate Required to Cover General Fund Personnel Increase

The property tax (M&O rate) supports only the **General Fund**.  Water/Sewer,
Drainage, and EMS personnel costs are funded by fees and service contracts,
not property taxes.

```{r gf-tax-calc}
# General Fund Personnel Services net change
gf_pers_increase <- gf_change
cents_needed     <- gf_pers_increase / rev_per_cent
rate_needed      <- cents_needed / 100

gtc_df <- data.frame(
  Item  = c("General Fund Personnel Services net increase",
            "Revenue per $0.01 per $100 of rate",
            "Tax rate increase needed (cents per $100)",
            "Tax rate increase needed ($/100)",
            "Rate increase as % of current M&O rate ($0.3326)"),
  Value = c(fmt_dollar(gf_pers_increase),
            fmt_dollar(rev_per_cent),
            base::sprintf("%.4f cents", cents_needed),
            base::sprintf("$%.4f per $100", rate_needed),
            base::sprintf("%.1f%%", rate_needed / 0.3326 * 100)),
  stringsAsFactors = FALSE
)

flextable::flextable(gtc_df) |>
  flextable::set_header_labels(Item = "Item", Value = "Value") |>
  flextable::align(j = "Value", align = "right", part = "all") |>
  flextable::bold(i = 4, part = "body") |>
  ds_style_flextable(title = "Tax Rate Required to Cover General Fund Personnel Services Increase")
```



```{r}
#| label: download-rate-reqd-cover-gf-personnel-increase

ds_download_data_btn(gtc_df, "25_26_budget_rate_req_cover_gf_pers_increase")

```

\
\


### Dollar Impact on Taxpayers

To translate the tax rate increase into a dollar cost per taxpayer, we need a
typical taxable home value.  The budget does not publish the median taxable
value directly, but we can derive an approximation.

```{r taxpayer-impact}
# Total taxable value from budget: $5,734,815,455
# Total homesteads approximated from the document:
#   2,616 properties use Disabled Veteran exemption (~23% of homesteads)
#   => total homesteads ≈ 2,616 / 0.23 ≈ 11,374
# Total taxable value includes commercial properties too.
# For illustrative purposes, compute per typical home values.

total_taxable      <- 5734815455
home_values        <- c(200000, 300000, 400000, 500000)
rate_increase      <- rate_needed
full_rate_increase <- 0.5064 - 0.4900

impact_df <- data.frame(
  Taxable_Value       = fmt_dollar(home_values),
  Personnel_Only      = fmt_dollar(home_values / 100 * rate_increase,      digits = 2),
  Full_Rate_Increase  = fmt_dollar(home_values / 100 * full_rate_increase,  digits = 2),
  stringsAsFactors = FALSE
)

flextable::flextable(impact_df) |>
  flextable::set_header_labels(
    Taxable_Value      = "Taxable Home Value",
    Personnel_Only     = "GF Personnel Increase Only ($/yr)",
    Full_Rate_Increase = "Full Proposed Rate Increase +$0.0164 ($/yr)"
  ) |>
  flextable::align(j = c("Taxable_Value","Personnel_Only","Full_Rate_Increase"),
                   align = "right", part = "all") |>
  ds_style_flextable(title = "Estimated Annual Property Tax Impact by Home Value")
```


```{r}
#| label: download-est-tax-impact-by-home-value

ds_download_data_btn(impact_df, "25_26_budget_est_tax_impact_by_home_value")

```

\
\



### Citywide Merit/Step Program — Standalone Rate Calculation

The budget explicitly states the merit/step program costs **\$784,000 citywide** for the half-year (mid-year implementation).  If we isolate only this policy cost and ask how much of it falls on property taxpayers, we need the General Fund share.

The document does not publish a GF-only merit/step figure, but we can estimate it proportionally using the ratio of GF Personnel Services to total personnel services:

```{r merit-step-calc}
gf_share      <- gf_proj / all_proj
gf_merit_step <- 784000 * gf_share
rate_merit_step <- (gf_merit_step / rev_per_cent) / 100

ms_df <- data.frame(
  Item  = c("Total merit/step program cost (half-year, all funds)",
            "General Fund share of total personnel services",
            "Estimated GF portion of merit/step cost (half-year)",
            "Tax rate to fund GF merit/step cost alone (half-year)",
            "Annualized GF merit/step cost (full year)",
            "Tax rate to fund annualized GF merit/step cost"),
  Value = c(fmt_dollar(784000),
            base::sprintf("%s (%.1f%%)", fmt_dollar(gf_proj), gf_share * 100),
            fmt_dollar(gf_merit_step),
            base::sprintf("$%.4f per $100", rate_merit_step),
            fmt_dollar(gf_merit_step * 2),
            base::sprintf("$%.4f per $100", rate_merit_step * 2)),
  stringsAsFactors = FALSE
)

flextable::flextable(ms_df) |>
  flextable::set_header_labels(Item = "Item", Value = "Value") |>
  flextable::align(j = "Value", align = "right", part = "all") |>
  flextable::bold(i = c(4, 6), part = "body") |>
  flextable::hline(i = 4, border = officer::fp_border(color = "navy", style = "dashed"),
                   part = "body") |>
  ds_style_flextable(title = "Merit/Step Program — General Fund Share and Tax Rate Equivalent")
```



```{r}
#| label: download-est-tax-impact-merit-step

ds_download_data_btn(ms_df, "25_26_budget_est_tax_impact_merit_step_prgrm")

```

\
\


---

## Question 4: Is the Proposed Tax Rate Sufficient to Cover the Compensation Increases?

```{r sufficiency-check}
# Proposed M&O rate increase revenue
mo_rate_increase    <- 0.3383 - 0.3326
mo_revenue_increase <- mo_rate_increase * 100 * rev_per_cent

gf_rev_2425  <- 52632269
gf_rev_2526  <- 52499128
gf_rev_change <- gf_rev_2526 - gf_rev_2425

# --- Table 1: GF Revenue summary ---
rev_df <- data.frame(
  Item  = c("GF Revenue FY 2024-25 (year-end estimate)",
            "GF Revenue FY 2025-26 (proposed budget)",
            "Net GF Revenue change"),
  Value = c(fmt_dollar(gf_rev_2425),
            fmt_dollar(gf_rev_2526),
            ifelse(gf_rev_change >= 0,
                   paste0("+", fmt_dollar(gf_rev_change)),
                   paste0("(", fmt_dollar(base::abs(gf_rev_change)), ")"))),
  stringsAsFactors = FALSE
)

flextable::flextable(rev_df) |>
  flextable::set_header_labels(Item = "Item", Value = "Value") |>
  flextable::align(j = "Value", align = "right", part = "all") |>
  flextable::bold(i = 3, part = "body") |>
  flextable::color(i = 3, j = "Value", color = "firebrick", part = "body") |>
  ds_style_flextable(title = "General Fund Revenue — FY 2024-25 vs. FY 2025-26")
```


```{r}
#| label: download-gf-rev-25-v-26

ds_download_data_btn(rev_df, "25_26_budget_gf_rev_25_v_26")

```

\
\



```{r}

# --- Table 2: Rate sufficiency test ---
covers <- mo_revenue_increase >= gf_change
suf_df <- data.frame(
  Item  = c("M&O rate increase ($0.0057) — additional revenue generated",
            "GF Personnel Services net increase (full year)",
            "Surplus / (Shortfall) of rate increase vs. personnel increase",
            "Rate increase DOES cover full-year GF personnel increase?",
            "Memo: merit/step cost as stated in budget (HALF-YEAR)",
            "Memo: annualized merit/step cost (full year)",
            "Rate increase revenue vs. annualized merit/step cost"),
  Value = c(fmt_dollar(mo_revenue_increase),
            fmt_dollar(gf_change),
            ifelse(covers,
                   paste0("+", fmt_dollar(mo_revenue_increase - gf_change)),
                   paste0("(", fmt_dollar(base::abs(mo_revenue_increase - gf_change)), ")")),
            ifelse(covers, "YES", "NO"),
            fmt_dollar(784000),
            fmt_dollar(784000 * 2),
            ifelse(mo_revenue_increase >= 784000 * 2,
                   paste0("+", fmt_dollar(mo_revenue_increase - 784000 * 2)),
                   paste0("(", fmt_dollar(base::abs(mo_revenue_increase - 784000 * 2)), ")"))),
  stringsAsFactors = FALSE
)

flextable::flextable(suf_df) |>
  flextable::set_header_labels(Item = "Item", Value = "Value") |>
  flextable::align(j = "Value", align = "right", part = "all") |>
  flextable::bold(i = c(3, 4), part = "body") |>
  flextable::color(i = 3, j = "Value",
                   color = ifelse(covers, "darkgreen", "firebrick"), part = "body") |>
  flextable::color(i = 4, j = "Value",
                   color = ifelse(covers, "darkgreen", "firebrick"), part = "body") |>
  flextable::hline(i = 4, border = officer::fp_border(color = "navy", style = "dashed"),
                   part = "body") |>
  flextable::italic(i = c(5, 6, 7), part = "body") |>
  ds_style_flextable(title = "Rate Sufficiency Test — Does M&O Increase Cover Personnel Increase?")
```


```{r}
#| label: rate-sufficiency-test

ds_download_data_btn(suf_df, "25_26_budget_rate_sufficiency_test")

```

\
\


### Deficit Budget Context

Despite the No-New-Revenue rate increase, the General Fund remains in
**deficit**:

```{r}

gf_deficit_df <- data.frame(
  Item   = c("FY 2025-26 GF Revenue (proposed)",
             "FY 2025-26 GF Expenditures (proposed)",
             "Surplus / (Deficit)",
             "Resulting Reserve Balance",
             "Reserve as % of Expenditures",
             "City Reserve Policy Maximum"),
  Amount = c(fmt_dollar(52499128),
             fmt_dollar(53754934),
             paste0("(", fmt_dollar(1255806), ")"),
             fmt_dollar(13936617),
             "26.8%",
             "≤ 26.0%"),
  stringsAsFactors = FALSE
)

flextable::flextable(gf_deficit_df) |>
  flextable::set_header_labels(Item = "Item", Amount = "Amount") |>
  flextable::align(j = "Amount", align = "right", part = "all") |>
  flextable::bold(i = 3, part = "body") |>
  flextable::color(i = 3, j = "Amount", color = "firebrick", part = "body") |>
  flextable::hline(i = 2, border = officer::fp_border(color = "navy", width = 1),
                   part = "body") |>
  ds_style_flextable(title = "General Fund — FY 2025-26 Budget Surplus / (Deficit)")
```


```{r}
#| label: budget-gf-surplus-or-deficit

ds_download_data_btn(gf_deficit_df, "25_26_budget_gf_surplus_or_deficit")

```

\
\





The proposed budget would draw down reserves by \$1,255,806, leaving the
reserve balance at \$13,936,617, which is *above* the City's own 26% maximum
policy target.  This means the budget proposes to overspend relative to
current revenue while still technically meeting the reserve floor — but the
reserve direction is downward.

Other funds are also in deficit:

```{r}

other_funds_df <- data.frame(
  Fund         = c("Water & Sewer", "Drainage", "SEDC", "EMS"),
  Revenue      = c(fmt_dollar(35391478), fmt_dollar(2316283),
                   fmt_dollar(8928608),  fmt_dollar(17504469)),
  Expenditures = c(fmt_dollar(36695153), fmt_dollar(2818831),
                   fmt_dollar(16701463), fmt_dollar(17149568)),
  Surplus_Def  = c(paste0("(", fmt_dollar(1303675), ")"),
                   paste0("(", fmt_dollar(502548),  ")"),
                   paste0("(", fmt_dollar(7772855),  ")"),
                   paste0("+", fmt_dollar(354901))),
  stringsAsFactors = FALSE
)

flextable::flextable(other_funds_df) |>
  flextable::set_header_labels(
    Fund = "Fund", Revenue = "Revenue",
    Expenditures = "Expenditures", Surplus_Def = "Surplus / (Deficit)"
  ) |>
  flextable::align(j = c("Revenue","Expenditures","Surplus_Def"),
                   align = "right", part = "all") |>
  flextable::color(i = c(1,2,3), j = "Surplus_Def", color = "firebrick", part = "body") |>
  flextable::color(i = 4,        j = "Surplus_Def", color = "darkgreen", part = "body") |>
  ds_style_flextable(title = "Other Funds — FY 2025-26 Budget Surplus / (Deficit)")
```

The EMS fund is the only major operating fund with a projected surplus, made possible by a 20% increase in contracted service revenues and a \$100/transport fee increase.


```{r}
#| label: budget-other-funds-surplus-or-deficit

ds_download_data_btn(other_funds_df, "25_26_budget_other_funds_surplus_or_deficit")

```

\
\


### A Note on Sales Tax as a Funding Source

Sales tax contributes approximately **\$15.3 million** to General Fund revenues
in FY 2025-26 — roughly **29% of total GF resources**.  It is the second-largest
revenue source after property taxes.

However, **sales tax is not a policy lever**.  The City's sales tax allocation
is fixed by state law at 1% to the General Fund; the City cannot increase it by
budget action.  Sales tax revenue fluctuates with economic conditions and retail
activity — making it less predictable than property tax and an unreliable basis
for funding permanent compensation commitments.

Any compensation shortfall — whether from the current deficit, from the
annualized merit/step run-rate hitting FY 2026-27, or from future increases —
can only be addressed through some combination of:

- **Property tax rate increases** (up to the voter-approval rate cap),
- **Utility and drainage fee increases** (for the fee-funded funds), or
- **Reserve drawdowns** (a one-time measure, not a recurring solution).

Sales tax growth may incidentally relieve future pressure, but it cannot be
counted on, and it cannot be deliberately increased by Council action.  It is
therefore excluded from the rate calculations in this analysis.

---

## Question 5: Annual Growth Rate and Doubling Time for Total Compensation

### Data Available

The budget worksheets present four fiscal years of Personnel Services data for every department:

```{r}
#| label: data-years

yr_df <- data.frame(
  Year        = c("FY 2022-23", "FY 2023-24", "FY 2024-25", "FY 2025-26"),
  Description = c("Actual expenditures", "Actual expenditures",
                  "Year-end projected (staff estimate)", "Proposed budget"),
  stringsAsFactors = FALSE
)

flextable::flextable(yr_df) |>
  flextable::set_header_labels(Year = "Year", Description = "Description") |>
  flextable::bold(i = 4, part = "body") |>
  flextable::bg(i = 4, bg = "#FFFACD", part = "body") |>
  ds_style_flextable(title = "Personnel Services Data Available — Four Fiscal Years")
```



```{r}
#| label: budget-describe-comp-data-available

ds_download_data_btn(yr_df, "25_26_budget_describe_comp_data_available")

```

\
\




This yields a **3-year span** for CAGR calculation (FY 2022-23 → FY 2025-26 proposed), with a 2-year cross-check using FY 2024-25 projected as the endpoint to avoid dependence on the proposed budget.

### Complete Four-Year Dataset

```{r}
#| tbl-cap: "Personnel Services — Four-Year History, All Departments"

# Values read from ExpCategory: 511 - Personnel Services Total rows,
# column 2 (FY22-23 Actual) and column 4 (FY23-24 Actual).

dept_data$FY2223_Act <- c(
  # General Fund (24 departments, same order as dept_data)
    65991.00,    # City Council (150)
  1268219.37,    # City Manager (151)
   349992.25,    # Municipal Court (152)
   130167.92,    # 311/Customer Relations (153)
   332998.93,    # Planning & Zoning (154)
         0.00,   # Economic Development (156) — dept created FY23-24
   195204.99,    # City Secretary (164)
   338315.64,    # Public Affairs (170)
   995485.25,    # Engineering (173)
  9667983.16,    # Police (256)
  6825077.73,    # Fire Rescue (257)
  1056532.12,    # Inspections (258)
   423553.18,    # Neighborhood Services (259)
   996132.16,    # Streets (359)
  1035692.62,    # Parks (460)
    17692.84,    # Swim Pool (463)
   440489.69,    # Event Facilities (464)
  1001807.06,    # Library (665)
  1147821.16,    # IT (853)
   643658.58,    # Human Resources (866)
   696231.81,    # Finance (868)
   267811.27,    # Purchasing & Asset Mgt (872)
   694009.96,    # Fleet Services (877)
   897424.16,    # Facility Services (878)
  # Water & Sewer
   601713.12,    # Utility Billing (176)
  1942709.33,    # W&S Administration (575)
  # Drainage
   634080.66,    # Drainage (579)
  # EMS
  7602824.08,    # EMS (203-280)
  # Special Revenue
      3558.32    # Historical Committee (615-609)
)

dept_data$FY2324_Act <- c(
  # General Fund
    70269.22,    # City Council (150)
  1133902.71,    # City Manager (151)
   365992.45,    # Municipal Court (152)
   150130.32,    # 311/Customer Relations (153)
   415440.81,    # Planning & Zoning (154)
   373804.15,    # Economic Development (156)
   218750.26,    # City Secretary (164)
   361689.77,    # Public Affairs (170)
  1050157.62,    # Engineering (173)
 11191198.57,    # Police (256)
  8283649.91,    # Fire Rescue (257)
  1261620.00,    # Inspections (258)
   506751.71,    # Neighborhood Services (259)
  1091718.73,    # Streets (359)
  1322151.35,    # Parks (460)
    15723.23,    # Swim Pool (463)
   543478.03,    # Event Facilities (464)
  1154459.87,    # Library (665)
  1343640.33,    # IT (853)
   713034.39,    # Human Resources (866)
   727058.93,    # Finance (868)
   234729.59,    # Purchasing & Asset Mgt (872)
   897046.05,    # Fleet Services (877)
  1115683.58,    # Facility Services (878)
  # Water & Sewer
   653083.93,    # Utility Billing (176)
  2210979.68,    # W&S Administration (575)
  # Drainage
   613853.27,    # Drainage (579)
  # EMS
  8986816.34,    # EMS (203-280)
  # Special Revenue
        0.00     # Historical Committee (615-609) — no activity FY23-24
)

# Four-year totals by scope
calc_totals <- function(fund_filter = NULL) {
  sub <- if (is.null(fund_filter)) dept_data else dept_data[dept_data$Fund %in% fund_filter, ]
  c(FY2223 = sum(sub$FY2223_Act),
    FY2324 = sum(sub$FY2324_Act),
    FY2425 = sum(sub$FY2425_Proj),
    FY2526 = sum(sub$FY2526_Budget))
}

gf_funds   <- "General Fund (101)"
gf_totals  <- calc_totals(gf_funds)
all_totals <- calc_totals()

hist_df <- data.frame(
  Scope         = c("General Fund", "All Funds Combined"),
  FY2223        = c(fmt_dollar(gf_totals["FY2223"]),  fmt_dollar(all_totals["FY2223"])),
  FY2324        = c(fmt_dollar(gf_totals["FY2324"]),  fmt_dollar(all_totals["FY2324"])),
  FY2425        = c(fmt_dollar(gf_totals["FY2425"]),  fmt_dollar(all_totals["FY2425"])),
  FY2526        = c(fmt_dollar(gf_totals["FY2526"]),  fmt_dollar(all_totals["FY2526"])),
  stringsAsFactors = FALSE
)

flextable::flextable(hist_df) |>
  flextable::set_header_labels(
    Scope  = "Scope",
    FY2223 = "FY 2022-23 Actual",
    FY2324 = "FY 2023-24 Actual",
    FY2425 = "FY 2024-25 Projected",
    FY2526 = "FY 2025-26 Proposed"
  ) |>
  flextable::align(j = c("FY2223","FY2324","FY2425","FY2526"),
                   align = "right", part = "all") |>
  ds_style_flextable(title = "Personnel Services — Four-Year History by Scope")
```


```{r}
#| label: budget-four-year-dataset

ds_download_data_btn(hist_df, "25_26_budget_4_yr_dataset")

```

\
\




### Annualized Adjustment for the Merit/Step Program

The FY 2025-26 budget **deliberately understates** the full cost of the
compensation increases because the merit/step program is implemented
**mid-year** — only half the annual cost appears in the budget.  The other
half hits the City in FY 2026-27 before any new raises are negotiated.

To project where compensation actually lands on a full-year run-rate basis,
the CAGR calculation below uses an **annualized FY 2025-26 endpoint**: the
budgeted Personnel Services totals with the second half of the merit/step cost
added back.  This is the correct basis for doubling-time projections because
doubling time is a forward-looking statement — it answers "how long until
compensation is twice what it is today?" — and "today" should mean the full
run-rate cost, not the discounted budget figure.

```{r}

# The budget states the merit/step program costs $784,000 citywide for the
# half-year.  The full annualized incremental cost is $784,000 × 2 = $1,568,000.
# The second half ($784,000) is not in the proposed budget at all.
# We allocate the adjustment proportionally by fund share of total personnel.

merit_half    <- 784000
merit_full    <- merit_half * 2
merit_addback <- merit_half

gf_pers_budget  <- base::sum(dept_data$FY2526_Budget[dept_data$Fund == "General Fund (101)"])
all_pers_budget <- base::sum(dept_data$FY2526_Budget)
gf_share_2526   <- gf_pers_budget / all_pers_budget

gf_2526_ann  <- gf_totals["FY2526"]  + merit_addback * gf_share_2526
all_2526_ann <- all_totals["FY2526"] + merit_addback

ann_df <- data.frame(
  Item  = c("Merit/step half-year cost (as stated in budget)",
            "Merit/step addback for annualization (2nd half, not in budget)",
            "As-budgeted GF Personnel Services (FY 2025-26)",
            "Annualized GF Personnel Services (full run-rate)",
            "Addback — additional GF cost arriving in FY 2026-27",
            "As-budgeted All-Funds Personnel Services (FY 2025-26)",
            "Annualized All-Funds Personnel Services (full run-rate)",
            "Addback — additional All-Funds cost arriving in FY 2026-27"),
  Value = c(fmt_dollar(merit_half),
            fmt_dollar(merit_addback),
            fmt_dollar(gf_totals["FY2526"]),
            fmt_dollar(gf_2526_ann),
            fmt_dollar(gf_2526_ann - gf_totals["FY2526"]),
            fmt_dollar(all_totals["FY2526"]),
            fmt_dollar(all_2526_ann),
            fmt_dollar(all_2526_ann - all_totals["FY2526"])),
  stringsAsFactors = FALSE
)

flextable::flextable(ann_df) |>
  flextable::set_header_labels(Item = "Item", Value = "Value") |>
  flextable::align(j = "Value", align = "right", part = "all") |>
  flextable::bold(i = c(4, 5, 7, 8), part = "body") |>
  flextable::color(i = c(5, 8), j = "Value", color = "firebrick", part = "body") |>
  flextable::hline(i = c(2, 5),
                   border = officer::fp_border(color = "navy", style = "dashed"),
                   part = "body") |>
  ds_style_flextable(title = "Merit/Step Annualization — Budget vs. Full Run-Rate")
```


```{r}
#| label: budget-merit-step-annualization

ds_download_data_btn(ann_df, "25_26_budget_merit_step_annualized")

```

\
\


### Compound Annual Growth Rates (CAGR)

$$\text{CAGR} = \left(\frac{V_n}{V_0}\right)^{1/n} - 1 \qquad t_{\text{double}} = \frac{\ln 2}{\ln(1 + \text{CAGR})}$$

Two sets of CAGR are presented: one using the **as-budgeted** FY 2025-26
figures (half-year merit/step), and one using the **annualized** figures
(full-year run-rate).  The annualized CAGR is the appropriate basis for
doubling-time projections.

```{r}
#| tbl-cap: "CAGR and Doubling Time — As-Budgeted vs. Annualized"

cagr          <- function(v0, vn, n) (vn / v0)^(1/n) - 1
doubling_time <- function(r) log(2) / log(1 + r)

scopes <- list(
  list(label = "General Fund — as budgeted",  v0 = gf_totals["FY2223"],
       v2425 = gf_totals["FY2425"], v2526 = gf_totals["FY2526"],
       vy1324 = gf_totals["FY2324"]),
  list(label = "General Fund — annualized",   v0 = gf_totals["FY2223"],
       v2425 = gf_totals["FY2425"], v2526 = gf_2526_ann,
       vy1324 = gf_totals["FY2324"]),
  list(label = "All Funds — as budgeted",     v0 = all_totals["FY2223"],
       v2425 = all_totals["FY2425"], v2526 = all_totals["FY2526"],
       vy1324 = all_totals["FY2324"]),
  list(label = "All Funds — annualized",      v0 = all_totals["FY2223"],
       v2425 = all_totals["FY2425"], v2526 = all_2526_ann,
       vy1324 = all_totals["FY2324"])
)

cagr_results <- do.call(rbind, lapply(scopes, function(s) {
  c3 <- cagr(s$v0,     s$v2526,  3)
  c2 <- cagr(s$v0,     s$v2425,  2)
  y1 <- cagr(s$vy1324, s$v2425,  1)
  y2 <- cagr(s$v2425,  s$v2526,  1)
  data.frame(
    Scope    = s$label,
    CAGR_3yr = fmt_pct(c3),
    DT_3yr   = base::sprintf("%.1f yrs", doubling_time(c3)),
    CAGR_2yr = fmt_pct(c2),
    DT_2yr   = base::sprintf("%.1f yrs", doubling_time(c2)),
    YoY_2425 = fmt_pct(y1),
    YoY_2526 = fmt_pct(y2),
    stringsAsFactors = FALSE
  )
}))

flextable::flextable(cagr_results) |>
  flextable::set_header_labels(
    Scope    = "Scope",
    CAGR_3yr = "3-yr CAGR (FY22-23→FY25-26)",
    DT_3yr   = "Doubling Time",
    CAGR_2yr = "2-yr CAGR (FY22-23→FY24-25)",
    DT_2yr   = "Doubling Time",
    YoY_2425 = "YoY FY24-25",
    YoY_2526 = "YoY FY25-26"
  ) |>
  flextable::align(j = c("CAGR_3yr","DT_3yr","CAGR_2yr","DT_2yr","YoY_2425","YoY_2526"),
                   align = "right", part = "all") |>
  flextable::bold(i = c(2, 4), part = "body") |>
  flextable::bg(i = c(2, 4), bg = "#FFF8DC", part = "body") |>
  flextable::hline(i = c(2),
                   border = officer::fp_border(color = "navy", style = "dashed"),
                   part = "body") |>
  ds_style_flextable(title = "CAGR and Doubling Time — As-Budgeted vs. Annualized (highlighted rows = annualized)")
```



```{r}
#| label: budget-cagr-doubling-time

ds_download_data_btn(cagr_results, "25_26_budget_cagr_doubling_times")

```

\
\


```{r}

gf_c3_ann   <- cagr(gf_totals["FY2223"],  gf_2526_ann,  3)
gf_dt3_ann  <- doubling_time(gf_c3_ann)
all_c3_ann  <- cagr(all_totals["FY2223"], all_2526_ann, 3)
all_dt3_ann <- doubling_time(all_c3_ann)

summary_cagr_df <- data.frame(
  Scope         = c("General Fund (annualized)", "All Funds (annualized)"),
  CAGR_3yr      = c(fmt_pct(gf_c3_ann),   fmt_pct(all_c3_ann)),
  Doubling_Time = c(base::sprintf("%.1f years", gf_dt3_ann),
                    base::sprintf("%.1f years", all_dt3_ann)),
  stringsAsFactors = FALSE
)

flextable::flextable(summary_cagr_df) |>
  flextable::set_header_labels(
    Scope = "Scope", CAGR_3yr = "3-Year CAGR (Annualized)",
    Doubling_Time = "Doubling Time"
  ) |>
  flextable::align(j = c("CAGR_3yr","Doubling_Time"), align = "right", part = "all") |>
  flextable::bold(part = "body") |>
  flextable::fontsize(i = 1:2, size = 16, part = "body") |>
  flextable::color(j = "Doubling_Time", color = "firebrick", part = "body") |>
  ds_style_flextable(title = "Key Result: Annualized Personnel Compensation Growth Rate")
```


```{r}
#| label: budget-key-cagr-results

ds_download_data_btn(summary_cagr_df, "25_26_budget_key_cagr_results")

```

\
\


### Interpretation

At the annualized 3-year CAGR, compare the compensation doubling time against:

- **Property tax revenue** — Schertz's long-run growth has averaged ~7% per year
  (doubling time ~10 years), constrained by the voter-approval rate cap.
- **Population growth** — approximately 2–3% per year in the Schertz area
  (doubling time 23–35 years).

If compensation costs compound faster than either revenue or population, the
gap grows exponentially and must eventually be closed by rate increases,
service cuts, or reserve drawdowns.  The year-over-year figures show whether
the growth rate is accelerating or decelerating.

---



## Supplemental: Full-Year (Annualized) Compensation Cost Impact

The proposed FY 2025-26 budget includes only **half a year** of merit/step
increases.  The City will face the **full annualized run-rate** beginning in
FY 2026-27, before any new compensation adjustments are negotiated.  This
section presents those forward-looking figures.

```{r}
#| tbl-cap: "Full-Year Run-Rate Compensation — FY 2026-27 Baseline Projection"

addback_gf     <- gf_2526_ann  - gf_totals["FY2526"]
addback_all    <- all_2526_ann - all_totals["FY2526"]
rev_per_cent_v <- av_revenue_2526 / rate_2526_cents
rate_addback   <- (addback_gf / rev_per_cent_v) / 100

# --- Table 1: Run-rate comparison ---
rr_df <- data.frame(
  Item  = c("As-budgeted GF Personnel Services (FY 2025-26)",
            "Annualized GF Personnel Services (full run-rate)",
            "Additional GF cost arriving in FY 2026-27",
            "As-budgeted All-Funds Personnel Services (FY 2025-26)",
            "Annualized All-Funds Personnel Services (full run-rate)",
            "Additional All-Funds cost arriving in FY 2026-27"),
  Value = c(fmt_dollar(gf_totals["FY2526"]),
            fmt_dollar(gf_2526_ann),
            fmt_dollar(addback_gf),
            fmt_dollar(all_totals["FY2526"]),
            fmt_dollar(all_2526_ann),
            fmt_dollar(addback_all)),
  stringsAsFactors = FALSE
)

flextable::flextable(rr_df) |>
  flextable::set_header_labels(Item = "Item", Value = "Value") |>
  flextable::align(j = "Value", align = "right", part = "all") |>
  flextable::bold(i = c(2, 3, 5, 6), part = "body") |>
  flextable::color(i = c(3, 6), j = "Value", color = "firebrick", part = "body") |>
  flextable::hline(i = 3, border = officer::fp_border(color = "navy", width = 1),
                   part = "body") |>
  ds_style_flextable(title = "Full-Year Run-Rate — As-Budgeted vs. Annualized Personnel Costs")

```


```{r}
#| label: budgeted-v-annualized-comp-costs

ds_download_data_btn(rr_df, "25_26_budgeted_v_annualized_comp_costs")

```

\
\

```{r}

# --- Table 2: Tax rate equivalent of the addback ---
ta_df <- data.frame(
  Item  = c("Additional GF revenue needed (merit/step addback)",
            "Revenue per $0.01 of tax rate",
            "Additional rate increase required to fund addback"),
  Value = c(fmt_dollar(addback_gf),
            fmt_dollar(rev_per_cent_v),
            base::sprintf("$%.4f per $100", rate_addback)),
  stringsAsFactors = FALSE
)

flextable::flextable(ta_df) |>
  flextable::set_header_labels(Item = "Item", Value = "Value") |>
  flextable::align(j = "Value", align = "right", part = "all") |>
  flextable::bold(i = 3, part = "body") |>
  flextable::color(i = 3, j = "Value", color = "firebrick", part = "body") |>
  ds_style_flextable(title = "Tax Rate Equivalent — FY 2026-27 Merit/Step Addback (GF Only)")

```


```{r}
#| label: budget-tax-rate-equiv

ds_download_data_btn(ta_df, "25_26_budget_tax_rate_equiv")

```

\
\


```{r}

# --- Table 3: Taxpayer impact ---
home_vals <- c(200000, 300000, 400000, 500000)
gf_proj_2425 <- base::sum(dept_data$FY2425_Proj[dept_data$Fund == "General Fund (101)"])

ti_df <- data.frame(
  Home_Value   = fmt_dollar(home_vals),
  Addback_Cost = fmt_dollar(home_vals / 100 * rate_addback, digits = 2),
  FullYr_Total = fmt_dollar(home_vals / 100 *
                               (gf_2526_ann - gf_proj_2425) / (rev_per_cent_v * 100) * 100,
                             digits = 2),
  stringsAsFactors = FALSE
)

flextable::flextable(ti_df) |>
  flextable::set_header_labels(
    Home_Value   = "Taxable Home Value",
    Addback_Cost = "Merit/Step Addback Cost (FY 2026-27) ($/yr)",
    FullYr_Total = "Total GF Personnel Increase vs FY24-25 — Annualized ($/yr)"
  ) |>
  flextable::align(j = c("Home_Value","Addback_Cost","FullYr_Total"),
                   align = "right", part = "all") |>
  ds_style_flextable(title = "Taxpayer Impact — Full-Year Merit/Step Cost by Home Value")
```


```{r}
#| label: budget-taxpayer-impact

ds_download_data_btn(ti_df, "25_26_budget_taxpayer_impact")

```

\
\

**Bottom line:** The FY 2025-26 budget presents the merit/step program as a
\$784,000 cost.  That figure is accurate for the current budget year only.
By FY 2026-27, the identical compensation structure costs **\$1,568,000
per year** before any additional raises — simply because employees will
receive the increase for a full 12 months instead of 6.  This is a structural
cost commitment, not a one-time event.

---

## Summary of Findings

```{r final-summary}
# --- Table 1: Compensation summary ---
comp_df <- data.frame(
  Item  = c("Merit/Step program — as stated in budget (half-year, all funds)",
            "Merit/Step program — annualized full-year run-rate (all funds)",
            "Net Personnel Services increase — General Fund (full year)",
            "Net Personnel Services increase — Fee-funded funds (full year)",
            "Net Personnel Services increase — ALL FUNDS (full year)",
            "Benefits included in above totals?"),
  Value = c(fmt_dollar(784000),
            fmt_dollar(784000 * 2),
            fmt_dollar(gf_change),
            fmt_dollar(non_gf_change),
            fmt_dollar(all_change),
            "YES — ExpCategory 511 includes all wages, benefits, and payroll costs"),
  stringsAsFactors = FALSE
)

flextable::flextable(comp_df) |>
  flextable::set_header_labels(Item = "Item", Value = "Value") |>
  flextable::align(j = "Value", align = "right", part = "all") |>
  flextable::bold(i = c(2, 5), part = "body") |>
  flextable::color(i = 2, j = "Value", color = "firebrick", part = "body") |>
  flextable::hline(i = 2, border = officer::fp_border(color = "navy", style = "dashed"),
                   part = "body") |>
  ds_style_flextable(title = "Summary — Compensation Increases")

# --- Table 2: Tax rate analysis ---
rate_df <- data.frame(
  Item  = c("Current M&O rate (FY 2024-25)",
            "Proposed M&O rate — No-New-Revenue (FY 2025-26)",
            "M&O rate increase proposed",
            "Rate needed solely for GF personnel increase (full year)",
            "Rate needed solely for GF merit/step — half-year (as budgeted)",
            "Rate needed solely for GF merit/step — annualized (full year)",
            "Revenue per $0.01 of rate (at FY 2025-26 taxable base)"),
  Value = c("$0.3326 per $100",
            "$0.3383 per $100",
            "+$0.0057 per $100",
            base::sprintf("$%.4f per $100", rate_needed),
            base::sprintf("$%.4f per $100", rate_merit_step),
            base::sprintf("$%.4f per $100", rate_merit_step * 2),
            fmt_dollar(rev_per_cent)),
  stringsAsFactors = FALSE
)

flextable::flextable(rate_df) |>
  flextable::set_header_labels(Item = "Item", Value = "Value") |>
  flextable::align(j = "Value", align = "right", part = "all") |>
  flextable::bold(i = c(3, 4), part = "body") |>
  ds_style_flextable(title = "Summary — Property Tax Rate Analysis")

# --- Table 3: Deficit and sufficiency ---
covers <- mo_revenue_increase >= gf_change
def_df <- data.frame(
  Item  = c("General Fund deficit (FY 2025-26 proposed)",
            "M&O rate increase revenue generated ($0.0057)",
            "Does rate increase cover the full-year GF personnel increase?",
            "Budget is structurally balanced (revenue ≥ expenditures)?",
            "Reserve direction"),
  Value = c(paste0("(", fmt_dollar(1255806), ")"),
            fmt_dollar(mo_revenue_increase),
            ifelse(covers, "YES", paste0("NO — shortfall of ",
                                          fmt_dollar(gf_change - mo_revenue_increase))),
            "NO",
            "Declining — reserves drawn down by $1,255,806"),
  stringsAsFactors = FALSE
)

flextable::flextable(def_df) |>
  flextable::set_header_labels(Item = "Item", Value = "Value") |>
  flextable::align(j = "Value", align = "right", part = "all") |>
  flextable::bold(i = c(1, 3, 4), part = "body") |>
  flextable::color(i = c(1, 3, 4, 5), j = "Value", color = "firebrick", part = "body") |>
  ds_style_flextable(title = "Summary — Deficit Context and Rate Sufficiency")
```

### Interpretation

1. **The explicit merit/step compensation policy costs \$784,000** across all
   funds for a half-year.  The annualized run-rate is approximately
   **\$1.57 million** citywide.

2. **The total net increase in Personnel Services across all funds is
   `r fmt_dollar(all_change)`**, inclusive of all wages, benefits, and
   employee-related costs.  Benefits are included in this figure.

3. **Only the General Fund increase (`r fmt_dollar(gf_change)`) is
   property-tax-supported.**  The fee-funded funds (Water/Sewer, Drainage, EMS)
   cover their own personnel increases through utility fees and service contract
   revenues.

4. **The proposed M&O rate increase of \$0.0057** (from \$0.3326 to \$0.3383)
   generates approximately \$`r format(round(mo_revenue_increase), big.mark=",")`
   in additional revenue — **not sufficient** to cover the General Fund
   personnel net increase of `r fmt_dollar(gf_change)` on its own, let alone
   the full General Fund deficit of \$1,255,806. The rate increase covers only
   a portion of the compensation increase; the remainder is absorbed by
   drawing down reserves.

5. **Important note on the \$784,000 merit/step figure vs. the budget totals.**
   The \$784,000 stated in the budget message is explicitly a **half-year cost**
   (the program takes effect mid-fiscal-year). All Personnel Services totals
   compiled in this analysis are **full fiscal-year** figures from the budget
   worksheets. These two figures cannot be directly compared without adjustment.
   The annualized (full-year) cost of the merit/step program is approximately
   **\$1,568,000 citywide**. Conversely, the full-year net General Fund
   Personnel Services change of `r fmt_dollar(gf_change)` is suppressed below
   the annualized merit/step cost because some departments show decreases
   (notably Police, whose FY 2024-25 projected year-end was inflated by
   mid-year overtime actuals that will not recur).

5. **Taxpayer cost decomposition — what are the raises actually costing you?**
   The full proposed rate increase of \$0.0164 per \$100 covers multiple drivers,
   not just compensation.  The table below decomposes the total additional tax
   for a range of home values into the personnel-driven portion and everything
   else (debt service increases, new facility operating costs, and other
   expenditure growth).  The "non-personnel" column is what the taxpayer would
   pay even if there had been **zero compensation increase**.

```{r}
#| label: taxpayer-decomp

home_vals_d  <- c(200000, 300000, 400000, 500000)
non_pers_rate <- full_rate_increase - rate_needed   # rate increase not driven by personnel

decomp_df <- data.frame(
  Home_Value  = fmt_dollar(home_vals_d),
  Total_Addl  = fmt_dollar(home_vals_d / 100 * full_rate_increase,  digits = 2),
  Personnel   = fmt_dollar(home_vals_d / 100 * rate_needed,         digits = 2),
  Non_Pers    = fmt_dollar(home_vals_d / 100 * non_pers_rate,       digits = 2),
  stringsAsFactors = FALSE
)

flextable::flextable(decomp_df) |>
  flextable::set_header_labels(
    Home_Value = "Taxable Home Value",
    Total_Addl = "Total Additional Tax\n(Full $0.0164 Rate Increase)",
    Personnel  = "Personnel-Driven Portion\n(GF Compensation Increase)",
    Non_Pers   = "Non-Personnel Portion\n(Would Pay With Zero Raises)"
  ) |>
  flextable::align(j = c("Home_Value","Total_Addl","Personnel","Non_Pers"),
                   align = "right", part = "all") |>
  flextable::bold(j = "Non_Pers", part = "body") |>
  flextable::color(j = "Personnel", color = "firebrick", part = "body") |>
  flextable::add_footer_lines(
    base::sprintf(
      "Rate decomposition: $%.4f total increase = $%.4f personnel + $%.4f other drivers (per $100 taxable value)",
      full_rate_increase, rate_needed, non_pers_rate
    )
  ) |>
  flextable::fontsize(part = "footer", size = 10) |>
  flextable::italic(part = "footer") |>
  ds_style_flextable(title = "Annual Property Tax Impact Decomposition — $300K Taxable Home Pays ~$X for Raises Alone")
```

---

*Analysis prepared by Dan Swart, CPA (ret.)| 
Sources: City of Schertz FY 2025-26 Proposed Budget
         (Excel budget document dated August 2025) | All figures read directly from budget document; no automated parsing routines were used.*
