---
title: "Schertz FY 2025–26: Compensation Growth, Tax Structure, and Doubling-Time Warning"
subtitle: ""
description: ""
authors: 
  - name: "Dan Swart, CPA (ret)"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true     # Set to 'false' for drafts. Override if you're NOT sharing output
    include-in-header:
      - text: 
          <link href="https://fonts.googleapis.com/css2?family=Lato&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Fira+Mono&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Montserrat&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Cabin&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Roboto+Slab&display=swap" rel="stylesheet">
          <link href="https://fonts.googleapis.com/css2?family=Schoolbell&display=swap" rel="stylesheet">
      - header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    fig-width: 13
    fig-height: 8
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 16pt
    mainfont: "Cabin"
  

    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: true
  message: false
  warning: false
  eval: true
  fig-width: 13
  fig-height: 8


# for .rmd files
knitr:
  opts_chunk:
    echo: true
    error: false
    warning: false
    message: false
    cache: false

prefer-html: true

params:
  data_updated: !expr format(Sys.Date(), "%B %Y")
---


```{r}
#| label: setup
#| include: false


# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# # load libraries - NOT NEEDED

# Download button helpers live here; knitr::is_html_output() guard is inside
# the helpers so they silently no-op when rendering to typst

# source(here::here("R", "download_button_helpers.R"))

# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Options
options(scipen = 999)
options(qic.clshade = T) # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.linecol = 'black') # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.signalcol = "firebrick") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(qic.targetcol = "purple") # NO LONGER NEEDED; CHARTS ALL PREPARED WITH GGPLOT2 ONLY
options(DT.options = list(dom = 'pBlfrti')) # Add buttons, filtering, and top (t) pagination controls
options(shiny.maxRequestSize = 50 * 1024^2) # Set upload maximum to 50 MB
options(tigris_use_cache = TRUE)
options(device = "RStudioGD") 


# Flextable defaults:
flextable::set_flextable_defaults(
  font.size = 14, 
  font.family = "Cabin",
  font.color = "black",
  table.layout = "fixed",
  border.color = "darkgray",
  padding.top = 3, padding.bottom = 3,
  padding.left = 4, padding.right = 4,
  line_spacing = 1.3,
  digits = 2,
  decimal.mark = ".",
  big.mark = " ",
  na_str = "<na>",
  theme_fun = base::identity,          # <-- add this
  post_process_html = base::identity,
  post_process_docx = base::identity
)

# Standard table styler: blue italic title row + palegreen column header row.
# Defined here so all QMDs that source this setup get the same presentation
# without depending on census_helpers.R for a purely cosmetic function.
ds_style_flextable <- function(ft, title = NULL) {
  if (!base::is.null(title)) {
    ft <- flextable::add_header_lines(ft, values = title)
    ft <- flextable::color(ft, i = 1, part = "header", color = "blue")
    ft <- flextable::italic(ft, i = 1, part = "header")
    ft <- flextable::align(ft, i = 1, part = "header", align = "left")
    ft <- flextable::fontsize(ft, i = 1, part = "header", size = 12)
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "white")
    ft <- flextable::bg(ft, i = 2, part = "header", bg = "palegreen")
  } else {
    ft <- flextable::bg(ft, i = 1, part = "header", bg = "palegreen")
  }
  ft <- flextable::autofit(ft)
  ft
}


# Flextable built-in themes:
  # flextable::theme_alafoli()	|>  # BLAH
  # flextable::theme_apa()  # THIS IS NICE
  # flextable::theme_booktabs() |>  # NICE, MORE COMPACT
  # flextable::theme_box() |>   # OK, INCLUDES CELL BORDERS
  # flextable::theme_tron() |>  # 'DARK MODE' BLUE TEXT
  # flextable::theme_tron_legacy() |>   # 'DARK MODE' YELLOW TEXT
  # flextable::theme_vader() |>    # 'DARK MODE' WHITE TEXT
  # flextable::theme_vanilla() |>   # NOT SPECIAL
  # flextable::theme_zebra()	|>
  #


# Set global theme for consistent plots
ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 20) +
    ggplot2::theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(
        family = "Cabin",
        face = "bold",
        color = "darkgreen",
        size = 28,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 0.0, 5.5),
        margin = ggplot2::margin(0, 0, 5.5, 0)
      ),
      plot.subtitle = ggtext::element_textbox_simple(
        family = "Cabin",
        color = "darkgreen",
        face = "bold",
        size = 22,
        fill = "yellow",
        lineheight = 0.9,
        padding = ggplot2::margin(5.5, 5.5, 5.5, 5.5),
        margin = ggplot2::margin(0.0, 0, 5.5, 0)
      ),
      plot.caption = ggtext::element_markdown(
        family = "Cabin",
        size = 18,
        hjust = 1,
        color = "darkblue",
        face = "italic",
        fill = "yellow",
        lineheight = 1.0
      ),
      axis.text.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.x = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18),
        # ggplot2::element_blank(),
      axis.text.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 18,
        angle = 45,
        hjust = 1
      ),
        # ggplot2::element_blank(),
      axis.title.y = ggtext::element_markdown(
        family = "Cabin",
        face = "bold",
        color = "blue",
        size = 12),
        # ggplot2::element_blank(),
      strip.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black",
        size = 18,
        face = "italic",
        margin = ggplot2::margin(2, 0, 0.5, 0, "lines")
      ),
      axis.text = ggtext::element_markdown(
        family = "Cabin",
        color = "black"),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      plot.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position = "none",
      panel.spacing.x = grid::unit(1.5, "cm"),
      panel.spacing.y = grid::unit(1.5, "cm"),
      plot.margin = ggplot2::margin(20, 20, 20, 20, "pt")
    )
)


# Set seed for reproducibility
base::set.seed(123)

```

\

#### **Filename:** 
schertz_comp_vs_tax_sustainability_FINAL_with_
warning_appendix_chatgpt.qmd

\


This analysis examines just one issue:

**How fast is total compensation growing, and how long before it doubles in size if current growth patterns continue?**

The purpose is conservative evaluation of compensation growth and warn if risk levels are high.


------------------------------------------------------------------------

# FY25 Verified Baseline (From City Payroll Spreadsheet)

Estimated General-Fund-type departments (excluding EMS, Water/Wastewater, Utility Billing, Drainage):

-   Total Earnings: \$25,334,548\
-   Benefits: \$9,824,008\
-   **Total Regular Compensation: \$35,010,997**

Benefits load ≈ 38.8% of regular earnings.\
Fully loaded multiplier ≈ 1.388.

**As of FY 24-25 citywide total compensation was ≈ \$48.49 million.**

**From FY20-FY25 compensation growth rate averaged 16.7%!**

------------------------------------------------------------------------

# Annualized Impact of FY26 Merit/Step Proposal

Budget states midyear cost:

> “These changes are planned to take effect midyear and cost \$784,000.”

Full-year wage impact: \$784,000 × 2 = \$1,568,000 - This is the full impact on FY 2026-27 budget for this one item

Fully loaded total compensation impact:

\$1,568,000 × 1.388 ≈ **\$2.18 million per year for just this one item**

This becomes a recurring structural cost added to the existing compensation growth.

------------------------------------------------------------------------

# Why This Matters

Recurring compensation growth compounds total compensation well beyond regular earnings growth.

If growth continues at 4–7% per year, compensation doubles in 10–18 years.

If it maintains average growth over the last five years of 16.7% total compensation will double in ***just 4.5 years.***  

Many of you will not have finished your terms yet!

Even if NO additional raises occur:

-   Benefits (healthcare, retirement, payroll taxes) naturally increase
-   Medical insurance “no increase” agreements are only one year (temporary)
-   Retirement contributions and healthcare costs trend strongly upward long-term

It is reasonable to assume a MINIMUM of **3–7% annual total compensation growth** **based on benefit increases alone**.

------------------------------------------------------------------------

# Doubling Time for Total Compensation Under Various Growth Rates

```{r}

growth_rates <- c(0.03,0.04,0.05,0.06,0.07,0.167)

doubling_years <- log(2) / log(1 + growth_rates)

sens_table <- data.frame(
  Annual_Growth_Rate = paste0(growth_rates*100,"%"),
  Years_to_Double = round(doubling_years,1)
)

flextable::flextable(sens_table) |>
  ds_style_flextable(title = "Doubling Time by Annual Growth Rate")
```

Interpretation:

-   3% growth → doubles in \~24 years\

-   4% growth → doubles in \~18 years\

-   5% growth → doubles in \~14 years\

-   6% growth → doubles in \~12 years\

-   7% growth → doubles in \~10 years\

-   16.7% growth → doubles in \~4.5 years

The current fully-loaded annual increase implies roughly 4.5–7% growth range.

------------------------------------------------------------------------

# Financial Warning

The General Fund is primarily funded by property tax and sales tax.

Sales tax is economically sensitive.\
Future property tax increases may require political approval.

When long-term recurring costs grow faster than stable revenue, large tax rate increases are ***unavoidable.***

This is not speculation, it is a mathematical fact.


It is a fundamental financial principle to **match long-term commitments with long-term reliable revenue sources.**

Recurring compensation increases are long-term commitments.

Temporary sales tax strength, sudden windfall property tax valuations, and one-year insurance freezes are not reliable long-term revenue sources.

------------------------------------------------------------------------

# APPENDIX: 3-Minute Public Comment Script (Must-Say Version)

Council, I want to focus on one number: doubling time.

The City’s own payroll data shows that General-Fund-type compensation is about \$35 million per year, fully loaded with benefits.

The merit and step program shown in the budget is listed as \$784,000 because it starts midyear. But annualized and fully loaded with benefits, that is about \$2.2 million per year in recurring cost.

That pushes total compensation growth into roughly a 4 to 6 percent annual range.

At 5 percent growth, compensation doubles in about 14 years. At 6 percent growth, it doubles in about 12 years.

Even if no further raises are approved, benefits alone typically grow 3 to 5 percent annually. A one-year negotiated freeze in medical insurance does not eliminate long-term healthcare inflation.

This is not about opposing compensation. It is about structural math.

If recurring compensation grows faster than stable long-term revenue, then tax rate increases are not optional — they become inevitable.

The financial principle is simple:

Long-term commitments must be matched with long-term reliable revenue.

If we do not align those two curves now, the pressure builds quietly — and then shows up later as dramatic tax increases.

Thank you.
