---
title: "Schertz FY 2025–26: Compensation Growth vs. Tax Rate (Sustainability Review)"
authors:
    - Dan Swart
    - ChatGPT
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

## Purpose and scope

This memo reviews one issue only: **whether recurring compensation growth is being supported by a stable revenue base, or whether it increases pressure for future property tax rate increases**.

It is written as a policy critique focused on long-term sustainability and tax impact.

## What the budget says (direct quotes)

The budget states the merit/step program begins midyear:

> “These changes are planned to take effect midyear and cost $784,000.” (Budget Highlights, p. 7)

The budget shows the adopted/proposed tax rates:

- FY 2024–25 rate: **$0.4900**  
- No-New-Revenue rate: **$0.5064**  
- Budget proposed rate: **$0.5064** (Tax Rates table, p. 11)

The budget also reports weakness in the existing tax base:

> “Without new value from annexations and development, the value of the existing properties decreased by $83,012,028.” (Property tax discussion, p. 11)

## Full-year correction for the merit/step program

Because the merit/step program starts midyear, the $784,000 amount is a **half-year** cost. For sustainability and “time-to-double” analysis, the recurring **full-year** amount should be used:

- Half-year shown in the budget: **$784,000**
- Full-year recurring amount: **$784,000 × 2 = $1,568,000**

This matters because recurring pay commitments do not end after one year. They become a permanent part of the payroll base.

## Key sustainability concern: short-term financing vs. long-term commitments

A fundamental financial principle is to match **short-term financing** with **short-term costs**, and **long-term revenue sources** with **long-term commitments**.

When a city supports a **long-term, recurring** payroll commitment using revenue gains that are not reliable (for example, gains that depend on a changing tax base or one-time conditions), the result is predictable: **future tax rate increases become not optional, but inevitable** unless services are reduced or hiring is frozen.

---

## Inputs and assumptions (transparent and editable)

```r
#| label: inputs-assumptions-table

assumptions <- data.frame(
  Item = c(
    "Merit/Step cost shown in budget (half-year)",
    "Full-year recurring Merit/Step cost",
    "FY 2024-25 tax rate",
    "FY 2025-26 No-New-Revenue / Proposed rate",
    "FY 2025-26 total taxable value"
  ),
  Value = c(
    "$784,000 (midyear start)",
    "$1,568,000 (784,000 × 2)",
    "$0.4900 per $100",
    "$0.5064 per $100",
    "$5,734,815,455"
  ),
  Source = c(
    "Budget Highlights (p. 7)",
    "Calculated from budget statement",
    "Tax Rates table (p. 11)",
    "Tax Rates table (p. 11)",
    "Property tax discussion (p. 11)"
  )
)

flextable::flextable(assumptions) |>
  flextable::autofit()
```

---

## Time-to-double and structural pressure

This chart shows the relationship between ongoing compensation growth (as a percent per year) and how quickly costs can double. It is a simple planning tool: higher ongoing growth rates lead to much faster doubling.

```{r}
#| label: time-to-double-plot

growth_rates <- data.frame(
  annual_growth_rate = seq(0.01, 0.08, by = 0.005)
) |>
  dplyr::mutate(
    annual_growth_pct = annual_growth_rate * 100,
    doubling_time_years = log(2) / log(1 + annual_growth_rate)
  )

ggplot2::ggplot(growth_rates, ggplot2::aes(x = annual_growth_pct, y = doubling_time_years)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::labs(
    title = "Time to Double vs. Ongoing Compensation Growth",
    x = "Ongoing compensation growth (% per year)",
    y = "Approximate years to double"
  ) +
  ggplot2::theme_minimal()
```

### Plain-language takeaway (10th grade level)

If payroll keeps rising every year, it eventually doubles.  
When payroll doubles, the city needs much more revenue just to keep services the same.

---

## 5-year projection model: payroll pressure vs. projected tax collected

### What this model does

This model compares:

1. **Payroll pressure** = starting payroll base + recurring annual increase (full-year $1.568M)  
2. **Projected tax collected** (simplified) = taxable value × assumed growth × tax rate

This is not a full revenue forecast. It is a focused stress test to keep the conversation on compensation growth vs. the property tax base.

### Growth rates used (shown in scenario names)

- **Optimistic:** 4.0% annual taxable value growth  
- **Baseline:** 2.0% annual taxable value growth  
- **Stress:** -0.5% annual taxable value decline  

### Model code (editable)



```{r}
#| label: five-year-model

# Inputs (edit if you want)
tax_rate_per_100 <- 0.5064
taxable_value_start <- 5734815455

# Merit/Step full-year recurring cost (from midyear $784k statement)
merit_step_full_year <- 784000 * 2

# Illustrative payroll base (replace with a confirmed payroll base if desired)
payroll_base_start <- 48623715

years <- 0:5

scenarios <- data.frame(
  scenario = c("Optimistic (4% Growth)",
               "Baseline (2% Growth)",
               "Stress (-0.5% Growth)"),
  taxable_value_growth = c(0.04, 0.02, -0.005)
)

proj <- tidyr::expand_grid(
  scenarios,
  year = years
) |>
  dplyr::mutate(
    taxable_value = taxable_value_start * (1 + taxable_value_growth) ^ year,
    projected_tax_collected = (taxable_value / 100) * tax_rate_per_100,
    payroll_pressure = payroll_base_start + (merit_step_full_year * year),
    gap = projected_tax_collected - payroll_pressure
  )
```

### Projection table (includes Gap)

Gap is defined as:

- **gap = projected_tax_collected − payroll_pressure**

So:
- **Positive gap** = more tax capacity than payroll pressure (less stress)
- **Negative gap** = payroll pressure is larger (more stress)

```{r}
#| label: projection-table-with-gap

proj_table <- proj |>
  dplyr::mutate(
    projected_tax_collected = round(projected_tax_collected, 0),
    payroll_pressure = round(payroll_pressure, 0),
    gap = round(gap, 0)
  ) |>
  dplyr::select(
    scenario,
    year,
    projected_tax_collected,
    payroll_pressure,
    gap
  )

flextable::flextable(proj_table) |>
  flextable::colformat_num(
    j = c("projected_tax_collected", "payroll_pressure", "gap"),
    big.mark = ",",
    digits = 0
  ) |>
  flextable::autofit()
```

### Projection plot


```{r}
#| label: five-year-plot

plot_df <- proj |>
  tidyr::pivot_longer(
    cols = c(projected_tax_collected, payroll_pressure),
    names_to = "metric",
    values_to = "value"
  ) |>
  dplyr::mutate(
    metric = dplyr::recode(
      metric,
      projected_tax_collected = "Projected Tax Collected",
      payroll_pressure = "Payroll Pressure (Base + $1.568M/year)"
    )
  )

ggplot2::ggplot(plot_df, ggplot2::aes(x = year, y = value, linetype = metric)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::facet_wrap(~ scenario, scales = "free_y") +
  ggplot2::scale_y_continuous(labels = scales::label_dollar()) +
  ggplot2::scale_x_continuous(breaks = years) +
  ggplot2::labs(
    title = "5-Year Projection: Projected Tax Collected vs Payroll Pressure",
    x = "Years after FY 2025–26",
    y = "Dollars (illustrative)",
    linetype = ""
  ) +
  ggplot2::theme_minimal()
```

---


## Bottom line (focused only on compensation vs tax rate)

- The merit/step program cost shown in the budget is **midyear** ($784,000), which becomes **$1.568M per year** when annualized. (p. 7)
- The property tax rate is proposed at the **No-New-Revenue** rate ($0.5064), and the budget reports the **existing** tax base declined without new growth. (p. 11)
- That combination creates a sustainability risk: **recurring payroll growth** can keep stacking up even when the tax base is not reliably growing.

If the city continues to add recurring compensation costs without matching them to long-term, stable funding, **large tax rate increases are not optional, but inevitable** unless staffing or services are reduced.
